---
title: "Additional Coding Examples"
author: "davegoblue"
date: "August 14, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background  
This document is for additional examples, using some DataCamp_Insights_v001 materials where applicable.  To start, this program looks at hands that can be made in a game of Three Card Poker (TCP).  Specifically, a TCP hand consists of three cards from a standard 52-card deck.  Suits do not matter (there is no master or trump suit), though three cards of the same suit form a flush.  Further, three cards of consecutive rank (A-K-Q and 3-2-A are both counted) are considered to be a straight.
  
## Simulations  
### Potential Hand Types  
The card deck will be considered to be 1-52, where 1-13, 14-26, 27-39, and 40-52 shall each be the suits.  Further, the lowest number of each suit shall be considered the Ace, and the highest shall be considered the King.  There are 52c3 or 22,100 possible hands that can be made in this game:  
```{r}
choose(52, 3)
```
  
First, each of these 22,100 hands is created using a for loop:  
```{r}
# matrix will be index 1, index 2, index 3
allHands <- matrix(data=0L, nrow=choose(52, 3), ncol=3)

intCtr <- 1
for (i in 1:50) {
    for (j in (i+1):51) {
        for (k in (j+1):52) {
           allHands[intCtr, ] <- c(i, j, k)
           intCtr <- intCtr + 1
        }
    }
}
```
  
Next, each of the hands is scored based on whether it meets the following criteria (a hand may fall in multiple categories, or in to none):  
  
* Three of a Kind  
* Straight  
* Flush  
* Pair (note that all Three of a Kind by definition will be part of Pair)  
  
```{r}
## Find the flushes
isFlush <- apply(allHands, 1, FUN = function(x) { diff(range( (x-1) %/% 13 )) == 0 })

## Find the trips
isTrips <- apply(allHands, 1, FUN = function(x) { diff(range( (x-1) %% 13 )) == 0 })

## Find the pairs (n.b. that all trips will also be defined as pairs)
isPair <- apply(allHands, 1, FUN = function(x) { !identical(x %% 13, unique(x %% 13)) })

## Find the posible straights (2-gappers using only Ace-low)
isStraightALow <- apply(allHands, 1, FUN = function(x) { diff(range( (x-1) %% 13 )) == 2 })

## Find the possible straights (2-gappers using only Ace-high)
isStraightAHigh <- apply(allHands, 1, FUN = function(x) { diff(range( (x-2) %% 13 )) == 2 })

## Consolidate (a straight is a 2-gapper that is not paired)
isStraight <- (isStraightALow | isStraightAHigh) & !isPair

```
  
Report on all of the potential hand types:  
```{r}
print(sum(isStraight & isFlush))  # Straight Flushes
print(sum(isTrips))  # Three of a Kind
print(sum(isStraight & !isFlush))  # Straight (but not SF)
print(sum(!isStraight & isFlush))  # Flush (but not SF)
print(sum(isPair & !isTrips))  # Pair (but not Trips)
print(sum(!isStraight & !isFlush & !isPair))  # Nothing

```
  
Lastly, a table of hand values is created:  
```{r}
handValues <- rep(-1, nrow(allHands))  # Default hand value is -1
handValues[isPair] <- 1  # Convert pairs to +1 (trips will be over-written later)
handValues[isFlush] <- 3  # Convert flushes to +3 (SF will be over-written later)
handValues[isStraight] <- 6  # Convert straights to +6 (SF will be over-written later)
handValues[isTrips] <- 30  # Convert trips to +30 (previous +1 for pair over-written)
handValues[isFlush & isStraight] <- 40  # Convert SF to +40 (previous +3/+6 for flush/straight over-written)

sum(handValues)
sum(handValues) / length(handValues)  # Overall return
table(handValues)
round(table(handValues) / length(handValues), 5)  # Probabilities
```
  
These values all match the published literature.  
  
Next, a similar process is run for the 52c5 example applicable to a typical poker game.  As before, all of the possible combinations are created and assessed, with caching due to moderate run times:  
```{r, cache=TRUE}
jbHands <- matrix(data=0L, nrow=choose(52, 5), ncol=5)

intCtr <- 1
for (i in 1:48) {
    for (j in (i+1):49) {
        for (k in (j+1):50) {
            for (m in (k+1):51) {
                for (n in (m+1):52) {
                               jbHands[intCtr, ] <- c(i, j, k, m, n)
                               intCtr <- intCtr + 1
                }
            }

        }
    }
}


## Find the flushes
jbFlush <- apply(jbHands, 1, FUN = function(x) { diff(range( (x-1) %/% 13 )) == 0 })

## Find the number of uniques
jbUniques <- apply(jbHands, 1, FUN = function(x) { length(unique(x %% 13)) })

## Find the range, specifically is it exactly 5
foo <- function(x) {
    min( diff(range( (x-1) %% 13 )) , diff(range( (x-2) %% 13 )) )
}
jbRange <- apply(jbHands, 1, FUN = foo)

# Is the maximum number of cards equal to 3 (only for jbUniques %in% c(2, 3))
foo <- function(x) {
    max(table(x %% 13))
}
jbMax <- apply(jbHands[jbUniques %in% c(2, 3), ], 1, FUN=foo)
```
  
Hand types are then declared, with this portion not cached since 1) it will need further refinement, and 2) it runs quickly (with the exception of the table function, since updated with dplyr/tidyr for speed):  
```{r}
jbType <- rep(-1L, nrow(jbHands))  # Default value is -1
jbType[jbUniques == 4] <- 0  # Treat all pairs as 0 for now
jbType[jbUniques == 4] <- 0  # Treat all pairs as 0 for now
jbType[jbFlush == FALSE & jbUniques == 5 & jbRange == 4] <- 3  # Straights
jbType[jbFlush == TRUE & jbRange > 4] <- 5  # Flushes

# Declare the straight flush and royal flush
rfVector <- c(1L, 10L, 11L, 12L, 0L)
foo <- function(x) {
    if (sum( x %% 13 == rfVector ) ==5 ) { return(799) }
    else { return(49) }
}
jbType[jbFlush == TRUE & jbRange == 4] <- apply(jbHands[jbFlush == TRUE & jbRange == 4, ], 1, FUN=foo)

# Look at the subset that composes Two Pair, Trips, Full House, and Quads
jbTemp <- jbUniques[jbUniques %in% c(2, 3)]
jbTempValue <- rep(0L, length(jbTemp))

jbTempValue[jbTemp == 3 & jbMax != 3] <- 1  # Two Pair
jbTempValue[jbTemp == 3 & jbMax == 3] <- 2  # Trips
jbTempValue[jbTemp == 2 & jbMax == 3] <- 8  # Full House
jbTempValue[jbTemp == 2 & jbMax != 3] <- 24  # Quads

jbType[jbUniques %in% c(2, 3)] <- jbTempValue


library(dplyr)
library(tidyr)

# Use dplyr and tidyr since they are MUCH faster than table() for the cross-tabs
data.frame(jbType=jbType) %>% group_by(jbType) %>% summarize(count=n())
data.frame(jbType=jbType, jbUniques=jbUniques) %>% 
    group_by(jbType, jbUniques) %>% 
    summarize(count=n()) %>%
    spread(jbUniques, count, fill=0)
data.frame(jbType=jbType, jbRange=jbRange) %>% 
    group_by(jbType, jbRange) %>% 
    summarize(count=n()) %>%
    spread(jbType, count, fill=0)

```
  
Potential additional refinements will include:  
  
* Only count some of the pairs (e.g., Jacks or Better)  
* Allow for rank and kickers to impact the value of the quads (e.g., Bonus or Double Double Bonus)  
  
Next, a routine is established to look-up the value of a specific hand in the existing tables:  
```{r}
# Create a function to make the numeric from a matrix with each row being a 5-integer vectors
handVec2Int <- function(x) {
    # Matricize an integer vector -- just for QC, should only pass matrices
    if (class(x) == "integer") {
        x <- matrix(data=x, nrow=1)
    }
    stopifnot(class(x) == "matrix")
    100000000 * x[,1] + 1000000 * x[,2] + 10000 * x[,3] + 100 * x[,4] + x[,5]
}
```
  
The actual hash table creation is hashed since this is one of the longer elements of the routine:  
```{r, cache=TRUE}
## jbHands is the 2,598,960 x 5 matrix of the cards (sorted ascending) in the hand
## jbType is the length 2,598,960 vector associated to the outcomes of those hands

# Set up for the looping to populate the hash table
b <- handVec2Int(jbHands)
stopifnot(all.equal(b, unique(b)))
c <- as.character(b)
intCtr <- 1

## Create a hash table for fast look-ups
myHandHash <- new.env(hash=TRUE, parent=emptyenv(), size=nrow(jbHands))

startTime <- proc.time()
for (eachNum in c) {
    
    myHandHash[[ eachNum ]] <- jbType[intCtr]
    intCtr <- intCtr + 1
}
proc.time() - startTime

intCtr
```
  
Next, the hash table is tested by passing example hands to it:  
```{r}

# Pull only from the four aces, middle spades (5-T), high hearts (9-K) and low diamonds (2-6) for speed
sampHands <- combn( sort(c(1, 14, 27, 40, 5:10, 22:26, 28:32)), 5)
sampHashLookup <- handVec2Int(t(sampHands))
sampHashChars <- as.character(sampHashLookup)

# Look-up the value for each of these hands
sampVals <- numeric(length(sampHashChars))

intCtr <- 0
startTime <- proc.time()
for (eachChar in sampHashChars) {
    
    intCtr <- intCtr + 1
    
    sampVals[intCtr] <- myHandHash[[ eachChar ]]
    
}
proc.time() - startTime

intCtr
table(sampVals)
```
  
Then, the beginning of a routine is written to look at the singe-card-draw process:  
```{r}
## Grab a sample of 20 hands from jbHands
set.seed(1608161356)
myKeys <- sample(1:nrow(jbHands), 20, replace=FALSE)
```
  
First, the routine is attempted using the hash table lookups:  
```{r}

startTime <- proc.time()
for (thisKey in myKeys) {
    keyHand <- jbHands[thisKey, , drop=TRUE]  # Convert to vector
    startValue <- jbType[thisKey]
    canReplace <- setdiff(1:52, as.vector(keyHand))  # Can only draw cards you do not already have
    
    possHands <- data.frame(canReplace, keyHand[2], keyHand[3], keyHand[4], keyHand[5])
    possHands <- matrix(data=apply(possHands, 1, FUN=sort), ncol=5, byrow=TRUE)
    
    bHash <- handVec2Int(possHands)
    stopifnot(all.equal(bHash, unique(bHash)))
    cHash <- as.character(bHash)

    valHands <- mget(cHash, envir=myHandHash)
}
proc.time() - startTime

```
  
Then, the routine is attempted using the data.table() library:  
```{r}
## Practice with data.table
library(data.table)
jbDT <- data.table(cbind(jbHands, 1:nrow(jbHands)))
setkey(jbDT, V1, V2, V3, V4, V5)

## For a given hand, replace the first card with each possible card
foo <- function(x, myDT) { myDT[ .(x[1], x[2], x[3], x[4], x[5]) ] }

startTime <- proc.time()
for (thisKey in myKeys) {
    keyHand <- jbHands[thisKey, , drop=TRUE]  # Convert to vector
    startValue <- jbType[thisKey]
    canReplace <- setdiff(1:52, as.vector(keyHand))  # Can only draw cards you do not already have
    
    possHands <- data.frame(canReplace, keyHand[2], keyHand[3], keyHand[4], keyHand[5])
    possHands <- apply(possHands, 1, FUN=sort)
    
    idxHands <- unlist(purrr::transpose(apply(possHands, 2, FUN=foo, myDT=jbDT))$V6)
    valHands <- jbType[idxHands]
}
proc.time() - startTime
```
  
The lookups still take FAR too long, although the hashed lookups run significantly faster for this small sample.  A much-improved algorithm will be needed, perhaps "working in reverse" as per material from the Wizard of Odds website.  
  
Sample plots are created using ggplot2 (cached, since they take a while to plot off the large frame):  
```{r, cache=TRUE}
library(ggplot2)

graphFrame <- data.frame(jbFlush=as.integer(jbFlush), jbRange=jbRange, 
                         jbType=jbType, jbUniques=jbUniques
                         )

# Bar chart for jbRange x jbUniques
ggplot(graphFrame, aes(x=as.factor(jbRange))) + geom_bar(aes(fill=as.factor(jbUniques)))

# Stacked bar chart for jbRange x jbUniques
ggplot(graphFrame, aes(x=as.factor(jbRange))) + 
    geom_bar(aes(fill=as.factor(jbUniques)), position="fill")

# Bar chart for jbRange x jbUniques, faceted by jbType
ggplot(graphFrame, aes(x=as.factor(jbRange))) + 
    geom_bar(aes(fill=as.factor(jbUniques))) + 
    facet_wrap(~ jbType, ncol=5)

# Stacked bar chart for jbRange x jbUniques, faceted by jbType
ggplot(graphFrame, aes(x=as.factor(jbRange))) + 
    geom_bar(aes(fill=as.factor(jbUniques)), position="fill") + 
    facet_wrap(~ jbType, ncol=5)

```
  
As per the tables, the graphs line up with expectations.  
  
A table is created to find the minimum row for each of the smallest cards:  
```{r}
# Index creation algorithm
idxTemp <- table(jbHands[, 1])

# Similar process, but using dplyr
as.data.table(jbHands) %>% group_by(V1) %>% summarize(ct=n())
```
  
Next, the program takes a stab at implementing the algorithm described by the Wizard of Odds at <http://wizardofodds.com/games/video-poker/methodology/> for vastly reducing run times for a program of this type.  The Wizard recommends an 11-step process, including:  
  
1.  Initialize array 0 (2,598,960) for the result of a single dealt hand with 0 draws, array 1 (270,725 x 16) for a 1-card draw, array 2 (22,100 x 16) for a 2-card draw, array 3 (1,326 x 16) for a 3-card draw, array 4 (52 x 16) for a 4-card draw, and array 5 (1 x 16) for a 5-card draw.  The 16 is for the maximum number of outcomes (e.g., -1 or +799) for the hand, and could be expanded if a game had more than 16 paying types.  
  
2.  Loop through every possible hand (these are the jbHands and jbType in this algorithm), and populate array0, array1, array2, array3, array4, and array5 based on possible permutations of cards held on the draw  
  
First, the 16-hand condition is tested, and the relevant arrays are set up:
```{r}

nrowArray0 <- data.frame(jbType=jbType) %>% 
    group_by(jbType) %>% 
    summarize(ct=n()) %>%
    arrange(jbType)
nrowArray0

if (nrow(nrowArray0) > 16) { 
    stop(paste0("There are too many hand types, cap is 16, this will have: ", nrow(nrowArray0)))
}
print("OK")

# Initialize array1-array5
array1 <- matrix(data=0L, nrow=choose(52, 4), ncol=16)
array2 <- matrix(data=0L, nrow=choose(52, 3), ncol=16)
array3 <- matrix(data=0L, nrow=choose(52, 2), ncol=16)
array4 <- matrix(data=0L, nrow=choose(52, 1), ncol=16)
array5 <- matrix(data=0L, nrow=choose(52, 0), ncol=16)

# Create array0 which will have an index number corresponding to the associated jbType
array0 <- match(jbType, nrowArray0[[1]])

data.frame(array0=array0) %>% 
    group_by(array0) %>%
    summarize(ct=n()) %>%
    arrange(array0)

```
    
Begin by populating the easiest arrays: array3, array4, and array5.  Basically, this is just looping through each of the jbHands, keeping 2 or 1 or 0 of the cards, and adding 1 to the appropriate column of array3, or array4, or array5:  
```{r}

idxKeep2 <- data.frame(card1=1:51, posn=cumsum( c(0, 51:2) ) )

startTime <- proc.time()

# Populate array5
a <- data.frame(array0=array0) %>% group_by(array0) %>% summarize(ct=n()) %>% arrange(array0)
array5[1, a$array0] <- a$ct

proc.time() - startTime


# Populate array4
a <- data.frame(keyCard=as.vector(jbHands), val=rep(array0, 5)) %>%
    group_by(keyCard, val) %>% summarize(ct=n()) %>% arrange(keyCard, val)
for (intCtr in 1:nrow(a)) { b <- a[intCtr, , drop=FALSE] ; array4[b$keyCard, b$val] <- b$ct }

proc.time() - startTime


# Populate array3
idx <- integer(0)
ct <- integer(0)
val <- integer(0)

for (c1 in 1:4) {
    for (c2 in (c1+1):5) {
        a <- data.frame(card1=jbHands[, c1], card2=jbHands[, c2], val=array0) %>% 
            group_by(card1, card2, val) %>% 
            summarize(ct=n()) %>% 
            arrange(card1, card2, val) %>% 
            mutate(idx = idxKeep2[card1, "posn"] + card2 - card1)
        idx <- c(idx, as.integer(a$idx))
        val <- c(val, a$val)
        ct <- c(ct, a$ct)
    }
}

for (intCtr in 1:length(ct)) {
    array3[idx[intCtr], val[intCtr]] <- ct[intCtr]
}

proc.time() - startTime

```
  
With array3 run in a more vectorized manner, the program runs fairly quickly.  The next steps are to create array2 and array1, each of which have a more complex indexing methodology.  