---
title: "Mapping and Plotting"
author: "davegoblue"
date: "3/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

This document is to provide some plotting examples for reference.

***

#### _Example #1: US State and County Maps_ 
The package usmap contains maps of US states and counties.  There is also some associated data available about state and county demographics.

Example code includes:  
```{r}

library(tidyverse)

# Population by county
data(countypop, package="usmap")
# Population by state
data(statepop, package="usmap")
# Poverty rate by county
data(countypov, package="usmap")
# Poverty rate by state
data(statepov, package="usmap")
# Population of largest city by state
data(citypop, package="usmap")
# Location of earthquakes
data(earthquakes, package="usmap")


# Included datasets
countypop
statepop
countypov
statepov
citypop
earthquakes

# Basic, empty US maps
usmap::plot_usmap(regions="states")
usmap::plot_usmap(regions="counties")

# Basic, empty US maps subsetted to an area
usmap::plot_usmap(regions="states", 
                  include=c("WA", "OR", "CA", "NV", "ID", "MT", "WY", "UT", "CO", "AZ", "NM")
                  )
usmap::plot_usmap(regions="counties", include=c("MN", "WI", "MI", "OH", "PA", "NY", "IN", "IL"))

# Basic, subsetted state map with poverty rates included
usmap::plot_usmap(regions="states", 
                  include=c("WA", "OR", "CA", "NV", "ID", "MT", "WY", "UT", "CO", "AZ", "NM"), 
                  values="pct_pov_2014", data=statepov, labels=TRUE
                  ) + 
    scale_fill_continuous(low="lightblue", high="darkblue", "Poverty Rate (%)") + 
    labs(title="Poverty Rates by Western and Mountain States")

# Basic, subsetted county map with poverty rates included
usmap::plot_usmap(regions="counties", include=c("MN", "WI", "MI", "OH", "PA", "NY", "IN", "IL"),
                  values="pct_pov_2014", data=countypov, labels=FALSE
                  ) + 
    scale_fill_continuous(low="lightblue", high="darkblue", "Poverty Rate (%)") + 
    labs(title="Poverty Rates by County in Great Lakes States")

```

#### _Example #2: Converting and adding lat/lon data_ 
The latitude and longitude data can be converted to a form suitable for usmap by using the usmap_transform function.

Example code includes:  
```{r}

# Transform the earthquakes data
trQuakes <- usmap::usmap_transform(earthquakes)
str(trQuakes)

# Add as a layer to the state map
usmap::plot_usmap(regions="states") + 
    geom_point(data=trQuakes, aes(x=lon.1, y=lat.1, size=mag), alpha=0.4) + 
    labs(title="Earthquakes of Magnitude 2.5+ (H1 2019)")

# Transform the largest city data
trCity <- usmap::usmap_transform(citypop)
str(trCity)

# Add as a layer to the state map
usmap::plot_usmap(regions="states") + 
    geom_point(data=trCity, aes(x=lon.1, y=lat.1, size=city_pop)) + 
    labs(title="Largest City by State")

```
  
#### _Example #3: Filtering and coloring by region_ 
The census region definitions are included, and can be used to filter or color the maps.

Example code includes:  
```{r}

# Filter the map to include only new_england, mid_atlantic, and south_atlantic
usmap::plot_usmap(regions="states", 
                  include=c(usmap::.new_england, usmap::.mid_atlantic, usmap::.south_atlantic)
                  )

# Create regions data for US states
regionData <- usmap::statepop %>%
    mutate(region=as.factor(ifelse(abbr %in% usmap::.midwest_region, 1, 0)))
usmap::plot_usmap(regions="states", data=regionData, values="region") + 
    scale_fill_discrete("Midwest") + 
    labs(title="Midwest Region US States")

# Enhanced Coloring and Labelling
usmap::plot_usmap(regions="states", data=regionData, values="region") + 
    scale_fill_manual(values=c("lightgray", "lightblue"), "", labels=c("Other", "Midwest")) + 
    labs(title="Midwest Region US States")

```

#### _Example #4: Labelling geographies_ 
Since usmap is built on ggplot2, the standard techniques from ggplot2 can be used to enhance the geography labelling.  Further, centroids for the geographies are available in loadable files.

Example code includes:  
```{r}

# Base state map labelled with defaults
usmap::plot_usmap(regions="states", labels=TRUE, label_color="red")

# Base county map labelled with defaults
usmap::plot_usmap(regions="counties", include=c("TX", "OK"), labels=TRUE, label_color="grey")

# Load state centroid data
stCenter <- utils::read.csv(system.file("extdata", 
                                        paste0("us_", "states", "_centroids.csv"), package = "usmap"
                                        ),
                            colClasses = c(rep("numeric", 2), rep("character", 3)), stringsAsFactors = FALSE
                            )

# Load county centroid data
ctCenter <- utils::read.csv(system.file("extdata", 
                                        paste0("us_", "counties", "_centroids.csv"), package = "usmap"
                                        ),
                            colClasses = c(rep("numeric", 2), rep("character", 4)), stringsAsFactors = FALSE
                            )

# Add state labels using geom_text
regionData <- usmap::statepop %>%
    mutate(region=as.factor(ifelse(abbr %in% usmap::.midwest_region, 1, 0))) %>%
    left_join(stCenter %>% select(x, y, full, fips) %>% rename(fname=full)) %>%
    mutate(fname=ifelse(fname=="District of Columbia", "DC", str_replace_all(fname, " ", "\n")))

usmap::plot_usmap(regions="states", data=regionData[, c("fips", "region")], values="region") + 
    scale_fill_manual(values=c("lightgray", "lightblue"), "", labels=c("Other", "Midwest")) + 
    labs(title="Midwest Region US States") + 
    geom_text(data=filter(regionData, region==1), aes(x=x, y=y, label=fname), size=2.5)


# Add county labels using geom_text
regionData <- usmap::countypop %>%
    mutate(region=as.factor(case_when(abbr=="OK" ~ 1, abbr=="TX" ~ 2, TRUE ~ 0))) %>%
    left_join(ctCenter %>% select(x, y, county, fips) %>% rename(cname=county)) %>%
    mutate(cname=str_replace_all(str_replace(cname, " County", ""), " ", "\n"))

usmap::plot_usmap(regions="counties", include=c("TX", "OK"), 
                  data=regionData[, c("fips", "region")], values="region") + 
    scale_fill_manual(values=c("red", "orange"), "", labels=c("Oklahoma", "Texas")) + 
    labs(title="Texas and Oklahoma Counties") + 
    geom_text(data=filter(regionData, abbr %in% c("TX", "OK")), 
              aes(x=x, y=y, label=cname), size=2.5, 
              color=ifelse(pull(filter(regionData, abbr %in% c("TX", "OK")), abbr)=="OK", "white", "black")
              )

```

#### _Example #5: Adding population centers_ 
Separate data exists for key population centers, which can be loaded and then added to maps.

Example code includes:  
```{r}

# Transform the largest city data
str(maps::us.cities)
trCity <- usmap::usmap_transform(select(maps::us.cities, long, lat, everything())) %>% 
    mutate(useName=str_replace_all(str_sub(name, 1, -4), " ", "\n"))
str(trCity)

# Define a key region for plotting
rgnPlot <- c(usmap::.west_south_central, usmap::.east_south_central)
popFilter <- 100000

# Add cities as a layer to the state map
usmap::plot_usmap(regions="states", include=rgnPlot, fill="lightblue") + 
    labs(title=paste0("South Central Cities with Population >= ", popFilter/1000, "k")) + 
    geom_point(data=filter(trCity, pop >= popFilter, country.etc %in% rgnPlot), 
               aes(x=long.1, y=lat.1, size=pop), alpha=0.5
               )

# Plot the full nation for cities of 250k +
rgnPlot <- c(usmap::.midwest_region, usmap::.northeast_region, 
             usmap::.south_region, usmap::.west_region
             )
popFilter <- 250000

# Add cities as a layer to the state map
usmap::plot_usmap(regions="states", include=rgnPlot, fill="lightblue") + 
    labs(title=paste0("US Cities with Population >= ", popFilter/1000, "k")) + 
    geom_point(data=filter(trCity, pop >= popFilter, country.etc %in% rgnPlot), 
               aes(x=long.1, y=lat.1, size=pop), alpha=0.5
               )

# Plot cities by name for the Four Corners region
rgnPlot <- c("UT", "CO", "NM", "AZ")
popFilter <- 125000

# Add cities as a layer to the state map (points)
usmap::plot_usmap(regions="counties", include=rgnPlot, fill="lightblue") + 
    labs(title=paste0("Four Corners Cities with Population >= ", popFilter/1000, "k")) + 
    geom_point(data=filter(trCity, pop >= popFilter, country.etc %in% rgnPlot), 
               aes(x=long.1, y=lat.1, size=pop), alpha=0.5
               )

# Add cities as a layer to the state map (text)
usmap::plot_usmap(regions="counties", include=rgnPlot, fill="lightblue") + 
    labs(title=paste0("Four Corners Cities with Population >= ", popFilter/1000, "k")) + 
    geom_text(data=filter(trCity, pop >= popFilter, country.etc %in% rgnPlot), 
               aes(x=long.1, y=lat.1, size=pop, label=useName)
               )

popFilter <- 50000
popFilter2 <- 250000

# Add cities as a layer to the state map (points and text)
usmap::plot_usmap(regions="counties", include=rgnPlot, fill="lightblue") + 
    labs(title=paste0("Four Corners Cities with Population >= ", popFilter/1000, "k")) + 
    geom_point(data=filter(trCity, pop >= popFilter, country.etc %in% rgnPlot), 
               aes(x=long.1, y=lat.1, size=pop), alpha=0.5
               ) + 
    geom_text(data=filter(trCity, pop >= popFilter2, country.etc %in% rgnPlot), 
               aes(x=long.1, y=lat.1, size=pop, label=useName), color="red", show.legend=FALSE
               )

```
  
#### _Example #6: Custom coloring geographies_  
Using scale_fill_manual(), custom colors can be created by geography.

Example code includes:  
```{r}

# Basic county population map with continuous colors
usmap::countypop %>% 
    filter(abbr %in% c("OH", "IN", "KY")) %>% 
    mutate(pop=pop_2015/1000, name=str_replace(str_replace(county, " County", ""), " ", "\n")) %>%
    usmap::plot_usmap(regions="counties", include=c("OH", "IN", "KY"), data=., values="pop") +
    scale_fill_continuous(low="lightblue", high="darkblue", "Pop. (000)") + 
    labs(title="Indiana, Ohio, and Kentucky - Population by County")

# Custom county population map with colors - red for Indiana, blue for Kentucky, grey for Ohio
popBucket <- c(0, 100, 500)
popLabels <- sapply(1:(length(popBucket)-1), FUN=function(x){paste0(popBucket[x], "-", popBucket[x+1])})
popLabels <- c(popLabels, paste0(popBucket[length(popBucket)], "+"))
guideLabels <- paste(rep(c("OH", "KY", "IN"), each=3), popLabels)

usmap::countypop %>% 
    filter(abbr %in% c("OH", "KY", "IN")) %>% 
    mutate(pop=pop_2015/1000, name=str_replace(str_replace(county, " County", ""), " ", "\n"), 
           pBucket=findInterval(pop, popBucket), 
           pColor=rgb(abbr=="IN", 0, abbr=="KY", pBucket/length(popBucket))
           ) %>%
    usmap::plot_usmap(regions="counties", include=c("OH", "IN", "KY"), data=., values="pColor") +
    scale_fill_identity(guide="legend", "Pop. (000)", labels=guideLabels) + 
    labs(title="Indiana, Ohio, and Kentucky - Population by County") + 
    theme(legend.position = "bottom") + 
    guides(fill=guide_legend(nrow=3))

# Custom county poverty rate map with colors - red for Indiana, blue for Kentucky, grey for Ohio
povBucket <- c(0, 15, 30)
povLabels <- sapply(1:(length(povBucket)-1), FUN=function(x){paste0(povBucket[x], "-", povBucket[x+1])})
povLabels <- c(povLabels, paste0(povBucket[length(povBucket)], "+"))
guideLabels <- paste(rep(c("OH", "KY", "IN"), each=3), povLabels)

usmap::countypov %>% 
    filter(abbr %in% c("OH", "KY", "IN")) %>% 
    mutate(name=str_replace(str_replace(county, " County", ""), " ", "\n"), 
           pBucket=findInterval(pct_pov_2014, povBucket), 
           pColor=rgb(abbr=="IN", 0, abbr=="KY", pBucket/length(povBucket))
           ) %>%
    usmap::plot_usmap(regions="counties", include=c("OH", "IN", "KY"), data=., values="pColor") +
    scale_fill_identity(guide="legend", "Poverty Rate (%)", labels=guideLabels) + 
    labs(title="Indiana, Ohio, and Kentucky - Poverty Rate by County") + 
    theme(legend.position = "bottom") + 
    guides(fill=guide_legend(nrow=3))

```

#### _Example #7: Custom labeling of key geographies_  
The above techniques can be combined for custom labeling of key geographies.

Example code includes:  
```{r}

# Basic state population data
stateData <- usmap::statepop %>% 
    mutate(pop=round(pop_2015/1000000, 1), 
           name=ifelse(full=="District of Columbia", "DC", str_replace(full, " ", "\n")), 
           lab=paste0(abbr, "\n(", pop, ")\n")
           )

# Load state centroid data
stCenter <- utils::read.csv(system.file("extdata", 
                                        paste0("us_", "states", "_centroids.csv"), package = "usmap"
                                        ),
                            colClasses = c(rep("numeric", 2), rep("character", 3)), stringsAsFactors = FALSE
                            )

# Grab centroids for top 5 states
top5State <- stateData %>%
    top_n(5, pop) %>% 
    left_join(select(stCenter, x, y, fips))

# Plot state population with continuous colors and custom labels
stateData %>% 
    usmap::plot_usmap(regions="states", data=., values="pop") +
    scale_fill_continuous(low="lightblue", high="darkblue", "Pop. (millions)") + 
    labs(title="Population by State", subtitle="Top 5 in millions") + 
    geom_text(data=top5State, aes(x=x, y=y, label=lab), color="white", size=4, fontface="bold")


# Load county centroid data
ctCenter <- utils::read.csv(system.file("extdata", 
                                        paste0("us_", "counties", "_centroids.csv"), package = "usmap"
                                        ),
                            colClasses = c(rep("numeric", 2), rep("character", 4)), stringsAsFactors = FALSE
                            )

# Custom county population map with colors - red for Wisconsin, blue for Michigan
popBucket <- c(0, 100, 500)
popLabels <- sapply(1:(length(popBucket)-1), FUN=function(x){paste0(popBucket[x], "-", popBucket[x+1])})
popLabels <- c(popLabels, paste0(popBucket[length(popBucket)], "+"))
guideLabels <- paste(rep(c("MI", "WI"), each=3), popLabels)

# Grab county data for counties exceeding the top popBucket
ctyData <- usmap::countypop %>%
    filter(abbr %in% c("MI", "WI")) %>%
    mutate(pop=round(pop_2015/1000, 0), 
           name=str_replace(str_replace(county, " County", ""), " ", "\n"), 
           lab=paste0(name, "\n(", pop, ")\n")
           )

topCounty <- ctyData %>%
    filter(pop >= max(popBucket)) %>%
    left_join(select(ctCenter, x, y, fips))

# Create county population map
usmap::countypop %>% 
    filter(abbr %in% c("MI", "WI")) %>% 
    mutate(pop=pop_2015/1000, name=str_replace(str_replace(county, " County", ""), " ", "\n"), 
           pBucket=findInterval(pop, popBucket), 
           pColor=rgb(abbr=="WI", 0, abbr=="MI", pBucket/length(popBucket))
           ) %>%
    usmap::plot_usmap(regions="counties", include=c("WI", "MI"), data=., values="pColor") +
    scale_fill_identity(guide="legend", "Pop. (000)", labels=guideLabels) + 
    geom_text(data=topCounty, aes(x=x, y=y, label=lab), size=3, fontface="bold", color="white") +
    labs(title="Michigan and Wisconsin - Population by County", subtitle="Labelled Pop. (000) for 500k+") + 
    theme(legend.position = "bottom") + 
    guides(fill=guide_legend(nrow=3)) + 
    theme(panel.background=element_rect(color="black", fill="lightgrey"))

```


