---
title: "Coronavirus US - All-Cause Death"
author: "davegoblue"
date: '2022-06-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background  
This module extends code contained in Coronavirus_Statistics_v005.Rmd to include sourcing of updated functions and parameters.  This file includes the latest code for analyzing all-cause death data from [CDC Weekly Deaths by Jurisdiction](https://catalog.data.gov/dataset/weekly-counts-of-deaths-by-jurisdiction-and-age-group).  CDC maintains data on deaths by week, age cohort, and state in the US.  Downloaded data are unique by state, epidemiological week, year, age, and type (actual vs. predicted/projected).

These data are known to have a lag between death and reporting, and the CDC back-correct to report deaths at the time the death occurred even if the death is reported in following weeks.  This means totals for recent weeks tend to run low (lag), and the CDC run a projection of the expected total number of deaths given the historical lag times.  Per other analysts on the internet, there is currently significant supra-lag, with lag times much longer than historical averages causing CDC projected deaths for recent weeks to be low.

The code leverages tidyverse and sourced functions throughout:  
```{r}

# All functions assume that tidyverse and its components are loaded and available
library(tidyverse)

# If the same function is in both files, use the version from the more specific source
source("./Generic_Added_Utility_Functions_202105_v001.R")
source("./Coronavirus_CDC_Excess_Functions_v001.R")

```
  
The basic process includes three data update steps:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

# STEP 0: Appropriate parameters for 2022 data
cdcExcessParams <- list(remapVars=c('Jurisdiction'='fullState', 
                                    'Week Ending Date'='weekEnding', 
                                    'State Abbreviation'='state', 
                                    'Age Group'='age', 
                                    'Number of Deaths'='deaths', 
                                    'Time Period'='period', 
                                    'Year'='year', 
                                    'Week'='week'
                                    ),
                        colTypes="ccciicdcccc",
                        ageLevels=c("Under 25 years", 
                                    "25-44 years", 
                                    "45-64 years", 
                                    "65-74 years", 
                                    "75-84 years", 
                                    "85 years and older"
                                    ),
                        periodLevels=c("2015-2019", "2020", "2021", "2022"),
                        periodKeep=c("2015-2019", "2020", "2021"),
                        yearLevels=2015:2022
                        )

# STEP 1: Latest CDC all-cause deaths data
cdcLoc <- "Weekly_counts_of_deaths_by_jurisdiction_and_age_group_downloaded_20220623.csv"
cdcList_20220623 <- readRunCDCAllCause(loc=cdcLoc, 
                                       weekThru=21, 
                                       lst=readFromRDS("cdc_daily_220602"), 
                                       stateNoCheck=c(), 
                                       pdfCluster=TRUE, 
                                       pdfAge=TRUE
                                       )
saveToRDS(cdcList_20220623, ovrWriteError=FALSE)

# STEP 2: Latest death bu location-cause data
allCause_220623 <- analyzeAllCause(loc="COvID_deaths_age_place_20220623.csv", 
                                   cdcDailyList=readFromRDS("cdc_daily_220602"), 
                                   compareThruDate="2022-05-31"
                                   )
saveToRDS(allCause_220623, ovrWriteError=FALSE)

# STEP 3: Facets for excess all-cause deaths
excessDeathFacets(lstCDC=cdcList_20220623, lstAll=allCause_220623, dateThru="2022-04-30", plotYLim=c(-200, 1200))

```
  
Updated with the latest data:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

# STEP 1: Latest CDC all-cause deaths data
cdcLoc <- "Weekly_counts_of_deaths_by_jurisdiction_and_age_group_downloaded_20220713.csv"
cdcList_20220713 <- readRunCDCAllCause(loc=cdcLoc, 
                                       weekThru=24, 
                                       lst=readFromRDS("cdc_daily_220704"), 
                                       stateNoCheck=c(), 
                                       pdfCluster=TRUE, 
                                       pdfAge=TRUE
                                       )
saveToRDS(cdcList_20220713, ovrWriteError=FALSE)

# STEP 2: Latest death bu location-cause data
allCause_220713 <- analyzeAllCause(loc="COvID_deaths_age_place_20220713.csv", 
                                   cdcDailyList=readFromRDS("cdc_daily_220704"), 
                                   compareThruDate="2022-06-30"
                                   )
saveToRDS(allCause_220713, ovrWriteError=FALSE)

# STEP 3: Facets for excess all-cause deaths
excessDeathFacets(lstCDC=cdcList_20220713, lstAll=allCause_220713, dateThru="2022-05-31", plotYLim=c(-200, 1200))

```
  
