---
title: "CDC Daily by State"
author: "davegoblue"
date: "2023-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
This file is designed to use CDC data to assess coronavirus disease burden by state, including creating and analyzing state-level clusters.

Through March 7, 2021, [The COVID Tracking Project](https://covidtracking.com/) collected and integrated data on tests, cases, hospitalizations, deaths, and the like by state and date.  The latest code for using this data is available in Coronavirus_Statistics_CTP_v004.Rmd.

The COVID Tracking Project suggest that [US federal data sources](https://covidtracking.com/analysis-updates/federal-covid-data-101-how-to-find-data) are now sufficiently robust to be used for analyses that previously relied on COVID Tracking Project.

The code in this module builds on code available in _v005, with function and mapping files updated:  
  
* Generic_Added_Utility_Functions_202105_v001.R - generic functions that can be used in other areas  
* Coronavirus_CDC_Daily_Functions_v002.R - functions specific to coronavirus daily data  
  
Broadly, the CDC data analyzed by this module includes:  
  
* CDC weekly case and death data by state and date are available for download on the [CDC website](https://data.cdc.gov/Case-Surveillance/Weekly-United-States-COVID-19-Cases-and-Deaths-by-/pwn4-m3yp)  
* CDC hospital data are available for download on the [healthdata.gov website](https://beta.healthdata.gov/api/views/g62h-syeh/rows.csv?accessType=DOWNLOAD)  
* CDC vaccines data are also available for download on the [CDC website](https://data.cdc.gov/api/views/unsk-b7fc/rows.csv?accessType=DOWNLOAD):  
  
## Functions and Mapping Files
The tidyverse package is loaded and functions are sourced:  
```{r}

# The tidyverse functions are routinely used without package::function format
library(tidyverse)
library(geofacet)

# Functions are available in source file
source("./Generic_Added_Utility_Functions_202105_v001.R")
source("./Coronavirus_CDC_Daily_Functions_v002.R")

```

A series of mapping files are also available to allow for parameterized processing.  Mappings include:  
  
* urlMapper - mapping file for urlType and url location to download data  
* renMapper - mapping file for renaming of variables in the raw data file  
* selfListMapper - mapping file for transformations by variable type  
* fullListMapper - mapping file for transformations across variable types  
* lstComboMapper - mapping file for elements to be combined by data type (most common is to combine NYC and NYS data to NY if the file provides them separately)  
* uqMapper - mapping file for fields that should combine to be unique keys for processed files  
* lstFilterMapper - mapping file for filtering to subset (most common is to keep 50 states and DC)  
* vecSelectMapper - mapping file for variables to keep  
* checkControlGroupMapper - mapping file for group_by() of control total checks  
* checkControlAggMapper - mapping file for numeric variables for control total checks  
* checkSimilarityMapper - mapping file for similarity checks to perform  
* plotSimilarityMaooer - mapping file for fields where differences in universe should be plotted
* keyAggMapper - mapping file for the aggregate-level control total checks to perform  
* perCapMapper - named vector that drives conversion from original field name to per capita field name  
* hhsMapper - named vector that drivers numerical variables to keep (and renaming) from HHS capacity data file
  
These default parameters are maintained in a separate .R file and can be sourced:  
```{r}

# These have been updated to _v003
source("./Coronavirus_CDC_Daily_Default_Mappings_v003.R")

```

## Example Code Processing
The function is run on previously downloaded data:

The latest data are downloaded and processed:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

readList <- list("cdcWeeklyBurden"="./RInputFiles/Coronavirus/CDC_dc_wkly_downloaded_230402.csv", 
                 "cdcHosp"="./RInputFiles/Coronavirus/CDC_h_downloaded_230402.csv", 
                 "vax"="./RInputFiles/Coronavirus/vaxData_downloaded_230402.csv"
                 )
compareList <- list("cdcWeeklyBurden"=readFromRDS("cdc_daily_230302")$dfRaw$cdcWeeklyBurden, 
                    "cdcHosp"=readFromRDS("cdc_daily_230302")$dfRaw$cdcHosp, 
                    "vax"=readFromRDS("cdc_daily_230302")$dfRaw$vax
                    )

cdc_daily_230402 <- readRunCDCDaily(thruLabel="Mar 31, 2023", 
                                    downloadTo=lapply(readList, FUN=function(x) if(file.exists(x)) NA else x), 
                                    readFrom=readList,
                                    compareFile=compareList, 
                                    writeLog=NULL, 
                                    useClusters=readFromRDS("cdc_daily_210528")$useClusters, 
                                    weightedMeanAggs=c("tcpm7", "tdpm7", "cpm7", "dpm7", "hpm7", 
                                                       "vxcpm7", "vxcgte65pct"
                                                       ),
                                    skipAssessmentPlots=FALSE, 
                                    brewPalette="Paired"
                                    )
saveToRDS(cdc_daily_230402, ovrWriteError=FALSE)

```
  
The latest hospitalization data is also downloaded and processed:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

# Run for latest data, save as RDS
indivHosp_20230403 <- downloadReadHospitalData(loc="./RInputFiles/Coronavirus/HHS_Hospital_20230403.csv")
saveToRDS(indivHosp_20230403, ovrWriteError=FALSE)

```
  
Post-processing is run, including hospital summaries:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

# Create pivoted burden data
burdenPivotList_230402 <- postProcessCDCDaily(cdc_daily_230402, 
                                              dataThruLabel="Mar 2023", 
                                              keyDatesBurden=c("2023-03-29", "2022-06-30", 
                                                               "2021-12-31", "2021-06-30"
                                                               ),
                                              keyDatesVaccine=c("2023-03-29", "2021-12-31", 
                                                                "2021-08-31", "2021-03-31"
                                                                ), 
                                              returnData=TRUE
                                              )

# Create hospitalized per capita data
hospPerCap_230402 <- hospAgePerCapita(readFromRDS("dfStateAgeBucket2019"), 
                                      lst=burdenPivotList_230402, 
                                      popVar="pop2019", 
                                      excludeState=c(), 
                                      cumStartDate="2020-07-15"
                                      )

burdenPivotList_230402$hospAge %>%
    group_by(adultPed, confSusp, age, name) %>%
    summarize(value=sum(value, na.rm=TRUE), n=n(), .groups="drop")

saveToRDS(burdenPivotList_230402, ovrWriteError=FALSE)
saveToRDS(hospPerCap_230402, ovrWriteError=FALSE)

```
  
Peaks and valleys of key metrics are also updated:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

peakValleyCDCDaily(cdc_daily_230402)

```
  
Hospital data are pieced together as needed:
```{r, fig.height=9, fig.width=9}

# Create modified hospital data
multiSourceHosp_20230402 <- multiSourceDataCombine(list(readFromRDS("indivHosp_20220704"),
                                                        readFromRDS("indivHosp_20230403")
                                                        ),
                                                   timeVec=as.Date("2022-01-01")
                                                   )

```
  
The updated hospital data are then plotted:
```{r, fig.height=9, fig.width=9, cache=TRUE}

# Run hospital plots
modStateHosp_20230402 <- hospitalCapacityCDCDaily(multiSourceHosp_20230402, 
                                                  plotSub="Aug 2020 to Mar 2023\nOld data used pre-2022"
                                                  )

```
  

## Data Refreshes
### _Data From 2023-05-02_
The latest data are downloaded and processed:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

readList <- list("cdcWeeklyBurden"="./RInputFiles/Coronavirus/CDC_dc_wkly_downloaded_230502.csv", 
                 "cdcHosp"="./RInputFiles/Coronavirus/CDC_h_downloaded_230502.csv", 
                 "vax"="./RInputFiles/Coronavirus/vaxData_downloaded_230502.csv"
                 )
compareList <- list("cdcWeeklyBurden"=readFromRDS("cdc_daily_230402")$dfRaw$cdcWeeklyBurden, 
                    "cdcHosp"=readFromRDS("cdc_daily_230402")$dfRaw$cdcHosp, 
                    "vax"=readFromRDS("cdc_daily_230402")$dfRaw$vax
                    )

cdc_daily_230502 <- readRunCDCDaily(thruLabel="Apr 30, 2023", 
                                    downloadTo=lapply(readList, FUN=function(x) if(file.exists(x)) NA else x), 
                                    readFrom=readList,
                                    compareFile=compareList, 
                                    writeLog=NULL, 
                                    useClusters=readFromRDS("cdc_daily_210528")$useClusters, 
                                    weightedMeanAggs=c("tcpm7", "tdpm7", "cpm7", "dpm7", "hpm7", 
                                                       "vxcpm7", "vxcgte65pct"
                                                       ),
                                    skipAssessmentPlots=FALSE, 
                                    brewPalette="Paired"
                                    )
saveToRDS(cdc_daily_230502, ovrWriteError=FALSE)

```
  
The latest hospitalization data is also downloaded and processed:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

# Run for latest data, save as RDS
indivHosp_20230503 <- downloadReadHospitalData(loc="./RInputFiles/Coronavirus/HHS_Hospital_20230503.csv")
saveToRDS(indivHosp_20230503, ovrWriteError=FALSE)

```
  
