---
title: "CDC Daily by State"
author: "davegoblue"
date: "8/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
This file is designed to use CDC data to assess coronavirus disease burden by state, including creating and analyzing state-level clusters.

Through March 7, 2021, [The COVID Tracking Project](https://covidtracking.com/) collected and integrated data on tests, cases, hospitalizations, deaths, and the like by state and date.  The latest code for using this data is available in Coronavirus_Statistics_CTP_v004.Rmd.

The COVID Tracking Project suggest that [US federal data sources](https://covidtracking.com/analysis-updates/federal-covid-data-101-how-to-find-data) are now sufficiently robust to be used for analyses that previously relied on COVID Tracking Project.  This code is an attempt to update modules in Coronavirus_Statistics_CTP_v004.Rmd to leverage US federal data.

The code in this module builds on code available in _v002 to include vaccines data:  
  
* Generic_Added_Utility_Functions_202105_v001.R - generic functions that can be used in other areas  
* Coronavirus_CDC_Daily_Functions_v001.R - functions specific to coronavirus daily data  
  
Broadly, the CDC data analyzed by this module includes:  
  
* CDC case and death data by state and date are available for download on the [CDC website](https://data.cdc.gov/api/views/9mfq-cb36/rows.csv?accessType=DOWNLOAD)  
* CDC hospital data are available for download on the [healthdata.gov website](https://beta.healthdata.gov/api/views/g62h-syeh/rows.csv?accessType=DOWNLOAD)  
* CDC vaccines data are also available for download on the [CDC website](https://data.cdc.gov/api/views/unsk-b7fc/rows.csv?accessType=DOWNLOAD):  
  
## Functions and Mapping Files
The tidyverse package is loaded and functions are sourced:  
```{r}

# The tidyverse functions are routinely used without package::function format
library(tidyverse)

# Functions are available in source file
source("./Generic_Added_Utility_Functions_202105_v001.R")
source("./Coronavirus_CDC_Daily_Functions_v001.R")

```

A series of mapping files are also available to allow for parameterized processing.  Mappings include:  
  
* urlMapper - mapping file for urlType and url location to download data  
* renMapper - mapping file for renaming of variables in the raw data file  
* selfListMapper - mapping file for transformations by variable type  
* fullListMapper - mapping file for transformations across variable types  
* lstComboMapper - mapping file for elements to be combined by data type (most common is to combine NYC and NYS data to NY if the file provides them separately)  
* uqMapper - mapping file for fields that should combine to be unique keys for processed files  
* lstFilterMapper - mapping file for filtering to subset (most common is to keep 50 states and DC)  
* vecSelectMapper - mapping file for variables to keep  
* checkControlGroupMapper - mapping file for group_by() of control total checks  
* checkControlAggMapper - mapping file for numeric variables for control total checks  
* checkSimilarityMapper - mapping file for similarity checks to perform  
* plotSimilarityMaooer - mapping file for fields where differences in universe should be plotted
* keyAggMapper - mapping file for the aggregate-level control total checks to perform  
* perCapMapper - named vector that drives conversion from original field name to per capita field name  
  
These default parameters are maintained in a separate .R file and can be sourced:  
```{r}

source("./Coronavirus_CDC_Daily_Default_Mappings_v002.R")

```

## Example for Comparison to Previous
The function is tested on existing, previously downloaded data:  
```{r, fig.height=9, fig.width=9}

readList <- list("cdcDaily"="./RInputFiles/Coronavirus/CDC_dc_downloaded_210801.csv", 
                 "cdcHosp"="./RInputFiles/Coronavirus/CDC_h_downloaded_210801.csv", 
                 "vax"="./RInputFiles/Coronavirus/vaxData_downloaded_210801.csv"
                 )
compareList <- list("cdcDaily"=readFromRDS("cdc_daily_210708")$dfRaw$cdcDaily, 
                    "cdcHosp"=readFromRDS("cdc_daily_210708")$dfRaw$cdcHosp, 
                    "vax"=readFromRDS("cdc_daily_210728_vaxonly")$dfRaw$vax
                    )

cdc_daily_210801_test <- readRunCDCDaily(thruLabel="Jul 31, 2021", 
                                         downloadTo=lapply(readList, FUN=function(x) if(file.exists(x)) NA else x), 
                                         readFrom=readList,
                                         compareFile=compareList, 
                                         writeLog=NULL, 
                                         useClusters=readFromRDS("cdc_daily_210528")$useClusters, 
                                         weightedMeanAggs=c("tcpm7", "tdpm7", "cpm7", "dpm7", "hpm7", 
                                                            "vxcpm7", "vxcgte65pct"
                                                            ),
                                         skipAssessmentPlots=FALSE, 
                                         brewPalette="Paired"
                                         )

all.equal(names(cdc_daily_210801_test), names(readFromRDS("cdc_daily_210801")))
sapply(names(cdc_daily_210801_test), FUN=function(x)
    identical(cdc_daily_210801_test[[x]], readFromRDS("cdc_daily_210801")[[x]])
    )
sapply(names(cdc_daily_210801_test$plotDataList), FUN=function(x)
    identical(cdc_daily_210801_test$plotDataList[[x]], readFromRDS("cdc_daily_210801")$plotDataList[[x]])
    )

```
  
As expected, all data elements are identical.  Plot environments change with each creation, so the plot objects are not identical.

The latest data are downloaded and processed, with caching to avoid multiple file downloads:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

readList <- list("cdcDaily"="./RInputFiles/Coronavirus/CDC_dc_downloaded_210804.csv", 
                 "cdcHosp"="./RInputFiles/Coronavirus/CDC_h_downloaded_210804.csv", 
                 "vax"="./RInputFiles/Coronavirus/vaxData_downloaded_210804.csv"
                 )
compareList <- list("cdcDaily"=readFromRDS("cdc_daily_210708")$dfRaw$cdcDaily, 
                    "cdcHosp"=readFromRDS("cdc_daily_210708")$dfRaw$cdcHosp, 
                    "vax"=readFromRDS("cdc_daily_210728_vaxonly")$dfRaw$vax
                    )

cdc_daily_210804 <- readRunCDCDaily(thruLabel="Aug 3, 2021", 
                                    downloadTo=lapply(readList, FUN=function(x) if(file.exists(x)) NA else x), 
                                    readFrom=readList,
                                    compareFile=compareList, 
                                    writeLog=NULL, 
                                    useClusters=readFromRDS("cdc_daily_210528")$useClusters, 
                                    weightedMeanAggs=c("tcpm7", "tdpm7", "cpm7", "dpm7", "hpm7", 
                                                       "vxcpm7", "vxcgte65pct"
                                                       ),
                                    skipAssessmentPlots=FALSE, 
                                    brewPalette="Paired"
                                    )

saveToRDS(cdc_daily_210804)

```

The raw hospital data is explored for admissions with confirmed/suspected coronavirus:  
```{r, fig.height=9, fig.width=9}

cdcHospAdmit_210804 <- cdc_daily_210804$dfRaw$cdcHosp %>%
    select(state, date, ends_with("confirmed"), ends_with("suspected")) %>%
    filter(state %in% c(state.abb, "DC")) %>%
    arrange(date, state) %>%
    pivot_longer(-c(state, date)) %>%
    mutate(name=stringr::str_replace(name, pattern="previous_day_admission_", replacement=""), 
           name=stringr::str_replace(name, pattern="_covid", replacement="")
           ) %>%
    left_join(getStateData(keepVars=c("state", "pop")), by="state") %>%
    mutate(vpm=1000000*value/pop) %>%
    group_by(state, name) %>%
    mutate(vpm7=zoo::rollmean(vpm, k=7, fill=NA)) %>%
    ungroup()

cdcHospAdmit_210804 %>%
    filter(!is.na(vpm7), state %in% state.abb) %>%
    mutate(div=as.character(state.division)[match(state, state.abb)]) %>%
    group_by(div, date, name) %>%
    summarize(wt_vpm7=sum(pop*vpm7)/sum(pop), tot_pop=sum(pop), .groups="drop") %>%
    ggplot(aes(x=date, y=wt_vpm7)) + 
    geom_col(aes(fill=name), position="stack") + 
    facet_wrap(~div) + 
    scale_fill_discrete("Metric") + 
    labs(x=NULL, 
         y="Newly admitted for COVID per million (rolling 7-day)", 
         title="Hospital admissions for COVID", 
         subtitle="All metrics divided by total population (all ages) for states reporting"
         )

```

Hospital coverage data became robust about a year ago.  The overwhelming majority of admissions are adult, split between confirmed and suspected cases.  Next steps are to explore changes in admissions by age groups:  
```{r, fig.height=9, fig.width=9}

hospAge_210804 <- cdc_daily_210804$dfRaw$cdcHosp %>%
    select(state, 
           date, 
           grep(x=names(.), pattern="ed_\\d.*[9+]$", value=TRUE), 
           grep(x=names(.), pattern="pediatric.*ed$", value=TRUE)
           ) %>% 
    pivot_longer(-c(state, date)) %>% 
    mutate(confSusp=ifelse(grepl(x=name, pattern="confirmed"), "confirmed", "suspected"), 
           adultPed=ifelse(grepl(x=name, pattern="adult"), "adult", "ped"), 
           age=ifelse(adultPed=="ped", "0-17", stringr::str_replace_all(string=name, pattern=".*_", replacement="")), 
           age=ifelse(age %in% c("0-17", "18-19"), "0-19", age), 
           div=as.character(state.division)[match(state, state.abb)]
           ) 
hospAge_210804

# Plot for overall trends by age group
p1 <- hospAge_210804 %>% 
    filter(state %in% c(state.abb, "DC"), !is.na(value)) %>% 
    mutate(ageBucket=age) %>% 
    group_by(date, ageBucket) %>% 
    summarize(value=sum(value), .groups="drop") %>% 
    arrange(date) %>%
    group_by(ageBucket) %>% 
    mutate(value7=zoo::rollmean(value, k=7, fill=NA)) %>% 
    filter(date >= "2020-08-01") %>% 
    ggplot(aes(x=date, y=value7)) + 
    labs(x=NULL, 
         y="Confirmed or suspected COVID admissions (rolling-7 mean)", 
         title="Hospital admissions for COVID by age bucket (Aug 2020 - Jul 2021)", 
         subtitle="50 states and DC (includes confirmed and suspected from CDC data)"
         ) + 
    lims(y=c(0, NA))

# Line plots by age group
p1 + 
    geom_line(aes(group=ageBucket, color=ageBucket), size=1) + 
    scale_color_discrete("Age\nbucket")
# Stacked bar plots by age group
p1 + 
    geom_col(aes(fill=ageBucket), position="stack") + 
    scale_color_discrete("Age\nbucket")
# Proportions by age group
p1 + 
    geom_col(aes(fill=ageBucket), position="fill") + 
    scale_color_discrete("Age\nbucket")

# Plot for overall trends by age group
hospAge_210804 %>% 
    filter(state %in% state.abb, !is.na(value)) %>% 
    mutate(ageBucket=ifelse(age >= "60", "60+", ifelse(age=="0-19", "0-19", "20-59"))) %>% 
    group_by(date, state, ageBucket) %>% 
    summarize(value=sum(value), .groups="drop") %>% 
    group_by(ageBucket, state) %>% 
    mutate(value7=zoo::rollmean(value, k=7, fill=NA)) %>% 
    filter(date >= "2020-06-01") %>% 
    ggplot(aes(x=date, y=value7)) + 
    geom_line(aes(color=ageBucket, group=ageBucket)) + 
    scale_color_discrete("Age\nbucket") + 
    labs(x=NULL, 
         y="Confirmed or suspected COVID admissions (rolling-7 mean)", 
         title="Hospital admissions for COVID by age bucket (Aug 2020 - Jul 2021)"
         ) + 
    lims(y=c(0, NA)) + 
    facet_wrap(~state, scales="free_y")

```

Next steps are to explore alignment of the hospitalization and case/death curves.  Michigan having had a distinct spring peak is used as an example:  
```{r, fig.height=9, fig.width=9}

allHosp_210804 <- hospAge_210804 %>%
    mutate(ageBucket=ifelse(age >= "60", "60+", ifelse(age=="0-19", "0-19", "20-59"))) %>% 
    group_by(date, state, ageBucket) %>% 
    summarize(value=sum(value), .groups="drop") %>% 
    group_by(ageBucket, state) %>% 
    mutate(value7=zoo::rollmean(value, k=7, fill=NA)) %>%
    ungroup() %>%
    left_join(getStateData(keepVars=c("state", "pop"))) %>%
    mutate(vpm7=1000000*value7/pop)
allHosp_210804

allCaseDeath_210804 <- cdc_daily_210804$dfPerCapita %>%
    select(state, date, new_cases, new_deaths, vxa, vxc, cpm7, dpm7, vxapm7, vxcpm7) %>%
    pivot_longer(-c(state, date))
allCaseDeath_210804

allPivot_210804 <- allHosp_210804 %>%
    select(state, date, name=ageBucket, value=vpm7) %>%
    bind_rows(allCaseDeath_210804) %>%
    checkUniqueRows(uniqueBy=c("state", "date", "name"))
allPivot_210804

# Plot Michigan data
typeMapper <- c("cases"="1. Cases per million per day (rolling 7 mean)", 
                "deaths"="2. Deaths per million per day (rolling 7 mean)", 
                "hosp"="3. Admitted to hospital per million per day (rolling 7 mean)", 
                "vax"="4. Vaccinated per capita (administered, completed)"
                )
allPivot_210804 %>%
    filter(state=="MI", !is.na(value)) %>%
    mutate(plotType=case_when(name %in% c("0-19", "20-59", "60+") ~ "hosp", 
                              name %in% c("vxapm7", "vxcpm7") ~ "vax", 
                              name=="cpm7" ~ "cases", 
                              name=="dpm7" ~ "deaths", 
                              TRUE ~ "notuse"
                              )
           ) %>%
    filter(plotType != "notuse") %>%
    ggplot(aes(x=date, y=value)) + 
    geom_line(data=~filter(., plotType=="cases"), color="blue", size=1) +
    geom_line(data=~filter(., plotType=="deaths"), color="red", size=1) +
    geom_line(data=~filter(., plotType=="vax"), aes(color=name, group=name, y=value/1000000)) +
    geom_col(data=~filter(., plotType=="hosp"), aes(fill=name), position="stack") + 
    scale_color_discrete("Vaccine\nMetric") + 
    scale_fill_discrete("Hospitalized\nby Age") +
    facet_wrap(~typeMapper[plotType], scales="free_y") + 
    labs(x=NULL, y=NULL) +
    lims(y=c(0, NA))

hospCase <- 10
allPivot_210804 %>%
    filter(state=="MI", !is.na(value)) %>%
    mutate(plotType=case_when(name %in% c("0-19", "20-59", "60+") ~ "hosp", 
                              name %in% c("vxapm7", "vxcpm7") ~ "vax", 
                              name=="cpm7" ~ "cases", 
                              name=="dpm7" ~ "deaths", 
                              TRUE ~ "notuse"
                              )
           ) %>%
    filter(plotType %in% c("cases", "hosp")) %>%
    ggplot(aes(x=date, y=value)) + 
    geom_col(data=~filter(., plotType=="hosp"), aes(fill=name, y=hospCase*value), position="stack") + 
    geom_line(data=~filter(., plotType=="cases"), color="black", size=1) +
    scale_fill_discrete("Hospital Admssions by Age") + 
    scale_y_continuous("Cases per million (rolling 7 mean per day)", 
                       sec.axis = sec_axis(~ . / hospCase, 
                                           name = "Hospital admissions per million (rolling 7 mean per day"
                                           )
                       ) +
    labs(x=NULL, 
         title="Alignment of Michigan cases and hospitalizations data", 
         subtitle=paste0("Ratio of ", 
                         hospCase, 
                         ":1 applied (cases are black line, hospital admissions are stacked bar)"
                         )
         ) + 
    theme(legend.position="bottom")

hospDeath <- 0.2
allPivot_210804 %>%
    filter(state=="MI", !is.na(value)) %>%
    mutate(plotType=case_when(name %in% c("0-19", "20-59", "60+") ~ "hosp", 
                              name %in% c("vxapm7", "vxcpm7") ~ "vax", 
                              name=="cpm7" ~ "cases", 
                              name=="dpm7" ~ "deaths", 
                              TRUE ~ "notuse"
                              )
           ) %>%
    filter(plotType %in% c("deaths", "hosp")) %>%
    ggplot(aes(x=date, y=value)) + 
    geom_col(data=~filter(., plotType=="hosp"), aes(fill=name, y=hospDeath*value), position="stack") + 
    geom_line(data=~filter(., plotType=="deaths"), color="black", size=1) +
    scale_fill_discrete("Hospital Admssions by Age") + 
    scale_y_continuous("Deaths per million (rolling 7 mean per day)", 
                       sec.axis = sec_axis(~ . / hospDeath, 
                                           name = "Hospital admissions per million (rolling 7 mean per day"
                                           )
                       ) +
    labs(x=NULL, 
         title="Alignment of Michigan deaths and hospitalizations data", 
         subtitle=paste0("Ratio of ", 
                         hospDeath, 
                         ":1 applied (deaths are black line, hospital admissions are stacked bar)"
                         )
         ) + 
    theme(legend.position="bottom")

```

At a glance, the Michigan data appear reasonably well aligned.  Hospital admissions run at ~10% of confirmed cases with a small delay.  Deaths run at ~20% of hospital admissions in the mid-winter wave and ~10% of admissions in the spring wave.  These are consistent with estimates of ~1.5% CFR and ~5 cases per diagnosed cases (fatality rate ~0.3%).

Next steps are to convert this process to functional form, automate selection of scaling parameters, and explore other states of interest:  
```{r}

# Function to create case-hospital-death file
makeCaseHospDeath <- function(dfHosp, dfCaseDeath) {

    # FUNCTION ARGUMENTS:
    # dfHosp: the tibble or data.frame containing the hospital data by date-state
    # dfCaseDeath: the tibble or data.frame containing the case and death data by date-state
    
    allHosp <- dfHosp %>%
        mutate(ageBucket=ifelse(age >= "60", "60+", ifelse(age=="0-19", "0-19", "20-59"))) %>% 
        group_by(date, state, ageBucket) %>% 
        summarize(value=sum(value), .groups="drop") %>% 
        group_by(ageBucket, state) %>% 
        mutate(value7=zoo::rollmean(value, k=7, fill=NA)) %>%
        ungroup() %>%
        left_join(getStateData(keepVars=c("state", "pop"))) %>%
        mutate(vpm7=1000000*value7/pop)

    allCaseDeath <- dfCaseDeath %>%
        select(state, date, new_cases, new_deaths, vxa, vxc, cpm7, dpm7, vxapm7, vxcpm7) %>%
        pivot_longer(-c(state, date))

    allPivot <- allHosp %>%
        select(state, date, name=ageBucket, value=vpm7) %>%
        bind_rows(allCaseDeath) %>%
        checkUniqueRows(uniqueBy=c("state", "date", "name"))
    
    allPivot
    
}

alignCaseHospDeath <- function(dfPivot=NULL, 
                               dfHosp=NULL, 
                               dfCaseDeath=NULL, 
                               typeMapper=c("cases"="1. Cases per million per day (rolling 7 mean)", 
                                            "deaths"="2. Deaths per million per day (rolling 7 mean)", 
                                            "hosp"="3. Admitted to hospital per million per day (rolling 7 mean)", 
                                            "vax"="4. Vaccinated per capita (administered, completed)"
                                            ), 
                               keyState="MI", 
                               hospCaseScalar=10, 
                               hospDeathScalar=0.2, 
                               returnPlots=FALSE
                               ) 
    {
    
    # FUNCTION ARGUMENTS:
    # dfPivot: the tibble or data.frame containing integrated case-hospital-death data 
    #          (if NULL, build from dfHosp and dfCaseDeath)
    # dfHosp: the tibble or data.frame containing the hospital data by date-state
    # dfCaseDeath: the tibble or data.frame containing the case and death data by date-state
    # typeMapper: mapping file for labelling facets
    # keyState: the state to explore
    # hospCaseScalar: the scalar to be applied for placing cases and hospitalizations on the same plot
    # hospDeathScalar: the scalar to be applied for placing hospitalizations and deaths on the same plot
    
    # Create the pivoted data if it was not passed
    if (is.null(dfPivot)) {
        if (is.null(dfHosp) | is.null(dfCaseDeath)) stop("\nMust pass dfPivot OR both of dfHosp and dfCaseDeath\n")
        dfPivot <- makeCaseHospDeath(dfHosp=dfHosp, dfCaseDeath=dfCaseDeath)
    }

    # Create the plotting data
    plotData <- dfPivot %>%
        filter(state==keyState, !is.na(value)) %>%
        mutate(plotType=case_when(name %in% c("0-19", "20-59", "60+") ~ "hosp", 
                                  name %in% c("vxapm7", "vxcpm7") ~ "vax", 
                                  name=="cpm7" ~ "cases", 
                                  name=="dpm7" ~ "deaths", 
                                  TRUE ~ "notuse"
                                  )
               )
    
    # Create overall plot
    p1 <- plotData %>%
        filter(plotType != "notuse") %>%
        ggplot(aes(x=date, y=value)) + 
        geom_line(data=~filter(., plotType=="cases"), color="blue", size=1) +
        geom_line(data=~filter(., plotType=="deaths"), color="red", size=1) +
        geom_line(data=~filter(., plotType=="vax"), aes(color=name, group=name, y=value/1000000)) +
        geom_col(data=~filter(., plotType=="hosp"), aes(fill=name), position="stack") + 
        scale_color_discrete("Vaccine\nMetric") + 
        scale_fill_discrete("Hospitalized\nby Age") +
        facet_wrap(~typeMapper[plotType], scales="free_y") + 
        labs(x=NULL, y=NULL, title=paste0("Key coronavirus metrics for state: ", keyState)) +
        lims(y=c(0, NA))
    print(p1)

    p2 <- plotData %>%
        filter(plotType %in% c("cases", "hosp")) %>%
        ggplot(aes(x=date, y=value)) + 
        geom_col(data=~filter(., plotType=="hosp"), aes(fill=name, y=hospCaseScalar*value), position="stack") + 
        geom_line(data=~filter(., plotType=="cases"), color="black", size=1) +
        scale_fill_discrete("Hospital Admssions by Age") + 
        scale_y_continuous("Cases per million (rolling 7 mean per day)", 
                           sec.axis = sec_axis(~ . / hospCaseScalar, 
                                               name = "Hospital admissions per million (rolling 7 mean per day"
                                               )
                           ) +
        labs(x=NULL, 
             title=paste0("Alignment of ", keyState, " cases and hospitalizations data"), 
             subtitle=paste0("Ratio of ", 
                             hospCaseScalar, 
                             ":1 applied (cases are black line, hospital admissions are stacked bar)"
                             )
             ) + 
        theme(legend.position="bottom")
    print(p2)

    p3 <- plotData %>%
        filter(plotType %in% c("deaths", "hosp")) %>%
        ggplot(aes(x=date, y=value)) + 
        geom_col(data=~filter(., plotType=="hosp"), aes(fill=name, y=hospDeathScalar*value), position="stack") + 
        geom_line(data=~filter(., plotType=="deaths"), color="black", size=1) +
        scale_fill_discrete("Hospital Admssions by Age") + 
        scale_y_continuous("Deaths per million (rolling 7 mean per day)", 
                           sec.axis = sec_axis(~ . / hospDeathScalar, 
                                               name = "Hospital admissions per million (rolling 7 mean per day"
                                               )
                           ) +
        labs(x=NULL, 
             title=paste0("Alignment of ", keyState, " deaths and hospitalizations data"), 
             subtitle=paste0("Ratio of ", 
                             hospDeathScalar, 
                             ":1 applied (deaths are black line, hospital admissions are stacked bar)"
                             )
             ) + 
        theme(legend.position="bottom")
    print(p3)
    
    if(isTRUE(returnPlots)) list(p1=p1, p2=p2, p3=p3)
    
}

```

The function is tested for MI, FL, and TX:  
```{r, fig.height=9, fig.width=9}

dfPivot_210804 <- makeCaseHospDeath(dfHosp=hospAge_210804, dfCaseDeath=cdc_daily_210804$dfPerCapita)
dfPivot_210804

miAlign <- alignCaseHospDeath(dfPivot_210804, keyState="MI", returnPlots=TRUE)
flAlign <- alignCaseHospDeath(dfPivot_210804, keyState="FL", returnPlots=TRUE, hospDeathScalar=0.14)
txAlign <- alignCaseHospDeath(dfPivot_210804, keyState="TX", returnPlots=TRUE, hospDeathScalar=0.14)

# Key plots on a single page
gridExtra::grid.arrange(miAlign$p1, flAlign$p1, txAlign$p1, nrow=2)
gridExtra::grid.arrange(miAlign$p2, flAlign$p2, txAlign$p2, nrow=1)
gridExtra::grid.arrange(miAlign$p3, flAlign$p3, txAlign$p3, nrow=1)

```

Next, an implied CFR (deaths per case) rate is calculated for inclusion:  
```{r, fig.height=9, fig.width=9}

# Filter to single-state data
miData <- dfPivot_210804 %>%
    filter(state=="MI", name %in% c("cpm7", "dpm7"), !is.na(value))

# Plot core metrics
miData %>%
    ggplot(aes(x=date, y=value)) + 
    geom_line(aes(group=name, color=name)) + 
    facet_wrap(~c("cpm7"="Cases per million", "dpm7"="Deaths per million")[name], scales="free_y") + 
    labs(x=NULL, y="Value per million (rolling 7-day mean)", title="Coronavirus burden in state: MI") + 
    theme(legend.position="none")

# Create correlation for assigned lag/lead and variables in a data frame
lagCorrCheck <- function(df, lagLead=0, varFix="dpm7", varMove="cpm7") {
    df %>%
        mutate(lagVar=if(lagLead >= 0) lag(get(varMove), lagLead) else lead(get(varMove), abs(lagLead))) %>%
        filter(!is.na(lagVar)) %>%
        summarize(correl=cor(lagVar, get(varFix))) %>%
        pull(correl)
}

miCorrDF <- miData %>%
    select(date, name, value) %>%
    pivot_wider(date)
miCorrDF

# Assessing best lag/lead for full dataset
lagLeads=-10:40
miRhoFull <- tibble::tibble(lagLead=lagLeads, 
                            rho=sapply(lagLeads, FUN=function(x) lagCorrCheck(miCorrDF, lagLead=x))
                            )
bestFull <- miRhoFull %>% filter(rho==max(rho))
miRhoFull %>% 
    ggplot(aes(x=lagLead, y=rho)) + 
    geom_point() + 
    geom_hline(data=bestFull, aes(yintercept=rho), lty=2) +
    geom_vline(data=bestFull, aes(xintercept=lagLead), lty=2) +
    labs(x="Lag or lead of cases", 
         y="Correlation to deaths", 
         title="All Michigan case and death data", 
         subtitle=paste0("Best correlation ", 
                         round(bestFull$rho, 3), 
                         " obtained at lag/lead of: ", 
                         bestFull$lagLead
                         )
         )

# Assessing best lag/lead for August 2020 and beyond
lagLeads=-10:40
miRhoLate <- tibble::tibble(lagLead=lagLeads, 
                            rho=sapply(lagLeads, 
                                       FUN=function(x) lagCorrCheck(miCorrDF %>% filter(date >= "2020-08-01"), 
                                                                    lagLead=x
                                                                    )
                                       )
                            )
bestLate <- miRhoLate %>% filter(rho==max(rho))
miRhoLate %>% 
    ggplot(aes(x=lagLead, y=rho)) + 
    geom_point() + 
    geom_hline(data=bestLate, aes(yintercept=rho), lty=2) +
    geom_vline(data=bestLate, aes(xintercept=lagLead), lty=2) +
    labs(x="Lag or lead of cases", 
         y="Correlation to deaths", 
         title="Michigan case and death data (August 2020 to present)", 
         subtitle=paste0("Best correlation ", 
                         round(bestLate$rho, 3), 
                         " obtained at lag/lead of: ", 
                         bestLate$lagLead
                         )
         )

miCorrDF %>%
    mutate(lag20=lag(cpm7, 20)) %>%
    filter(!is.na(lag20), lag20 > 0) %>%
    ggplot(aes(x=date)) + 
    geom_line(aes(y=cpm7), color="navy") + 
    geom_line(aes(y=lag20), color="navy", lty=2) + 
    geom_line(aes(y=8000*pmin(0.1, dpm7/lag20)), color="red") + 
    scale_y_continuous("Cases per million (rolling 7 mean per day) - actual and 20-day lag", 
                       sec.axis = sec_axis(~ . / 8000, 
                                           name = "Implied CFR (capped at 10%)"
                                           )
                       ) +
    labs(x=NULL, title="Implied case fatality rate using 20-day lag in Michigan")

miCorrDF %>%
    mutate(lag20=lag(cpm7, 20)) %>%
    filter(!is.na(lag20)) %>%
    ggplot(aes(x=date)) + 
    geom_line(aes(y=cpm7), color="navy") + 
    geom_line(aes(y=lag20), color="navy", lty=2) + 
    geom_line(aes(y=50*dpm7), color="red") + 
    labs(x=NULL, 
         y="Per million (7-day rolling mean)\nCases, Lag-20 cases, and 50*Deaths", 
         title="Burden metrics for Michigan"
         )

```
  
Michigan appears to fairly consistently have a lag of 2-3 weeks between cases and deaths.  The CFR during spikes has consistently declined from over 10% to around 2% to around 1%.  The estimated CFR outside of spikes appears noisy.  Next steps are to convert to functional form and run for other states:  
```{r, fig.height=9, fig.width=9}

# Create correlation for assigned lag/lead and variables in a data frame
lagCorrCheck <- function(df, lagLead=0, varFix="dpm7", varMove="cpm7") {
    df %>%
        mutate(lagVar=if(lagLead >= 0) lag(get(varMove), lagLead) else lead(get(varMove), abs(lagLead))) %>%
        filter(!is.na(lagVar)) %>%
        summarize(correl=cor(lagVar, get(varFix))) %>%
        pull(correl)
}


findCorrAlign <- function(df, 
                          keyState, 
                          varFix="dpm7", 
                          varMove="cpm7", 
                          lagLeads=-10:40,
                          minDate=NULL, 
                          maxDate=NULL,
                          varMapper=c("cpm7"="Cases per million", "dpm7"="Deaths per million"), 
                          yLab="Value per million (rolling 7-day mean)", 
                          printPlots=TRUE, 
                          returnPlots=FALSE,
                          returnData=FALSE
                          ) {
    
    # FUNCTION ARGUMENTS
    # df: pivoted data frame with state-date-name-value
    # keyState: state to include
    # varFix: metric to be held constant
    # varMove: metric to be lagged/led
    # lagLeads: lags and leads for the variable that moves
    # minDate: minimum date for lag/lead (NULL means data-driven)
    # maxDate: maximum date for lag/lead (NULL means data-driven)
    # varMapper: mapping file for varFix and varMove to descriptive labels
    # yLab: label for the y-axis in the first plot
    # printPlots: boolean, should the plots be printed?
    # returnPlots: boolean, should the plots be returned?
    # returnData: boolean, should the data frames be returned as a list?
    
    # Set minDate and maxDate to the actual minmax if passed as NULL 
    if (is.null(minDate)) minDate <- df %>% summarize(date=min(date)) %>% pull(date)
    if (is.null(maxDate)) maxDate <- df %>% summarize(date=max(date)) %>% pull(date)
    
    # Filter to relevant data
    df <- df %>%
        filter(state %in% all_of(keyState), name %in% all_of(c(varFix, varMove)), !is.na(value))

    # Plot core metrics for requested states
    p1 <- df %>%
        ggplot(aes(x=date, y=value)) + 
        geom_line(aes(group=name, color=name)) + 
        facet_wrap(~varMapper[name], scales="free_y") + 
        labs(x=NULL, y=yLab, title=paste0("Metrics by state for: ", paste0(keyState, collapse=", "))) + 
        theme(legend.position="none")
    if(isTRUE(printPlots)) print(p1)

    # Create dataset for correlations
    dfCorr <- df %>%
        select(date, name, value) %>%
        pivot_wider(date)

    # Find correlation by lag/lead for dataset
    dfRho <- tibble::tibble(lagLead=lagLeads, 
                            rho=sapply(lagLeads, 
                                       FUN=function(x) {
                                           lagCorrCheck(dfCorr %>% filter(date >= minDate, date <= maxDate), 
                                                        lagLead=x
                                                        )
                                           }
                                       )
                            )
    
    # Find best correlation and lag/lead
    bestRho <- dfRho %>% 
        filter(rho==max(rho))

    # Plot correlations by lag/lead
    p2 <- dfRho %>% 
        ggplot(aes(x=lagLead, y=rho)) + 
        geom_point() + 
        geom_hline(data=bestRho, aes(yintercept=rho), lty=2) +
        geom_vline(data=bestRho, aes(xintercept=lagLead), lty=2) +
        labs(x=paste0("Lag or lead of ", varMapper[varMove]), 
             y=paste0("Correlation to ", varMapper[varFix]), 
             title=paste0("Correlations by lag/lead for state: ", keyState), 
             subtitle=paste0("Best correlation ", 
                             round(bestRho$rho, 3), 
                             " obtained at lag/lead of: ", 
                             bestRho$lagLead
                             )
             )
    if(isTRUE(printPlots)) print(p2)

    if (isTRUE(returnData) | isTRUE(returnPlots)) {
        list(dfRho=if(isTRUE(returnData)) dfRho else NULL, 
             bestRho=if(isTRUE(returnData)) bestRho else NULL, 
             dfCorr=if(isTRUE(returnData)) dfCorr else NULL, 
             p1=if(isTRUE(returnPlots)) p1 else NULL, 
             p2=if(isTRUE(returnPlots)) p2 else NULL
             )
    }
    
}

findCorrAlign(dfPivot_210804, keyState="MI")
findCorrAlign(dfPivot_210804, keyState="MI", minDate="2020-08-01", returnData=TRUE)
findCorrAlign(dfPivot_210804, keyState="FL")
findCorrAlign(dfPivot_210804, keyState="TX", minDate="2020-10-01")

```

Next steps are to create functional form for calculating IFR and aligning plots:  
```{r, fig.height=9, fig.width=9}

plotCFRLag <- function(lst, 
                       lagUse=NULL, 
                       scaleUse=NULL, 
                       cfrCap=0.06, 
                       multDeath=50,
                       mainTitle="Coronavirus data for selected geography: ", 
                       printPlots=TRUE, 
                       returnPlots=FALSE
                       ) {
    
    # FUNCTION ARGUMENTS:
    # lst: data frame with date-cpm7-dpm7 OR list with both dfCorr and bestRho
    # lagUse: the lag to use (if NULL, use the value in bestRho$lagLead)
    # scaleUse: scalar for secondary y-axis (NULL means calculate from data)
    # cfrCap: the cap for all values of CFR
    # multDeath: multiplier for death data in plot 2
    # mainTitle: main title for plots
    # printPlots: boolean, should the plots be printed?
    # returnPlots: boolean, should the plots be returned?
    
    # Create dfCorr and lagUse
    if ("list" %in% class(lst)) {
        dfCorr <- lst[["dfCorr"]]
        if (is.null(lagUse)) lagUse <- lst[["bestRho"]]$lagLead
    } else {
        dfCorr <- lst
    }
    
    # Check that dfCorr is a data frame with date-cpm7-dpm7 and lagUse is not NULL
    if (!("data.frame" %in% class(dfCorr))) stop("\nMust have a data frame for lst/dfCorr\n")
    if (!(all(c("date", "cpm7", "dpm7") %in% names(dfCorr)))) stop("\ndfCorr must have date-cpm7-dpm7\n")
    if (is.null(lagUse)) stop("\nMust have a value for lagUse\n")
    
    # Create scaleUse if not passed
    if (is.null(scaleUse)) scaleUse <- 500*ceiling(max(dfCorr$cpm7)/cfrCap/500)
    
    # Create plot of CFR by date, showing lagged cases
    basePlot <- dfCorr %>%
        mutate(lagData=if(lagUse >= 0) lag(cpm7, lagUse) else lead(cpm7, -lagUse)) %>%
        filter(!is.na(lagData), lagData > 0) %>%
        ggplot(aes(x=date)) + 
        geom_line(aes(y=cpm7), color="navy") + 
        geom_line(aes(y=lagData), color="navy", lty=2)
    p1 <- basePlot + 
        geom_line(aes(y=scaleUse*pmin(cfrCap, dpm7/lagData)), color="red") + 
        scale_y_continuous(paste0("Cases per million\n(rolling 7-day mean)"), 
                           sec.axis = sec_axis(~ . / scaleUse, 
                                               name = paste0("Implied CFR (capped at ", 
                                                             round(100*cfrCap, 1), 
                                                             "%)"
                                                             )
                                               )
                           ) +
        labs(x=NULL, 
             title=mainTitle, 
             subtitle=paste0("Red line (right axis) is implied fatality rate\n", 
                             "Blue line is cases with and without ", 
                             abs(lagUse), 
                             "-day ", 
                             if(lagUse > 0) "lag" else "lead"
                             )
             )
    if (isTRUE(printPlots)) print(p1)
    
    # Apply a CFR to the data and show alignment
    p2 <- basePlot +
        geom_line(aes(y=multDeath*dpm7), color="red") + 
        labs(x=NULL, 
             y="Per million (7-day rolling mean)", 
             title=mainTitle, 
             subtitle=paste0("Red line is ", 
                             multDeath, 
                             "*deaths\n", 
                             "Blue line is cases with and without ", 
                             abs(lagUse), 
                             "-day ", 
                             if(lagUse > 0) "lag" else "lead"
                             )
         )
    if (isTRUE(printPlots))print(p2)
    
    if (isTRUE(returnPlots)) list(p1=p1, p2=p2)
    
}

# Run for Michigan, Florida, California
findCorrAlign(dfPivot_210804, keyState="MI", minDate="2020-08-01", returnData=TRUE) %>%
    plotCFRLag(mainTitle=paste0(formals(plotCFRLag)$mainTitle, "MI"))
findCorrAlign(dfPivot_210804, keyState="FL", returnData=TRUE) %>%
    plotCFRLag(mainTitle=paste0(formals(plotCFRLag)$mainTitle, "FL"), multDeath=100)
findCorrAlign(dfPivot_210804, keyState="CA", returnData=TRUE) %>%
    plotCFRLag(mainTitle=paste0(formals(plotCFRLag)$mainTitle, "CA"), multDeath=70)

```

Functions have been updated so that plots can be returned and/or printed.  A new function allows for creating all plots for state on a single page:  
```{r, fig.height=9, fig.width=9}

# Function to plot all states on the same page
onePageCFRPlot <- function(df, keyState, multDeath=100, cfrCap=0.06, ...) {
    
    # FUNCTION ARGUMENTS:
    # df: the data frame containing state-date-name-value
    # keyState: the key state to be analyzed
    # multDeath: multiplier for death in the death/lagged cases chart of plotCFRLag()
    # ...: other arguments to be passed to findCorrAlign()
    
    # Find the correlations data
    corrData <- findCorrAlign(df, 
                              keyState=keyState, 
                              yLab="Value per million\n(rolling 7-day mean)", 
                              printPlots=FALSE, 
                              returnPlots=TRUE, 
                              returnData=TRUE, 
                              ...
                              )
    
    # Find CFR
    cfrData <- plotCFRLag(corrData, 
                          cfrCap=cfrCap, 
                          multDeath=multDeath, 
                          mainTitle=paste0(formals(plotCFRLag)$mainTitle, keyState), 
                          printPlots=FALSE, 
                          returnPlots=TRUE
                          )
    
    # Create single-page summary
    gridExtra::grid.arrange(corrData$p1, corrData$p2, cfrData$p1, cfrData$p2, nrow=2)
    
}

# Run for Michigan, Florida, California, Texas, New York, South Dakota
onePageCFRPlot(dfPivot_210804, keyState="MI", minDate="2020-08-01")
onePageCFRPlot(dfPivot_210804, keyState="FL", multDeath=100)
onePageCFRPlot(dfPivot_210804, keyState="CA", multDeath=70)
onePageCFRPlot(dfPivot_210804, keyState="TX", multDeath=70)
onePageCFRPlot(dfPivot_210804, keyState="NY", multDeath=10, cfrCap=0.1)
onePageCFRPlot(dfPivot_210804, keyState="SD", multDeath=70)

```

Summaries are created for vaccination status by age cohort and state:  
```{r, fig.height=9, fig.width=9}

cdc_daily_210815 <- readFromRDS("cdc_daily_210815")

# Example using hard coding and wide data
cdc_daily_210815$dfRaw$vax %>% 
    filter(date==max(date), state %in% c(state.abb, "DC")) %>% 
    select(state, contains("Administered_Dose1")) %>% 
    ggplot(aes(x=fct_reorder(state, Administered_Dose1_Recip_65PlusPop_Pct))) + 
    geom_col(aes(fill="65+", y=Administered_Dose1_Recip_65PlusPop_Pct)) + 
    geom_col(aes(y=Administered_Dose1_Recip_18PlusPop_Pct, fill="18+")) + 
    geom_col(aes(y=Administered_Dose1_Pop_Pct, fill="All")) + 
    geom_text(aes(y=Administered_Dose1_Recip_65PlusPop_Pct, 
                  label=paste0(Administered_Dose1_Recip_65PlusPop_Pct, "% (", state, ")")
                  ), 
              hjust=0, 
              size=3
              ) + 
    geom_text(aes(y=Administered_Dose1_Recip_18PlusPop_Pct+0.5, 
                  label=paste0(Administered_Dose1_Recip_18PlusPop_Pct, "%")
                  ), 
              hjust=0, 
              size=3
              ) + 
    coord_flip() + 
    labs(x=NULL, 
         y="% Fully Vaccinated", 
         title="First-dose vaccinated by age cohort and state (mid-Aug 2020)", 
         subtitle="Black: all population, Yellow: 18+, Blue: 65+"
         ) + 
    scale_fill_manual("Cohort", 
                      breaks=c("All", "18+", "65+"), 
                      values=c("65+"="lightblue", "18+"="yellow", "All"="black" )
                      ) + 
    geom_text(aes(y=Administered_Dose1_Pop_Pct+0.5, 
                  label=paste0(Administered_Dose1_Pop_Pct, "%")
                  ), 
              hjust=0, 
              size=3
              ) + 
    theme(legend.position="bottom")



# Function for reproducibility
tempStackPlot <- function(df, 
                          yVars, 
                          xVar="state", 
                          yLab="", 
                          plotTitle="", 
                          colorVector=c("lightblue", "grey", "orange", "black")
                          ) {

    # FUNCTION ARGUMENTS:
    # df: data frame or tibble
    # yVars: named vector with c("variable"="name")
    # xVar: the x variable
    # yLab: the y-axis label for the plot
    # plotTitle: the title for the plot
    # colorVector: colors to use for filled bars (sequentially, can have more, but not less, than length(yVars))
    
    colorVector <- colorVector[1:length(yVars)]
    names(colorVector) <- names(yVars)
    
    p1 <- df %>% 
        select(all_of(xVar), all_of(names(yVars))) %>%
        pivot_longer(-c(all_of(xVar))) %>%
        ggplot(aes(x=fct_reorder(get(xVar[1]), value, max))) + 
        coord_flip() + 
        labs(x=NULL, y=yLab, title=plotTitle) + 
        geom_col(aes(y=value, fill=name), position="identity") + 
        geom_text(aes(y=value+0.5, 
                      label=paste0(value, "%", ifelse(name==names(yVars)[1], paste0(" (", state, ")"), ""))
                      ), 
                  hjust=0, 
                  size=3
                  ) + 
        scale_fill_manual("Cohort", 
                          breaks=rev(names(yVars)), 
                          labels=rev(unname(yVars)),
                          values=colorVector
                          ) +
        theme(legend.position="bottom")

    p1
    
}

# Run for fully vaccinated
tempStackPlot(cdc_daily_210815$dfRaw$vax %>% filter(date==max(date), state %in% c(state.abb, "DC")), 
              yVars=c("vxcgte65pct"="65+", 
                      "vxcgte18pct"="18+", 
                      "vxcpoppct"="All"
                      ), 
              yLab="% Fully vaccinated", 
              plotTitle="Fully vaccinated by age cohort and state (mid-Aug 2020)"
              )

# Run for first dose
tempStackPlot(cdc_daily_210815$dfRaw$vax %>% filter(date==max(date), state %in% c(state.abb, "DC")), 
              yVars=c("Administered_Dose1_Recip_65PlusPop_Pct"="65+", 
                      "Administered_Dose1_Recip_18PlusPop_Pct"="18+", 
                      "Administered_Dose1_Pop_Pct"="All"
                      ), 
              yLab="% Receiving First Dose", 
              plotTitle="First-dose vaccinated by age cohort and state (mid-Aug 2020)"
              )

```

The plotting is extended to show the evolution of vaccination over time:  
```{r, fig.height=9, fig.width=9}

# Create data
cdcVaxGrowth <- cdc_daily_210815$dfRaw$vax %>% 
    filter(date %in% c(as.Date(max(date)-lubridate::dmonths(c(0, 2, 4)), origin="1970-01-01")), 
           state %in% c(state.abb, "DC")
           )

# Run for fuly vaccinated
p1 <- cdcVaxGrowth %>%
    select(state, date, vxcpoppct) %>%
    pivot_wider(state, names_from="date", values_from="vxcpoppct") %>%
    tempStackPlot(yVars=c("2021-08-15"="2021-08-15", 
                          "2021-06-15"="2021-06-15", 
                          "2021-04-15"="2021-04-15"
                          ), 
                  yLab="% Fully Vaccinated (all population)", 
                  plotTitle="Evolution of fully vaccinated rate by state"
                  )

p2 <- cdcVaxGrowth %>%
    select(state, date, vxcgte65pct) %>%
    pivot_wider(state, names_from="date", values_from="vxcgte65pct") %>%
    tempStackPlot(yVars=c("2021-08-15"="2021-08-15", 
                          "2021-06-15"="2021-06-15", 
                          "2021-04-15"="2021-04-15"
                          ), 
                  yLab="% Fully Vaccinated (65+)", 
                  plotTitle="Evolution of fully vaccinated rate by state"
                  )

gridExtra::grid.arrange(p1, p2, nrow=1)



# Run for first dose
p1 <- cdcVaxGrowth %>%
    select(state, date, Administered_Dose1_Pop_Pct) %>%
    pivot_wider(state, names_from="date", values_from="Administered_Dose1_Pop_Pct") %>%
    tempStackPlot(yVars=c("2021-08-15"="2021-08-15", 
                          "2021-06-15"="2021-06-15", 
                          "2021-04-15"="2021-04-15"
                          ), 
                  yLab="% First-dose (all population)", 
                  plotTitle="Evolution of first dose rate by state"
                  )

p2 <- cdcVaxGrowth %>%
    select(state, date, Administered_Dose1_Recip_65PlusPop_Pct) %>%
    pivot_wider(state, names_from="date", values_from="Administered_Dose1_Recip_65PlusPop_Pct") %>%
    tempStackPlot(yVars=c("2021-08-15"="2021-08-15", 
                          "2021-06-15"="2021-06-15", 
                          "2021-04-15"="2021-04-15"
                          ), 
                  yLab="% First-dose (65+)", 
                  plotTitle="Evolution of first dose rate by state"
                  )

gridExtra::grid.arrange(p1, p2, nrow=1)

```

Similar plots are created for the evolution of cases per thousand and deaths per milion:  
```{r, fig.height=9, fig.width=9}

# Updated function for reproducibility
tempStackPlot <- function(df, 
                          yVars, 
                          xVar="state", 
                          yLab=NULL, 
                          plotTitle=NULL, 
                          colorVector=c("lightblue", "grey", "orange", "black"), 
                          addSuffix="%", 
                          scaleName="Cohort", 
                          textBuffer=0.5
                          ) {

    # FUNCTION ARGUMENTS:
    # df: data frame or tibble
    # yVars: named vector with c("variable"="name"), in the desired order from right-most to left-most
    # xVar: the x variable
    # yLab: the y-axis label for the plot
    # plotTitle: the title for the plot
    # colorVector: colors to use for filled bars (sequentially, can have more, but not less, than length(yVars))
    # addSuffix: value to be appended to all values in plots (e.g., 96 would show as 96% in the text label)
    # scaleName: the name to use for the legend
    # textBuffer: distance from bar to text
    
    colorVector <- colorVector[1:length(yVars)]
    names(colorVector) <- names(yVars)
    
    p1 <- df %>% 
        select(all_of(xVar), all_of(names(yVars))) %>%
        pivot_longer(-c(all_of(xVar))) %>%
        ggplot(aes(x=fct_reorder(get(xVar[1]), value, max))) + 
        coord_flip() + 
        labs(x=NULL, y=yLab, title=plotTitle) + 
        geom_col(aes(y=value, fill=name), position="identity") + 
        geom_text(aes(y=value+textBuffer, 
                      label=paste0(value, 
                                   addSuffix, 
                                   ifelse(name==names(yVars)[1], paste0(" (", get(xVar[1]), ")"), "")
                                   )
                      ), 
                  hjust=0, 
                  size=3
                  ) + 
        scale_fill_manual(scaleName, 
                          breaks=rev(names(yVars)), 
                          labels=rev(unname(yVars)),
                          values=colorVector
                          ) +
        theme(legend.position="bottom")

    p1
    
}


# Create data
cdcBurdenGrowth <- cdc_daily_210815$dfPerCapita %>% 
    filter(date %in% c(as.Date(max(date)-2-lubridate::dmonths(c(0, 6, 12)), origin="1970-01-01")), 
           state %in% c(state.abb, "DC")
           )

# Run for cases
p1 <- cdcBurdenGrowth %>%
    select(state, date, tcpm) %>% 
    mutate(tcpm=round(tcpm/1000)) %>%
    pivot_wider(state, names_from="date", values_from="tcpm") %>%
    tempStackPlot(yVars=c("2021-08-13"="2021-08-13", 
                          "2021-02-11"="2021-02-11", 
                          "2020-08-12"="2020-08-12"
                          ), 
                  yLab="Cumulative cases per thousand", 
                  plotTitle="Evolution of cumulative cases per thousand by state", 
                  addSuffix="",
                  scaleName="Date"
                  )

# Run for deaths
p2 <- cdcBurdenGrowth %>%
    select(state, date, tdpm) %>% 
    mutate(tdpm=round(tdpm)) %>%
    pivot_wider(state, names_from="date", values_from="tdpm") %>%
    tempStackPlot(yVars=c("2021-08-13"="2021-08-13", 
                          "2021-02-11"="2021-02-11", 
                          "2020-08-12"="2020-08-12"
                          ), 
                  yLab="Cumulative deaths per million", 
                  plotTitle="Evolution of cumulative deaths per million by state", 
                  addSuffix="",
                  scaleName="Date"
                  )

gridExtra::grid.arrange(p1, p2, nrow=1)

```

