---
title: "Coronavirus US - CDC"
author: "davegoblue"
date: "10/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
## Background  
This is a subset of the code contained in Coronavirus_Statistics_v002.Rmd.  This file includes the latest code for analyzing all-cause death data from [CDC Weekly Deaths by Jurisdiction](https://catalog.data.gov/dataset/weekly-counts-of-deaths-by-jurisdiction-and-age-group).  CDC maintains data on deaths by week, age cohort, and state in the US.  Downloaded data are unique by state, epidemiological week, year, age, and type (actual vs. predicted/projected).

These data are known to have a lag between death and reporting, and the CDC back-correct to report deaths at the time the death occurred even if the death is reported in following weeks.  This means totals for recent weeks tend to run low (lag), and the CDC run a projection of the expected total number of deaths given the historical lag times.  Per other analysts on the internet, there is currently significant supra-lag, with lag times much longer than historical averages causing CDC projected deaths for recent weeks to be low.

Functions are available in Coronavirus_Statistics_Functions_CDC_v003.R and Coronavirus_Statistics_Functions_Shared_v003.R.  The code leverages tidyverse and a variable mapping file throughout:  
```{r}

# All functions assume that tidyverse and its components are loaded and available
# Other functions are declared in the sourcing files or use library::function()
library(tidyverse)

# If the same function is in both files, use the version from the more specific source
source("./Coronavirus_Statistics_Functions_Shared_v003.R")
source("./Coronavirus_Statistics_Functions_CDC_v003.R")

# Create a variable mapping file
varMapper <- c("cases"="Cases", 
               "newCases"="Increase in cases, most recent 30 days",
               "casesroll7"="Rolling 7-day mean cases", 
               "deaths"="Deaths", 
               "newDeaths"="Increase in deaths, most recent 30 days",
               "deathsroll7"="Rolling 7-day mean deaths", 
               "cpm"="Cases per million",
               "cpm7"="Cases per day (7-day rolling mean) per million", 
               "newcpm"="Increase in cases, most recent 30 days, per million",
               "dpm"="Deaths per million", 
               "dpm7"="Deaths per day (7-day rolling mean) per million", 
               "newdpm"="Increase in deaths, most recent 30 days, per million", 
               "hpm7"="Currently Hospitalized per million (7-day rolling mean)", 
               "tpm"="Tests per million", 
               "tpm7"="Tests per million per day (7-day rolling mean)"
               )

```
  
## Running Code  
The main function is readRunCDCAllCause(), which performs multiple tasks:  
  
STEP 0: Optionally, downloads the latest data file from CDC
STEP 1: Reads and processes a data file has been downloaded from CDC to local  
STEP 2: Extract relevant data from a processed state-level COVID Tracking Project list  
STEP 3: Basic plots of the CDC data  
STEP 4: Basic excess-deaths analysis  
STEP 5: Create cluster-level aggregate plots  
STEP 6: Create state-level aggregate plots  
STEP 7: Create age-cohort aggregate plots  
STEP 8: Returns a list of key data frames, modeling objects, named cluster vectors, etc.  
  
The main use case is to run the full CDC data download and analysis.  An example is created, with the caveat that data are not repeatedly downloaded (process is cached) to avoid unnecessary calls to the CDC server.

Further, files can be saved in RDS format so they can be loaded and used later.

#### _Use Case: Download new data, analyze all-cause excess deaths_  
The full process downloads data, and analyzes all-cause excess deaths:  
```{r cache=TRUE}

# Use data that have previously been downloaded
cdcLoc <- "Weekly_counts_of_deaths_by_jurisdiction_and_age_group_downloaded_20201027.csv"
cdcList_20201027 <- readRunCDCAllCause(loc=cdcLoc, 
                                       startYear=2015, 
                                       curYear=2020,
                                       weekThru=35, 
                                       startWeek=9, 
                                       lst=readFromRDS("test_hier5_201025"), 
                                       epiMap=readFromRDS("epiMonth"), 
                                       agePopData=readFromRDS("usPopBucket2020"), 
                                       cvDeathThru="2020-08-29", 
                                       cdcPlotStartWeek=10, 
                                       dlData=FALSE
                                       )

```
  
#### _Saving key files as RDS_  
```{r cache=TRUE}

saveToRDS(cdcList_20201027, ovrWriteError=FALSE)

```

