---
title: "Coronavirus US - USA Facts"
author: "davegoblue"
date: "2023-02-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
This module builds on code contained in Coronavirus_Statistics_USAF_v007.Rmd.  This file includes the latest code for analyzing data from [USA Facts](https://usafacts.org/visualizations/coronavirus-covid-19-spread-map/).  USA Facts maintains data on cases and deaths by county for coronavirus in the US.  Downloaded data are unique by county with date as a column and a separate file for each of cases, deaths, and population.

The intent of this code is to move updated functions to sourcing files and to better manage memory.

## Sourcing Functions  
The tidyverse library is loaded, and the functions used for CDC daily processing are sourced.  Additionally, specific functions for USA Facts are also sourced:  
```{r}

library(tidyverse)

# Functions are available in source file
source("./Generic_Added_Utility_Functions_202105_v001.R")
source("./Coronavirus_CDC_Daily_Functions_v002.R")
source("./Coronavirus_USAF_Functions_v002.R")

```
  
Further, the mapping file specific to USA Facts is sourced:  
```{r}

# Updated to handle length-zero inputs - also in CDC Daily v005
# Generic function to rename columns in a file using an input vector
colRenamer <- function(df, 
                       vecRename=c(), 
                       ...
                       ) {
    
    # FUNCTION ARGUMENTS:
    # df: the data frame or tibble
    # vecRename: vector for renaming c('existing name'='new name'), can be any length from 0 to ncol(df)
    # ...: additional arguments to be passed to rename_with
    
    # Rename the columns as requested
    if(length(vecRename)>0) dplyr::rename_with(df, .fn=function(x) vecRename[x], .cols=names(vecRename), ...)
    else df
    
}

source("./Coronavirus_USAF_Default_Mappings_v002.R")

```
  
## Data Updates  
The latest county-level burden data are downloaded:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

readList <- list("usafCase"="./RInputFiles/Coronavirus/covid_confirmed_usafacts_downloaded_20230208.csv", 
                 "usafDeath"="./RInputFiles/Coronavirus/covid_deaths_usafacts_downloaded_20230208.csv"
                 )
compareList <- list("usafCase"=readFromRDS("cty_newdata_20230108")$dfRaw$usafCase, 
                    "usafDeath"=readFromRDS("cty_newdata_20230108")$dfRaw$usafDeath
                    )

# Use existing clusters
cty_newdata_20230208 <- readRunUSAFacts(maxDate="2023-02-06", 
                                        downloadTo=lapply(readList, 
                                                          FUN=function(x) if(file.exists(x)) NA else x
                                                          ),
                                        readFrom=readList, 
                                        compareFile=compareList, 
                                        writeLog="./RInputFiles/Coronavirus/USAF_NewData_20230208_chk_v005.log", 
                                        ovrwriteLog=TRUE,
                                        useClusters=readFromRDS("cty_newdata_20210813")$useClusters,
                                        skipAssessmentPlots=FALSE,
                                        brewPalette="Paired"
                                        )

# Plot all counties based on closest cluster
sparseCountyClusterMap(cty_newdata_20230208$useClusters, 
                       caption="Includes only counties with 25k+ population",
                       brewPalette="viridis"
                       )

# Save the refreshed file
saveToRDS(cty_newdata_20230208, ovrWriteError=FALSE)

```
  
Vaccines data are also updated:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

cty_vaxdata_20230209 <- processCountyVaccines(loc="./RInputFiles/Coronavirus/county_vaccine_20230209.csv", 
                                              ctyList=readFromRDS("cty_newdata_20230208"), 
                                              minDateCD=c("2022-06-09", "2022-06-09"),
                                              maxDateCD="2023-01-26"
                                              )

# Save the refreshed file
saveToRDS(cty_vaxdata_20230209, ovrWriteError=FALSE)

```
  
County-level data are post-processed:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

cty_postdata_20230208 <- postProcessCountyData(lstCtyBurden=cty_newdata_20230208$dfPerCapita, 
                                               lstCtyVax=cty_vaxdata_20230209$vaxFix, 
                                               lstState=readFromRDS("cdc_daily_230202")$dfPerCapita, 
                                               excludeStates="AK"
                                               )

```
  
Additional post-processing steps are run:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

# Step 1a: Burden comparisons for aggregated states
additionalCountyPostProcess(cty_postdata_20230208, p1CompareStates=c(state.abb, "DC"), p1AggData=TRUE)

# Step 1: Burden aggregation for key states
# Step 2: vaccine comparisons
# Step 3: Scoring updates (and errors)
# Step 4: New rolling data (28-day default with ceilings 50000 CPM, 500 DPM)
additionalCountyPostProcess(cty_postdata_20230208, 
                            p1CompareStates=c("GA", "FL", "NE", "IL", "OR"), 
                            p2VaxStates=c("MA", "HI", "VA", "VT", "RI", "NE"), 
                            p3VaxTimes=sort(c("2022-01-01", "2023-01-25")),
                            p4DF=cty_newdata_20230208$dfPerCapita, 
                            excludeStates=c("AK")
                            )

```
  
Memory is cleaned:  
```{r}

# List of files
sapply(ls(), FUN=function(x) object.size(get(x))) %>% sort(decreasing=FALSE)

# Clean large objects
largeObjs <- c("cty_newdata_20230208", "cty_vaxdata_20230209")
cleanMem(largeObjs, delObjs=TRUE)

# List of files
sapply(ls(), FUN=function(x) object.size(get(x))) %>% sort(decreasing=FALSE)

```
  
The latest county-level burden data are downloaded:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

readList <- list("usafCase"="./RInputFiles/Coronavirus/covid_confirmed_usafacts_downloaded_20230308.csv", 
                 "usafDeath"="./RInputFiles/Coronavirus/covid_deaths_usafacts_downloaded_20230308.csv"
                 )
compareList <- list("usafCase"=readFromRDS("cty_newdata_20230208")$dfRaw$usafCase, 
                    "usafDeath"=readFromRDS("cty_newdata_20230208")$dfRaw$usafDeath
                    )

# Use existing clusters
cty_newdata_20230308 <- readRunUSAFacts(maxDate="2023-03-06", 
                                        downloadTo=lapply(readList, 
                                                          FUN=function(x) if(file.exists(x)) NA else x
                                                          ),
                                        readFrom=readList, 
                                        compareFile=compareList, 
                                        writeLog="./RInputFiles/Coronavirus/USAF_NewData_20230308_chk_v005.log", 
                                        ovrwriteLog=TRUE,
                                        useClusters=readFromRDS("cty_newdata_20210813")$useClusters,
                                        skipAssessmentPlots=FALSE,
                                        brewPalette="Paired"
                                        )

# Plot all counties based on closest cluster
sparseCountyClusterMap(cty_newdata_20230308$useClusters, 
                       caption="Includes only counties with 25k+ population",
                       brewPalette="viridis"
                       )

# Save the refreshed file
saveToRDS(cty_newdata_20230308, ovrWriteError=FALSE)

```
  
Data on USA Facts have not been updated since February 5, so this process may have run its course