---
title: "CDC Daily by State"
author: "davegoblue"
date: "8/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
This file is designed to use CDC data to assess coronavirus disease burden by state, including creating and analyzing state-level clusters.

Through March 7, 2021, [The COVID Tracking Project](https://covidtracking.com/) collected and integrated data on tests, cases, hospitalizations, deaths, and the like by state and date.  The latest code for using this data is available in Coronavirus_Statistics_CTP_v004.Rmd.

The COVID Tracking Project suggest that [US federal data sources](https://covidtracking.com/analysis-updates/federal-covid-data-101-how-to-find-data) are now sufficiently robust to be used for analyses that previously relied on COVID Tracking Project.  This code is an attempt to update modules in Coronavirus_Statistics_CTP_v004.Rmd to leverage US federal data.

The code in this module builds on code available in _v002 to include vaccines data:  
  
* Generic_Added_Utility_Functions_202105_v001.R - generic functions that can be used in other areas  
* Coronavirus_CDC_Daily_Functions_v001.R - functions specific to coronavirus daily data  
  
Broadly, the CDC data analyzed by this module includes:  
  
* CDC case and death data by state and date are available for download on the [CDC website](https://data.cdc.gov/api/views/9mfq-cb36/rows.csv?accessType=DOWNLOAD)  
* CDC hospital data are available for download on the [healthdata.gov website](https://beta.healthdata.gov/api/views/g62h-syeh/rows.csv?accessType=DOWNLOAD)  
* CDC vaccines data are also available for download on the [CDC website](https://data.cdc.gov/api/views/unsk-b7fc/rows.csv?accessType=DOWNLOAD):  
  
## Functions and Mapping Files
The tidyverse package is loaded and functions are sourced:  
```{r}

# The tidyverse functions are routinely used without package::function format
library(tidyverse)

# Functions are available in source file
source("./Generic_Added_Utility_Functions_202105_v001.R")
source("./Coronavirus_CDC_Daily_Functions_v001.R")

```

A series of mapping files are also available to allow for parameterized processing.  Mappings include:  
  
* urlMapper - mapping file for urlType and url location to download data  
* renMapper - mapping file for renaming of variables in the raw data file  
* selfListMapper - mapping file for transformations by variable type  
* fullListMapper - mapping file for transformations across variable types  
* lstComboMapper - mapping file for elements to be combined by data type (most common is to combine NYC and NYS data to NY if the file provides them separately)  
* uqMapper - mapping file for fields that should combine to be unique keys for processed files  
* lstFilterMapper - mapping file for filtering to subset (most common is to keep 50 states and DC)  
* vecSelectMapper - mapping file for variables to keep  
* checkControlGroupMapper - mapping file for group_by() of control total checks  
* checkControlAggMapper - mapping file for numeric variables for control total checks  
* checkSimilarityMapper - mapping file for similarity checks to perform  
* plotSimilarityMaooer - mapping file for fields where differences in universe should be plotted
* keyAggMapper - mapping file for the aggregate-level control total checks to perform  
* perCapMapper - named vector that drives conversion from original field name to per capita field name  
  
These default parameters are maintained in a separate .R file and can be sourced:  
```{r}

source("./Coronavirus_CDC_Daily_Default_Mappings_v002.R")

```

## Example for Comparison to Previous
The function is tested on existing, previously downloaded data:  
```{r, fig.height=9, fig.width=9}

readList <- list("cdcDaily"="./RInputFiles/Coronavirus/CDC_dc_downloaded_210801.csv", 
                 "cdcHosp"="./RInputFiles/Coronavirus/CDC_h_downloaded_210801.csv", 
                 "vax"="./RInputFiles/Coronavirus/vaxData_downloaded_210801.csv"
                 )
compareList <- list("cdcDaily"=readFromRDS("cdc_daily_210708")$dfRaw$cdcDaily, 
                    "cdcHosp"=readFromRDS("cdc_daily_210708")$dfRaw$cdcHosp, 
                    "vax"=readFromRDS("cdc_daily_210728_vaxonly")$dfRaw$vax
                    )

cdc_daily_210801_test <- readRunCDCDaily(thruLabel="Jul 31, 2021", 
                                         downloadTo=lapply(readList, FUN=function(x) if(file.exists(x)) NA else x), 
                                         readFrom=readList,
                                         compareFile=compareList, 
                                         writeLog=NULL, 
                                         useClusters=readFromRDS("cdc_daily_210528")$useClusters, 
                                         weightedMeanAggs=c("tcpm7", "tdpm7", "cpm7", "dpm7", "hpm7", 
                                                            "vxcpm7", "vxcgte65pct"
                                                            ),
                                         skipAssessmentPlots=FALSE, 
                                         brewPalette="Paired"
                                         )

all.equal(names(cdc_daily_210801_test), names(readFromRDS("cdc_daily_210801")))
sapply(names(cdc_daily_210801_test), FUN=function(x)
    identical(cdc_daily_210801_test[[x]], readFromRDS("cdc_daily_210801")[[x]])
    )
sapply(names(cdc_daily_210801_test$plotDataList), FUN=function(x)
    identical(cdc_daily_210801_test$plotDataList[[x]], readFromRDS("cdc_daily_210801")$plotDataList[[x]])
    )

```
  
As expected, all data elements are identical.  Plot environments change with each creation, so the plot objects are not identical.

The latest data are downloaded and processed, with caching to avoid multiple file downloads:  
```{r, fig.height=9, fig.width=9, cache=TRUE}

readList <- list("cdcDaily"="./RInputFiles/Coronavirus/CDC_dc_downloaded_210804.csv", 
                 "cdcHosp"="./RInputFiles/Coronavirus/CDC_h_downloaded_210804.csv", 
                 "vax"="./RInputFiles/Coronavirus/vaxData_downloaded_210804.csv"
                 )
compareList <- list("cdcDaily"=readFromRDS("cdc_daily_210708")$dfRaw$cdcDaily, 
                    "cdcHosp"=readFromRDS("cdc_daily_210708")$dfRaw$cdcHosp, 
                    "vax"=readFromRDS("cdc_daily_210728_vaxonly")$dfRaw$vax
                    )

cdc_daily_210804 <- readRunCDCDaily(thruLabel="Aug 3, 2021", 
                                    downloadTo=lapply(readList, FUN=function(x) if(file.exists(x)) NA else x), 
                                    readFrom=readList,
                                    compareFile=compareList, 
                                    writeLog=NULL, 
                                    useClusters=readFromRDS("cdc_daily_210528")$useClusters, 
                                    weightedMeanAggs=c("tcpm7", "tdpm7", "cpm7", "dpm7", "hpm7", 
                                                       "vxcpm7", "vxcgte65pct"
                                                       ),
                                    skipAssessmentPlots=FALSE, 
                                    brewPalette="Paired"
                                    )

saveToRDS(cdc_daily_210804)

```

The raw hospital data is explored for admissions with confirmed/suspected coronavirus:  
```{r, fig.height=9, fig.width=9}

cdcHospAdmit_210804 <- cdc_daily_210804$dfRaw$cdcHosp %>%
    select(state, date, ends_with("confirmed"), ends_with("suspected")) %>%
    filter(state %in% c(state.abb, "DC")) %>%
    arrange(date, state) %>%
    pivot_longer(-c(state, date)) %>%
    mutate(name=stringr::str_replace(name, pattern="previous_day_admission_", replacement=""), 
           name=stringr::str_replace(name, pattern="_covid", replacement="")
           ) %>%
    left_join(getStateData(keepVars=c("state", "pop")), by="state") %>%
    mutate(vpm=1000000*value/pop) %>%
    group_by(state, name) %>%
    mutate(vpm7=zoo::rollmean(vpm, k=7, fill=NA)) %>%
    ungroup()

cdcHospAdmit_210804 %>%
    filter(!is.na(vpm7), state %in% state.abb) %>%
    mutate(div=as.character(state.division)[match(state, state.abb)]) %>%
    group_by(div, date, name) %>%
    summarize(wt_vpm7=sum(pop*vpm7)/sum(pop), tot_pop=sum(pop), .groups="drop") %>%
    ggplot(aes(x=date, y=wt_vpm7)) + 
    geom_col(aes(fill=name), position="stack") + 
    facet_wrap(~div) + 
    scale_fill_discrete("Metric") + 
    labs(x=NULL, 
         y="Newly admitted for COVID per million (rolling 7-day)", 
         title="Hospital admissions for COVID", 
         subtitle="All metrics divided by total population (all ages) for states reporting"
         )

```

Hospital coverage data became robust about a year ago.  The overwhelming majority of admissions are adult, split between confirmed and suspected cases.  Next steps are to explore changes in admissions by age groups:  
```{r, fig.height=9, fig.width=9}

hospAge_210804 <- cdc_daily_210804$dfRaw$cdcHosp %>%
    select(state, 
           date, 
           grep(x=names(.), pattern="ed_\\d.*[9+]$", value=TRUE), 
           grep(x=names(.), pattern="pediatric.*ed$", value=TRUE)
           ) %>% 
    pivot_longer(-c(state, date)) %>% 
    mutate(confSusp=ifelse(grepl(x=name, pattern="confirmed"), "confirmed", "suspected"), 
           adultPed=ifelse(grepl(x=name, pattern="adult"), "adult", "ped"), 
           age=ifelse(adultPed=="ped", "0-17", stringr::str_replace_all(string=name, pattern=".*_", replacement="")), 
           age=ifelse(age %in% c("0-17", "18-19"), "0-19", age), 
           div=as.character(state.division)[match(state, state.abb)]
           ) 
hospAge_210804

# Plot for overall trends by age group
p1 <- hospAge_210804 %>% 
    filter(state %in% c(state.abb, "DC"), !is.na(value)) %>% 
    mutate(ageBucket=age) %>% 
    group_by(date, ageBucket) %>% 
    summarize(value=sum(value), .groups="drop") %>% 
    arrange(date) %>%
    group_by(ageBucket) %>% 
    mutate(value7=zoo::rollmean(value, k=7, fill=NA)) %>% 
    filter(date >= "2020-08-01") %>% 
    ggplot(aes(x=date, y=value7)) + 
    labs(x=NULL, 
         y="Confirmed or suspected COVID admissions (rolling-7 mean)", 
         title="Hospital admissions for COVID by age bucket (Aug 2020 - Jul 2021)", 
         subtitle="50 states and DC (includes confirmed and suspected from CDC data)"
         ) + 
    lims(y=c(0, NA))

# Line plots by age group
p1 + 
    geom_line(aes(group=ageBucket, color=ageBucket), size=1) + 
    scale_color_discrete("Age\nbucket")
# Stacked bar plots by age group
p1 + 
    geom_col(aes(fill=ageBucket), position="stack") + 
    scale_color_discrete("Age\nbucket")
# Proportions by age group
p1 + 
    geom_col(aes(fill=ageBucket), position="fill") + 
    scale_color_discrete("Age\nbucket")

# Plot for overall trends by age group
hospAge_210804 %>% 
    filter(state %in% state.abb, !is.na(value)) %>% 
    mutate(ageBucket=ifelse(age >= "60", "60+", ifelse(age=="0-19", "0-19", "20-59"))) %>% 
    group_by(date, state, ageBucket) %>% 
    summarize(value=sum(value), .groups="drop") %>% 
    group_by(ageBucket, state) %>% 
    mutate(value7=zoo::rollmean(value, k=7, fill=NA)) %>% 
    filter(date >= "2020-06-01") %>% 
    ggplot(aes(x=date, y=value7)) + 
    geom_line(aes(color=ageBucket, group=ageBucket)) + 
    scale_color_discrete("Age\nbucket") + 
    labs(x=NULL, 
         y="Confirmed or suspected COVID admissions (rolling-7 mean)", 
         title="Hospital admissions for COVID by age bucket (Aug 2020 - Jul 2021)"
         ) + 
    lims(y=c(0, NA)) + 
    facet_wrap(~state, scales="free_y")

```

Next steps are to explore alignment of the hospitalization and case/death curves.
