---
title: "Coronavirus US - CDC All-Cause Death"
author: "davegoblue"
date: "7/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background  
This module extends code contained in Coronavirus_Statistics_v004.Rmd to include sourcing of all key functions and parameters.  This file includes the latest code for analyzing all-cause death data from [CDC Weekly Deaths by Jurisdiction](https://catalog.data.gov/dataset/weekly-counts-of-deaths-by-jurisdiction-and-age-group).  CDC maintains data on deaths by week, age cohort, and state in the US.  Downloaded data are unique by state, epidemiological week, year, age, and type (actual vs. predicted/projected).

These data are known to have a lag between death and reporting, and the CDC back-correct to report deaths at the time the death occurred even if the death is reported in following weeks.  This means totals for recent weeks tend to run low (lag), and the CDC run a projection of the expected total number of deaths given the historical lag times.  Per other analysts on the internet, there is currently significant supra-lag, with lag times much longer than historical averages causing CDC projected deaths for recent weeks to be low.

The code leverages tidyverse and sourced functions throughout:  
```{r}

# All functions assume that tidyverse and its components are loaded and available
library(tidyverse)

# If the same function is in both files, use the version from the more specific source
source("./Generic_Added_Utility_Functions_202105_v001.R")
source("./Coronavirus_CDC_Excess_Functions_v001.R")

```
  
## Running Code  
The main function is readRunCDCAllCause(), which performs multiple tasks:  
  
STEP 0: Optionally, downloads the latest data file from CDC
STEP 1: Reads and processes a data file has been downloaded from CDC to local  
STEP 2: Extract relevant data from a processed state-level COVID Tracking Project list  
STEP 3: Basic plots of the CDC data  
STEP 4: Basic excess-deaths analysis  
STEP 5: Create cluster-level aggregate plots  
STEP 6: Create state-level aggregate plots  
STEP 7: Create age-cohort aggregate plots  
STEP 8: Returns a list of key data frames, modeling objects, named cluster vectors, etc.  
  
The functions are tested on previously downloaded data:  
```{r, fig.height=9, fig.width=9}

cdcLoc <- "Weekly_counts_of_deaths_by_jurisdiction_and_age_group_downloaded_20210623.csv"
cdcList_20210703 <- readRunCDCAllCause(loc=cdcLoc, 
                                       weekThru=17, 
                                       lst=readFromRDS("cdc_daily_210528"), 
                                       dlData=FALSE, 
                                       stateNoCheck=c("NC"), 
                                       pdfCluster=TRUE, 
                                       pdfAge=TRUE
                                       )

```
  
