---
title: "Data Camp Insights"
author: "davegoblue"
date: "June 1, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(tidyverse)

```

## Background and Overview  
DataCamp offer interactive courses related to R Programming.  While some is review, it is helpful to see other perspectives on material.  As well, DataCamp has some interesting materials on packages that I want to learn better (ggplot2, dplyr, ggvis, etc.).  This document summarizes a few key insights from:  

This document is currently split between _v003 and _v003_a and _v003_b and _v003_c due to the need to keep the number of DLL that it opens below the hard-coded maximum.  This introductory section needs to be re-written, and the contents consolidated, at a future date.
  
* R Programming (Introductory R, Intermediate R, Writing Functions in R, Object Oriented Programming in R, Introduction to Tidyverse)  
* Importing and Cleaning Data (Cleaning Data in R, Importing Data in to R)  
* Data Manipulation (dplyr, data.table, xts/zoo, dplyr joins, cases for EDA/Time Series/Pitch Data)  
* Data Visualization (base, ggplot2 parts I/II/III, ggvis, geospatial)  
* Statistics (8 refresher modules)  
* Machine Learning (3 modules + 1 text mining case)  
* R Studio (2 parts) and R Markdown (1 module)  
  
The original DataCamp_Insights_v001 and DataCamp_Insights_v002 documents have been split for this document:  
  
* This DataCamp_Insights_v003 document contains evolving sections on R Programming, Machine Learning, and RStudio / R Markdown  
* Importing and Cleaning Data components have been moved to DataCamp_ImportClean_v002  
* Data Manipulation components have been moved to DataCamp_DataManipulation_v002  
* Visualization components have been moved to DataCamp_Visualization_v002  
* Statistics components have been moved to DataCamp_Statistics_v002  
  
  
***
  

###_Hierarchical and Mixed Effects Models_  
  
Chapter 1 - Overview and Introduction  
  
What is a hierarchical model?  
  
* Hierarchical data is nested within itself, and can be analyzed using the lme package  
	* Example of students in a classroom - may not all be independent of each other due to teacher quality, building conditions, etc.  
    * Hierarchical models can help with pooling means across small sample sizes  
    * Repeated measurements (test scores each year) are also a common example of data that are not truly independent  
* Hierarchical models can include nested models and multi-level models  
* Regression frameworks can include pool information and random effects (vs. fixed effects) and mixed-effects and linear mixed-effects  
* Repeated sampling can have repeated measures modeling  
  
Parts of a regression:  
  
* Linear regression and linear model can be used interchangeably for this course - epsilon is the error term, assumed to be normal with zero mean and constant variance  
* The linear model in R is closely related to analysis of variance (ANOVA)  
	* lm (y ~ x, myData)  
    * anova( lm (y ~ x, myData) )  
* The most basic regression has an intercept, a slope, a single predictor, and an error term  
	* The concept can be extended to multiple regression with additional predictors  
* There are some limitations to the multiple regression approach  
	* Parameter estimates can be very sensitive to other variables - Simpson's paradox and the like  
    * Need to note that the regression coefficient is "after controlling for . . . " (all the other variables)  
    * Interaction terms can be important as well  
* Regressions in R for an intercept for every group are called as lm(y ~ x - 1)  
* The interaction term x1*x2 is the same as x1 + x2 + x1:x2  
  
Random effects in regression:  
  
* Nested relationships tend to be hierarchical in nature - students are part of classes are part of schools and the like  
	* Mathematically, this is referred to as a mapping among the distributions  
* The algebraic representation is that y ~ B*x + eps, with B ~ N(mu, sigma**2)  
	* library(lme4) is the best packages for this in R  
    * lme4::lmer(y ~ x + (1|randomGroup), data=myData)  
    * lme4::lmer(y ~ x + (randomSlope|randomGroup), data=myData)  
  
School data:  
  
* Appliciation of multi-level models to school data - influence of sex, teacher training, plotting parameter estmates  
  
Example code includes:  
```{r}

rawStudent <- read.csv("./RInputFiles/classroom.csv")

studentData <- rawStudent %>%
    mutate(sex=factor(sex, labels=c("male", "female")), minority=factor(minority, labels=c("no", "yes")))


# Plot the data
ggplot(data = studentData, aes(x = housepov, y = mathgain)) +
    geom_point() +
    geom_smooth(method = 'lm')

# Fit a linear model
summary( lm(mathgain ~ housepov , data = studentData))


# I have aggregated the data for you into two new datasets at the classroom- and school-levels (As a side note, if you want to learn how to aggregate data, the dplyr or data.table courses teach these skills)
# We will also compare the model outputs across all three outputs
# Note: how we aggregate the data is important
# I aggregated the data by taking the mean across the student data (in pseudo-code: mean(mathgain) by school or mean(mathgain) by classroom), 
# but another reasonable method for aggregating the data would be to aggregate by classroom first and school second

classData <- studentData %>%
    group_by(schoolid, classid) %>%
    summarize_at(vars(mathgain, mathprep, housepov, yearstea), mean, na.rm=TRUE)
str(classData)

schoolData <- studentData %>%
    group_by(schoolid) %>%
    summarize_at(vars(mathgain, mathprep, housepov, yearstea), mean, na.rm=TRUE)
str(schoolData)


# First, plot the hosepov and mathgain at the classroom-level from the classData data.frame
ggplot(data = classData, aes(x = housepov, y = mathgain)) +
    geom_point() +
    geom_smooth(method = 'lm')

# Second, plot the hosepov and mathgain at the school-level from the schoolData data.frame
ggplot(data = schoolData, aes(x = housepov, y = mathgain)) +
    geom_point() +
    geom_smooth(method = 'lm')

# Third, compare your liner regression results from the previous expercise to the two new models
summary( lm(mathgain ~ housepov, data = studentData)) ## student-level data
summary( lm(mathgain ~ housepov, data = classData)) ## class-level data
summary( lm(mathgain ~ housepov, data = schoolData)) ## school-level data


# Plot the means of your data, predictor is your x-variable, response is your y-variable, and intDemo is your data.frame
intDemo <- data.frame(predictor=factor(c(rep("a", 5), rep("b", 5), rep("c", 5))), 
                      response=c(-1.207, 0.277, 1.084, -2.346, 0.429, 5.759, 4.138, 4.18, 4.153, 3.665, 9.046, 8.003, 8.447, 10.129, 11.919)
                      )
str(intDemo)


ggIntDemo <- ggplot(intDemo, aes(x = predictor, y = response) ) +
    geom_point() +
    theme_minimal() + stat_summary(fun.y = "mean", color = "red",
                                   size = 3, geom = "point") +
    xlab("Intercept groups")
print(ggIntDemo)

# Fit a linear model to your data where response is "predicted by"(~) predictor
intModel <- lm( response ~ predictor - 1 , data = intDemo)
summary(intModel)


extractAndPlotResults <- function(intModel){
    intCoefPlot <- broom::tidy(intModel)
    intCoefPlot$term <- factor(gsub("predictor", "", intCoefPlot$term))

    plotOut <- ggIntDemo + geom_point(data = intCoefPlot,
                           aes(x = term, y = estimate),
                           position = position_dodge(width = 0.4),
                           color = 'blue', size = 8, alpha = 0.25)
    print(plotOut)
}


# Run the next code that extracts out the model's coeffiecents and plots them 
extractAndPlotResults(intModel)


multIntDemo <- data.frame(group=factor(c(rep("a", 5), rep("b", 5), rep("c", 5))), 
                          x=rep(0:4, times=3), 
                          intercept=c(4.11, -1.69, 1.09, 1.9, 1.21, 4.63, 10.29, 4.67, 12.06, 4.78, 15.22, 19.15, 4.44, 8.88, 9.47), 
                          response=c(4.11, 2.31, 9.09, 13.9, 17.21, 4.63, 14.29, 12.67, 24.06, 20.78, 15.22, 23.15, 12.44, 20.88, 25.47)
                          )
str(multIntDemo)

plot_output1 <- function(out1){
    ggmultIntgDemo <- ggplot( multIntDemo, aes(x = x, y = response) ) +
        geom_point(aes(color = group)) +
        theme_minimal() +
        scale_color_manual(values = c("blue", "black", "red")) +
        stat_smooth(method = 'lm', fill = NA, color = 'orange', size = 3)
    print(ggmultIntgDemo)
}

plot_output2 <- function(out2){
    out2Tidy <- broom::tidy(out2)
    out2Tidy$term <- gsub("group", "", out2Tidy$term)
    out2Plot <- data.frame(group = out2Tidy[ -1, 1],
                           slope = out2Tidy[ 1, 2],
                           intercept = out2Tidy[ -1, 2])
    ggmultIntgDemo2 <- ggplot( multIntDemo, aes(x = x, y = response) ) +
        geom_point(aes(color = group))+
        theme_minimal() +
        scale_color_manual(values = c("blue", "black", "red")) +
        geom_abline(data = out2Plot,
                    aes(intercept = intercept, slope = slope, color = group))
    print(ggmultIntgDemo2)
}

plot_output3 <- function(out3){
    ggmultIntgDemo3 <- ggplot( multIntDemo, aes(x = x, y = response) ) +
        geom_point(aes(color = group)) +
        theme_minimal() +
        scale_color_manual(values = c("blue", "black", "red")) +
        stat_smooth(method = 'lm', aes(color = group), fill = NA)
    print(ggmultIntgDemo3)
}

# First, run a model without considering different intercept for each group
out1 <- lm( response ~ x, data=multIntDemo )
summary(out1)
plot_output1(out1)

# Considering same slope but different intercepts
out2 <- lm( response ~ x + group - 1, data=multIntDemo )
summary(out2)
plot_output2(out2)

# Consdering different slope and intercept for each group (i.e., an interaction)
out3 <- lm( response ~ x + group - 1 + x:group, multIntDemo)
summary(out3)
plot_output3(out3)


multIntDemo$intercept <- c(-0.87, 3.35, 1.25, 0.88, -1.05, 4.55, 1.22, 3.34, 1.26, 3.75, 7.71, 9.59, 2.28, 1.9, 13.35)
multIntDemo$response <- c(-0.87, 6.35, 7.25, 9.88, 10.95, 4.55, 4.22, 9.34, 10.26, 15.75, 7.71, 12.59, 8.28, 10.9, 25.35)

# Run model
outLmer <- lme4::lmer( response ~ x + ( 1 | group), multIntDemo)

# Look at model outputs 
summary( outLmer )
broom::tidy( outLmer )


extractAndPlotOutput <- function(outLmer, slope=3){
    multIntDemo$lmerPredict <- predict(outLmer)
    ggmultIntgDemo2 <- ggplot( multIntDemo, aes(x = x, y = response) ) +
        geom_point(aes(color = group))+
        theme_minimal() +
        scale_color_manual(values = c("blue", "black", "red")) +
        geom_abline(data = multIntDemo,
                    aes(intercept = intercept, slope = slope, color = group))
    outPlot <-  ggmultIntgDemo2 +
                geom_line( data =  multIntDemo,
                      aes(x = x, y = lmerPredict, color = group),
                      linetype = 2)
    print(outPlot)
}


# Extract predictor variables and plot
extractAndPlotOutput(outLmer)


# Random effect slopes
multIntDemo$response <- c(-0.72, 1.5, 4.81, 6.61, 13.62, 10.21, 9.64, 11.91, 16.39, 16.97, 8.76, 14.79, 15.83, 15.27, 17.36)
multIntDemo$intercept <- c(-0.72, -1.5, -1.19, -2.39, 1.62, 10.21, 6.64, 5.91, 7.39, 4.97, 8.76, 11.79, 9.83, 6.27, 5.36)

outLmer2 <- lme4::lmer( response ~ ( x|group ), multIntDemo)
summary(outLmer2)
broom::tidy(outLmer2)


plotOutput <- function(outLmer2){
    multIntDemo$lmerPredict2 <- predict(outLmer2)
    ggmultIntgDemo3 <- ggplot( multIntDemo, aes(x = x, y = response) ) +
        geom_point(aes(color = group)) +
        theme_minimal() +
        scale_color_manual(values = c("blue", "black", "red")) +
        stat_smooth(method = 'lm', aes(color = group), fill = NA)
    plotOut <- ggmultIntgDemo3 +
            geom_line( data =  multIntDemo,
                      aes(x = x, y = lmerPredict2, color = group),
                      linetype = 2)
    print(plotOut)
}


# Extract and plot
plotOutput(outLmer2)


# Mixed effect model
lmerModel <- lme4::lmer(mathgain ~ sex + 
                  mathprep + mathknow + (1|classid) +
                  (1|schoolid), data = studentData, na.action = "na.omit",
                  REML = TRUE)
summary(lmerModel)


extractAndPlot <- function(lmerModel){
    modelOutPlot <- broom::tidy(lmerModel, conf.int = TRUE)
    modelOutPlot <- modelOutPlot[ modelOutPlot$group =="fixed" &
                               modelOutPlot$term != "(Intercept)", ]
    plotOut <- ggplot(modelOutPlot, aes(x = term, y = estimate,
                             ymin = conf.low,
                             ymax = conf.high)) +
            theme_minimal() +
            geom_hline(yintercept = 0.0, color = 'red', size = 2.0) +
            geom_point() +
            geom_linerange() + coord_flip()
    print(plotOut)
}


# Extract and plot 
extractAndPlot(lmerModel)

```
  
  
  
***
  
Chapter 2 - Linear Mixed-Effect Models  
  
Linear mixed effect model - Birth rates data:  
  
* Small populations are highly sensitive to stochastic effects - if the mean is 1, a group of 5 might have 0 or 10  
* Questions about how counties may impact birth rates, over and above other demographic factors  
	* Example of plotting birth rate vs. county - will see both the highest and lowest birth rates in the smallest counties  
* Random effect syntax for the lmer model includes  
	* (1 | group) - random intercept with fixed mean  
    * (1 | g1/g2) - intercepts vary among g1 and g2 within g2  
    * (1 | g1) + (1 | g2) - random intercepts for two variables  
    * x + (x | g) - correlated random slope and intercept  
    * x + (x || g) - uncorrelated random slope and intercept  
    * See lme4 documentation for additional details  
  
Understanding and reporting the outputs of lmer:  
  
* The output from lmer is similar to the output from lm, but with some key differences - if using print(), will see  
	* The method used is REML - restricted maximum likelihood - which tends to solve better than maximum likelihood for these problems  
    * There is an REML convergence criteria, which can be a helpful diagnostic  
    * Can see the standard deviations for both the state and the residual, along with the number of observations  
    * Get the fixed effects coefficients in a similar form as lm()  
* The summary() call on lmer produces several additional outputs  
	* Residuals summary  
    * Fixed effects estimates include SE and t-values (but not p-values)  
    * Correlations of fixed effects  
* Can grab only the fixed effects using fixef(myLMERObject)  
	* Can grab only the random effects using ranef(myLMERObject), though these will not have confidence intervals  
    * The random effects confidence intervals could be estimated using bootstrapping or Bayesian methods per the author of lme4 - but actual random effects are just unobserved random variables rather than parameters  
* Can grab only the confidence intervals using confint(myLMERObject)  
* Need to be careful in reporting the results - figures vs. tables vs. in-line descriptions  
  
Statistical inference with Maryland crime data:  
  
* The Maryland crime data is available on data.gov - interesting for many public and private purposes  
* The null hypothsis test can be used with LMER - frequentist approach  
	* By default, lmer does not provide p-values, as there is ongoing debate as to the degrees of freedom and impact on reported results  
    * Can use lmerTest package to calculate and report on the p-values  
* Can use ANOVA to look at the variability explained by one model versus another model, and the associated degrees of freedom needed  
  
Example code includes:  
```{r}

# Read in births data
rawBirths <- read.csv("./RInputFiles/countyBirthsDataUse.csv")
countyBirthsData <- rawBirths
str(countyBirthsData)


# First, build a lmer with state as a random effect. Then look at the model's summary and the plot of residuals. 
birthRateStateModel <- lme4::lmer(BirthRate ~ (1|State), data=countyBirthsData)
summary(birthRateStateModel)
plot(birthRateStateModel)

# Next, plot the predicted values from the model ontop of the plot shown during the video.
countyBirthsData$birthPredictState <- predict(birthRateStateModel, countyBirthsData)
ggplot() + theme_minimal() +
    geom_point(data =countyBirthsData, aes(x = TotalPopulation, y = BirthRate)) + 
    geom_point(data = countyBirthsData, aes(x = TotalPopulation, y = birthPredictState),
               color = 'blue', alpha = 0.5
               )

# Include the AverageAgeofMother as a fixed effect within the lmer and state as a random effect
ageMotherModel <- lme4::lmer( BirthRate ~ AverageAgeofMother + (1|State), data=countyBirthsData)
summary(ageMotherModel)

# Compare the random-effect model to the linear effect model 
summary(lm(BirthRate ~ AverageAgeofMother, data = countyBirthsData))


# Include the AverageAgeofMother as a correlated random-effect slope parameter
ageMotherModelRandomCorrelated <- lme4::lmer(BirthRate ~ AverageAgeofMother + (AverageAgeofMother|State),
                       countyBirthsData)
summary(ageMotherModelRandomCorrelated)


# Include the AverageAgeofMother as a correlated random-effect slope parameter
ageMotherModelRandomUncorrelated <- lme4::lmer(BirthRate ~ AverageAgeofMother + 
                                                    (AverageAgeofMother || State), data=countyBirthsData
                                               )
summary(ageMotherModelRandomUncorrelated)


out <- ageMotherModelRandomUncorrelated

# Extract the fixed-effect coefficients
lme4::fixef(out)

# Extract the random-effect coefficients
lme4::ranef(out)

# Estimate the confidence intervals 
(ciOut <- confint(out))


# Technical note: Extracting out the regression coefficients from lmer is tricky (see discussion between the lmer and broom authors development)
# Extract out the parameter estimates and confidence intervals and manipulate the data
dataPlot <- data.frame(cbind( lme4::fixef(out), ciOut[ 4:5, ]))
rownames(dataPlot)[1] <- "Intercept"
colnames(dataPlot) <- c("mean", "l95", "u95")
dataPlot$parameter <- rownames(dataPlot)

# Print the new dataframe
print(dataPlot)

# Plot the results using ggplot2
ggplot(dataPlot, aes(x = parameter, y = mean,
                     ymin = l95, ymax = u95)) +
    geom_hline( yintercept = 0, color = 'red' ) +
    geom_linerange() + geom_point() + coord_flip() + theme_minimal()



# Read in crime data
rawCrime <- read.csv("./RInputFiles/MDCrime.csv")
MDCrime <- rawCrime
str(MDCrime)


plot1 <- ggplot(data = MDCrime, aes(x = Year, y = Crime, group = County)) +
    geom_line() + theme_minimal() +
    ylab("Major crimes reported per county")
print(plot1)

plot1 + geom_smooth(method = 'lm')


# Null hypothesis testing uses p-values to see if a variable is "significant"
# Recently, the abuse and overuse of null hypothesis testing and p-values has caused the American Statistical Association to issue a statement about the use of p-values
# Because of these criticisms and other numerical challenges, Doug Bates (the creator of the lme4 package) does not include p-values as part of his package
# However, you may still want to estimate p-values, because p-values are sill commonly used. Several packages exist, including the lmerTest package
# https://www.amstat.org/asa/files/pdfs/P-ValueStatement.pdf

# Load lmerTest
# library(lmerTest)

# Fit the model with Year as both a fixed and random-effect
lme4::lmer(Crime ~ Year + (1 + Year | County) , data = MDCrime)

# Fit the model with Year2 rather than Year
out <- lme4::lmer(Crime ~ Year2 + (1 + Year2 | County) , data = MDCrime)

# Examine the model's output
summary(out)


## Build the Null model with only County as a random-effect
null_model <- lme4::lmer(Crime ~ (1 | County) , data = MDCrime)

## Build the Year2 model with Year2 as a fixed and random slope and County as the random-effect
year_model <- lme4::lmer(Crime ~ Year2 + (1 + Year2 | County) , data = MDCrime)

## Compare the two models using an anova
anova(null_model, year_model)

```
  
  
  
***
  
Chapter 3 - Generalized Linear Mixed-Effect Models  
  
Crash course on GLMs - relaxing the assumptions around normality of the residuals:  
  
* Non-normal data can be transformed using arcsin or the like  
* However, with advances in methodology, it is possible to more directly model the data using binomial and poisson distributions  
* The basic glm call is glm(y ~ x, family="")  # default is family="gaussian", which same as the lm()  
* The Poisson distribution is frequently best for count data, such as website visitors per hour - mean equals variance (generally best for small counts less than 30; can use normals for large counts)  
* For logistic regression, data can be entered in any of three formats  
	* Binary (y=0 or 1) - glm(y ~ x, family="binomial")  
    * Wilkinson-Rogers - glm(cbind(success, failure) ~ x, family="binomial")  
    * Weighted - glm(y ~ x, weights=weights, family="binomial")  
    * These methods differ primarily in the degrees of freedom (and thus deviance)  
  
Binomial data - modeling data with only two outcomes:  
  
* Traditional method for analysis includes looking at proportion of successes  
* The GLM allows for direct looks at the data - logistic regression (logit)  
* Binomial data can be fit using glmer(y ~ x + (1/group), family="error term")  
* The regression coefficients can be difficult to explain, sometimes leading to the use of odds ratios instead  
	* The odds ratio of 2.0 would mean 2:1 odds for that specific group  
  
Count data:  
  
* Examples like number of events per hour (website hits) or counts per area (birds)  
* The count data differs from the binomial in that there is no pre-specified upper boundary  
* While Chi-squared is often used for goodness of fit or test of association for count data, the Poisson GLM can be a nice alternative  
	* glm(y ~ x, family="poisson")  
    * glmer(y ~ x + (1|group), family="poisson")  
  
Example code includes:  
```{r cache=TRUE}

# In this case study, we will be working with simulated dose-response data
# The response is mortality (1) or survival (0) at the end of a study. During this exercise, we will fit a logistic regression using all three methods described in the video
# You have been given two datasets. dfLong has the data in a "long" format with each row corresponding to an observation (i.e., a 0 or 1)
# dfShort has the data in an aggregated format with each row corresponding to a treatment (e.g., 6 successes, 4 failures, number of replicates = 10, proportion = 0.6)

dfLong <- data.frame(dose=c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10), 
                     mortality=c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 1, 0, 0, 1, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 1, 0, 1, 0, 0, 1, 0, 1, 1, 0, 1, 1, 0, 1, 0, 1, 1, 0, 0, 0, 1, 1, 0, 0, 1, 1, 1, 0, 0, 1, 0, 0, 1, 0, 1, 0, 1, 0, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 0, 0, 0, 1, 0, 1, 0, 1, 1, 0, 0, 1, 1)
                     )
str(dfLong)

dfShort <- dfLong %>% 
    group_by(dose) %>%
    summarize(mortality=sum(mortality), nReps=n()) %>%
    mutate(survival=nReps-mortality, mortalityP=mortality/nReps)
dfShort


# Fit a glm using data in a long format
fitLong <- glm(mortality ~ dose, data = dfLong, family = "binomial")
summary(fitLong)

# Fit a glm using data in a short format with two columns
fitShort <- glm( cbind(mortality , survival ) ~ dose , data = dfShort, family = "binomial")
summary(fitShort)

# Fit a glm using data in a short format with weights
fitShortP <- glm( mortalityP ~ dose , data = dfShort, weights = nReps , family = "binomial")
summary(fitShortP)


y <- c(0, 1, 0, 1, 0, 1, 0, 1, 0, 2, 1, 2, 0, 1, 1, 0, 1, 5, 1, 1)
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20)


# Fit the linear model
summary(lm(y ~ x))

# Fit the generalized linear model
summary(glm(y ~ x, family = "poisson"))


# Often, we want to "look" at our data and trends in our data
# ggplot2 allows us to add trend lines to our data
# The defult lines are created using a technique called local regression
# However, we can specify different models, including GLMs
# During this exercise, we'll see how to plot a GLM

# Plot the data using jittered points and the default stat_smooth
ggplot(data = dfLong, aes(x = dose, y = mortality)) + 
    geom_jitter(height = 0.05, width = 0.1) +
    stat_smooth(fill = 'pink', color = 'red') 

# Plot the data using jittered points and the the glm stat_smooth
ggplot(data = dfLong, aes(x = dose, y = mortality)) + 
    geom_jitter(height = 0.05, width = 0.1) +
    stat_smooth(method = 'glm',  method.args = list(family = "binomial"))


# library(lmerTest)

df <- data.frame(dose=rep(rep(c(0, 2, 4, 6, 8, 10), each=20), times=3), 
                 mortality=c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 0, 1, 0, 0, 1, 0, 1, 0, 0, 1, 0, 1, 1, 0, 0, 1, 0, 1, 0, 1, 1, 1, 0, 0, 0, 1, 1, 0, 1, 1, 1, 1, 1, 0, 0, 1, 0, 0, 1, 1, 0, 0, 1, 0, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 0, 1, 0, 1, 1, 0, 1, 0, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 0, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 1, 1, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 1, 0, 1, 1, 0, 1, 1, 1, 1, 0, 0, 1, 1, 0, 1, 0, 0, 1, 0, 1, 0, 1, 0, 0, 1, 0, 0, 1, 0, 0, 1, 0, 1, 0, 1, 1, 0, 1, 1, 0, 1, 0, 1, 1, 0, 0, 1, 1, 1, 1, 1, 1, 1, 0, 1, 1, 0, 1, 1, 0, 0, 0, 0, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 0, 1, 1, 0, 0, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 0, 0, 0, 1, 0, 0, 0, 1, 1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 1, 1, 1, 0, 0, 1, 1, 0, 1, 1, 1, 0, 0, 0, 0, 0, 0, 1, 1, 1, 0, 0, 0, 1, 1, 1, 0, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1), 
                 replicate=factor(rep(letters[1:3], each=120))
                 )
str(df)


glmerOut <- lme4::glmer(mortality ~ dose + (1|replicate), family = 'binomial', data = df)
summary(glmerOut)


# library(lmerTest)
# Fit the model and look at its summary 
# modelOut <- lme4::glmer( cbind(Purchases, Pass) ~ friend + ranking + (1|city), data = allData, family = 'binomial')
# summary( modelOut) 

# Compare outputs to a lmer model
# summary(lme4::lmer( Purchases/( Purchases + Pass) ~ friend + ranking + (1|city), data = allData))


# Run the code to see how to calculate odds ratios
# summary(modelOut) 
# exp(fixef(modelOut)[2])
# exp(confint(modelOut)[3, ])


# Load lmerTest
# library(lmerTest)


userGroups <- data.frame(group=factor(rep(rep(LETTERS[1:4], each=10), times=2)), 
                         webpage=factor(rep(c("old", "new"), each=40)), 
                         clicks=c(0, 0, 0, 0, 0, 0, 2, 0, 0, 0, 1, 0, 0, 1, 1, 1, 2, 0, 1, 1, 0, 3, 2, 3, 1, 2, 4, 2, 1, 0, 2, 0, 1, 2, 0, 2, 1, 1, 2, 4, 2, 8, 1, 1, 1, 2, 1, 1, 0, 0, 3, 0, 1, 4, 1, 2, 0, 1, 1, 0, 0, 3, 2, 0, 3, 1, 2, 2, 0, 2, 3, 1, 3, 2, 4, 4, 2, 1, 5, 2)
                         )
str(userGroups)


# Fit a Poisson glmer
summary( lme4::glmer(clicks ~ webpage + (1|group), family = 'poisson', data = userGroups))


# library(lmerTest)


rawIL <- read.csv("./RInputFiles/ILData.csv")
ILdata <- rawIL
str(ILdata)

# Age goes before year
modelOut <- lme4::glmer(count ~ age + year + (year|county), family = 'poisson', data = ILdata)
summary(modelOut)


# Extract out fixed effects
lme4::fixef(modelOut)

# Extract out random effects 
lme4::ranef(modelOut)


# Run code to see one method for plotting the data
ggplot(data = ILdata, aes(x = year, y = count, group = county)) +
    geom_line() +
    facet_grid(age ~ . ) +
    stat_smooth( method = 'glm',
                method.args = list( family = "poisson"), se = FALSE,
                alpha = 0.5) +
    theme_minimal()

```
  
  
  
***
  
Chapter 4 - Repeated Measures  
  
An introduction to repeated measures:  
  
* Sampling the same thing over time is a repeated measure, a specific example of a mixed effects model  
	* Follow the same individual through time - cohorts allow for controlling for individuality  
    * The paired t-test is often used for assessing a repeated measures dataset - t.test(x1, x2, paired=TRUE)  # x1 and x2 need to be the same length and each element needs to be the same individual  
* Repeated measures ANOVA is a conceptual extension of the paired t-test - are the means constant over time  
	* anova(lmer(y ~ time + (1|individual)))  
    * Can be used with glmer() also  
    * Note that degrees of freedom is still an open question - different packages calculate this differently  
  
Sleep study:  
  
* Applying LMER to the sleep study dataset - impact of drugs on sleep patterns for 10 patients followed over time  
	* This is the classic "Student" dataset due to Guinness at the time not allowing its employees to publish  
    * Ho will be that the amount of sleep does not vary with the treatments  
    * Modeling will be done using a linear mixed model  
* Modeling approach - iteratively;  
	* EDA  
    * Simple regression  
    * Model of interest  
    * Extract information from model  
    * Visualize final data  
  
Hate in NY state?  
  
* Change in rate of hate crimes over time by county - available from data.gov for 2010-2016  
* Level of technical details in reporting should vary significantly by audience - blend data in to story for wider audiences, while being reporducible/technical for a scientifc audience  
  
Wrap up:  
  
* Hiearchical data, mixed effects models, case studies  
* Start with the LME4 documentation for additional explorations and details  
  
Example code includes:  
```{r}

y <- c(0.23, 2.735, -0.038, 6.327, -0.643, 1.69, -1.378, -1.228, -0.252, 2.014, -0.073, 6.101, 0.213, 3.127, -0.29, 8.395, -0.33, 2.735, 0.223, 1.301)
treat <- rep(c("before", "after"), times=10)
x <- rep(letters[1:10], each=2)

# Run a standard, non-paired t-test
t.test(y[treat == "before"], y[treat == "after"], paired = FALSE)

# Run a standard, paired t-test
t.test(y[treat == "before"], y[treat == "after"], paired = TRUE)


library(lmerTest)
library(lme4)

# Run the paired-test like before
t.test(y[treat == "before"], y[treat == "after"], paired = TRUE)

# Run a repeated-measures ANOVA
anova(lmer( y ~ treat + (1|x)))


data(sleepstudy, package="lme4")
str(sleepstudy)

# Plot the data
ggplot(data = sleepstudy) +
    geom_line(aes(x = Days, y = Reaction, group = Subject)) +
    stat_smooth(aes(x = Days, y = Reaction), method = 'lm', size = 3, se = FALSE)

# Build a lm 
lm( Reaction ~ Days, data = sleepstudy)

# Build a lmer
(lmerOut <- lmer( Reaction ~ Days + (1|Subject), data = sleepstudy))


# The lmer model you built during the previous exercise has been saved as lmerOut
# During this exercise, you will examine the effects of drug type using both an ANOVA framework and a regression framework

# Run an anova
anova(lmerOut)

# Look at the regression coefficients
summary(lmerOut)


# Read in NY hate data
rawHate <- read.csv("./RInputFiles/hateNY.csv")
hate <- rawHate
str(hate)


ggplot( data = hate, aes(x = Year, y = TotalIncidents, group = County)) +
    geom_line() + 
    geom_smooth(method = 'lm', se = FALSE)


# During this exercise, you will build a glmer
# Because most of the incidents are small count values, use a Poisson (R function family poisson) error term
# First, build a model using the actually year (variable Year, such as 2006, 2007, etc) - this model will fail
# Second, build a model using the rescaled year (variable Year2, such as 0, 1, 2, etc)
# This demonstrates the importance of considering where the intercept is located when building regression models
# Recall that a variable x can be both a fixed and random effect in a lmer() or glmer(): for example lmer(y ~ x + (x| group) demonstrates this syntax

# glmer with raw Year
glmer(TotalIncidents ~ Year + (Year|County), data = hate, family = "poisson")

# glmer with scaled Year, Year2
glmerOut <- glmer(TotalIncidents ~ Year2 + (Year2|County), data = hate, family = "poisson")
summary(glmerOut)


# Extract and manipulate data
countyTrend <- ranef(glmerOut)$County
countyTrend$county <- factor(row.names(countyTrend), levels =row.names(countyTrend)[order(countyTrend$Year2)])

# Plot results 
ggplot(data = countyTrend, aes(x = county, y = Year2)) + geom_point() +
    coord_flip() + 
    ylab("Change in hate crimes per year")  +
    xlab("County")

```
  
  
  
***
  
###_Forecasting Product Demand in R_  
  
Chapter 1 - Forecasting Demand with Time Series  
  
Loading data in to an xts object:  
  
* The xts object will be the buidling block for the course - extensible time series (xts) is an extension of the zoo package - basically, a time index attached to the data matrix  
* Can create dates using dates=seq(as.Date("MM-DD-YYYY"), length=, by="weeks")  # to create weekly data  
	* xts(myData, order.by=dates)  # will create an XTS using dates as the index  
  
ARIMA Time Series 101:  
  
* AR - AutoRegressive (lags help to determine today's values - "long memory models")  
* MA - Moving Average (shocks/errors help to determine today's shocks/errors - "short memory models" due to dissipation)  
* I - Integrated (does the data have a dependency across time, and how long does it last) - make the time series stationary  
	* Stationarity is the idea that effects disipate over time - today has more impact on tomorrow than on time periods in the future  
    * Differencing (monthly, seasonal, etc.) the data can be a useful approach for data with stationarity  
* Begin by creating training dataset and valiadation training dataset  
* The auto.arima() function tries to estimate the best ARIMA for a given data series  
	* ARIMA(p, d, q) is ARIMA(AR, Differencing, MA)  
  
Forecasting and Evaluating:  
  
* Can use the ARIMA data to forecast the data forward - extrapolating the signal (forecasting) and estimating the amount of noise (error or CI)  
* The forecast() function in R simplifies the process - forecast(myModel, h=) which will forecast forward h time periods  
* Two common error measurements include MAE (mean average error) and MAPE (mean average percentage error)  
	* MAPE is better at putting things on a common scale  
  
Example code includes:  
```{r}


# Read in beverages data
rawBev <- read.csv("./RInputFiles/Bev.csv")
bev <- rawBev
str(bev)


# Load xts package 
library(xts)
library(forecast)


# Create the dates object as an index for your xts object
dates <- seq(as.Date("2014-01-19"), length = 176, by = "weeks")

# Create an xts object called bev_xts
bev_xts <- xts(bev, order.by = dates)


# Create the individual region sales as their own objects
MET_hi <- bev_xts[,"MET.hi"]
MET_lo <- bev_xts[,"MET.lo"]
MET_sp <- bev_xts[,"MET.sp"]

# Sum the region sales together
MET_t <- MET_hi + MET_lo + MET_sp

# Plot the metropolitan region total sales
plot(MET_t)


# Split the data into training and validation
MET_t_train <- MET_t[index(MET_t) < "2017-01-01"]
MET_t_valid <- MET_t[index(MET_t) >= "2017-01-01"]

# Use auto.arima() function for metropolitan sales
MET_t_model <- auto.arima(MET_t_train)


# Forecast the first 22 weeks of 2017
forecast_MET_t <- forecast(MET_t_model, h = 22)

# Plot this forecast #
plot(forecast_MET_t)


# Convert to numeric for ease
for_MET_t <- as.numeric(forecast_MET_t$mean)
v_MET_t <- as.numeric(MET_t_valid)

# Calculate the MAE
MAE <- mean(abs(for_MET_t - v_MET_t))

# Calculate the MAPE
MAPE <- 100*mean(abs(for_MET_t - v_MET_t)/v_MET_t)

# Print to see how good your forecast is!
print(MAE)
print(MAPE)


# Convert your forecast to an xts object
for_dates <- seq(as.Date("2017-01-01"), length = 22, by = "weeks")
for_MET_t_xts <- xts(forecast_MET_t$mean, order.by = for_dates)

# Plot the validation data set
plot(for_MET_t_xts, main = 'Forecast Comparison', ylim = c(4000, 8500))

# Overlay the forecast of 2017
lines(MET_t_valid, col = "blue")


# Plot the validation data set
plot(MET_t_valid, main = 'Forecast Comparison', ylim = c(4000, 8500))

# Overlay the forecast of 2017
lines(for_MET_t_xts, col = "blue")

# Convert the limits to xts objects
lower <- xts(forecast_MET_t$lower[, 2], order.by = for_dates)
upper <- xts(forecast_MET_t$upper[, 2], order.by = for_dates)

# Adding confidence intervals of forecast to plot
lines(lower, col = "blue", lty = "dashed")
lines(upper, col = "blue", lty = "dashed")

```
  
  
  
***
  
Chapter 2 - Components of Demand  
  
Price elasticity:  
  
* Price is one of the obvious factors that impacts demand, with the relationship called price elasticity (% dDemand / % dPrice)  
	* Elastic goods have elasticity > 1, meaning demand changes more quickly (percentage wise) than price  
    * Inelastic goods have elasticity < 1, for example gasoline  
    * Unit elastic goods have elasticity = 1, meaning that X% increase in price drives X% decrease in demand  
    * Linear regression can be employed to estimate the elasticity for a given product - the log-log transform helps get the % vs % coefficients  
  
Seasonal/holiday/promotional effects:  
  
* Seasonal products are common - can be bought any time of the year, though certain seasons have higher demand (holidays are a common example)  
* Promotions are attempts by companies to influence demand  
* Linear regression can help determine relationships between demand and many other factors  
	* If an xts vector has been created for key dates, can merge(train, holiday, fill=0) and the holiday column will be 0 wherever there is no match to holiday  
  
Forecasting with regression:  
  
* Forecasting with time series is straightforward due to the lag nature of the models - tomorrow forecasts today and today forecasts tomorrow and etc.  
* Forecasting with regression can be more tricky, particularly since we need the future inputs (such as price) in order to predict the future demand  
	* Even when there are contractually fixed prices, promotions can effectively create a de facto price change anyways  
* Need to have the same column names in the test/validation dataset as were used in the modeling  
	* Then, can use predict(myModel, myData)  
    * May need to exponentiate in case the data are currently on the log scale rather than the absolute scale  
  
Example code includes:  
```{r}

bev_xts_train <- bev_xts[index(bev_xts) < "2017-01-01"]
bev_xts_valid <- bev_xts[index(bev_xts) >= "2017-01-01"]

# Save the prices of each product
l_MET_hi_p <- log(as.vector(bev_xts_train[, "MET.hi.p"]))

# Save as a data frame
MET_hi_train <- data.frame(as.vector(log(MET_hi[index(MET_hi) < "2017-01-01"])), l_MET_hi_p)
colnames(MET_hi_train) <- c("log_sales", "log_price")

# Calculate the regression
model_MET_hi <- lm(log_sales ~ log_price, data = MET_hi_train)


# Plot the product's sales
plot(MET_hi)

# Plot the product's price
MET_hi_p <- bev_xts_train[, "MET.hi.p"]
plot(MET_hi_p)


# Create date indices for New Year's week
n.dates <- as.Date(c("2014-12-28", "2015-12-27", "2016-12-25"))

# Create xts objects for New Year's
newyear <- as.xts(rep(1, 3), order.by = n.dates)

# Create sequence of dates for merging
dates_train <- seq(as.Date("2014-01-19"), length = 154, by = "weeks")

# Merge training dates into New Year's object
newyear <- merge(newyear, dates_train, fill = 0)


# Add newyear variable to your data frame
MET_hi_train <- data.frame(MET_hi_train, newyear=as.vector(newyear))

# Build regressions for the product
model_MET_hi_full <- lm(log_sales ~ log_price + newyear, data = MET_hi_train)


# Subset the validation prices #
l_MET_hi_p_valid <- log(as.vector(bev_xts_valid[, "MET.hi.p"]))

# Create a validation data frame #
MET_hi_valid <- data.frame(l_MET_hi_p_valid)
colnames(MET_hi_valid) <- "log_price"


# Predict the log of sales for your high end product
pred_MET_hi <- predict(model_MET_hi, MET_hi_valid)

# Convert predictions out of log scale
pred_MET_hi <- exp(pred_MET_hi)


# Convert to an xts object
dates_valid <- seq(as.Date("2017-01-01"), length = 22, by = "weeks")
pred_MET_hi_xts <- xts(pred_MET_hi, order.by = dates_valid)

# Plot the forecast
plot(pred_MET_hi_xts)

# Calculate and print the MAPE
MET_hi_v <- bev_xts_valid[,"MET.hi"]

MAPE <- 100*mean(abs((pred_MET_hi_xts - MET_hi_v)/MET_hi_v))
print(MAPE)

```
  
  
  
***
  
Chapter 3 - Blending Regression with Time Series  
  
Residuals from regression model:  
  
* The residuals from the regression models can be used for further modeling - see if the residuals are related over time, and model them with time series if so  
* Need to start by gathering the residuals and then converting them to an XTS object - explore for patterns in this XTS object  
  
Forecasting residuals:  
  
* When the residuals are related across time, we can use time series to model the residuals - basically, patterns to the errors provide an opportunity for further modeling  
* Can use auto.arima() on the residuals data, to see what the best ARIMA model for the residuals is  
	* Can then forecast the residuals in to the future using forecast(myModel, h=) # h being the time periods to predict forward  
  
Transfer functions and ensembling:  
  
* Techniques for combining forecasts - single model (transfer function) or averaging of models (ensembling)  
* Demand can be based on both regression (modeling external factors) and time series (residuals)  
* Ensembling is a combination (blend) of the forecasts, with simple averaging being the simplest approach  
	* Basically, build a stand-alone time series model and a stand-alone regression model  
    * The ensemble forecast can be better or worse than any of the stand-alone models  
  
Example code includes:  
```{r}

# Calculate the residuals from the model
MET_hi_full_res <- resid(model_MET_hi_full)

# Convert the residuals to an xts object
MET_hi_full_res <- xts(MET_hi_full_res, order.by = dates_train)


# Plot the histogram of the residuals
hist(MET_hi_full_res)

# Plot the residuals over time
plot(MET_hi_full_res)


# Build an ARIMA model on the residuals: MET_hi_arima
MET_hi_arima <- auto.arima(MET_hi_full_res)

# Look at a summary of the model
summary(MET_hi_arima)


# Forecast 22 weeks with your model: for_MET_hi_arima
for_MET_hi_arima <- forecast(MET_hi_arima, h=22)

# Print first 10 observations
head(for_MET_hi_arima$mean, n = 10)


# Convert your forecasts into an xts object
dates_valid <- seq(as.Date("2017-01-01"), length = 22, by = "weeks")
for_MET_hi_arima <- xts(for_MET_hi_arima$mean, order.by = dates_valid)

# Plot the forecast
plot(for_MET_hi_arima)


# Convert your residual forecast to the exponential version
for_MET_hi_arima <- exp(for_MET_hi_arima)

# Multiply your forecasts together!
for_MET_hi_final <- for_MET_hi_arima * pred_MET_hi_xts


# Plot the final forecast - don't touch the options!
plot(for_MET_hi_final, ylim = c(1000, 4300))

# Overlay the validation data set
lines(MET_hi_v, col = "blue")


# Calculate the MAE
MAE <- mean(abs(for_MET_hi_final - MET_hi_v))
print(MAE)

# Calculate the MAPE
MAPE <- 100 * mean(abs(for_MET_hi_final - MET_hi_v)/MET_hi_v)
print(MAPE)


# Build an ARIMA model using the auto.arima function
MET_hi_model_arima <- auto.arima(MET_hi)

# Forecast the ARIMA model
for_MET_hi <- forecast(MET_hi_model_arima, h = length(MET_hi_v))

# Save the forecast as an xts object
dates_valid <- seq(as.Date("2017-01-01"), length = 22, by = "weeks")
for_MET_hi_xts <- xts(for_MET_hi$mean, order.by = dates_valid)

# Calculate the MAPE of the forecast
MAPE <- 100 * mean(abs(for_MET_hi_xts - MET_hi_v) / MET_hi_v)
print(MAPE)


# Ensemble the two forecasts together
for_MET_hi_en <- 0.5 * (for_MET_hi_xts + pred_MET_hi_xts)

# Calculate the MAE and MAPE
MAE <- mean(abs(for_MET_hi_en - MET_hi_v))
print(MAE)

MAPE <- 100 * mean(abs(for_MET_hi_en - MET_hi_v) / MET_hi_v)
print(MAPE)

```
  
  
  
***
  
Chapter 4 - Hierarchical Forecasting  
  
Bottom-Up Hierarchical Forecasting:  
  
* The hierarchical data structuring can be an advantage in forecasting, provided that the data has a natural hierarchy  
* The sum of all the lower-level forecasts should equal the higher-level forecasts  
	* Bottom-up: Forecast at the lowest level and aggregate (easiest but requires the most number of forecasts)  
    * Top-down: Forecast at the top level and the apply downwards  
    * Middle-out: Forecast at the middle levels and then apply both upwards and downwards  
  
Top-Down Hierarchical Forecasting:  
  
* The top-down forecasting process is typically quicker but less accurate than the bottom-up forecasting process  
* Two techniques available for top-down reconciliation  
	* Average of historical proportions - mean percentage that each component contributes to the total (calculated by sub-component such as week)  
    * Proportion of historical averages - mean percentage that each component contributes to the total (calculated by aggregate)  
* Reconciled forecasts at lower levels are typically less accurate than the direct forecast of the lower levels  
  
Middle-Out Hierarchical Forecasting:  
  
* Bottom-up forecasting is higher quality but more time-consuming than top-down forecasting  
* The middle-out forecasting method is a sometimes successful blend of the methods, getting decent accuracy at a lesser time commitment  
  
Wrap up:  
  
* Using time series to forecast demand forward  
* Incorporating external factors using linear regression  
* Blending time series and regression approaches  
* Top-down, bottom-up, middle-out approaches to aggregation and forecasting at various levels (hierarchical)  
* Can extend by looking at cross-elasticities (impact of competitor pricing)  
* Can better forecast proportions using time series analysis  
* Additional demand forecasting models include neural networks, exponential smoothing, etc.  
  
Example code includes:  
```{r}

# Build a time series model 
MET_sp_model_arima <- auto.arima(MET_sp)

# Forecast the time series model for 22 periods
for_MET_sp <- forecast(MET_sp_model_arima, h=22)

# Create an xts object
for_MET_sp_xts <- xts(for_MET_sp$mean, order.by=dates_valid)

MET_sp_v <- MET_sp["2017"]

# Calculate the MAPE
MAPE <- 100 * mean(abs(for_MET_sp_xts - MET_sp_v) / MET_sp_v)
print(MAPE)


MET_sp_train <- bev_xts_train %>%
    transform(log_sales = log(MET.sp), log_price=log(MET.sp.p))
MET_sp_train <- MET_sp_train[, c("log_sales", "log_price")]
MET_sp_train$newyear <- 0
MET_sp_train$valentine <- 0
MET_sp_train$christmas <- 0
MET_sp_train$mother <- 0

MET_sp_train[index(MET_sp_train) %in% as.Date(c("2014-12-28", "2015-12-27", "2016-12-25")), "newyear"] <- 1
MET_sp_train[index(MET_sp_train) %in% as.Date(c("2014-02-09", "2015-02-08", "2016-02-07")), "valentine"] <- 1
MET_sp_train[index(MET_sp_train) %in% as.Date(c("2014-12-21", "2015-12-20", "2016-12-18")), "christmas"] <- 1
MET_sp_train[index(MET_sp_train) %in% as.Date(c("2014-05-04", "2015-05-03", "2016-05-01")), "mother"] <- 1


# THE BELOW IS TOTAL NONSENSE
# Build a regression model
model_MET_sp <- lm(log_sales ~ log_price + newyear + valentine + christmas + mother, data = MET_sp_train)


MET_sp_valid <- as.data.frame(bev_xts_valid) %>%
    mutate(log_sales = log(MET.sp), log_price=log(MET.sp.p)) %>%
    select("log_sales", "log_price")
MET_sp_valid$newyear <- 0
MET_sp_valid$valentine <- 0
MET_sp_valid$christmas <- 0
MET_sp_valid$mother <- 0  

MET_sp_valid[7, "valentine"] <- 1
MET_sp_valid[19, "mother"] <- 1
MET_sp_valid$log_sales <- NULL


# Forecast the regression model using the predict function 
pred_MET_sp <- predict(model_MET_sp, MET_sp_valid)

# Exponentiate your predictions and create an xts object
pred_MET_sp <- exp(pred_MET_sp)
pred_MET_sp_xts <- xts(pred_MET_sp, order.by = dates_valid)

# Calculate MAPE
MAPE <- 100*mean(abs((pred_MET_sp_xts - MET_sp_v)/MET_sp_v))
print(MAPE)


# Ensemble the two forecasts
for_MET_sp_en <- 0.5 * (for_MET_sp_xts + pred_MET_sp_xts)

# Calculate the MAPE
MAPE <- 100 * mean(abs(for_MET_sp_en - MET_sp_v) / MET_sp_v)
print(MAPE)


# Copy over pred_MET_lo_xts
pred_MET_lo_xts <- xts(c(2960.6, 2974.1, 2943.2, 2948.6, 2915.6, 2736.4, 2953.9, 3199.4, 2934, 2898.7, 3027.7, 3165.9, 3073.1, 2842.7, 2928.7, 3070.2, 2982.2, 3018, 3031.9, 2879.4, 2993.2, 2974.1), order.by=dates_valid)


# Calculate the metropolitan regional sales forecast
for_MET_total <- pred_MET_hi_xts + for_MET_sp_en + pred_MET_lo_xts

# Calculate a validation data set 
MET_t_v <- bev_xts_valid[,"MET.hi"] + bev_xts_valid[,"MET.lo"] + bev_xts_valid[,"MET.sp"]

# Calculate the MAPE
MAPE <- 100 * mean(abs(for_MET_total - MET_t_v) / MET_t_v)
print(MAPE)


# Create the MET_total data
MET_total <- xts(data.frame(MET.hi=c(5942, 5600, 5541, 6892, 5586, 5943, 6329, 6693, 6938, 6138, 6361, 6378, 5423, 5097, 4937, 5496, 6870, 6626, 6356, 5657, 6577, 7202, 7381, 7404, 7204, 6667, 6153, 6035, 5633, 5283, 5178, 4758, 5058, 5254, 5954, 6166, 6247, 6304, 7202, 6662, 6814, 6174, 5412, 5380, 5674, 6472, 6912, 7404, 8614, 8849, 7174, 6489, 7174, 6555, 6402, 7671, 5012, 4790, 5075, 5238, 5615, 6113, 7706, 7811, 7898, 7232, 6585, 5870, 7084, 5125, 5330, 5553, 6349, 6195, 6271, 5851, 5333, 5854, 5609, 5649, 6051, 6409, 5786, 5190, 5085, 4949, 5151, 5147, 5426, 5509, 6956, 7870, 8224, 6685, 6153, 5802, 5244, 5162, 5036, 5025, 8378, 8944, 7109, 7605, 7846, 7598, 8012, 9551, 6102, 5366, 4932, 4962, 5392, 6194, 7239, 7621, 7460, 7097, 6596, 5848, 8306, 5344, 5848, 6341, 7364, 7269, 7053, 6682, 6971, 7521, 7063, 6298, 6003, 5227, 5047, 4877, 4851, 4628, 4516, 4442, 4935, 5181, 5431, 5866, 5919, 5704, 5957, 6019, 5962, 6021, 5880, 5674, 7439, 7415)),
                 order.by=dates_train
                 )

# Build a regional time series model
MET_t_model_arima <- auto.arima(MET_total)

# Calculate a 2017 forecast for 22 periods
for_MET_t <- forecast(MET_t_model_arima, h=22)

# Make an xts object from your forecast
for_MET_t_xts <- xts(for_MET_t$mean, order.by=dates_valid)

# Calculate the MAPE
MAPE <- 100 * mean(abs(for_MET_t_xts - MET_t_v) / MET_t_v)
print(MAPE)


# Calculate the average historical proportions
prop_hi <- mean(MET_hi/MET_total)
prop_lo <- mean(MET_lo/MET_total)
prop_sp <- mean(MET_sp/MET_total)

# Distribute out your forecast to each product
for_prop_hi <- prop_hi*for_MET_t_xts
for_prop_lo <- prop_lo*for_MET_t_xts
for_prop_sp <- prop_sp*for_MET_t_xts

# Calculate the MAPE's for each product
MAPE_hi <- 100 * mean(abs(for_prop_hi - MET_hi_v) / MET_hi_v)
print(MAPE_hi)

MET_lo_v <- bev_xts_valid[,"MET.lo"]
MAPE_lo <- 100 * mean(abs(for_prop_lo - MET_lo_v) / MET_lo_v)
print(MAPE_lo)

MAPE_sp <- 100 * mean(abs(for_prop_sp - MET_sp_v) / MET_sp_v)
print(MAPE_sp)


# Calculate the average historical proportions
prop_hi_2 <- mean(MET_hi) / mean(MET_total)
prop_lo_2 <- mean(MET_lo) / mean(MET_total)
prop_sp_2 <- mean(MET_sp) / mean(MET_total)

# Distribute out your forecast to each product
for_prop_hi_2 <- prop_hi_2 * for_MET_t_xts
for_prop_lo_2 <- prop_lo_2 * for_MET_t_xts
for_prop_sp_2 <- prop_sp_2 * for_MET_t_xts

# Calculate the MAPE's for each product
MAPE_hi <- 100 * mean(abs(for_prop_hi_2 - MET_hi_v) / MET_hi_v)
print(MAPE_hi)
MAPE_lo <- 100 * mean(abs(for_prop_lo_2 - MET_lo_v) / MET_lo_v)
print(MAPE_lo)
MAPE_sp <- 100 * mean(abs(for_prop_sp_2 - MET_sp_v) / MET_sp_v)
print(MAPE_sp)


SEC_total <- xts(data.frame(SEC.hi=c(700, 775, 789, 863, 765, 759, 757, 747, 746, 709, 749, 786, 796, 726, 727, 723, 778, 755, 739, 740, 723, 695, 727, 707, 725, 684, 667, 698, 727, 722, 748, 695, 742, 739, 715, 724, 686, 671, 688, 682, 710, 700, 672, 680, 695, 780, 751, 693, 809, 881, 703, 712, 768, 796, 808, 904, 641, 662, 693, 725, 719, 736, 715, 722, 732, 745, 689, 705, 811, 739, 744, 700, 745, 735, 732, 722, 721, 732, 750, 714, 752, 677, 731, 674, 720, 675, 741, 722, 715, 719, 649, 697, 743, 733, 772, 698, 690, 734, 713, 644, 788, 833, 749, 731, 670, 675, 675, 993, 773, 751, 697, 677, 750, 723, 780, 763, 721, 701, 704, 684, 985, 791, 731, 714, 704, 694, 685, 652, 708, 754, 747, 705, 711, 699, 712, 745, 706, 665, 666, 692, 676, 696, 689, 697, 689, 717, 697, 708, 660, 707, 715, 680, 922, 888)), order.by=dates_train
                 )

# Build a time series model for the region
SEC_t_model_arima <- auto.arima(SEC_total)

# Forecast the time series model
for_SEC_t <- forecast(SEC_t_model_arima, h=22)

# Make into an xts object
for_SEC_t_xts <- xts(for_SEC_t$mean, order.by=dates_valid)

SEC_t_v <- bev_xts_valid$SEC.hi + bev_xts_valid$SEC.lo
# Calculate the MAPE
MAPE <- 100 * mean(abs(for_SEC_t_xts - SEC_t_v) / SEC_t_v)
print(MAPE)


SEC_hi <- bev_xts_train[, "SEC.hi"]
SEC_lo <- bev_xts_train[, "SEC.lo"]
SEC_hi_v <- bev_xts_valid[, "SEC.hi"]
SEC_lo_v <- bev_xts_valid[, "SEC.lo"]

# Calculate the average of historical proportions
prop_hi <- mean(SEC_hi / SEC_total)
prop_lo <- mean(SEC_lo / SEC_total)

# Distribute the forecast
for_prop_hi <- prop_hi * for_SEC_t_xts
for_prop_lo <- prop_lo * for_SEC_t_xts

# Calculate a MAPE for each product
MAPE_hi <- 100 * mean(abs(for_prop_hi - SEC_hi_v) / SEC_hi_v)
print(MAPE_hi)

MAPE_lo <- 100 * mean(abs(for_prop_lo - SEC_lo_v) / SEC_lo_v)
print(MAPE_lo)


# Copy over for_M_t_xts data
for_M_t_xts <- xts(c(2207, 2021, 2010, 2052, 2075, 2074, 2065, 2058, 2056, 2055, 2053, 2052, 2050, 2049, 2048, 2047, 2046, 2045, 2044, 2043, 2043, 2042), order.by=dates_valid)

# Calculate the state sales forecast: for_state
for_state = for_SEC_t_xts + for_MET_t_xts + for_M_t_xts

# See the forecasts
for_state

```
  
  
  
***
  
###_HR Analytics in R: Exploring Employee Data_  
  
Chapter 1 - Identifying the Best Recruiting Source  
  
Introduction - Ben Teusch, HR Analytics Consultant:  
  
* HR analytics has many other names - people analytics, workforce analytics, etc.  
* Identify groups for comparison - high vs. low performers, groups with high vs. low turnover, etc.  
	* Exploratory analysis and statistics for each group, including plots of key differences  
* Course is outlines as a series of case studies, with one case per chapter  
  
Recruiting and quality of hire:  
  
* Where are the best hires coming from, and how can you get more of them  
	* Defining quality of hire is challenging - some mix of productivity, satisfaction, retention, performance reviews, etc.  
    * Attrition can be defined as the mean of a 1, 0 vector of "did the person leave in the time period T"  
  
Visualizing recruiting data:  
  
* Helpful for communicating findings to decision makers  
* The geom_col() in ggplot will make a bar chart, with the y aestehtic being the bar height  
  
Example code includes:  
```{r}

# Import the recruitment data
recruitment <- readr::read_csv("./RInputFiles/recruitment_data.csv")

# Look at the first few rows of the dataset
head(recruitment)

# Get an overview of the recruitment data
summary(recruitment)

# See which recruiting sources the company has been using
recruitment %>% 
  count(recruiting_source)


# Find the average sales quota attainment for each recruiting source
avg_sales <- recruitment %>% 
  group_by(recruiting_source) %>% 
  summarize(avg_sales_quota_pct=mean(sales_quota_pct))

# Display the result
avg_sales


# Find the average attrition for the sales team, by recruiting source, sorted from lowest attrition rate to highest
avg_attrition <- recruitment %>%
  group_by(recruiting_source) %>% 
  summarize(attrition_rate=mean(attrition)) %>%
  arrange(attrition_rate)

# Display the result
avg_attrition

# Plot the bar chart
avg_sales %>% ggplot(aes(x=recruiting_source, y=avg_sales_quota_pct)) + geom_col()

# Plot the bar chart
avg_attrition %>% ggplot(aes(x=recruiting_source, y=attrition_rate)) + geom_col()

```
  
  
  
***
  
Chapter 2 - What is driving low employee engagement  
  
Analyzing employee engagement:  
  
* Gallup defines engaged employees as those who are involved in, enthusiastic about, and committed to their workplace  
* Survey data are available in the example case study  
	* Will use both mutate() and ifelse()  
    * The ifelse() is needed for vectors of length > 1 since it can work in a vectorized manner (and is thus OK inside the mutate call)  
  
Visualizing the engagement data:  
  
* Multiple attributes in a single place can make for a more compelling report  
* The tidyr package is part of the tidyverse, and hslps arrange the data properly for plotting  
	* tidyr::gather(columns, key="key", value="value") will be the package used in this example - pull the data from the columns down to the rows  
    * ggplot(survey_gathered, aes(x = key, y = value, fill = department)) + geom_col(position = "dodge")  
    * ggplot(survey_gathered, aes(x = key, y = value, fill = department)) + geom_col(position = "dodge") + facet_wrap(~ key, scales = "free")  
  
Are differences meaningful?  
  
* Can use significance testing to assess likelhood (p-value) that the second sample could have come from the same population as the first sample  
	* This course will use t-test (continuous variables) and chi-squared test (categorical variables)  
    * t.test(tenure ~ is_manager, data = survey)  
    * chisq.test(survey$left_company, survey$is_manager)  # no data= argument is available in the function  
  
Example code includes:  
```{r}

# Import the data
survey <- readr::read_csv("./RInputFiles/survey_data.csv")

# Get an overview of the data
summary(survey)

# Examine the counts of the department variable
survey %>% count(department)


# Output the average engagement score for each department, sorted
survey %>%
  group_by(department) %>%
  summarize(avg_engagement=mean(engagement)) %>%
  arrange(avg_engagement)


# Create the disengaged variable and assign the result to survey
survey_disengaged <- survey %>% 
  mutate(disengaged = ifelse(engagement <= 2, 1, 0)) 

survey_disengaged

# Summarize the three variables by department
survey_summary <- survey_disengaged %>%
  group_by(department) %>%
  summarize(pct_disengaged=mean(disengaged), 
            avg_salary=mean(salary), 
            avg_vacation_taken=mean(vacation_days_taken)
            )

survey_summary


# Gather data for plotting
survey_gathered <- survey_summary %>% 
  gather(key = "measure", value = "value",
         pct_disengaged, avg_salary, avg_vacation_taken)

# Create three bar charts
ggplot(survey_gathered, aes(x=measure, y=value, fill=department)) +
  geom_col(position="dodge") + 
  facet_wrap(~ measure, scales="free")


# Add the in_sales variable
survey_sales <- survey %>%
  mutate(in_sales = ifelse(department == "Sales", "Sales", "Other"), 
         disengaged = ifelse(engagement < 3, 1L, 0L)
         )

# Test the hypothesis using survey_sales
chisq.test(survey_sales$disengaged, survey_sales$in_sales)
t.test(disengaged ~ in_sales, data=survey_sales)


# Test the hypothesis using the survey_sales data
t.test(vacation_days_taken ~ in_sales, data = survey_sales)

```
  
  
  
***
  
Chapter 3 - Are new hires getting paid too much?  
  
Paying new hires fairly:  
  
* Sometimes, current employees get paid less than new employees, which can drive low engagement and turnover  
* Case study will have a simulated pay dataset available for analysis  
* Can use broom::tidy() to return the outputs in a nicely formatted data frame  
	* chisq.test(survey$in_sales, survey$disengaged) %>% tidy()  
  
Omitted variable bias:  
  
* Key assumption of the tests is that the groups are the same, with the exception of the variables being tested  
* Omitted variable bias occurs when both 1) the omitted variable is correlated with the dependent variable, and 2) the omitted variable is correlated with an explanatory variable  
	* Omitted variables are often known as confounders  
    * Plotting can help to identify the issue, particularly with a stacked (to 100%) bar chart  
    * pay %>% ggplot(aes(x = new_hire, fill = department)) + geom_bar(position = "fill")  
    * The geom_bar() object has height that is fully dependent on x, in contrast to geom_col() which has a y-aestehtic  
  
Linear regression helps to test the multivariate impacts of variables:  
  
* lm(salary ~ new_hire, data = pay) %>% tidy()  # single dependent variable  
* lm(salary ~ new_hire + department, data = pay) %>% tidy()  # multiple dependent variables  
* lm(salary ~ new_hire + department, data = pay) %>% summary()  # more detailed summary of the linear regression  
  
Example code includes:  
```{r}

# Import the data
pay <- readr::read_csv("./RInputFiles/fair_pay_data.csv")

# Get an overview of the data
summary(pay)

# Check average salary of new hires and non-new hires
pay %>% 
  group_by(new_hire) %>%
  summarize(avg_salary=mean(salary))


# Perform the correct statistical test
t.test(salary ~ new_hire, data = pay)
t.test(salary ~ new_hire, data = pay) %>%
  broom::tidy()


# Create a stacked bar chart
pay %>%
  ggplot(aes(x=new_hire, fill=job_level)) + 
  geom_bar(position="fill")

# Calculate the average salary for each group of interest
pay_grouped <- pay %>% 
  group_by(new_hire, job_level) %>% 
  summarize(avg_salary = mean(salary))
  
# Graph the results using facet_wrap()  
pay_grouped %>%
  ggplot(aes(x=new_hire, y=avg_salary)) + 
  geom_col() + 
  facet_wrap(~ job_level)


# Filter the data to include only hourly employees
pay_filter <- pay %>%
  filter(job_level == "Hourly")

# Test the difference in pay
t.test(salary ~ new_hire, data=pay_filter) %>%
  broom::tidy()


# Run the simple regression
model_simple <- lm(salary ~ new_hire, data = pay)

# Display the summary of model_simple
model_simple %>% 
  summary()

# Display a tidy summary
model_simple %>% 
  broom::tidy()


# Run the multiple regression
model_multiple <- lm(salary ~ new_hire + job_level, data = pay)

# Display the summary of model_multiple
model_multiple %>% 
  summary()

# Display a tidy summary
model_multiple %>% 
  broom::tidy()

```
  
  
  
***
  
Chapter 4 - Are performance ratings being given consistently?  
  
Joining HR data:  
  
* Employee data tend to be stored in different locations, requiring joins (merges) prior to running analyses  
	* dplyr::left_join(hr_data, bonus_pay_data, by = "employee_id")  
    * All employees in hr_data will be kept, even if there is no matching record in bonus_pay_data  
    * Employee ID (or similar) is by far the best way to join data - names tend to be non-unique and can differ in different systems  
  
Performance ratings and fairness:  
  
* Performance ratings are inherently subjective and thus prone to bias  
* Unconscious bias is based on the brain's heuristics, and may include preferences for members of various groups (biases, as reflected in hiring, promotion, etc.)  
  
Logistic regression is especially helpful for modeling binary response variables:  
  
* glm(high_performer ~ salary, data = hr, family = "binomial") %>% tidy()  
* glm(high_performer ~ salary + department, data = hr, family = "binomial") %>% tidy()  
  
Example code includes:  
```{r}

# Import the data
hr_data <- readr::read_csv("./RInputFiles/hr_data.csv")
performance_data <- readr::read_csv("./RInputFiles/performance_data.csv")

# Examine the datasets
summary(hr_data)
summary(performance_data)


# Join the two tables
joined_data <- left_join(hr_data, performance_data, by = "employee_id")

# Examine the result
summary(joined_data)

# Check whether the average performance rating differs by gender 
joined_data %>%
  group_by(gender) %>%
  summarize(avg_rating = mean(rating))


# Add the high_performer column
performance <- joined_data %>%  
  mutate(high_performer = ifelse(rating >= 4, 1, 0))

# Test whether one gender is more likely to be a high performer
chisq.test(performance$gender, performance$high_performer)   
 
# Do the same test, and tidy the output
chisq.test(performance$gender, performance$high_performer) %>% broom::tidy()


# Visualize the distribution of high_performer by gender
performance %>%
  ggplot(aes(x=gender, fill=factor(high_performer))) + 
  geom_bar(position="fill")

# Visualize the distribution of all ratings by gender
performance %>%
  ggplot(aes(x=gender, fill=factor(rating))) + 
  geom_bar(position="fill")

# Visualize the distribution of job_level by gender
performance %>%
  ggplot(aes(x = gender, fill = job_level)) +
  geom_bar(position = "fill")
 
# Test whether men and women have different job level distributions
chisq.test(performance$gender, performance$job_level) 


# Visualize the distribution of high_performer by gender, faceted by job level
performance %>%
  ggplot(aes(x = gender, fill = factor(high_performer))) +
  geom_bar(position = "fill") + 
  facet_wrap(~ job_level)


# Run a simple logistic regression
logistic_simple <- glm(high_performer ~ gender, family = "binomial", data = performance) 

# View the result with summary()
logistic_simple %>%
  summary()

# View a tidy version of the result
logistic_simple %>%
  broom::tidy()


# Run a multiple logistic regression
logistic_multiple <- glm(high_performer ~ gender + job_level, family = "binomial", data = performance)

# View the result with summary() or tidy()
logistic_multiple %>% broom::tidy()

```
  
  
  
***
  
Chapter 5 - Improving employee safety with data  
  
Employee safety - looking at accident rates and drivers:  
  
* Requires joining data on multiple variables  
	* joined_data <- left_join(hr_data, safety_data, by = c("year", "employee_id"))  
    * joined_data %>% filter(is.na(accident_time)) # use is.na() instead  
  
Focusing on the location of interest:  
  
* May want to run comparisons of the same location over time  
* May want to assess differences by locations to see if they may be explanatory variables  
  
Explaining the increase in accidents:  
  
* Can use multiple regression to help test for explanatory variables that impact the accident rate  
  
Wrap up:  
  
* Key tools from the Tidyverse (ggplot2, broom, dplyr, etc.) to assess HR data  
* Analytics usage within HR, including differences in HR and other data  
* Can apply additional data science techniques on HR data  
  
Example code includes:  
```{r}

# Import the data 
hr_data <- readr::read_csv("./RInputFiles/hr_data_2.csv")
accident_data <- readr::read_csv("./RInputFiles/accident_data.csv")

# Create hr_joined with left_join() and mutate()
hr_joined <- left_join(hr_data, accident_data, by=c("year", "employee_id")) %>% 
  mutate(had_accident=ifelse(is.na(accident_type), 0, 1))
  
hr_joined


# Find accident rate for each year
hr_joined %>% 
  group_by(year) %>% 
  summarize(accident_rate = mean(had_accident))

# Test difference in accident rate between years
chisq.test(hr_joined$year, hr_joined$had_accident)

# Which location had the highest acccident rate?
hr_joined %>%
  group_by(location) %>%
  summarize(accident_rate=mean(had_accident)) %>%
  arrange(-accident_rate)


# Compare annual accident rates by location
accident_rates <- hr_joined %>% 
  group_by(location, year) %>% 
  summarize(accident_rate = mean(had_accident))
  
accident_rates

# Graph it
accident_rates %>% 
  ggplot(aes(factor(year), accident_rate)) +
  geom_col() +
  facet_wrap(~location)


# Filter out the other locations
southfield <- hr_joined %>% 
  filter(location == "Southfield")

# Find the average overtime hours worked by year
southfield %>%
  group_by(year) %>% 
  summarize(average_overtime_hours = mean(overtime_hours))

# Test difference in Southfield's overtime hours between years
t.test(overtime_hours ~ year, data=southfield) 


# Import the survey data
survey_data <- readr::read_csv("./RInputFiles/survey_data_2.csv")

# Create the safety dataset
safety <- left_join(hr_joined, survey_data, by=c("employee_id", "year")) %>%
  mutate(disengaged=ifelse(engagement <= 2, 1, 0), year=factor(year))


# Visualize the difference in % disengaged by year in Southfield
safety %>% 
    filter(location=="Southfield") %>%
    ggplot(aes(x = year, fill = factor(disengaged))) +
    geom_bar(position = "fill")
 
# Test whether one year had significantly more disengaged employees
southSafety <- safety %>% 
    filter(location=="Southfield")
chisq.test(southSafety$disengaged, southSafety$year)


# Filter out Southfield
other_locs <- safety %>% 
  filter(location != "Southfield")

# Test whether one year had significantly more overtime hours worked
t.test(overtime_hours ~ year, data = other_locs) 

# Test whether one year had significantly more disengaged employees
chisq.test(other_locs$year, other_locs$disengaged)


# Use multiple regression to test the impact of year and disengaged on accident rate in Southfield
regression <- glm(had_accident ~ year + disengaged, family = "binomial", data = southSafety)

# Examine the output
regression %>% broom::tidy()

```
  
  
  
***
  
###_Supervised Learning in R: Case Studies_  
  
Chapter 1 - Cars Data  
  
Making predictions using machine learning:  
  
* Course focuses on applied skills from predictive learning, using regression and classification as well as EDA  
	* Regression tends to be for predicting continuous, numeric variables  
    * Classification tends to be for predicting categorical variables  
* Case studies include 1) fuel efficiency, 2) Stack Overflow developer survey, 3) voter turnout, and 4) ages of nuns  
* The fuel efficiency data is stored in cars2018 and is based on data from the US Department of Energy  
	* Variables names with spaces can be handled by surrounding them with backticks  
    * Tidyverse includes tibble, readr, ggplot2, dplyr, tidyr, purrr, etc. - can be loaded as a package using library(tidyverse)  
  
Getting started with caret:  
  
* The caret package is useful for predictive modeling - full process including the test/train split for the raw dataset  
	* in_train <- createDataPartition(cars_vars$Aspiration, p = 0.8, list = FALSE)  # will stratify on 'Aspiration' variable  
    * training <- cars_vars[in_train,]  
    * testing <- cars_vars[-in_train,]  
* Can then train the model using only the training dataset  
	* fit_lm <- train(log(MPG) ~ ., method = "lm", data=training, trControl=trainControl(method = "none"))  
    * Can then use the yardstick package to assess the quality of the model  
  
Sampling data:  
  
* Bootstrap resampling means sampling with replacement, and then fitting on the resampled dataset (run multiple times)  
	* cars_rf_bt <- train(log(MPG) ~ ., method = "rf", data = training, trControl = trainControl(method = "boot"))  # default 25 resamples  
    * Can both visualize the models and assess the model statistically  
  
Example code includes:  
```{r cache=TRUE}

cars2018 <- readr::read_csv("./RInputFiles/cars2018.csv")
str(cars2018, give.attr = FALSE)
summary(cars2018)

# Print the cars2018 object
cars2018

# Plot the histogram
ggplot(cars2018, aes(x = MPG)) +
    geom_histogram(bins = 25) +
    labs(y = "Number of cars",
         x = "Fuel efficiency (mpg)")


# Deselect the 2 columns to create cars_vars
cars_vars <- cars2018 %>%
    select(-Model, -`Model Index`)

# Fit a linear model
fit_all <- lm(MPG ~ ., data = cars_vars)

# Print the summary of the model
summary(fit_all)


# Load caret
library(caret)

# Split the data into training and test sets
set.seed(1234)
in_train <- createDataPartition(cars_vars$Transmission, p = 0.8, list = FALSE)
training <- cars_vars[in_train, ]
testing <- cars_vars[-in_train, ]

# Train a linear regression model
fit_lm <- train(log(MPG) ~ ., method = "lm", data = training,
                trControl = trainControl(method = "none"))

# Print the model object
fit_lm


# Train a random forest model
fit_rf <- train(log(MPG) ~ ., method = "rf", data = training,
                trControl = trainControl(method = "none"))

# Print the model object
fit_rf


# Create the new columns
results <- training %>%
    mutate(`Linear regression` = predict(fit_lm, training),
           `Random forest` = predict(fit_rf, training))

# Evaluate the performance
yardstick::metrics(results, truth = MPG, estimate = `Linear regression`)
yardstick::metrics(results, truth = MPG, estimate = `Random forest`)


# Create the new columns
results <- testing %>%
    mutate(`Linear regression` = predict(fit_lm, testing),
           `Random forest` = predict(fit_rf, testing))

# Evaluate the performance
yardstick::metrics(results, truth = MPG, estimate = `Linear regression`)
yardstick::metrics(results, truth = MPG, estimate = `Random forest`)


# Fit the models with bootstrap resampling
cars_lm_bt <- train(log(MPG) ~ ., method = "lm", data = training,
                   trControl = trainControl(method = "boot"))
cars_rf_bt <- train(log(MPG) ~ ., method = "rf", data = training,
                   trControl = trainControl(method = "boot"))
                   
# Quick look at the models
cars_lm_bt
cars_rf_bt


results <- testing %>%
    mutate(`Linear regression` = predict(cars_lm_bt, testing),
           `Random forest` = predict(cars_rf_bt, testing))

yardstick::metrics(results, truth = MPG, estimate = `Linear regression`)
yardstick::metrics(results, truth = MPG, estimate = `Random forest`)

results %>%
    gather(Method, Result, `Linear regression`:`Random forest`) %>%
    ggplot(aes(log(MPG), Result, color = Method)) +
    geom_point(size = 1.5, alpha = 0.5) +
    facet_wrap(~Method) +
    geom_abline(lty = 2, color = "gray50") +
    geom_smooth(method = "lm")

```
  
  
  
***
  
Chapter 2 - Stack Overflow Developer Data  
  
Essential copying and pasting from Stack Overflow (largest and most trusted developer community):  
  
* Annual survey of developer perspectives on Stack Overflow - can be used for predictive modeling  
* Data is made available publicly at insights.stackoverflow.com/survey  
* Key question is "what makes a developer more likely to work remotely" (size of company, geography of employee, etc.)  
	* Data are calss imbalanced, with many more Non-Remote employees than Remote employees  
    * Best first step is the simplest model - logit, without any tricks  
    * simple_glm <- stackoverflow %>% select(-Respondent) %>% glm(Remote ~ ., family = "binomial", + data = .)  # Remote ~ . Means "all variables" while data=. Means from the piped dataset  
  
Dealing with imbalanced data:  
  
* Class imbalance is a common problem that can negatively impact model performance  
	* This dataset has 10x the number of non-remote, which can influence models to just start predicting non-remote in all cases  
* One approach to class imbalance is upsampling, basically running resampling with replacement on the small class until it is the same size as the large class  
	* Simple to implement, but with the risk of over-fitting  
    * up_train <- upSample(x = select(training, -Remote), y = training$Remote, yname = "Remote") %>% as_tibble()  
    * stack_glm <- train(Remote ~ ., method = "glm", family = "binomial", data = training, trControl = trainControl(method = "boot", sampling = "up"))  
  
Predicting remote status:  
  
* Classification models can include logistic regression and random forests  
	* stack_glm <- train(Remote ~ ., method = "glm", family = "binomial", data = training, trControl = trainControl(method = "boot", sampling = "up"))  
    * stack_rf <- train(Remote ~ ., method = "rf", data = training, trControl = trainControl(method = "boot", sampling = "up"))  
* Classification models can be evaluated using the confusion matrix  
	* confusionMatrix(predict(stack_glm, testing), testing$Remote)  
    * yardstick::accuracy(testing_results, truth = Remote, estimate = `Logistic regression`)  
    * yardstick::ppv(testing_results, truth = Remote, estimate = `Logistic regression`)  
    * yardstick::npv(testing_results, truth = Remote, estimate = `Logistic regression`)  
  
Example code includes:  
```{r cache=TRUE}

stackoverflow <- readr::read_csv("./RInputFiles/stackoverflow.csv")
stackoverflow$Remote <- factor(stackoverflow$Remote, levels=c("Not remote", "Remote"))
str(stackoverflow, give.attr = FALSE)


# Print stackoverflow
stackoverflow

# First count for Remote
stackoverflow %>% 
    count(Remote, sort = TRUE)

# then count for Country
stackoverflow %>% 
    count(Country, sort = TRUE)


ggplot(stackoverflow, aes(x=Remote, y=YearsCodedJob)) +
    geom_boxplot() +
    labs(x = NULL,
         y = "Years of professional coding experience") 


# Build a simple logistic regression model
simple_glm <- stackoverflow %>%
        select(-Respondent) %>%
        glm(Remote ~ .,
            family = "binomial",
            data = .)

# Print the summary of the model
summary(simple_glm)


stack_select <- stackoverflow %>%
    select(-Respondent)

# Split the data into training and testing sets
set.seed(1234)
in_train <- caret::createDataPartition(stack_select$Remote, p=0.8, list = FALSE)
training <- stack_select[in_train,]
testing <- stack_select[-in_train,]


up_train <- caret::upSample(x = select(training, -Remote), y = training$Remote, yname = "Remote") %>%
    as_tibble()

up_train %>%
    count(Remote)


# Sub-sample to 5% of original
inUse <- sample(1:nrow(training), round(0.05*nrow(training)), replace=FALSE)
useTrain <- training[sort(inUse), ]

# Build a logistic regression model
stack_glm <- caret::train(Remote ~ ., method="glm", family="binomial", data = training, 
                          trControl = trainControl(method = "boot", sampling = "up")
                          )

# Print the model object 
stack_glm


# Build a random forest model
stack_rf <- caret::train(Remote ~ ., method="rf", data = useTrain, 
                         trControl = trainControl(method = "boot", sampling="up")
                         )

# Print the model object
stack_rf


# Confusion matrix for logistic regression model
caret::confusionMatrix(predict(stack_glm, testing), testing$Remote)

# Confusion matrix for random forest model
caret::confusionMatrix(predict(stack_rf, testing), testing$Remote)


# Predict values
testing_results <- testing %>%
    mutate(`Logistic regression` = predict(stack_glm, testing), `Random forest` = predict(stack_rf, testing))

## Calculate accuracy
yardstick::accuracy(testing_results, truth = Remote, estimate = `Logistic regression`)
yardstick::accuracy(testing_results, truth = Remote, estimate = `Random forest`)

## Calculate positive predict value
yardstick::ppv(testing_results, truth = Remote, estimate = `Logistic regression`)
yardstick::ppv(testing_results, truth = Remote, estimate = `Random forest`)

```
  
  
  
***
  
Chapter 3 - Voting  
  
Predicting voter turnout from survey data:  
  
* Survey data available from https://www.voterstudygroup.org/publications/2016-elections/data  
	* Opinions about political and economic topics  
    * Includes whether the voter turned out (voted), based on self-reporting, in the 2016 election  
    * Data are coded as integers, requiring a data dictionary to map the questions and responses to what they mean  
  
Vote 2016:  
  
* Exploratory data analysis will help with learning about the underlying dataset  
	* There are differences on many of the individual dimensions between voters and non-voters  
    * A good first step can be to start with the very simplest model, Dependent ~ .  
  
Cross-validation is the process of sub-dividing the data into folds, with each fold used once as the validation set:  
  
* Allows for more accurate estimates of model performance on out-of-sample error  
* Each process of CV will work through the data k times (assuming there are k folds)  
	* Repeated CV is the process of running CV multiple times (this is particularly well suited to parallel processing)  
  
Comparing model performance:  
  
* Random forest models tend to be more powerful and capable of classifying the training data (and thus subject to risk of overfits and associated poor quality of test set predictions)  
  
Example code includes:  
```{r cache=TRUE}

voters <- readr::read_csv("./RInputFiles/voters.csv")
voters$turnout16_2016 <- factor(voters$turnout16_2016, levels=c("Did not vote", "Voted"))
str(voters, give.attr = FALSE)

# Print voters
voters

# How many people voted?
voters %>%
    count(turnout16_2016)


# How do the reponses on the survey vary with voting behavior?
voters %>%
    group_by(turnout16_2016) %>%
    summarize(`Elections don't matter` = mean(RIGGED_SYSTEM_1_2016 <= 2),
              `Economy is getting better` = mean(econtrend_2016 == 1),
              `Crime is very important` = mean(imiss_a_2016 == 2))


## Visualize difference by voter turnout
voters %>%
    ggplot(aes(econtrend_2016, ..density.., fill = turnout16_2016)) +
    geom_histogram(alpha = 0.5, position = "identity", binwidth = 1) +
    labs(title = "Overall, is the economy getting better or worse?")


# Remove the case_indetifier column
voters_select <- voters %>%
        select(-case_identifier)

# Build a simple logistic regression model
simple_glm <- glm(turnout16_2016 ~ .,  family = "binomial", 
                  data = voters_select)

# Print the summary                  
summary(simple_glm)


# Split data into training and testing sets
set.seed(1234)
in_train <- caret::createDataPartition(voters_select$turnout16_2016, p = 0.8, list = FALSE)
training <- voters_select[in_train, ]
testing <- voters_select[-in_train, ]


# Perform logistic regression with upsampling and no resampling
vote_glm_1 <- caret::train(turnout16_2016 ~ ., method = "glm", family = "binomial", data = training,
                           trControl = trainControl(method = "none", sampling = "up")
                           )

# Print vote_glm
vote_glm_1


useSmall <- sort(sample(1:nrow(training), round(0.1*nrow(training)), replace=FALSE))
trainSmall <- training[useSmall, ]

# Logistic regression
vote_glm <- caret::train(turnout16_2016 ~ ., method = "glm", family = "binomial", data = trainSmall,
                         trControl = trainControl(method = "repeatedcv", repeats = 2, sampling = "up")
                         )

# Print vote_glm
vote_glm


# Random forest
vote_rf <- caret::train(turnout16_2016 ~ ., method = "rf", data = trainSmall,
                        trControl = trainControl(method="repeatedcv", repeats=2, sampling = "up")
                        )

# Print vote_rf
vote_rf


# Confusion matrix for logistic regression model on training data
caret::confusionMatrix(predict(vote_glm, trainSmall), trainSmall$turnout16_2016)

# Confusion matrix for random forest model on training data
caret::confusionMatrix(predict(vote_rf, trainSmall), trainSmall$turnout16_2016)

# Confusion matrix for logistic regression model on testing data
caret::confusionMatrix(predict(vote_glm, testing), testing$turnout16_2016)

# Confusion matrix for random forest model on testing data
caret::confusionMatrix(predict(vote_rf, testing), testing$turnout16_2016)

```
  
  
  
***
  
Chapter 4 - Nuns  
  
Catholic sisters survey from 1967 - https://curate.nd.edu/show/0r967368551 with codebook at https://curate.nd.edu/downloads/0v838051f6x	 
	
* Responses from 130,000 sisters in ~400 congergations  
* There was significant change occuring during this time period, both in society at large and within the community of nuns  
* Age has been binned in groups of 10 years (has been recoded as a numeric at the top of the range, so 20 will mean 11-20 and 30 will mean 21-30 and the like)  
* Historical dataset, centered in the context of nuns in 1967  
* Good first step is to tidy the data, so that it is easier for exploratory data analysis  
	* sisters67 %>% select(-sister) %>% gather(key, value, -age)  
  
Exploratory data analysis with tidy data:  
  
* Easy to see levels of agreement (overall) using dplyr::count()  
* Agreement with specific questions by age  
	* tidy_sisters %>% filter(key %in% paste0("v", 153:170)) %>% group_by(key, value) %>% summarise(age = mean(age)) %>% ggplot(aes(value, age, color = key)) + geom_line(alpha = 0.5, size = 1.5) + geom_point(size = 2) + facet_wrap(~key)  
	* Can use the mix of responses to make estimates about the ages of the nuns  
* Data will be split in to training, validation, and test sets  
	* The validation set will be used for model selection  
  
Predicting age with supervised learning:  
  
* "rpart" - building a tree-based (CART) model  
* "xgbLinear" - extreme gradient boosting  
* "gbm" - gradient boosted ensembles  
* Validation datasets are useful for assessing hyper-parameters and model choices, leaving the test dataset pure for a final out-of-sample error estimate  
  
Wrap up:  
  
* Train-Validation-Test to select the best models, tune the parameters, and estimate the out-of-sample error rates  
* Dealing with class imbalances; improving performance with resamples (bootstraps, cross-validation, etc.)  
* Hyper-parameter tuning can be valuable, but the time investment in other areas can often generate a greater return  
* Gradient boosting and random forests tend to perform very well, but there is always value in trying out multiple models  
	* Start with EDA and begin with a very simple model  
  
Example code includes:  
```{r cache=TRUE}

sisters67 <- readr::read_csv("./RInputFiles/sisters.csv")
str(sisters67, give.attr = FALSE)


# View sisters67
glimpse(sisters67)

# Plot the histogram
ggplot(sisters67, aes(x = age)) +
    geom_histogram(binwidth = 10)


# Tidy the data set
tidy_sisters <- sisters67 %>%
    select(-sister) %>%
    gather(key, value, -age)

# Print the structure of tidy_sisters
glimpse(tidy_sisters)


# Overall agreement with all questions varied by age
tidy_sisters %>%
    group_by(age) %>%
    summarize(value = mean(value, na.rm = TRUE))

# Number of respondents agreed or disagreed overall
tidy_sisters %>%
    count(value)


# Visualize agreement with age
tidy_sisters %>%
    filter(key %in% paste0("v", 153:170)) %>%
    group_by(key, value) %>%
    summarize(age = mean(age, na.rm = TRUE)) %>%
    ggplot(aes(value, age, color = key)) +
    geom_line(show.legend = FALSE) +
    facet_wrap(~key, nrow = 3)


# Remove the sister column
sisters_select <- sisters67 %>% 
    select(-sister)

# Build a simple linear regression model
simple_lm <- lm(age ~ ., 
                data = sisters_select)

# Print the summary of the model
summary(simple_lm)


# Split the data into training and validation/test sets
set.seed(1234)
in_train <- caret::createDataPartition(sisters_select$age, p = 0.6, list = FALSE)
training <- sisters_select[in_train, ]
validation_test <- sisters_select[-in_train, ]

# Split the validation and test sets
set.seed(1234)
in_test <- caret::createDataPartition(validation_test$age, p = 0.5, list = FALSE)
testing <- validation_test[in_test, ]
validation <- validation_test[-in_test, ]


# Fit a CART model
sisters_cart <- caret::train(age ~ ., method = "rpart", data = training)

# Print the CART model
sisters_cart


inSmall <- sample(1:nrow(training), 500, replace=FALSE)
smallSisters <- training[sort(inSmall), ]

sisters_xgb <- caret::train(age ~ ., method = "xgbTree", data = smallSisters)
sisters_gbm <- caret::train(age ~ ., method = "gbm", data = smallSisters, verbose=FALSE)

# Make predictions on the three models
modeling_results <- validation %>%
    mutate(CART = predict(sisters_cart, validation),
           XGB = predict(sisters_xgb, validation),
           GBM = predict(sisters_gbm, validation))

# View the predictions
modeling_results %>% 
    select(CART, XGB, GBM)


# Compare performace
yardstick::metrics(modeling_results, truth = age, estimate = CART)
yardstick::metrics(modeling_results, truth = age, estimate = XGB)
yardstick::metrics(modeling_results, truth = age, estimate = GBM)


# Calculate RMSE
testing %>%
    mutate(prediction = predict(sisters_gbm, testing)) %>%
    yardstick::rmse(truth = age, estimate = prediction)

```
  
  
  
***
  
###_Business Process Analytics in R_  
  
Chapter 1 - Introduction to Process Analysis  
  
Introduction and overview:  
  
* Efficient processes are core to many businesses, and improved data makes further analysis possible  
* The "internet of things" has created significant amounts of event data - why, what, and who  
	* Why is the purpose  
    * What is the steps in the process  
    * Who is the person responsible for the activity (can be machines or IS or the like; referred to as "resources")  
* Process workflow is iterative across Extraction-Processing-Analysis  
  
Activities as cornerstones of processes:  
  
* Data from an online learning platform; activities are captured and can be used for further analysis  
* Activities describe the flow of the process, and are one of the most important components of the process  
	* bupaR::activities_labels() is like names() for activities data  
    * bupaR::activities() is like summary() for activities data  
* Each case is described by the sequence of activities, known as its "trace"  
	* bupaR::traces() will create a frequency table of the traces  
    * bupaR::trace_explorer() will visualize the cases  
  
Components of process data:  
  
* Cases are the objects flowing through the process, while activities are the actions performed on them  
	* An activity instance is the occurrence of an activity (which can be a series of events) - specific action, case, time, etc.  
    * The "lifecycle status" is an area like Scheduled, Started, Completed, and the like  
    * The "event log" is the journal of the events  
    * The "resources" are the actors in the process  
* Can create an event log using the eventlog() function  
	* event_data %>% eventlog(case_id = "patient", activity_id = "handling", activity_instance_id = "handling_id", timestamp = "time", lifecycle_id = "registration_type", resource = "employee")  
  
Example code includes:  
```{r}

# Load the processmapR package using library
library(processmapR)
library(bupaR)


handling <- c('Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Blood test', 'Blood test', 'Blood test', 'Blood test', 'Blood test', 'Blood test', 'Blood test', 'Blood test', 'MRI SCAN', 'MRI SCAN', 'MRI SCAN', 'MRI SCAN', 'MRI SCAN', 'MRI SCAN', 'MRI SCAN', 'MRI SCAN', 'X-Ray', 'X-Ray', 'X-Ray', 'X-Ray', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Registration', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Triage and Assessment', 'Blood test', 'Blood test', 'Blood test', 'Blood test', 'Blood test', 'Blood test', 'Blood test', 'Blood test', 'MRI SCAN', 'MRI SCAN', 'MRI SCAN', 'MRI SCAN', 'MRI SCAN', 'MRI SCAN', 'MRI SCAN', 'MRI SCAN', 'X-Ray', 'X-Ray', 'X-Ray', 'X-Ray', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Discuss Results', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out', 'Check-out')
patient <- c('43', '155', '156', '170', '172', '184', '221', '278', '348', '420', '455', '493', '43', '155', '156', '170', '172', '184', '221', '278', '348', '420', '455', '493', '43', '156', '170', '172', '184', '278', '348', '420', '43', '156', '170', '172', '184', '278', '348', '420', '155', '221', '455', '493', '43', '155', '156', '170', '172', '184', '221', '278', '348', '420', '455', '493', '43', '155', '156', '170', '172', '184', '221', '278', '348', '420', '455', '493', '43', '155', '156', '170', '172', '184', '221', '278', '348', '420', '455', '493', '43', '155', '156', '170', '172', '184', '221', '278', '348', '420', '455', '493', '43', '156', '170', '172', '184', '278', '348', '420', '43', '156', '170', '172', '184', '278', '348', '420', '155', '221', '455', '493', '43', '155', '156', '170', '172', '184', '221', '278', '348', '420', '455', '493', '43', '155', '156', '170', '172', '184', '221', '278', '348', '420', '455', '493')
employee <- c('r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r3', 'r3', 'r3', 'r3', 'r3', 'r3', 'r3', 'r3', 'r4', 'r4', 'r4', 'r4', 'r4', 'r4', 'r4', 'r4', 'r5', 'r5', 'r5', 'r5', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r1', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r2', 'r3', 'r3', 'r3', 'r3', 'r3', 'r3', 'r3', 'r3', 'r4', 'r4', 'r4', 'r4', 'r4', 'r4', 'r4', 'r4', 'r5', 'r5', 'r5', 'r5', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r6', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7', 'r7')
handling_id <- c('43', '155', '156', '170', '172', '184', '221', '278', '348', '420', '455', '493', '543', '655', '656', '670', '672', '684', '721', '778', '848', '920', '955', '993', '1020', '1072', '1081', '1082', '1088', '1127', '1163', '1199', '1257', '1309', '1318', '1319', '1325', '1364', '1400', '1436', '1557', '1587', '1710', '1730', '1777', '1889', '1890', '1904', '1906', '1918', '1955', '2012', '2082', '2154', '2189', '2227', '2272', '2384', '2385', '2399', '2401', '2413', '2450', '2507', '2577', '2649', '2684', '2720', '43', '155', '156', '170', '172', '184', '221', '278', '348', '420', '455', '493', '543', '655', '656', '670', '672', '684', '721', '778', '848', '920', '955', '993', '1020', '1072', '1081', '1082', '1088', '1127', '1163', '1199', '1257', '1309', '1318', '1319', '1325', '1364', '1400', '1436', '1557', '1587', '1710', '1730', '1777', '1889', '1890', '1904', '1906', '1918', '1955', '2012', '2082', '2154', '2189', '2227', '2272', '2384', '2385', '2399', '2401', '2413', '2450', '2507', '2577', '2649', '2684', '2720')
registration_type <- c('start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'start', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete', 'complete')
rTime <- c('2017-02-19 04:38:51', '2017-06-03 10:05:28', '2017-06-03 10:05:28', '2017-06-17 15:10:30', '2017-06-17 23:00:33', '2017-06-27 07:48:22', '2017-08-03 17:05:27', '2017-09-26 20:22:49', '2017-11-24 08:28:44', '2018-02-08 03:39:21', '2018-03-14 21:04:28', '2018-04-29 04:55:10', '2017-02-19 07:28:53', '2017-06-04 06:27:00', '2017-06-03 13:23:14', '2017-06-17 16:31:58', '2017-06-18 18:29:13', '2017-06-28 00:14:50', '2017-08-04 07:22:06', '2017-09-27 22:57:03', '2017-11-24 10:33:00', '2018-02-08 17:33:12', '2018-03-15 15:12:41', '2018-04-30 19:40:22', '2017-02-20 19:59:18', '2017-06-04 15:18:50', '2017-06-18 22:51:07', '2017-06-21 02:43:27', '2017-07-01 23:55:10', '2017-09-28 22:58:23', '2017-11-25 12:06:18', '2018-02-12 09:01:38', '2017-02-21 06:49:49', '2017-06-04 23:23:28', '2017-06-19 06:44:30', '2017-06-21 11:16:30', '2017-07-02 11:16:08', '2017-09-29 07:28:10', '2017-11-25 21:54:56', '2018-02-12 19:43:42', '2017-06-05 00:12:24', '2017-08-05 08:25:17', '2018-03-17 10:30:24', '2018-05-02 07:32:45', '2017-02-21 14:50:43', '2017-06-05 14:03:19', '2017-06-05 10:26:16', '2017-06-19 22:46:10', '2017-06-22 04:39:35', '2017-07-03 01:28:49', '2017-08-05 22:06:23', '2017-09-29 19:13:51', '2017-11-26 06:52:23', '2018-02-17 02:44:58', '2018-03-18 00:20:51', '2018-05-02 18:14:11', '2017-02-24 14:58:43', '2017-06-05 15:58:53', '2017-06-05 15:58:53', '2017-06-20 03:48:37', '2017-06-22 08:40:55', '2017-07-03 03:39:51', '2017-08-08 23:17:45', '2017-09-29 21:16:01', '2017-11-27 04:56:53', '2018-02-20 09:49:29', '2018-03-18 08:12:07', '2018-05-03 00:11:10', '2017-02-19 07:28:53', '2017-06-03 14:19:00', '2017-06-03 13:23:14', '2017-06-17 16:31:58', '2017-06-18 01:07:42', '2017-06-27 12:22:51', '2017-08-03 19:25:12', '2017-09-26 22:17:18', '2017-11-24 10:33:00', '2018-02-08 06:01:38', '2018-03-15 00:34:01', '2018-04-29 07:39:14', '2017-02-19 21:58:08', '2017-06-04 14:23:26', '2017-06-04 06:27:00', '2017-06-18 04:14:55', '2017-06-19 00:40:19', '2017-06-28 12:48:20', '2017-08-04 21:09:17', '2017-09-28 12:00:12', '2017-11-25 00:44:30', '2018-02-09 07:05:52', '2018-03-16 04:08:03', '2018-05-01 10:37:51', '2017-02-21 03:12:26', '2017-06-04 19:35:51', '2017-06-19 03:01:11', '2017-06-21 08:02:20', '2017-07-02 07:43:48', '2017-09-29 04:58:49', '2017-11-25 18:30:43', '2018-02-12 13:57:13', '2017-02-21 09:57:05', '2017-06-05 02:46:59', '2017-06-19 11:40:53', '2017-06-21 16:09:26', '2017-07-02 16:03:16', '2017-09-29 12:44:39', '2017-11-26 02:40:30', '2018-02-12 23:53:46', '2017-06-05 04:39:38', '2017-08-05 13:56:39', '2018-03-17 14:09:40', '2018-05-02 12:24:41', '2017-02-21 17:57:58', '2017-06-05 15:58:53', '2017-06-05 14:03:19', '2017-06-20 01:44:29', '2017-06-22 08:40:55', '2017-07-03 03:39:51', '2017-08-05 23:53:27', '2017-09-29 21:16:01', '2017-11-26 09:44:37', '2018-02-17 06:17:57', '2018-03-18 03:22:17', '2018-05-02 21:17:12', '2017-02-24 16:03:49', '2017-06-05 17:22:16', '2017-06-05 17:15:30', '2017-06-20 05:36:40', '2017-06-22 10:59:58', '2017-07-03 05:00:48', '2017-08-09 00:13:39', '2017-09-29 23:42:48', '2017-11-27 06:53:23', '2018-02-20 12:04:00', '2018-03-18 10:48:34', '2018-05-03 02:11:42')
rOrder <- c(43, 155, 156, 170, 172, 184, 221, 278, 348, 420, 455, 493, 543, 655, 656, 670, 672, 684, 721, 778, 848, 920, 955, 993, 1020, 1072, 1081, 1082, 1088, 1127, 1163, 1199, 1257, 1309, 1318, 1319, 1325, 1364, 1400, 1436, 1557, 1587, 1710, 1730, 1777, 1889, 1890, 1904, 1906, 1918, 1955, 2012, 2082, 2154, 2189, 2227, 2272, 2384, 2385, 2399, 2401, 2413, 2450, 2507, 2577, 2649, 2684, 2720, 2764, 2876, 2877, 2891, 2893, 2905, 2942, 2999, 3069, 3141, 3176, 3214, 3264, 3376, 3377, 3391, 3393, 3405, 3442, 3499, 3569, 3641, 3676, 3714, 3741, 3793, 3802, 3803, 3809, 3848, 3884, 3920, 3978, 4030, 4039, 4040, 4046, 4085, 4121, 4157, 4278, 4308, 4431, 4451, 4498, 4610, 4611, 4625, 4627, 4639, 4676, 4733, 4803, 4875, 4910, 4948, 4993, 5105, 5106, 5120, 5122, 5134, 5171, 5228, 5298, 5370, 5405, 5441)

pFrame <- tibble(handling=factor(handling, levels=c('Blood test', 'Check-out', 'Discuss Results', 'MRI SCAN', 'Registration', 'Triage and Assessment', 'X-Ray')), 
                 patient=patient, 
                 employee=factor(employee, levels=c('r1', 'r2', 'r3', 'r4', 'r5', 'r6', 'r7')), 
                 handling_id=handling_id, 
                 registration_type=factor(registration_type, levels=c("complete", "start")), 
                 time=as.POSIXct(rTime), 
                 .order=rOrder
                 )

patients <- eventlog(pFrame,
    case_id = "patient",
    activity_id = "handling",
    activity_instance_id = "handling_id",
    lifecycle_id = "registration_type",
    timestamp = "time",
    resource_id = "employee")


# The function slice can be used to take a slice of cases out of the eventdata. slice(1:10) will select the first ten cases in the event log, where first is defined by the current ordering of the data.

# How many patients are there?
n_cases(patients)

# Print the summary of the data
summary(patients)

# Show the journey of the first patient
slice(patients, 1)


# How many distinct activities are there?
n_activities(patients)

# What are the names of the activities?
activity_labels(patients)

# Create a list of activities
activities(patients)


# Have a look at the different traces
traces(patients)

# How many are there?
n_traces(patients)

# Visualize the traces using trace_explorer
trace_explorer(patients, coverage=1)

# Draw process map
process_map(patients)


claims <- tibble(id=c("claim1", "claim1", "claim2", "claim2", "claim2"), 
                 action=c(10002L, 10011L, 10015L, 10024L, 10024L), 
                 action_type=c("Check Contract", "Pay Back Decision", "Check Contract", "Pay Back Decision", "Pay Back Decision"), 
                 date=as.Date(c("2008-01-12", "2008-03-22", "2008-01-13", "2008-03-23", "2008-04-14")), 
                 originator=c("Assistant 1", "Manager 2", "Assistant 6", "Manager 2", "Manager 2"), 
                 status=as.factor(c("start", "start", "start", "start", "complete"))
                 )
claims


#create eventlog claims_log 
claims_log <- eventlog(claims,
    case_id = "id",
    activity_id = "action_type",
    activity_instance_id = "action",
    lifecycle_id = "status",
    timestamp = "date",
    resource_id = "originator")

# Print summary
summary(claims_log)

# Check activity labels
activity_labels(claims_log)

# Once you have an eventlog, you can access its complete metadata using the function mapping or the functions case_id, activity_id etc., to inspect individual identifiers.

```
  
  
  
***
  
Chapter 2 - Analysis Techniques  
  
Organizational analysis:  
  
* Processes are always dependent on resources, even if automated (machines and algorithms can be resources)  
	* Who executes the task, how specialized is the knowledge, etc.  
    * resource_labels(log_hospital)  # will pull out the resources  
    * resources(log_hospital)  # will pull out frequencies by resource  
* Can create a resource-activity matrix  
	* A person who performs only a few activities is considered to be specialized in that activity  
    * If only one person ever performs a specific activity, then there is a high risk of "brain drain"  
    * The plot() function, applied to an event_log, will create the resource-activity matrix  
    * resource_map(log_hospital)  # shows arrows between the work flows  
  
Structuredness:  
  
* Control-flow refers to the succession of activities  
	* Each unique flow is referred to as a trace  
    * Metrics include entry/exit points, length of cases, presence of activities, rework, etc.  
    * log_healthcare %>% start_activities("activity") %>% plot()  
    * log_healthcare %>% end_activities("activity") %>% plot()  
* Rework is when the same activity is done multiple times for the same case  
	* Repetitions are when the activity is repeated after some intervening steps  
    * Sel-loops are when the activity is repeated immediately after itself  
* The precedence matrix shows the relationships between the activities in a more structured manner  
	* eventlog %>% precedence_matrix(type = "absolute") %>% plot  # can be type="relative" also  
  
Performance analysis:  
  
* Visuals can include performance process maps and dotted charts; metrics can include throughput time, processing time, idle time  
	* eventlog %>% process_map(type = frequency())  # normal process map  
    * eventlog %>% process_map(type = performance())  # performance process map  
* The dotted chart shows the freqency of activities over time; basically, a form of scatter plot  
	* throughput_time is total time, processing_time is the sum of activity time, idle_time is the sume of when nothing is happening  
  
Linking perspectives:  
  
* Granularity can help give the statistics at the desired levels  
	* <process_metric>(level = "log", "trace", "case", "activity", "resource", "resource-activity")  
* Categorical data can be leveraged using the group_by() functionality - each group will then be calculated separately  
	* eventlog %>% group_by(priority) %>% number_of_repetitions(level = "resource") %>% plot()  
  
Example code includes:  
```{r eval=FALSE}

# Print list of resources
resource_frequency(sepsis, level="resource")

# Number of resources per activity
resource_frequency(sepsis, level = "activity")

# Plot Number of executions per resource-activity
resource_frequency(sepsis, level = "resource-activity") %>% plot()


# Calculate resource involvement
resource_involvement(sepsis, level="resource")

# Show graphically 
sepsis %>% resource_involvement(level = "resource") %>% plot

# Compare with resource frequency
resource_frequency(sepsis, level="resource")


# Min, max and average number of repetitions
sepsis %>% number_of_repetitions(level = "log")

# Plot repetitions per activity
sepsis %>% number_of_repetitions(level = "activity") %>% plot

# Number of repetitions per resources
sepsis %>% number_of_repetitions(level = "resource")


# Create performance map
eat_patterns %>% process_map(type = performance(FUN = median, units = "hours"))

# Inspect variation in activity durations graphically
eat_patterns %>% processing_time(level = "activity") %>% plot()

# Draw dotted chart
eat_patterns %>% dotted_chart(x = "relative_day", sort = "start_day", units = "secs")


# Time per activity
daily_activities %>% processing_time(level = "activity") %>% plot

# Average duration of recordings
daily_activities %>% throughput_time(level="log", units = "hours")

# Missing activities
daily_activities %>% idle_time(level="log", units = "hours")


# Distribution throughput time
vacancies %>% throughput_time(units="days")

# Distribution throughput time per department
vacancies %>% group_by(vacancy_department) %>% throughput_time(units="days") %>% plot()

# Repetitions of activities
vacancies %>% number_of_repetitions(level = "activity") %>% arrange(-relative)

```
  
  
  
***
  
Chapter 3 - Event Data Processing  
  


















