---
title: "Data Camp Insights"
author: "davegoblue"
date: "June 1, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(tidyverse)

```

## Background and Overview  
DataCamp offer interactive courses related to R Programming.  While some is review, it is helpful to see other perspectives on material.  As well, DataCamp has some interesting materials on packages that I want to learn better (ggplot2, dplyr, ggvis, etc.).  This document summarizes a few key insights from:  

This document is currently split between _v003 and _v003_a and _v003_b and _v003_c due to the need to keep the number of DLL that it opens below the hard-coded maximum.  This introductory section needs to be re-written, and the contents consolidated, at a future date.
  
* R Programming (Introductory R, Intermediate R, Writing Functions in R, Object Oriented Programming in R, Introduction to Tidyverse)  
* Importing and Cleaning Data (Cleaning Data in R, Importing Data in to R)  
* Data Manipulation (dplyr, data.table, xts/zoo, dplyr joins, cases for EDA/Time Series/Pitch Data)  
* Data Visualization (base, ggplot2 parts I/II/III, ggvis, geospatial)  
* Statistics (8 refresher modules)  
* Machine Learning (3 modules + 1 text mining case)  
* R Studio (2 parts) and R Markdown (1 module)  
  
The original DataCamp_Insights_v001 and DataCamp_Insights_v002 documents have been split for this document:  
  
* This DataCamp_Insights_v003 document contains evolving sections on R Programming, Machine Learning, and RStudio / R Markdown  
* Importing and Cleaning Data components have been moved to DataCamp_ImportClean_v002  
* Data Manipulation components have been moved to DataCamp_DataManipulation_v002  
* Visualization components have been moved to DataCamp_Visualization_v002  
* Statistics components have been moved to DataCamp_Statistics_v002  
  
  
***
  

###_Hierarchical and Mixed Effects Models_  
  
Chapter 1 - Overview and Introduction  
  
What is a hierarchical model?  
  
* Hierarchical data is nested within itself, and can be analyzed using the lme package  
	* Example of students in a classroom - may not all be independent of each other due to teacher quality, building conditions, etc.  
    * Hierarchical models can help with pooling means across small sample sizes  
    * Repeated measurements (test scores each year) are also a common example of data that are not truly independent  
* Hierarchical models can include nested models and multi-level models  
* Regression frameworks can include pool information and random effects (vs. fixed effects) and mixed-effects and linear mixed-effects  
* Repeated sampling can have repeated measures modeling  
  
Parts of a regression:  
  
* Linear regression and linear model can be used interchangeably for this course - epsilon is the error term, assumed to be normal with zero mean and constant variance  
* The linear model in R is closely related to analysis of variance (ANOVA)  
	* lm (y ~ x, myData)  
    * anova( lm (y ~ x, myData) )  
* The most basic regression has an intercept, a slope, a single predictor, and an error term  
	* The concept can be extended to multiple regression with additional predictors  
* There are some limitations to the multiple regression approach  
	* Parameter estimates can be very sensitive to other variables - Simpson's paradox and the like  
    * Need to note that the regression coefficient is "after controlling for . . . " (all the other variables)  
    * Interaction terms can be important as well  
* Regressions in R for an intercept for every group are called as lm(y ~ x - 1)  
* The interaction term x1*x2 is the same as x1 + x2 + x1:x2  
  
Random effects in regression:  
  
* Nested relationships tend to be hierarchical in nature - students are part of classes are part of schools and the like  
	* Mathematically, this is referred to as a mapping among the distributions  
* The algebraic representation is that y ~ B*x + eps, with B ~ N(mu, sigma**2)  
	* library(lme4) is the best packages for this in R  
    * lme4::lmer(y ~ x + (1|randomGroup), data=myData)  
    * lme4::lmer(y ~ x + (randomSlope|randomGroup), data=myData)  
  
School data:  
  
* Appliciation of multi-level models to school data - influence of sex, teacher training, plotting parameter estmates  
  
Example code includes:  
```{r}

rawStudent <- read.csv("./RInputFiles/classroom.csv")

studentData <- rawStudent %>%
    mutate(sex=factor(sex, labels=c("male", "female")), minority=factor(minority, labels=c("no", "yes")))


# Plot the data
ggplot(data = studentData, aes(x = housepov, y = mathgain)) +
    geom_point() +
    geom_smooth(method = 'lm')

# Fit a linear model
summary( lm(mathgain ~ housepov , data = studentData))


# I have aggregated the data for you into two new datasets at the classroom- and school-levels (As a side note, if you want to learn how to aggregate data, the dplyr or data.table courses teach these skills)
# We will also compare the model outputs across all three outputs
# Note: how we aggregate the data is important
# I aggregated the data by taking the mean across the student data (in pseudo-code: mean(mathgain) by school or mean(mathgain) by classroom), 
# but another reasonable method for aggregating the data would be to aggregate by classroom first and school second

classData <- studentData %>%
    group_by(schoolid, classid) %>%
    summarize_at(vars(mathgain, mathprep, housepov, yearstea), mean, na.rm=TRUE)
str(classData)

schoolData <- studentData %>%
    group_by(schoolid) %>%
    summarize_at(vars(mathgain, mathprep, housepov, yearstea), mean, na.rm=TRUE)
str(schoolData)


# First, plot the hosepov and mathgain at the classroom-level from the classData data.frame
ggplot(data = classData, aes(x = housepov, y = mathgain)) +
    geom_point() +
    geom_smooth(method = 'lm')

# Second, plot the hosepov and mathgain at the school-level from the schoolData data.frame
ggplot(data = schoolData, aes(x = housepov, y = mathgain)) +
    geom_point() +
    geom_smooth(method = 'lm')

# Third, compare your liner regression results from the previous expercise to the two new models
summary( lm(mathgain ~ housepov, data = studentData)) ## student-level data
summary( lm(mathgain ~ housepov, data = classData)) ## class-level data
summary( lm(mathgain ~ housepov, data = schoolData)) ## school-level data


# Plot the means of your data, predictor is your x-variable, response is your y-variable, and intDemo is your data.frame
intDemo <- data.frame(predictor=factor(c(rep("a", 5), rep("b", 5), rep("c", 5))), 
                      response=c(-1.207, 0.277, 1.084, -2.346, 0.429, 5.759, 4.138, 4.18, 4.153, 3.665, 9.046, 8.003, 8.447, 10.129, 11.919)
                      )
str(intDemo)


ggIntDemo <- ggplot(intDemo, aes(x = predictor, y = response) ) +
    geom_point() +
    theme_minimal() + stat_summary(fun.y = "mean", color = "red",
                                   size = 3, geom = "point") +
    xlab("Intercept groups")
print(ggIntDemo)

# Fit a linear model to your data where response is "predicted by"(~) predictor
intModel <- lm( response ~ predictor - 1 , data = intDemo)
summary(intModel)


extractAndPlotResults <- function(intModel){
    intCoefPlot <- broom::tidy(intModel)
    intCoefPlot$term <- factor(gsub("predictor", "", intCoefPlot$term))

    plotOut <- ggIntDemo + geom_point(data = intCoefPlot,
                           aes(x = term, y = estimate),
                           position = position_dodge(width = 0.4),
                           color = 'blue', size = 8, alpha = 0.25)
    print(plotOut)
}


# Run the next code that extracts out the model's coeffiecents and plots them 
extractAndPlotResults(intModel)


multIntDemo <- data.frame(group=factor(c(rep("a", 5), rep("b", 5), rep("c", 5))), 
                          x=rep(0:4, times=3), 
                          intercept=c(4.11, -1.69, 1.09, 1.9, 1.21, 4.63, 10.29, 4.67, 12.06, 4.78, 15.22, 19.15, 4.44, 8.88, 9.47), 
                          response=c(4.11, 2.31, 9.09, 13.9, 17.21, 4.63, 14.29, 12.67, 24.06, 20.78, 15.22, 23.15, 12.44, 20.88, 25.47)
                          )
str(multIntDemo)

plot_output1 <- function(out1){
    ggmultIntgDemo <- ggplot( multIntDemo, aes(x = x, y = response) ) +
        geom_point(aes(color = group)) +
        theme_minimal() +
        scale_color_manual(values = c("blue", "black", "red")) +
        stat_smooth(method = 'lm', fill = NA, color = 'orange', size = 3)
    print(ggmultIntgDemo)
}

plot_output2 <- function(out2){
    out2Tidy <- broom::tidy(out2)
    out2Tidy$term <- gsub("group", "", out2Tidy$term)
    out2Plot <- data.frame(group = out2Tidy[ -1, 1],
                           slope = out2Tidy[ 1, 2],
                           intercept = out2Tidy[ -1, 2])
    ggmultIntgDemo2 <- ggplot( multIntDemo, aes(x = x, y = response) ) +
        geom_point(aes(color = group))+
        theme_minimal() +
        scale_color_manual(values = c("blue", "black", "red")) +
        geom_abline(data = out2Plot,
                    aes(intercept = intercept, slope = slope, color = group))
    print(ggmultIntgDemo2)
}

plot_output3 <- function(out3){
    ggmultIntgDemo3 <- ggplot( multIntDemo, aes(x = x, y = response) ) +
        geom_point(aes(color = group)) +
        theme_minimal() +
        scale_color_manual(values = c("blue", "black", "red")) +
        stat_smooth(method = 'lm', aes(color = group), fill = NA)
    print(ggmultIntgDemo3)
}

# First, run a model without considering different intercept for each group
out1 <- lm( response ~ x, data=multIntDemo )
summary(out1)
plot_output1(out1)

# Considering same slope but different intercepts
out2 <- lm( response ~ x + group - 1, data=multIntDemo )
summary(out2)
plot_output2(out2)

# Consdering different slope and intercept for each group (i.e., an interaction)
out3 <- lm( response ~ x + group - 1 + x:group, multIntDemo)
summary(out3)
plot_output3(out3)


multIntDemo$intercept <- c(-0.87, 3.35, 1.25, 0.88, -1.05, 4.55, 1.22, 3.34, 1.26, 3.75, 7.71, 9.59, 2.28, 1.9, 13.35)
multIntDemo$response <- c(-0.87, 6.35, 7.25, 9.88, 10.95, 4.55, 4.22, 9.34, 10.26, 15.75, 7.71, 12.59, 8.28, 10.9, 25.35)

# Run model
outLmer <- lme4::lmer( response ~ x + ( 1 | group), multIntDemo)

# Look at model outputs 
summary( outLmer )
broom::tidy( outLmer )


extractAndPlotOutput <- function(outLmer, slope=3){
    multIntDemo$lmerPredict <- predict(outLmer)
    ggmultIntgDemo2 <- ggplot( multIntDemo, aes(x = x, y = response) ) +
        geom_point(aes(color = group))+
        theme_minimal() +
        scale_color_manual(values = c("blue", "black", "red")) +
        geom_abline(data = multIntDemo,
                    aes(intercept = intercept, slope = slope, color = group))
    outPlot <-  ggmultIntgDemo2 +
                geom_line( data =  multIntDemo,
                      aes(x = x, y = lmerPredict, color = group),
                      linetype = 2)
    print(outPlot)
}


# Extract predictor variables and plot
extractAndPlotOutput(outLmer)


# Random effect slopes
multIntDemo$response <- c(-0.72, 1.5, 4.81, 6.61, 13.62, 10.21, 9.64, 11.91, 16.39, 16.97, 8.76, 14.79, 15.83, 15.27, 17.36)
multIntDemo$intercept <- c(-0.72, -1.5, -1.19, -2.39, 1.62, 10.21, 6.64, 5.91, 7.39, 4.97, 8.76, 11.79, 9.83, 6.27, 5.36)

outLmer2 <- lme4::lmer( response ~ ( x|group ), multIntDemo)
summary(outLmer2)
broom::tidy(outLmer2)


plotOutput <- function(outLmer2){
    multIntDemo$lmerPredict2 <- predict(outLmer2)
    ggmultIntgDemo3 <- ggplot( multIntDemo, aes(x = x, y = response) ) +
        geom_point(aes(color = group)) +
        theme_minimal() +
        scale_color_manual(values = c("blue", "black", "red")) +
        stat_smooth(method = 'lm', aes(color = group), fill = NA)
    plotOut <- ggmultIntgDemo3 +
            geom_line( data =  multIntDemo,
                      aes(x = x, y = lmerPredict2, color = group),
                      linetype = 2)
    print(plotOut)
}


# Extract and plot
plotOutput(outLmer2)


# Mixed effect model
lmerModel <- lme4::lmer(mathgain ~ sex + 
                  mathprep + mathknow + (1|classid) +
                  (1|schoolid), data = studentData, na.action = "na.omit",
                  REML = TRUE)
summary(lmerModel)


extractAndPlot <- function(lmerModel){
    modelOutPlot <- broom::tidy(lmerModel, conf.int = TRUE)
    modelOutPlot <- modelOutPlot[ modelOutPlot$group =="fixed" &
                               modelOutPlot$term != "(Intercept)", ]
    plotOut <- ggplot(modelOutPlot, aes(x = term, y = estimate,
                             ymin = conf.low,
                             ymax = conf.high)) +
            theme_minimal() +
            geom_hline(yintercept = 0.0, color = 'red', size = 2.0) +
            geom_point() +
            geom_linerange() + coord_flip()
    print(plotOut)
}


# Extract and plot 
extractAndPlot(lmerModel)

```
  
  
  
***
  
Chapter 2 - Linear Mixed-Effect Models  
  
Linear mixed effect model - Birth rates data:  
  
* Small populations are highly sensitive to stochastic effects - if the mean is 1, a group of 5 might have 0 or 10  
* Questions about how counties may impact birth rates, over and above other demographic factors  
	* Example of plotting birth rate vs. county - will see both the highest and lowest birth rates in the smallest counties  
* Random effect syntax for the lmer model includes  
	* (1 | group) - random intercept with fixed mean  
    * (1 | g1/g2) - intercepts vary among g1 and g2 within g2  
    * (1 | g1) + (1 | g2) - random intercepts for two variables  
    * x + (x | g) - correlated random slope and intercept  
    * x + (x || g) - uncorrelated random slope and intercept  
    * See lme4 documentation for additional details  
  
Understanding and reporting the outputs of lmer:  
  
* The output from lmer is similar to the output from lm, but with some key differences - if using print(), will see  
	* The method used is REML - restricted maximum likelihood - which tends to solve better than maximum likelihood for these problems  
    * There is an REML convergence criteria, which can be a helpful diagnostic  
    * Can see the standard deviations for both the state and the residual, along with the number of observations  
    * Get the fixed effects coefficients in a similar form as lm()  
* The summary() call on lmer produces several additional outputs  
	* Residuals summary  
    * Fixed effects estimates include SE and t-values (but not p-values)  
    * Correlations of fixed effects  
* Can grab only the fixed effects using fixef(myLMERObject)  
	* Can grab only the random effects using ranef(myLMERObject), though these will not have confidence intervals  
    * The random effects confidence intervals could be estimated using bootstrapping or Bayesian methods per the author of lme4 - but actual random effects are just unobserved random variables rather than parameters  
* Can grab only the confidence intervals using confint(myLMERObject)  
* Need to be careful in reporting the results - figures vs. tables vs. in-line descriptions  
  
Statistical inference with Maryland crime data:  
  
* The Maryland crime data is available on data.gov - interesting for many public and private purposes  
* The null hypothsis test can be used with LMER - frequentist approach  
	* By default, lmer does not provide p-values, as there is ongoing debate as to the degrees of freedom and impact on reported results  
    * Can use lmerTest package to calculate and report on the p-values  
* Can use ANOVA to look at the variability explained by one model versus another model, and the associated degrees of freedom needed  
  
Example code includes:  
```{r}

# Read in births data
rawBirths <- read.csv("./RInputFiles/countyBirthsDataUse.csv")
countyBirthsData <- rawBirths
str(countyBirthsData)


# First, build a lmer with state as a random effect. Then look at the model's summary and the plot of residuals. 
birthRateStateModel <- lme4::lmer(BirthRate ~ (1|State), data=countyBirthsData)
summary(birthRateStateModel)
plot(birthRateStateModel)

# Next, plot the predicted values from the model ontop of the plot shown during the video.
countyBirthsData$birthPredictState <- predict(birthRateStateModel, countyBirthsData)
ggplot() + theme_minimal() +
    geom_point(data =countyBirthsData, aes(x = TotalPopulation, y = BirthRate)) + 
    geom_point(data = countyBirthsData, aes(x = TotalPopulation, y = birthPredictState),
               color = 'blue', alpha = 0.5
               )

# Include the AverageAgeofMother as a fixed effect within the lmer and state as a random effect
ageMotherModel <- lme4::lmer( BirthRate ~ AverageAgeofMother + (1|State), data=countyBirthsData)
summary(ageMotherModel)

# Compare the random-effect model to the linear effect model 
summary(lm(BirthRate ~ AverageAgeofMother, data = countyBirthsData))


# Include the AverageAgeofMother as a correlated random-effect slope parameter
ageMotherModelRandomCorrelated <- lme4::lmer(BirthRate ~ AverageAgeofMother + (AverageAgeofMother|State),
                       countyBirthsData)
summary(ageMotherModelRandomCorrelated)


# Include the AverageAgeofMother as a correlated random-effect slope parameter
ageMotherModelRandomUncorrelated <- lme4::lmer(BirthRate ~ AverageAgeofMother + 
                                                    (AverageAgeofMother || State), data=countyBirthsData
                                               )
summary(ageMotherModelRandomUncorrelated)


out <- ageMotherModelRandomUncorrelated

# Extract the fixed-effect coefficients
lme4::fixef(out)

# Extract the random-effect coefficients
lme4::ranef(out)

# Estimate the confidence intervals 
(ciOut <- confint(out))


# Technical note: Extracting out the regression coefficients from lmer is tricky (see discussion between the lmer and broom authors development)
# Extract out the parameter estimates and confidence intervals and manipulate the data
dataPlot <- data.frame(cbind( lme4::fixef(out), ciOut[ 4:5, ]))
rownames(dataPlot)[1] <- "Intercept"
colnames(dataPlot) <- c("mean", "l95", "u95")
dataPlot$parameter <- rownames(dataPlot)

# Print the new dataframe
print(dataPlot)

# Plot the results using ggplot2
ggplot(dataPlot, aes(x = parameter, y = mean,
                     ymin = l95, ymax = u95)) +
    geom_hline( yintercept = 0, color = 'red' ) +
    geom_linerange() + geom_point() + coord_flip() + theme_minimal()



# Read in crime data
rawCrime <- read.csv("./RInputFiles/MDCrime.csv")
MDCrime <- rawCrime
str(MDCrime)


plot1 <- ggplot(data = MDCrime, aes(x = Year, y = Crime, group = County)) +
    geom_line() + theme_minimal() +
    ylab("Major crimes reported per county")
print(plot1)

plot1 + geom_smooth(method = 'lm')


# Null hypothesis testing uses p-values to see if a variable is "significant"
# Recently, the abuse and overuse of null hypothesis testing and p-values has caused the American Statistical Association to issue a statement about the use of p-values
# Because of these criticisms and other numerical challenges, Doug Bates (the creator of the lme4 package) does not include p-values as part of his package
# However, you may still want to estimate p-values, because p-values are sill commonly used. Several packages exist, including the lmerTest package
# https://www.amstat.org/asa/files/pdfs/P-ValueStatement.pdf

# Load lmerTest
# library(lmerTest)

# Fit the model with Year as both a fixed and random-effect
lme4::lmer(Crime ~ Year + (1 + Year | County) , data = MDCrime)

# Fit the model with Year2 rather than Year
out <- lme4::lmer(Crime ~ Year2 + (1 + Year2 | County) , data = MDCrime)

# Examine the model's output
summary(out)


## Build the Null model with only County as a random-effect
null_model <- lme4::lmer(Crime ~ (1 | County) , data = MDCrime)

## Build the Year2 model with Year2 as a fixed and random slope and County as the random-effect
year_model <- lme4::lmer(Crime ~ Year2 + (1 + Year2 | County) , data = MDCrime)

## Compare the two models using an anova
anova(null_model, year_model)

```
  
  
  
***
  
Chapter 3 - Generalized Linear Mixed-Effect Models  
  
Crash course on GLMs - relaxing the assumptions around normality of the residuals:  
  
* Non-normal data can be transformed using arcsin or the like  
* However, with advances in methodology, it is possible to more directly model the data using binomial and poisson distributions  
* The basic glm call is glm(y ~ x, family="")  # default is family="gaussian", which same as the lm()  
* The Poisson distribution is frequently best for count data, such as website visitors per hour - mean equals variance (generally best for small counts less than 30; can use normals for large counts)  
* For logistic regression, data can be entered in any of three formats  
	* Binary (y=0 or 1) - glm(y ~ x, family="binomial")  
    * Wilkinson-Rogers - glm(cbind(success, failure) ~ x, family="binomial")  
    * Weighted - glm(y ~ x, weights=weights, family="binomial")  
    * These methods differ primarily in the degrees of freedom (and thus deviance)  
  
Binomial data - modeling data with only two outcomes:  
  
* Traditional method for analysis includes looking at proportion of successes  
* The GLM allows for direct looks at the data - logistic regression (logit)  
* Binomial data can be fit using glmer(y ~ x + (1/group), family="error term")  
* The regression coefficients can be difficult to explain, sometimes leading to the use of odds ratios instead  
	* The odds ratio of 2.0 would mean 2:1 odds for that specific group  
  
Count data:  
  
* Examples like number of events per hour (website hits) or counts per area (birds)  
* The count data differs from the binomial in that there is no pre-specified upper boundary  
* While Chi-squared is often used for goodness of fit or test of association for count data, the Poisson GLM can be a nice alternative  
	* glm(y ~ x, family="poisson")  
    * glmer(y ~ x + (1|group), family="poisson")  
  
Example code includes:  
```{r cache=TRUE}

# In this case study, we will be working with simulated dose-response data
# The response is mortality (1) or survival (0) at the end of a study. During this exercise, we will fit a logistic regression using all three methods described in the video
# You have been given two datasets. dfLong has the data in a "long" format with each row corresponding to an observation (i.e., a 0 or 1)
# dfShort has the data in an aggregated format with each row corresponding to a treatment (e.g., 6 successes, 4 failures, number of replicates = 10, proportion = 0.6)

dfLong <- data.frame(dose=c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10), 
                     mortality=c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 1, 0, 0, 1, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 1, 0, 1, 0, 0, 1, 0, 1, 1, 0, 1, 1, 0, 1, 0, 1, 1, 0, 0, 0, 1, 1, 0, 0, 1, 1, 1, 0, 0, 1, 0, 0, 1, 0, 1, 0, 1, 0, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 0, 0, 0, 1, 0, 1, 0, 1, 1, 0, 0, 1, 1)
                     )
str(dfLong)

dfShort <- dfLong %>% 
    group_by(dose) %>%
    summarize(mortality=sum(mortality), nReps=n()) %>%
    mutate(survival=nReps-mortality, mortalityP=mortality/nReps)
dfShort


# Fit a glm using data in a long format
fitLong <- glm(mortality ~ dose, data = dfLong, family = "binomial")
summary(fitLong)

# Fit a glm using data in a short format with two columns
fitShort <- glm( cbind(mortality , survival ) ~ dose , data = dfShort, family = "binomial")
summary(fitShort)

# Fit a glm using data in a short format with weights
fitShortP <- glm( mortalityP ~ dose , data = dfShort, weights = nReps , family = "binomial")
summary(fitShortP)


y <- c(0, 1, 0, 1, 0, 1, 0, 1, 0, 2, 1, 2, 0, 1, 1, 0, 1, 5, 1, 1)
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20)


# Fit the linear model
summary(lm(y ~ x))

# Fit the generalized linear model
summary(glm(y ~ x, family = "poisson"))


# Often, we want to "look" at our data and trends in our data
# ggplot2 allows us to add trend lines to our data
# The defult lines are created using a technique called local regression
# However, we can specify different models, including GLMs
# During this exercise, we'll see how to plot a GLM

# Plot the data using jittered points and the default stat_smooth
ggplot(data = dfLong, aes(x = dose, y = mortality)) + 
    geom_jitter(height = 0.05, width = 0.1) +
    stat_smooth(fill = 'pink', color = 'red') 

# Plot the data using jittered points and the the glm stat_smooth
ggplot(data = dfLong, aes(x = dose, y = mortality)) + 
    geom_jitter(height = 0.05, width = 0.1) +
    stat_smooth(method = 'glm',  method.args = list(family = "binomial"))


# library(lmerTest)

df <- data.frame(dose=rep(rep(c(0, 2, 4, 6, 8, 10), each=20), times=3), 
                 mortality=c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 0, 1, 0, 0, 1, 0, 1, 0, 0, 1, 0, 1, 1, 0, 0, 1, 0, 1, 0, 1, 1, 1, 0, 0, 0, 1, 1, 0, 1, 1, 1, 1, 1, 0, 0, 1, 0, 0, 1, 1, 0, 0, 1, 0, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 0, 1, 0, 1, 1, 0, 1, 0, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 0, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 1, 1, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 1, 0, 1, 1, 0, 1, 1, 1, 1, 0, 0, 1, 1, 0, 1, 0, 0, 1, 0, 1, 0, 1, 0, 0, 1, 0, 0, 1, 0, 0, 1, 0, 1, 0, 1, 1, 0, 1, 1, 0, 1, 0, 1, 1, 0, 0, 1, 1, 1, 1, 1, 1, 1, 0, 1, 1, 0, 1, 1, 0, 0, 0, 0, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 0, 1, 1, 0, 0, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 0, 0, 0, 1, 0, 0, 0, 1, 1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 1, 1, 1, 0, 0, 1, 1, 0, 1, 1, 1, 0, 0, 0, 0, 0, 0, 1, 1, 1, 0, 0, 0, 1, 1, 1, 0, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1), 
                 replicate=factor(rep(letters[1:3], each=120))
                 )
str(df)


glmerOut <- lme4::glmer(mortality ~ dose + (1|replicate), family = 'binomial', data = df)
summary(glmerOut)


# library(lmerTest)
# Fit the model and look at its summary 
# modelOut <- lme4::glmer( cbind(Purchases, Pass) ~ friend + ranking + (1|city), data = allData, family = 'binomial')
# summary( modelOut) 

# Compare outputs to a lmer model
# summary(lme4::lmer( Purchases/( Purchases + Pass) ~ friend + ranking + (1|city), data = allData))


# Run the code to see how to calculate odds ratios
# summary(modelOut) 
# exp(fixef(modelOut)[2])
# exp(confint(modelOut)[3, ])


# Load lmerTest
# library(lmerTest)


userGroups <- data.frame(group=factor(rep(rep(LETTERS[1:4], each=10), times=2)), 
                         webpage=factor(rep(c("old", "new"), each=40)), 
                         clicks=c(0, 0, 0, 0, 0, 0, 2, 0, 0, 0, 1, 0, 0, 1, 1, 1, 2, 0, 1, 1, 0, 3, 2, 3, 1, 2, 4, 2, 1, 0, 2, 0, 1, 2, 0, 2, 1, 1, 2, 4, 2, 8, 1, 1, 1, 2, 1, 1, 0, 0, 3, 0, 1, 4, 1, 2, 0, 1, 1, 0, 0, 3, 2, 0, 3, 1, 2, 2, 0, 2, 3, 1, 3, 2, 4, 4, 2, 1, 5, 2)
                         )
str(userGroups)


# Fit a Poisson glmer
summary( lme4::glmer(clicks ~ webpage + (1|group), family = 'poisson', data = userGroups))


# library(lmerTest)


rawIL <- read.csv("./RInputFiles/ILData.csv")
ILdata <- rawIL
str(ILdata)

# Age goes before year
modelOut <- lme4::glmer(count ~ age + year + (year|county), family = 'poisson', data = ILdata)
summary(modelOut)


# Extract out fixed effects
lme4::fixef(modelOut)

# Extract out random effects 
lme4::ranef(modelOut)


# Run code to see one method for plotting the data
ggplot(data = ILdata, aes(x = year, y = count, group = county)) +
    geom_line() +
    facet_grid(age ~ . ) +
    stat_smooth( method = 'glm',
                method.args = list( family = "poisson"), se = FALSE,
                alpha = 0.5) +
    theme_minimal()

```
  
  
  
***
  
Chapter 4 - Repeated Measures  
  
An introduction to repeated measures:  
  
* Sampling the same thing over time is a repeated measure, a specific example of a mixed effects model  
	* Follow the same individual through time - cohorts allow for controlling for individuality  
    * The paired t-test is often used for assessing a repeated measures dataset - t.test(x1, x2, paired=TRUE)  # x1 and x2 need to be the same length and each element needs to be the same individual  
* Repeated measures ANOVA is a conceptual extension of the paired t-test - are the means constant over time  
	* anova(lmer(y ~ time + (1|individual)))  
    * Can be used with glmer() also  
    * Note that degrees of freedom is still an open question - different packages calculate this differently  
  
Sleep study:  
  
* Applying LMER to the sleep study dataset - impact of drugs on sleep patterns for 10 patients followed over time  
	* This is the classic "Student" dataset due to Guinness at the time not allowing its employees to publish  
    * Ho will be that the amount of sleep does not vary with the treatments  
    * Modeling will be done using a linear mixed model  
* Modeling approach - iteratively;  
	* EDA  
    * Simple regression  
    * Model of interest  
    * Extract information from model  
    * Visualize final data  
  
Hate in NY state?  
  
* Change in rate of hate crimes over time by county - available from data.gov for 2010-2016  
* Level of technical details in reporting should vary significantly by audience - blend data in to story for wider audiences, while being reporducible/technical for a scientifc audience  
  
Wrap up:  
  
* Hiearchical data, mixed effects models, case studies  
* Start with the LME4 documentation for additional explorations and details  
  
Example code includes:  
```{r}

y <- c(0.23, 2.735, -0.038, 6.327, -0.643, 1.69, -1.378, -1.228, -0.252, 2.014, -0.073, 6.101, 0.213, 3.127, -0.29, 8.395, -0.33, 2.735, 0.223, 1.301)
treat <- rep(c("before", "after"), times=10)
x <- rep(letters[1:10], each=2)

# Run a standard, non-paired t-test
t.test(y[treat == "before"], y[treat == "after"], paired = FALSE)

# Run a standard, paired t-test
t.test(y[treat == "before"], y[treat == "after"], paired = TRUE)


library(lmerTest)
library(lme4)

# Run the paired-test like before
t.test(y[treat == "before"], y[treat == "after"], paired = TRUE)

# Run a repeated-measures ANOVA
anova(lmer( y ~ treat + (1|x)))


data(sleepstudy, package="lme4")
str(sleepstudy)

# Plot the data
ggplot(data = sleepstudy) +
    geom_line(aes(x = Days, y = Reaction, group = Subject)) +
    stat_smooth(aes(x = Days, y = Reaction), method = 'lm', size = 3, se = FALSE)

# Build a lm 
lm( Reaction ~ Days, data = sleepstudy)

# Build a lmer
(lmerOut <- lmer( Reaction ~ Days + (1|Subject), data = sleepstudy))


# The lmer model you built during the previous exercise has been saved as lmerOut
# During this exercise, you will examine the effects of drug type using both an ANOVA framework and a regression framework

# Run an anova
anova(lmerOut)

# Look at the regression coefficients
summary(lmerOut)


# Read in NY hate data
rawHate <- read.csv("./RInputFiles/hateNY.csv")
hate <- rawHate
str(hate)


ggplot( data = hate, aes(x = Year, y = TotalIncidents, group = County)) +
    geom_line() + 
    geom_smooth(method = 'lm', se = FALSE)


# During this exercise, you will build a glmer
# Because most of the incidents are small count values, use a Poisson (R function family poisson) error term
# First, build a model using the actually year (variable Year, such as 2006, 2007, etc) - this model will fail
# Second, build a model using the rescaled year (variable Year2, such as 0, 1, 2, etc)
# This demonstrates the importance of considering where the intercept is located when building regression models
# Recall that a variable x can be both a fixed and random effect in a lmer() or glmer(): for example lmer(y ~ x + (x| group) demonstrates this syntax

# glmer with raw Year
glmer(TotalIncidents ~ Year + (Year|County), data = hate, family = "poisson")

# glmer with scaled Year, Year2
glmerOut <- glmer(TotalIncidents ~ Year2 + (Year2|County), data = hate, family = "poisson")
summary(glmerOut)


# Extract and manipulate data
countyTrend <- ranef(glmerOut)$County
countyTrend$county <- factor(row.names(countyTrend), levels =row.names(countyTrend)[order(countyTrend$Year2)])

# Plot results 
ggplot(data = countyTrend, aes(x = county, y = Year2)) + geom_point() +
    coord_flip() + 
    ylab("Change in hate crimes per year")  +
    xlab("County")

```
  
  
  
***
  
###_Forecasting Product Demand in R_  
  
Chapter 1 - Forecasting Demand with Time Series  
  
Loading data in to an xts object:  
  
* The xts object will be the buidling block for the course - extensible time series (xts) is an extension of the zoo package - basically, a time index attached to the data matrix  
* Can create dates using dates=seq(as.Date("MM-DD-YYYY"), length=, by="weeks")  # to create weekly data  
	* xts(myData, order.by=dates)  # will create an XTS using dates as the index  
  
ARIMA Time Series 101:  
  
* AR - AutoRegressive (lags help to determine today's values - "long memory models")  
* MA - Moving Average (shocks/errors help to determine today's shocks/errors - "short memory models" due to dissipation)  
* I - Integrated (does the data have a dependency across time, and how long does it last) - make the time series stationary  
	* Stationarity is the idea that effects disipate over time - today has more impact on tomorrow than on time periods in the future  
    * Differencing (monthly, seasonal, etc.) the data can be a useful approach for data with stationarity  
* Begin by creating training dataset and valiadation training dataset  
* The auto.arima() function tries to estimate the best ARIMA for a given data series  
	* ARIMA(p, d, q) is ARIMA(AR, Differencing, MA)  
  
Forecasting and Evaluating:  
  
* Can use the ARIMA data to forecast the data forward - extrapolating the signal (forecasting) and estimating the amount of noise (error or CI)  
* The forecast() function in R simplifies the process - forecast(myModel, h=) which will forecast forward h time periods  
* Two common error measurements include MAE (mean average error) and MAPE (mean average percentage error)  
	* MAPE is better at putting things on a common scale  
  
Example code includes:  
```{r}


# Read in beverages data
rawBev <- read.csv("./RInputFiles/Bev.csv")
bev <- rawBev
str(bev)


# Load xts package 
library(xts)
library(forecast)


# Create the dates object as an index for your xts object
dates <- seq(as.Date("2014-01-19"), length = 176, by = "weeks")

# Create an xts object called bev_xts
bev_xts <- xts(bev, order.by = dates)


# Create the individual region sales as their own objects
MET_hi <- bev_xts[,"MET.hi"]
MET_lo <- bev_xts[,"MET.lo"]
MET_sp <- bev_xts[,"MET.sp"]

# Sum the region sales together
MET_t <- MET_hi + MET_lo + MET_sp

# Plot the metropolitan region total sales
plot(MET_t)


# Split the data into training and validation
MET_t_train <- MET_t[index(MET_t) < "2017-01-01"]
MET_t_valid <- MET_t[index(MET_t) >= "2017-01-01"]

# Use auto.arima() function for metropolitan sales
MET_t_model <- auto.arima(MET_t_train)


# Forecast the first 22 weeks of 2017
forecast_MET_t <- forecast(MET_t_model, h = 22)

# Plot this forecast #
plot(forecast_MET_t)


# Convert to numeric for ease
for_MET_t <- as.numeric(forecast_MET_t$mean)
v_MET_t <- as.numeric(MET_t_valid)

# Calculate the MAE
MAE <- mean(abs(for_MET_t - v_MET_t))

# Calculate the MAPE
MAPE <- 100*mean(abs(for_MET_t - v_MET_t)/v_MET_t)

# Print to see how good your forecast is!
print(MAE)
print(MAPE)


# Convert your forecast to an xts object
for_dates <- seq(as.Date("2017-01-01"), length = 22, by = "weeks")
for_MET_t_xts <- xts(forecast_MET_t$mean, order.by = for_dates)

# Plot the validation data set
plot(for_MET_t_xts, main = 'Forecast Comparison', ylim = c(4000, 8500))

# Overlay the forecast of 2017
lines(MET_t_valid, col = "blue")


# Plot the validation data set
plot(MET_t_valid, main = 'Forecast Comparison', ylim = c(4000, 8500))

# Overlay the forecast of 2017
lines(for_MET_t_xts, col = "blue")

# Convert the limits to xts objects
lower <- xts(forecast_MET_t$lower[, 2], order.by = for_dates)
upper <- xts(forecast_MET_t$upper[, 2], order.by = for_dates)

# Adding confidence intervals of forecast to plot
lines(lower, col = "blue", lty = "dashed")
lines(upper, col = "blue", lty = "dashed")

```
  
  
  
***
  
Chapter 2 - Components of Demand  
  
Price elasticity:  
  
* Price is one of the obvious factors that impacts demand, with the relationship called price elasticity (% dDemand / % dPrice)  
	* Elastic goods have elasticity > 1, meaning demand changes more quickly (percentage wise) than price  
    * Inelastic goods have elasticity < 1, for example gasoline  
    * Unit elastic goods have elasticity = 1, meaning that X% increase in price drives X% decrease in demand  
    * Linear regression can be employed to estimate the elasticity for a given product - the log-log transform helps get the % vs % coefficients  
  
Seasonal/holiday/promotional effects:  
  
* Seasonal products are common - can be bought any time of the year, though certain seasons have higher demand (holidays are a common example)  
* Promotions are attempts by companies to influence demand  
* Linear regression can help determine relationships between demand and many other factors  
	* If an xts vector has been created for key dates, can merge(train, holiday, fill=0) and the holiday column will be 0 wherever there is no match to holiday  
  
Forecasting with regression:  
  
* Forecasting with time series is straightforward due to the lag nature of the models - tomorrow forecasts today and today forecasts tomorrow and etc.  
* Forecasting with regression can be more tricky, particularly since we need the future inputs (such as price) in order to predict the future demand  
	* Even when there are contractually fixed prices, promotions can effectively create a de facto price change anyways  
* Need to have the same column names in the test/validation dataset as were used in the modeling  
	* Then, can use predict(myModel, myData)  
    * May need to exponentiate in case the data are currently on the log scale rather than the absolute scale  
  
Example code includes:  
```{r}

bev_xts_train <- bev_xts[index(bev_xts) < "2017-01-01"]
bev_xts_valid <- bev_xts[index(bev_xts) >= "2017-01-01"]

# Save the prices of each product
l_MET_hi_p <- log(as.vector(bev_xts_train[, "MET.hi.p"]))

# Save as a data frame
MET_hi_train <- data.frame(as.vector(log(MET_hi[index(MET_hi) < "2017-01-01"])), l_MET_hi_p)
colnames(MET_hi_train) <- c("log_sales", "log_price")

# Calculate the regression
model_MET_hi <- lm(log_sales ~ log_price, data = MET_hi_train)


# Plot the product's sales
plot(MET_hi)

# Plot the product's price
MET_hi_p <- bev_xts_train[, "MET.hi.p"]
plot(MET_hi_p)


# Create date indices for New Year's week
n.dates <- as.Date(c("2014-12-28", "2015-12-27", "2016-12-25"))

# Create xts objects for New Year's
newyear <- as.xts(rep(1, 3), order.by = n.dates)

# Create sequence of dates for merging
dates_train <- seq(as.Date("2014-01-19"), length = 154, by = "weeks")

# Merge training dates into New Year's object
newyear <- merge(newyear, dates_train, fill = 0)


# Add newyear variable to your data frame
MET_hi_train <- data.frame(MET_hi_train, newyear=as.vector(newyear))

# Build regressions for the product
model_MET_hi_full <- lm(log_sales ~ log_price + newyear, data = MET_hi_train)


# Subset the validation prices #
l_MET_hi_p_valid <- log(as.vector(bev_xts_valid[, "MET.hi.p"]))

# Create a validation data frame #
MET_hi_valid <- data.frame(l_MET_hi_p_valid)
colnames(MET_hi_valid) <- "log_price"


# Predict the log of sales for your high end product
pred_MET_hi <- predict(model_MET_hi, MET_hi_valid)

# Convert predictions out of log scale
pred_MET_hi <- exp(pred_MET_hi)


# Convert to an xts object
dates_valid <- seq(as.Date("2017-01-01"), length = 22, by = "weeks")
pred_MET_hi_xts <- xts(pred_MET_hi, order.by = dates_valid)

# Plot the forecast
plot(pred_MET_hi_xts)

# Calculate and print the MAPE
MET_hi_v <- bev_xts_valid[,"MET.hi"]

MAPE <- 100*mean(abs((pred_MET_hi_xts - MET_hi_v)/MET_hi_v))
print(MAPE)

```
  
  
  
***
  
Chapter 3 - Blending Regression with Time Series  
  
Residuals from regression model:  
  
* The residuals from the regression models can be used for further modeling - see if the residuals are related over time, and model them with time series if so  
* Need to start by gathering the residuals and then converting them to an XTS object - explore for patterns in this XTS object  
  
Forecasting residuals:  
  
* When the residuals are related across time, we can use time series to model the residuals - basically, patterns to the errors provide an opportunity for further modeling  
* Can use auto.arima() on the residuals data, to see what the best ARIMA model for the residuals is  
	* Can then forecast the residuals in to the future using forecast(myModel, h=) # h being the time periods to predict forward  
  
Transfer functions and ensembling:  
  
* Techniques for combining forecasts - single model (transfer function) or averaging of models (ensembling)  
* Demand can be based on both regression (modeling external factors) and time series (residuals)  
* Ensembling is a combination (blend) of the forecasts, with simple averaging being the simplest approach  
	* Basically, build a stand-alone time series model and a stand-alone regression model  
    * The ensemble forecast can be better or worse than any of the stand-alone models  
  
Example code includes:  
```{r}

# Calculate the residuals from the model
MET_hi_full_res <- resid(model_MET_hi_full)

# Convert the residuals to an xts object
MET_hi_full_res <- xts(MET_hi_full_res, order.by = dates_train)


# Plot the histogram of the residuals
hist(MET_hi_full_res)

# Plot the residuals over time
plot(MET_hi_full_res)


# Build an ARIMA model on the residuals: MET_hi_arima
MET_hi_arima <- auto.arima(MET_hi_full_res)

# Look at a summary of the model
summary(MET_hi_arima)


# Forecast 22 weeks with your model: for_MET_hi_arima
for_MET_hi_arima <- forecast(MET_hi_arima, h=22)

# Print first 10 observations
head(for_MET_hi_arima$mean, n = 10)


# Convert your forecasts into an xts object
dates_valid <- seq(as.Date("2017-01-01"), length = 22, by = "weeks")
for_MET_hi_arima <- xts(for_MET_hi_arima$mean, order.by = dates_valid)

# Plot the forecast
plot(for_MET_hi_arima)


# Convert your residual forecast to the exponential version
for_MET_hi_arima <- exp(for_MET_hi_arima)

# Multiply your forecasts together!
for_MET_hi_final <- for_MET_hi_arima * pred_MET_hi_xts


# Plot the final forecast - don't touch the options!
plot(for_MET_hi_final, ylim = c(1000, 4300))

# Overlay the validation data set
lines(MET_hi_v, col = "blue")


# Calculate the MAE
MAE <- mean(abs(for_MET_hi_final - MET_hi_v))
print(MAE)

# Calculate the MAPE
MAPE <- 100 * mean(abs(for_MET_hi_final - MET_hi_v)/MET_hi_v)
print(MAPE)


# Build an ARIMA model using the auto.arima function
MET_hi_model_arima <- auto.arima(MET_hi)

# Forecast the ARIMA model
for_MET_hi <- forecast(MET_hi_model_arima, h = length(MET_hi_v))

# Save the forecast as an xts object
dates_valid <- seq(as.Date("2017-01-01"), length = 22, by = "weeks")
for_MET_hi_xts <- xts(for_MET_hi$mean, order.by = dates_valid)

# Calculate the MAPE of the forecast
MAPE <- 100 * mean(abs(for_MET_hi_xts - MET_hi_v) / MET_hi_v)
print(MAPE)


# Ensemble the two forecasts together
for_MET_hi_en <- 0.5 * (for_MET_hi_xts + pred_MET_hi_xts)

# Calculate the MAE and MAPE
MAE <- mean(abs(for_MET_hi_en - MET_hi_v))
print(MAE)

MAPE <- 100 * mean(abs(for_MET_hi_en - MET_hi_v) / MET_hi_v)
print(MAPE)

```
  
  
  
***
  
Chapter 4 - Hierarchical Forecasting  
  
Bottom-Up Hierarchical Forecasting:  
  
* The hierarchical data structuring can be an advantage in forecasting, provided that the data has a natural hierarchy  
* The sum of all the lower-level forecasts should equal the higher-level forecasts  
	* Bottom-up: Forecast at the lowest level and aggregate (easiest but requires the most number of forecasts)  
    * Top-down: Forecast at the top level and the apply downwards  
    * Middle-out: Forecast at the middle levels and then apply both upwards and downwards  
  
Top-Down Hierarchical Forecasting:  
  
* The top-down forecasting process is typically quicker but less accurate than the bottom-up forecasting process  
* Two techniques available for top-down reconciliation  
	* Average of historical proportions - mean percentage that each component contributes to the total (calculated by sub-component such as week)  
    * Proportion of historical averages - mean percentage that each component contributes to the total (calculated by aggregate)  
* Reconciled forecasts at lower levels are typically less accurate than the direct forecast of the lower levels  
  
Middle-Out Hierarchical Forecasting:  
  
* Bottom-up forecasting is higher quality but more time-consuming than top-down forecasting  
* The middle-out forecasting method is a sometimes successful blend of the methods, getting decent accuracy at a lesser time commitment  
  
Wrap up:  
  
* Using time series to forecast demand forward  
* Incorporating external factors using linear regression  
* Blending time series and regression approaches  
* Top-down, bottom-up, middle-out approaches to aggregation and forecasting at various levels (hierarchical)  
* Can extend by looking at cross-elasticities (impact of competitor pricing)  
* Can better forecast proportions using time series analysis  
* Additional demand forecasting models include neural networks, exponential smoothing, etc.  
  
Example code includes:  
```{r}

# Build a time series model 
MET_sp_model_arima <- auto.arima(MET_sp)

# Forecast the time series model for 22 periods
for_MET_sp <- forecast(MET_sp_model_arima, h=22)

# Create an xts object
for_MET_sp_xts <- xts(for_MET_sp$mean, order.by=dates_valid)

MET_sp_v <- MET_sp["2017"]

# Calculate the MAPE
MAPE <- 100 * mean(abs(for_MET_sp_xts - MET_sp_v) / MET_sp_v)
print(MAPE)


MET_sp_train <- bev_xts_train %>%
    transform(log_sales = log(MET.sp), log_price=log(MET.sp.p))
MET_sp_train <- MET_sp_train[, c("log_sales", "log_price")]
MET_sp_train$newyear <- 0
MET_sp_train$valentine <- 0
MET_sp_train$christmas <- 0
MET_sp_train$mother <- 0

MET_sp_train[index(MET_sp_train) %in% as.Date(c("2014-12-28", "2015-12-27", "2016-12-25")), "newyear"] <- 1
MET_sp_train[index(MET_sp_train) %in% as.Date(c("2014-02-09", "2015-02-08", "2016-02-07")), "valentine"] <- 1
MET_sp_train[index(MET_sp_train) %in% as.Date(c("2014-12-21", "2015-12-20", "2016-12-18")), "christmas"] <- 1
MET_sp_train[index(MET_sp_train) %in% as.Date(c("2014-05-04", "2015-05-03", "2016-05-01")), "mother"] <- 1


# THE BELOW IS TOTAL NONSENSE
# Build a regression model
model_MET_sp <- lm(log_sales ~ log_price + newyear + valentine + christmas + mother, data = MET_sp_train)


MET_sp_valid <- as.data.frame(bev_xts_valid) %>%
    mutate(log_sales = log(MET.sp), log_price=log(MET.sp.p)) %>%
    select("log_sales", "log_price")
MET_sp_valid$newyear <- 0
MET_sp_valid$valentine <- 0
MET_sp_valid$christmas <- 0
MET_sp_valid$mother <- 0  

MET_sp_valid[7, "valentine"] <- 1
MET_sp_valid[19, "mother"] <- 1
MET_sp_valid$log_sales <- NULL


# Forecast the regression model using the predict function 
pred_MET_sp <- predict(model_MET_sp, MET_sp_valid)

# Exponentiate your predictions and create an xts object
pred_MET_sp <- exp(pred_MET_sp)
pred_MET_sp_xts <- xts(pred_MET_sp, order.by = dates_valid)

# Calculate MAPE
MAPE <- 100*mean(abs((pred_MET_sp_xts - MET_sp_v)/MET_sp_v))
print(MAPE)


# Ensemble the two forecasts
for_MET_sp_en <- 0.5 * (for_MET_sp_xts + pred_MET_sp_xts)

# Calculate the MAPE
MAPE <- 100 * mean(abs(for_MET_sp_en - MET_sp_v) / MET_sp_v)
print(MAPE)


# Copy over pred_MET_lo_xts
pred_MET_lo_xts <- xts(c(2960.6, 2974.1, 2943.2, 2948.6, 2915.6, 2736.4, 2953.9, 3199.4, 2934, 2898.7, 3027.7, 3165.9, 3073.1, 2842.7, 2928.7, 3070.2, 2982.2, 3018, 3031.9, 2879.4, 2993.2, 2974.1), order.by=dates_valid)


# Calculate the metropolitan regional sales forecast
for_MET_total <- pred_MET_hi_xts + for_MET_sp_en + pred_MET_lo_xts

# Calculate a validation data set 
MET_t_v <- bev_xts_valid[,"MET.hi"] + bev_xts_valid[,"MET.lo"] + bev_xts_valid[,"MET.sp"]

# Calculate the MAPE
MAPE <- 100 * mean(abs(for_MET_total - MET_t_v) / MET_t_v)
print(MAPE)


# Create the MET_total data
MET_total <- xts(data.frame(MET.hi=c(5942, 5600, 5541, 6892, 5586, 5943, 6329, 6693, 6938, 6138, 6361, 6378, 5423, 5097, 4937, 5496, 6870, 6626, 6356, 5657, 6577, 7202, 7381, 7404, 7204, 6667, 6153, 6035, 5633, 5283, 5178, 4758, 5058, 5254, 5954, 6166, 6247, 6304, 7202, 6662, 6814, 6174, 5412, 5380, 5674, 6472, 6912, 7404, 8614, 8849, 7174, 6489, 7174, 6555, 6402, 7671, 5012, 4790, 5075, 5238, 5615, 6113, 7706, 7811, 7898, 7232, 6585, 5870, 7084, 5125, 5330, 5553, 6349, 6195, 6271, 5851, 5333, 5854, 5609, 5649, 6051, 6409, 5786, 5190, 5085, 4949, 5151, 5147, 5426, 5509, 6956, 7870, 8224, 6685, 6153, 5802, 5244, 5162, 5036, 5025, 8378, 8944, 7109, 7605, 7846, 7598, 8012, 9551, 6102, 5366, 4932, 4962, 5392, 6194, 7239, 7621, 7460, 7097, 6596, 5848, 8306, 5344, 5848, 6341, 7364, 7269, 7053, 6682, 6971, 7521, 7063, 6298, 6003, 5227, 5047, 4877, 4851, 4628, 4516, 4442, 4935, 5181, 5431, 5866, 5919, 5704, 5957, 6019, 5962, 6021, 5880, 5674, 7439, 7415)),
                 order.by=dates_train
                 )

# Build a regional time series model
MET_t_model_arima <- auto.arima(MET_total)

# Calculate a 2017 forecast for 22 periods
for_MET_t <- forecast(MET_t_model_arima, h=22)

# Make an xts object from your forecast
for_MET_t_xts <- xts(for_MET_t$mean, order.by=dates_valid)

# Calculate the MAPE
MAPE <- 100 * mean(abs(for_MET_t_xts - MET_t_v) / MET_t_v)
print(MAPE)


# Calculate the average historical proportions
prop_hi <- mean(MET_hi/MET_total)
prop_lo <- mean(MET_lo/MET_total)
prop_sp <- mean(MET_sp/MET_total)

# Distribute out your forecast to each product
for_prop_hi <- prop_hi*for_MET_t_xts
for_prop_lo <- prop_lo*for_MET_t_xts
for_prop_sp <- prop_sp*for_MET_t_xts

# Calculate the MAPE's for each product
MAPE_hi <- 100 * mean(abs(for_prop_hi - MET_hi_v) / MET_hi_v)
print(MAPE_hi)

MET_lo_v <- bev_xts_valid[,"MET.lo"]
MAPE_lo <- 100 * mean(abs(for_prop_lo - MET_lo_v) / MET_lo_v)
print(MAPE_lo)

MAPE_sp <- 100 * mean(abs(for_prop_sp - MET_sp_v) / MET_sp_v)
print(MAPE_sp)


# Calculate the average historical proportions
prop_hi_2 <- mean(MET_hi) / mean(MET_total)
prop_lo_2 <- mean(MET_lo) / mean(MET_total)
prop_sp_2 <- mean(MET_sp) / mean(MET_total)

# Distribute out your forecast to each product
for_prop_hi_2 <- prop_hi_2 * for_MET_t_xts
for_prop_lo_2 <- prop_lo_2 * for_MET_t_xts
for_prop_sp_2 <- prop_sp_2 * for_MET_t_xts

# Calculate the MAPE's for each product
MAPE_hi <- 100 * mean(abs(for_prop_hi_2 - MET_hi_v) / MET_hi_v)
print(MAPE_hi)
MAPE_lo <- 100 * mean(abs(for_prop_lo_2 - MET_lo_v) / MET_lo_v)
print(MAPE_lo)
MAPE_sp <- 100 * mean(abs(for_prop_sp_2 - MET_sp_v) / MET_sp_v)
print(MAPE_sp)


SEC_total <- xts(data.frame(SEC.hi=c(700, 775, 789, 863, 765, 759, 757, 747, 746, 709, 749, 786, 796, 726, 727, 723, 778, 755, 739, 740, 723, 695, 727, 707, 725, 684, 667, 698, 727, 722, 748, 695, 742, 739, 715, 724, 686, 671, 688, 682, 710, 700, 672, 680, 695, 780, 751, 693, 809, 881, 703, 712, 768, 796, 808, 904, 641, 662, 693, 725, 719, 736, 715, 722, 732, 745, 689, 705, 811, 739, 744, 700, 745, 735, 732, 722, 721, 732, 750, 714, 752, 677, 731, 674, 720, 675, 741, 722, 715, 719, 649, 697, 743, 733, 772, 698, 690, 734, 713, 644, 788, 833, 749, 731, 670, 675, 675, 993, 773, 751, 697, 677, 750, 723, 780, 763, 721, 701, 704, 684, 985, 791, 731, 714, 704, 694, 685, 652, 708, 754, 747, 705, 711, 699, 712, 745, 706, 665, 666, 692, 676, 696, 689, 697, 689, 717, 697, 708, 660, 707, 715, 680, 922, 888)), order.by=dates_train
                 )

# Build a time series model for the region
SEC_t_model_arima <- auto.arima(SEC_total)

# Forecast the time series model
for_SEC_t <- forecast(SEC_t_model_arima, h=22)

# Make into an xts object
for_SEC_t_xts <- xts(for_SEC_t$mean, order.by=dates_valid)

SEC_t_v <- bev_xts_valid$SEC.hi + bev_xts_valid$SEC.lo
# Calculate the MAPE
MAPE <- 100 * mean(abs(for_SEC_t_xts - SEC_t_v) / SEC_t_v)
print(MAPE)


SEC_hi <- bev_xts_train[, "SEC.hi"]
SEC_lo <- bev_xts_train[, "SEC.lo"]
SEC_hi_v <- bev_xts_valid[, "SEC.hi"]
SEC_lo_v <- bev_xts_valid[, "SEC.lo"]

# Calculate the average of historical proportions
prop_hi <- mean(SEC_hi / SEC_total)
prop_lo <- mean(SEC_lo / SEC_total)

# Distribute the forecast
for_prop_hi <- prop_hi * for_SEC_t_xts
for_prop_lo <- prop_lo * for_SEC_t_xts

# Calculate a MAPE for each product
MAPE_hi <- 100 * mean(abs(for_prop_hi - SEC_hi_v) / SEC_hi_v)
print(MAPE_hi)

MAPE_lo <- 100 * mean(abs(for_prop_lo - SEC_lo_v) / SEC_lo_v)
print(MAPE_lo)


# Copy over for_M_t_xts data
for_M_t_xts <- xts(c(2207, 2021, 2010, 2052, 2075, 2074, 2065, 2058, 2056, 2055, 2053, 2052, 2050, 2049, 2048, 2047, 2046, 2045, 2044, 2043, 2043, 2042), order.by=dates_valid)

# Calculate the state sales forecast: for_state
for_state = for_SEC_t_xts + for_MET_t_xts + for_M_t_xts

# See the forecasts
for_state

```
  
  
  
***
  
###_HR Analytics in R: Exploring Employee Data_  
  
Chapter 1 - Identifying the Best Recruiting Source  
  
Introduction - Ben Teusch, HR Analytics Consultant:  
  
* HR analytics has many other names - people analytics, workforce analytics, etc.  
* Identify groups for comparison - high vs. low performers, groups with high vs. low turnover, etc.  
	* Exploratory analysis and statistics for each group, including plots of key differences  
* Course is outlines as a series of case studies, with one case per chapter  
  
Recruiting and quality of hire:  
  
* Where are the best hires coming from, and how can you get more of them  
	* Defining quality of hire is challenging - some mix of productivity, satisfaction, retention, performance reviews, etc.  
    * Attrition can be defined as the mean of a 1, 0 vector of "did the person leave in the time period T"  
  
Visualizing recruiting data:  
  
* Helpful for communicating findings to decision makers  
* The geom_col() in ggplot will make a bar chart, with the y aestehtic being the bar height  
  
Example code includes:  
```{r}

# Import the recruitment data
recruitment <- readr::read_csv("./RInputFiles/recruitment_data.csv")

# Look at the first few rows of the dataset
head(recruitment)

# Get an overview of the recruitment data
summary(recruitment)

# See which recruiting sources the company has been using
recruitment %>% 
  count(recruiting_source)


# Find the average sales quota attainment for each recruiting source
avg_sales <- recruitment %>% 
  group_by(recruiting_source) %>% 
  summarize(avg_sales_quota_pct=mean(sales_quota_pct))

# Display the result
avg_sales


# Find the average attrition for the sales team, by recruiting source, sorted from lowest attrition rate to highest
avg_attrition <- recruitment %>%
  group_by(recruiting_source) %>% 
  summarize(attrition_rate=mean(attrition)) %>%
  arrange(attrition_rate)

# Display the result
avg_attrition

# Plot the bar chart
avg_sales %>% ggplot(aes(x=recruiting_source, y=avg_sales_quota_pct)) + geom_col()

# Plot the bar chart
avg_attrition %>% ggplot(aes(x=recruiting_source, y=attrition_rate)) + geom_col()

```
  
  
  
***
  
Chapter 2 - What is driving low employee engagement  
  
Analyzing employee engagement:  
  
* Gallup defines engaged employees as those who are involved in, enthusiastic about, and committed to their workplace  
* Survey data are available in the example case study  
	* Will use both mutate() and ifelse()  
    * The ifelse() is needed for vectors of length > 1 since it can work in a vectorized manner (and is thus OK inside the mutate call)  
  
Visualizing the engagement data:  
  
* Multiple attributes in a single place can make for a more compelling report  
* The tidyr package is part of the tidyverse, and hslps arrange the data properly for plotting  
	* tidyr::gather(columns, key="key", value="value") will be the package used in this example - pull the data from the columns down to the rows  
    * ggplot(survey_gathered, aes(x = key, y = value, fill = department)) + geom_col(position = "dodge")  
    * ggplot(survey_gathered, aes(x = key, y = value, fill = department)) + geom_col(position = "dodge") + facet_wrap(~ key, scales = "free")  
  
Are differences meaningful?  
  
* Can use significance testing to assess likelhood (p-value) that the second sample could have come from the same population as the first sample  
	* This course will use t-test (continuous variables) and chi-squared test (categorical variables)  
    * t.test(tenure ~ is_manager, data = survey)  
    * chisq.test(survey$left_company, survey$is_manager)  # no data= argument is available in the function  
  
Example code includes:  
```{r}

# Import the data
survey <- readr::read_csv("./RInputFiles/survey_data.csv")

# Get an overview of the data
summary(survey)

# Examine the counts of the department variable
survey %>% count(department)


# Output the average engagement score for each department, sorted
survey %>%
  group_by(department) %>%
  summarize(avg_engagement=mean(engagement)) %>%
  arrange(avg_engagement)


# Create the disengaged variable and assign the result to survey
survey_disengaged <- survey %>% 
  mutate(disengaged = ifelse(engagement <= 2, 1, 0)) 

survey_disengaged

# Summarize the three variables by department
survey_summary <- survey_disengaged %>%
  group_by(department) %>%
  summarize(pct_disengaged=mean(disengaged), 
            avg_salary=mean(salary), 
            avg_vacation_taken=mean(vacation_days_taken)
            )

survey_summary


# Gather data for plotting
survey_gathered <- survey_summary %>% 
  gather(key = "measure", value = "value",
         pct_disengaged, avg_salary, avg_vacation_taken)

# Create three bar charts
ggplot(survey_gathered, aes(x=measure, y=value, fill=department)) +
  geom_col(position="dodge") + 
  facet_wrap(~ measure, scales="free")


# Add the in_sales variable
survey_sales <- survey %>%
  mutate(in_sales = ifelse(department == "Sales", "Sales", "Other"), 
         disengaged = ifelse(engagement < 3, 1L, 0L)
         )

# Test the hypothesis using survey_sales
chisq.test(survey_sales$disengaged, survey_sales$in_sales)
t.test(disengaged ~ in_sales, data=survey_sales)


# Test the hypothesis using the survey_sales data
t.test(vacation_days_taken ~ in_sales, data = survey_sales)

```
  
  
  
***
  
Chapter 3 - Are new hires getting paid too much?  
  

	
	



	


	



	




	
