---
title: "Data Camp Insights"
author: "davegoblue"
date: "July 7, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background and Overview  
DataCamp offers several interactive courses related to R Programming.  While much of it is review, it is always helpful to see other perspectives on material.  As well, DataCamp has some interesting materials on packages that I want to learn better (ggplot2, dplyr, ggvis, etc.).  This document summarizes a few key insights from:  
  
* Introduction to R  
* Intermediate R  
* Writing Functions in R  
  
## Key Insights and Findings  
###_Introduction to R and Intermediate R_  
There are a few nuggest from within these beginning modules, including:  
  
####_Generic statements_  
* factor(x, ordered=TRUE, levels=c(myLevels)) creates ordinal factors (e.g., a > b > c)  
* subset(a, b) is functionally the same as a[a$b, ] but easier to read  
* & looks at each element while && looks only at the first element (same for | and ||)  
* Inside of a for loop, break kills the loop entirely while next moves back to the top for the next item  
* args(function) shows the arguments (with defaults) for function  
* search() shows the current search path (all auto-load packages and all attached packages)  
* cat("expression") will print the expression or direct it to a file; this is a way to allow \n and \t to take effect in a print statement  
* unique() keeps only the non-duplicated elements of a vector  
* unlist() converts a list back to a vector, somewhat similar to as.vector() on a matrix  
* sort() will sort a vector, but not a data frame  
* rep(a, times=m, each=n) replicates each element of a n times, and then the whole string m times  
* append(x, values, after=length(x)) will insert values in to vector x after point after  
* rev() reverses a vector  
* Inside a grep, "\\1" captures what is inside the ()  
    
####_Apply usages_  
* lapply() operates on a vector/list and always returns a list  
* sapply() is lapply but converted to a vector/array when possible (same as lapply if not possible); if USE.NAMES=FALSE then the vector will be unnamed, though the default is USE.NAMES=TRUE for a named vector  
* vapply(X, FUN, FUN.VALUE, ... , USE.NAMES=TRUE) is safer than sapply in that you specify what type of vector each iteration should produce; e.g., FUN.VALUE=character(1) or FUN.VALUE=numeric(3), with an error if the vector produced by an iteration is not exactly that  
  
####_Dates and times_  
* Sys.Date() grabs the system date as class "Date", with units of days  
* Sys.time() grabs the system time as class "POSIXct", with units of seconds  
* Sys.timezone() shows the system timezone  
* Years are formatted as %Y (4-digit) or %y (2-digit)  
* Months are formatted as %m (2-digit) or %B (full character) or %b (3-character)  
* Days are formatted as %d (2-digit)  
* Weekdays are formatted as %A (full name) or %a (partial name)  
* Times include %H (24-hour hour), %M (minutes), %S (seconds)  
* ?strptime will provide a lot more detail on the formats  
  
Below is some sample code showing examples for the generic statements:  
```{r}
# Factors
xRaw = c("High", "High", "Low", "Low", "Medium", "Very High", "Low")

xFactorNon = factor(xRaw, levels=c("Low", "Medium", "High", "Very High"))
xFactorNon
xFactorNon[xFactorNon == "High"] > xFactorNon[xFactorNon == "Low"][1]

xFactorOrder = factor(xRaw, ordered=TRUE, levels=c("Low", "Medium", "High", "Very High"))
xFactorOrder
xFactorOrder[xFactorOrder == "High"] > xFactorOrder[xFactorOrder == "Low"][1]


# Subsets
data(mtcars)
subset(mtcars, mpg>=25)
identical(subset(mtcars, mpg>=25), mtcars[mtcars$mpg>=25, ])
subset(mtcars, mpg>25, select=c("mpg", "cyl", "disp"))


# & and && (same as | and ||)
compA <- c(2, 3, 4, 1, 2, 3)
compB <- c(1, 2, 3, 4, 5, 6)
(compA > compB) & (compA + compB < 6)
(compA > compB) | (compA + compB < 6)
(compA > compB) && (compA + compB < 6)
(compA > compB) || (compA + compB < 6)


# Loops and cat()
# for (a in b) {
#     do stuff
#     if (exitCond) { break }
#     if (nextCond) { next }
#     do some more stuff
# }
for (myVal in compA*compB) {
    print(paste0("myVal is: ", myVal))
    if ((myVal %% 3) == 0) { cat("Divisible by 3, not happy about that\n\n"); next }
    print("That is not divisible by 3")
    if ((myVal %% 5) == 0) { cat("Exiting due to divisible by 5 but not divisible by 3\n\n"); break }
    cat("Onwards and upwards\n\n")
}


# args() and search()
args(plot.default)
search()


# unique()
compA
unique(compA)


# unlist()
listA <- as.list(compA)
unlist(listA)
identical(compA, unlist(listA))


# sort()
sort(mtcars$mpg)
sort(mtcars$mpg, decreasing=TRUE)

# rep()
rep(1:6, times=2)  # 1:6 followed by 1:6
rep(1:6, each=2)  # 1 1 2 2 3 3 4 4 5 5 6 6
rep(1:6, times=2, each=3)  # 1 1 1 2 2 2 3 3 3 4 4 4 5 5 5 6 6 6 repeated twice (each comes first)
rep(1:6, times=6:1)  # 1 1 1 1 1 1 2 2 2 2 2 3 3 3 3 4 4 4 5 5 6


# append()
myWords <- c("The", "cat", "in", "the", "hat")
paste(append(myWords, c("is", "fun", "to", "read")), collapse=" ")
paste(append(myWords, "funny", 4), collapse=" ")

# grep("//1")
sampMsg <- "This is from myname@subdomain.mydomain.com again"
gsub("(^.*\\w*[a-zA-Z0-9]+@)([a-zA-Z0-9]+\\.[a-zA-Z0-9.]+)(.*$)", "\\1", sampMsg)
gsub("(^.*\\w*[a-zA-Z0-9]+@)([a-zA-Z0-9]+\\.[a-zA-Z0-9.]+)(.*$)", "\\2", sampMsg)
gsub("(^.*\\w*[a-zA-Z0-9]+@)([a-zA-Z0-9]+\\.[a-zA-Z0-9.]+)(.*$)", "\\3", sampMsg)

# rev()
compA
rev(compA)

```
  
Below is some sample code showing examples for the apply statements:  
```{r}
# lapply
args(lapply)
lapply(1:5, FUN=sqrt)
lapply(1:5, FUN=function(x, y=2) { c(x=x, y=y, pow=x^y) }, y=3)
lapply(1:5, FUN=function(x, y=2) { if (x <= 3) {c(x=x, y=y, pow=x^y) } else { c(pow=x^y) } }, y=3)

# sapply (defaults to returning a named vector/array if possible; is lapply otherwise)
args(sapply)
args(simplify2array)
sapply(1:5, FUN=sqrt)
sapply(1:5, FUN=function(x, y=2) { c(x=x, y=y, pow=x^y) }, y=3)
sapply(1:5, FUN=function(x, y=2) { if (x <= 3) {c(x=x, y=y, pow=x^y) } else { c(pow=x^y) } }, y=3)

# vapply (tells sapply exactly what should be returned; errors out otherwise)
args(vapply)
vapply(1:5, FUN=sqrt, FUN.VALUE=numeric(1))
vapply(1:5, FUN=function(x, y=2) { c(x=x, y=y, pow=x^y) }, FUN.VALUE=numeric(3), y=3)

```
  
Below is some sample code for handing dates and times in R:  
```{r}
Sys.Date()
Sys.time()
args(strptime)

rightNow <- as.POSIXct(Sys.time())
format(rightNow, "%Y**%M-%d %H hours and %M minutes", usetz=TRUE)

lastChristmasNoon <- as.POSIXct("2015-12-25 12:00:00", format="%Y-%m-%d %X")
rightNow - lastChristmasNoon

nextUMHomeGame <- as.POSIXct("16/SEP/3 12:00:00", format="%y/%b/%d %H:%M:%S", tz="America/Detroit")
nextUMHomeGame - rightNow

# Time zones available in R
OlsonNames()

# From ?strptime (excerpted)
#
# ** General formats **
# %c Date and time. Locale-specific on output, "%a %b %e %H:%M:%S %Y" on input.
# %F Equivalent to %Y-%m-%d (the ISO 8601 date format).
# %T Equivalent to %H:%M:%S.
# %D Date format such as %m/%d/%y: the C99 standard says it should be that exact format
# %x Date. Locale-specific on output, "%y/%m/%d" on input.
# %X Time. Locale-specific on output, "%H:%M:%S" on input.
# 
# ** Key Components **
# %y Year without century (00-99). On input, values 00 to 68 are prefixed by 20 and 69 to 99 by 19
# %Y Year with century
# %m Month as decimal number (01-12).
# %b Abbreviated month name in the current locale on this platform.
# %B Full month name in the current locale.
# %d Day of the month as decimal number (01-31).
# %e Day of the month as decimal number (1-31), with a leading space for a single-digit number.
# %a Abbreviated weekday name in the current locale on this platform.
# %A Full weekday name in the current locale.
# %H Hours as decimal number (00-23)
# %I Hours as decimal number (01-12)
# %M Minute as decimal number (00-59).
# %S Second as integer (00-61), allowing for up to two leap-seconds (but POSIX-compliant implementations will ignore leap seconds).
# 
# ** Additional Options **
# %C Century (00-99): the integer part of the year divided by 100.
# 
# %g The last two digits of the week-based year (see %V). (Accepted but ignored on input.)
# %G The week-based year (see %V) as a decimal number. (Accepted but ignored on input.)
# 
# %h Equivalent to %b.
# 
# %j Day of year as decimal number (001-366).
# 
# %n Newline on output, arbitrary whitespace on input.
# 
# %p AM/PM indicator in the locale. Used in conjunction with %I and not with %H. An empty string in some locales (and the behaviour is undefined if used for input in such a locale).  Some platforms accept %P for output, which uses a lower-case version: others will output P.
# 
# %r The 12-hour clock time (using the locale's AM or PM). Only defined in some locales.
# 
# %R Equivalent to %H:%M.
# 
# %t Tab on output, arbitrary whitespace on input.
# 
# %u Weekday as a decimal number (1-7, Monday is 1).
# 
# %U Week of the year as decimal number (00-53) using Sunday as the first day 1 of the week (and typically with the first Sunday of the year as day 1 of week 1). The US convention.
# 
# %V Week of the year as decimal number (01-53) as defined in ISO 8601. If the week (starting on Monday) containing 1 January has four or more days in the new year, then it is considered week 1. Otherwise, it is the last week of the previous year, and the next week is week 1. (Accepted but ignored on input.)
# 
# %w Weekday as decimal number (0-6, Sunday is 0).
# 
# %W Week of the year as decimal number (00-53) using Monday as the first day of week (and typically with the first Monday of the year as day 1 of week 1). The UK convention.
# 
# For input, only years 0:9999 are accepted.
# 
# %z Signed offset in hours and minutes from UTC, so -0800 is 8 hours behind UTC. Values up to +1400 are accepted as from R 3.1.1: previous versions only accepted up to +1200. (Standard only for output.)
# 
# %Z (Output only.) Time zone abbreviation as a character string (empty if not available). This may not be reliable when a time zone has changed abbreviations over the years.

```
  
