---
title: "Data Camp Python Notes"
author: "davegoblue"
date: "May 11, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(engine.path=list(python="C:\\Users\\Dave\\AppData\\Local\\Programs\\Python\\Python36-32\\python.exe"))
```

## Background and Overview  
DataCamp offer interactive courses related to Python Programming.  Since R Markdown documents can run simple Python code chunks (though the data is not accessible to future chunks, a large difference from R Markdown for R), this document attempts to summarize notes from the first module.
  
## Python Programming  
###_Intro to Python for Data Science_#
  
Chapter 1 - Python Basics  
  
Hello Python! - focusing on Python specific to data science:  
  
* Designed by Guido Van Rossum (started as a hobby), but has become a general purpose language that can build anything  
* Python is open-source, free, and has packages for data science  
* This course will focus on Python 3.x given that support for Python 2.7 has been (and will continue to) decreasing  
	* Available at https://www.python.org/downloads (have downloaded Python 3.6.1 for Windows)  
    * The DataCamp module uses interactive Python (iPython)  
* Python scripts are simply text files with a .py extension - must use print() inside scripts in order to force printing  
  
Variables and Types - variables names are case-sensitive in Python:  
  
* The single equals sign is the assignment operator  
* The type(myVar) call will return the type of the variable - float, integer ("int"), string ("str"), boolean ("bool"), etc.  
	* The booleans are represented as proper-noun capitalization - True and False  
* String summation is concatenation without spacing (roughly the same as paste0() in R) -- "ab" + "cd" = "abcd" ; note also that "ab" * 2 = "abab"  
	* In general, different types of data will respond differently to the same function  
  
Example code includes:  
```{r engine='python'}

# Example, do not modify!
print(5 / 8)

# Put code below here
print(7 + 10)

# Recall that commented lines are marked by the hash-sign, same as R
# Exponentiation is ** and modulo division is %

# Addition and subtraction
print(5 + 5)
print(5 - 5)

# Multiplication and division
print(3 * 5)
print(10 / 2)

# Exponentiation
print(4 ** 2)

# Modulo
print(18 % 7)

# How much is your $100 worth after 7 years?
print(100 * 1.1**7)


# Create a variable savings
savings = 100

# Print out savings
print(savings)


# Create a variable savings
savings = 100

# Create a variable factor
factor = 1.10

# Calculate result
result = savings * factor ** 7

# Print out result
print(result)


# Create a variable desc
desc = "compound interest"

# Create a variable profitable
profitable = True


# Several variables to experiment with
savings = 100
factor = 1.1
desc = "compound interest"

# Assign product of factor and savings to year1
year1 = savings * factor

# Print the type of year1
print(type(year1))

# Assign sum of desc and desc to doubledesc
doubledesc = desc + desc

# Print out doubledesc
print(doubledesc)


# Definition of savings and result
savings = 100
result = 100 * 1.10 ** 7

# Fix the printout
print("I started with $" + str(savings) + " and now have $" + str(result) + ". Awesome!")

# Definition of pi_string
pi_string = "3.1415926"

# Convert pi_string into float: pi_float
pi_float = float(pi_string)


```
  
The output all comes at once, another difference from R Markdown for R.  In combination with being unable to access any of the variables later in the same document, there are tangible limitations to this approach.

Using Python within R Markdown may be more useful if I install "feather" for both Python and R.  Feather allows for running code in Python, then quick-saving pandas in a way that is quick-readable as frames for the next R chunk.  See https://blog.rstudio.org/2016/03/29/feather/.  

Getting feather for R took just a few seconds using install.packages().  Getting feather for Python 3.6 using Windows seems to require a C++ 14.0 compiler from MS Visual Studio.  So far, that is easier said than done.
  

Chapter 2 - Lists  
  

