---
title: "VAERS Exploration"
author: "davegoblue"
date: '2022-06-24'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background  
This module explores vaccine adverse events (VAERS) in the United States, based on data available at [HHS - VAERS](https://vaers.hhs.gov/data/datasets.html).  Data are maintained on reported adverse events by vaccine, symptom or sign, year, etc. Because VAERS is a reporting system, it may include over-reporting of adverse events (e.g., false reports, errors, spurious correlation, etc.) and under-reporting of adverse events (e.g., patient or doctor is not aware of system or chooses not to spend time reporting). While there is under-reporting and over-reporting in VAERS, it is widely believed that VAERS significantly underestimates adverse events. This under-reporting ratio may differ by attributes such as vaccine, symptom, etc.

The code leverages tidyverse and sourced functions throughout:  
```{r}

# All functions assume that tidyverse and its components are loaded and available
library(tidyverse)

# Generic added R functions will be available
source("./Generic_Added_Utility_Functions_202105_v001.R")

```
  
## Initial Exploration  
An initial exploration is made using the 2020 data. There are three datasets provided in a ZIP file:  
  
* 2020VAERSDATA - patient information (blinded) unique by VAERS_ID such as age, sex, vaccine date, outcome, etc.  
* 2020VAERSSYMPTOMS - additional detail on symptoms, keyed by VAERS_ID (not necessarily unique if many symptoms)  
* 2020VAERSVAX - vaccine type, unique by VAERS_ID  
  
Data are in CSV format and are loaded accordingly:  
```{r, fig.height=9, fig.width=9}

# FILE #1: VAERS DATA (check for uniqueness by VAERS_ID)
vaersData2020 <- readr::read_csv("./RInputFiles/Coronavirus/VAERS/Explore2020/2020VAERSDATA.csv", 
                                 col_types="cccdddccccccccdcccccdccccccccdccccc"
                                 ) %>%
    mutate(across(c("RECVDATE", "RPT_DATE", "DATEDIED", "VAX_DATE", "ONSET_DATE", "TODAYS_DATE"), 
                  .fns=function(x) lubridate::mdy(x)
                  )
           ) %>%
    checkUniqueRows(uniqueBy="VAERS_ID")
glimpse(vaersData2020)

```
  
```{r, fig.height=9, fig.width=9}

# FILE #2: VAERS SYMPTOMS
vaersSymptoms2020 <- readr::read_csv("./RInputFiles/Coronavirus/VAERS/Explore2020/2020VAERSSYMPTOMS.csv", 
                                     col_types="ccccccccccc"
                                     )
glimpse(vaersSymptoms2020)

```
  
```{r, fig.height=9, fig.width=9}  

# FILE #3: VAERS VAX
vaersVax2020 <- readr::read_csv("./RInputFiles/Coronavirus/VAERS/Explore2020/2020VAERSVAX.csv")
glimpse(vaersVax2020)

```
  
```{r, fig.height=9, fig.width=9}

# Check for overlaps and uniqueness of VAERS_ID
vaersOverlap2020 <- vaersData2020 %>%
    select(VAERS_ID) %>%
    mutate(nData=1) %>%
    checkUniqueRows() %>%
    full_join(count(vaersSymptoms2020, VAERS_ID, name="nSymptoms"), by="VAERS_ID") %>%
    full_join(count(vaersVax2020, VAERS_ID, name="nVax"), by="VAERS_ID")
vaersOverlap2020 %>%
    count(nData, nSymptoms, nVax) %>%
    as.data.frame()
vaersOverlap2020 %>% ggplot(aes(x=nSymptoms)) + geom_histogram(binwidth=1)
vaersOverlap2020 %>% ggplot(aes(x=nVax)) + geom_histogram(binwidth=1)


# Pull outlier VAERS_ID (roughly 1% of data has very many vaccines or very many symptoms)
vaersOutlier2020 <- vaersOverlap2020 %>%
    filter(nVax > 4 | nSymptoms > 4)
vaersOutlier2020

# Example data
set.seed(22062422)
vaersCheck <- vaersOutlier2020 %>% sample_n(size=5)
vaersCheck

vaersSymptoms2020 %>%
    right_join(vaersCheck, by="VAERS_ID")
vaersVax2020 %>%
    right_join(vaersCheck, by="VAERS_ID")

```
  
The VAERS symptoms data can be pivoted longer so that it is unique by VAERS_ID and SYMPTOM:  
```{r, fig.height=9, fig.width=9}

vaersSymptomsPivot2020 <- vaersSymptoms2020 %>%
    pivot_longer(-VAERS_ID) %>%
    filter(!is.na(value)) %>%
    mutate(type=ifelse(stringr::str_detect(name, pattern="VERSION"), "code", "symptom")) %>%
    group_by(VAERS_ID, type) %>%
    mutate(number=row_number()) %>%
    ungroup() %>%
    pivot_wider(c(VAERS_ID, number), names_from="type", values_from="value") %>%
    checkUniqueRows(uniqueBy=c("VAERS_ID", "symptom"))
vaersSymptomsPivot2020
colSums(is.na(vaersSymptomsPivot2020))
vaersSymptomsPivot2020 %>% count(symptom, sort=TRUE)
vaersSymptomsPivot2020 %>% count(code, sort=TRUE)

```
  
Data are limited to vaccines received in 2020 (some reports in 2020 are for events in previous years, some events in 2020 will not be reported unitl 2021 or later), and evolution by month plotted:  
```{r, fig.height=9, fig.width=9}

vaersData2020 %>%
    filter(lubridate::year(VAX_DATE)==2020) %>%
    select(VAX_DATE, CAGE_YR, SEX, DIED, L_THREAT, ER_VISIT, HOSPITAL, DISABLE) %>%
    mutate(REPORT="Y", month=customYYYYMM(VAX_DATE)) %>%
    pivot_longer(-c(VAX_DATE, month, CAGE_YR, SEX)) %>%
    group_by(month, SEX, name) %>%
    summarize(nEvent=sum(ifelse(is.na(value), 0, 1)), n=n(), .groups="drop") %>%
    ggplot(aes(x=month, y=nEvent)) + 
    geom_col(aes(group=SEX, fill=SEX), position="stack") + 
    facet_wrap(~name, scales="free_y") + 
    labs(title="Adverse events by severity, sex, and month based on 2020 VAERS data", 
         subtitle="Some events reported in multiple categories", 
         x=NULL, 
         y="Events in 2020 reported in 2020 data"
         ) + 
    theme(axis.text.x = element_text(angle=90))

```
  
The raw data loading processes are converted to functional form:  
```{r, fig.height=9, fig.width=9}

readRawVAERS <- function(fileName, 
                         fnRead=readr::read_csv,
                         colTypes=NULL, 
                         convertDate=c(), 
                         checkUnique=FALSE,
                         glimpseFile=TRUE, 
                         ...
                         ) {
    
    # FUNCTION ARGUMENTS:
    # fileName: location of the CSV file to be read
    # fnRead: the function to be used in reading the file
    # colTypes: character vector of data types by column (NULL means let read_csv guess)
    # convertDate: columns to be converted to date using lubridate::mdy
    # checkUnique: confirm that file is unique by this variable (NULL means check full record; FALSE means do not check)
    # glimpseFile: boolean, should the file be glimpsed?
    # ...: other arguments passed to fileRead (which will be further passed to read_csv)
    
    # Read the VAERS data file and convert dates
    df <- fileRead(fileName, fnRead=fnRead, col_types=colTypes, ...) %>%
        mutate(across(all_of(convertDate), .fns=function(x) lubridate::mdy(x)))
    
    # Check for uniqueness if requested
    if(!isFALSE(checkUnique)) df %>% checkUniqueRows(uniqueBy=checkUnique, returnDF=FALSE)
    
    # Glimpse the file if requested
    if(isTRUE(glimpseFile)) glimpse(df)
    
    # Return the file
    df
    
}

# Individual functions to read each of three files
readRawVAERSData <- function(fileName, glimpseFile=TRUE, ...) {
    
    # FUNCTION ARGUMENTS:
    # fileName: file to be read
    # glimpseFile: boolean, should the file be glimpsed?
    # ...: other arguments to be passed to read_csv
    
    readRawVAERS(fileName, 
                 colTypes="cccdddccccccccdcccccdccccccccdccccc", 
                 convertDate=c("RECVDATE", "RPT_DATE", "DATEDIED", "VAX_DATE", "ONSET_DATE", "TODAYS_DATE"),
                 checkUnique="VAERS_ID", 
                 glimpseFile=glimpseFile, 
                 ...
                 )
    
}

readRawVAERSSymptoms <- function(fileName, glimpseFile=TRUE, ...) {
    
    # FUNCTION ARGUMENTS:
    # fileName: file to be read
    # glimpseFile: boolean, should the file be glimpsed?
    # ...: other arguments to be passed to read_csv
    
    readRawVAERS(fileName, 
                 colTypes="ccccccccccc",
                 checkUnique=NULL, 
                 glimpseFile=glimpseFile, 
                 ...
                 )
}

readRawVAERSVax <- function(fileName, glimpseFile=TRUE, ...) {
    
    # FUNCTION ARGUMENTS:
    # fileName: file to be read
    # glimpseFile: boolean, should the file be glimpsed?
    # ...: other arguments to be passed to read_csv
    
    readRawVAERS(fileName, 
                 checkUnique=FALSE, 
                 glimpseFile=glimpseFile, 
                 ...
                 )
}

# Check that same data is returned
identical(readRawVAERSData("./RInputFiles/Coronavirus/VAERS/Explore2020/2020VAERSDATA.csv", glimpseFile=FALSE), 
          vaersData2020
          )
identical(readRawVAERSSymptoms("./RInputFiles/Coronavirus/VAERS/Explore2020/2020VAERSSYMPTOMS.csv", glimpseFile=FALSE), 
          vaersSymptoms2020
          )
identical(readRawVAERSVax("./RInputFiles/Coronavirus/VAERS/Explore2020/2020VAERSVAX.csv", glimpseFile=FALSE), 
          vaersVax2020
          )

```
  
Additional processing can be run to ensure uniqueness of the vaccines data and convert the symptoms data so there is one row for each symptom-patient:  
```{r, fig.height=9, fig.width=9}

processRawVAERS <- function(dfMain=NULL, 
                            dfSymptoms=NULL, 
                            dfVax=NULL
                            ) {
    
    # FUNCTION ARGUMENTS:
    # dfMain: the main VAERS data file (NULL means do not use)
    # dfSymptoms: the symptoms data file (NULL means do not use)
    # dfVax: the vaccines data file (NULL means do not use)
    
    # Convert the symptoms to be unique by VAERS_ID-SYMPTOM
    if(!is.null(dfSymptoms)) {
        dfSymptoms <- dfSymptoms %>%
            pivot_longer(-VAERS_ID) %>%
            filter(!is.na(value)) %>%
            mutate(type=ifelse(stringr::str_detect(name, pattern="VERSION"), "code", "symptom")) %>%
            group_by(VAERS_ID, type) %>%
            mutate(number=row_number()) %>%
            ungroup() %>%
            pivot_wider(c(VAERS_ID, number), names_from="type", values_from="value") %>%
            checkUniqueRows(uniqueBy=c("VAERS_ID", "symptom"))
        glimpse(dfSymptoms)
        cat("\nNA record counts in pivoted symptoms data:\n")
        colSums(is.na(dfSymptoms)) %>% print()
        cat("\nCounts of top symptoms, codes, and number of symptoms in pivoted symptoms data:\n")
        dfSymptoms %>% count(symptom, sort=TRUE) %>% print()
        dfSymptoms %>% count(code, sort=TRUE) %>% print()
        dfSymptoms %>% count(number, sort=TRUE) %>% print()
    }
    
    # Convert the vaccines data to be unique
    if(!is.null(dfVax)) {
        dfVax <- dfVax %>%
            group_by_at(all_of(names(dfVax))) %>%
            summarize(nRecords=n(), .groups="drop") %>%
            checkUniqueRows(uniqueBy=names(dfVax))
        glimpse(dfVax)
        cat("\nOccurences per vaccine record in raw VAERS vaccines data")
        dfVax %>% count(nRecords, sort=TRUE) %>% print()
    }

    # Check overlap of data if all datasets passed
    if(!is.null(dfMain) & !is.null(dfSymptoms) & !is.null(dfVax)) {
        # Create overlap database
        dfOverlap <- dfMain %>%
            select(VAERS_ID) %>%
            mutate(nData=1) %>%
            full_join(dfSymptoms %>% group_by(VAERS_ID) %>% summarize(nSymptoms=max(number)), by="VAERS_ID") %>%
            full_join(count(dfVax, VAERS_ID, name="nVax"), by="VAERS_ID") %>%
            checkUniqueRows(uniqueBy="VAERS_ID")
        # Confirm that every VAERS_ID is in each file
        cat("\nChecking that overlap of main-symptoms-vax has zero NA (each VAERS_ID is in each file)\n")
        colSums(is.na(dfOverlap)) %>% print()
        cat("\n")
    }

    # Return a list of the processed data (or pass-through of NULL for data not provided)
    list(dfMain=dfMain, dfSymptoms=dfSymptoms, dfVax=dfVax)
    
}

processVAERS2020 <- processRawVAERS(dfMain=vaersData2020, dfSymptoms=vaersSymptoms2020, dfVax=vaersVax2020)
identical(vaersData2020, processVAERS2020$dfMain)
identical(vaersSymptomsPivot2020, processVAERS2020$dfSymptoms)
# Making vaccines data unique changes sort order of some components for records with multiple vaccines
all.equal(unique(vaersVax2020) %>% arrange(VAERS_ID, VAX_DOSE_SERIES, VAX_LOT, VAX_ROUTE, VAX_SITE), 
          select(processVAERS2020$dfVax, -nRecords) %>% arrange(VAERS_ID, VAX_DOSE_SERIES, VAX_LOT, VAX_ROUTE, VAX_SITE)
          )

```
  
A function is created for basic facetted bar plots of evolution in monthly VAERS:  
```{r, fig.height=9, fig.width=9}

plotVAERSMonthly <- function(lst, 
                             vaxYears=NULL, 
                             plotVars=c("DIED", "L_THREAT", "ER_VISIT", "HOSPITAL", "DISABLE", "REPORT"), 
                             segVars=c(),
                             monthAsChar=if(length(segVars)>0) TRUE else FALSE,
                             plotTitle=NULL, 
                             plotSubtitle=if(length(plotVars) > 1) "One event may be in multiple categories" else NULL, 
                             plotYLabel=NULL, 
                             plotXLabel=NULL, 
                             printPlot=TRUE,
                             returnDF=FALSE
                             ) {
    
    # FUNCTION ARGUMENTS:
    # lst: list of processed VAERS data
    # vaxYears: which years of vaccination should be included (NULL means all)
    # plotVars: variables of interest (facets if length greater than 1)
    # segVars: demographic variables of interest
    # monthAsChar: boolean, should month be treated as a character variable (rather than converted to date)?
    # plotTitle: title for the plot
    # plotSubTitle: subtitle for the plot
    # plotYLabel: y-label for the plot
    # plotXLabel: x-label for the plot
    # printPlot: boolean, should the plot be printed?
    # returnDF: boolean, should the data frame be returned?
    
    # Extract data
    df <- lst[["dfMain"]]
    
    # Filter for years if requested
    if(!is.null(vaxYears)) df <- df %>% filter(lubridate::year(VAX_DATE) %in% all_of(vaxYears))
    
    # Create the custom date and reporting flag, keep only relevant variables, pivot, and count
    df <- df %>% 
        mutate(month=customYYYYMM(VAX_DATE), REPORT="Y") %>%
        select(all_of(c("month", plotVars, segVars))) %>%
        pivot_longer(-all_of(c("month", segVars))) %>%
        group_by_at(all_of(c("month", segVars, "name"))) %>%
        summarize(nEvent=sum(ifelse(is.na(value), 0, 1)), n=n(), .groups="drop")

    # Create the plot
    p1 <- df %>% 
        ggplot(aes(x=if(isTRUE(monthAsChar)) month else lubridate::ym(month), y=nEvent)) + 
        labs(title=plotTitle, 
             subtitle=plotSubtitle, 
             x=plotXLabel, 
             y=plotYLabel
             ) + 
        theme(axis.text.x = element_text(angle=90))
    
    # Add a line plot if there are no segments, bar plot otherwise
    if(length(segVars)==0) p1 <- p1 + geom_line()
    else p1 <- p1 + 
        geom_col(aes(group=get(segVars), fill=get(segVars)), position="stack") + 
        scale_fill_discrete(segVars)
    
    # Add facets if relevant
    if(length(plotVars) > 1) p1 <- p1 + facet_wrap(~name, scales="free_y")
    
    # Print the plot
    if(isTRUE(printPlot)) print(p1)

    # Return the data frame
    if(isTRUE(returnDF)) return(df)
    
}

plotVAERSMonthly(processVAERS2020, 
                 plotTitle="Adverse events by severity and month based on 2020 VAERS data", 
                 plotYLabel="Events reported in 2020 data"
                 )
plotVAERSMonthly(processVAERS2020, 
                 vaxYears=2020, 
                 plotTitle="Adverse events by severity and month based on 2020 VAERS data", 
                 plotYLabel="Events in 2020 reported in 2020 data"
                 )
plotVAERSMonthly(processVAERS2020, 
                 vaxYears=2020, 
                 segVars="SEX",
                 plotTitle="Adverse events by severity, sex, and month based on 2020 VAERS data", 
                 plotYLabel="Events in 2020 reported in 2020 data", 
                 returnDF=TRUE
                 )

```
  
A function is written to create a skinnier version of processed data:  
```{r, fig.height=9, fig.width=9}

makeSkinnyVAERSList <- function(lst, 
                                excludeMain=c("SYMPTOM_TEXT"), 
                                excludeVax=c(), 
                                excludeSymptoms=c(), 
                                makeSymptomMap=TRUE
                                ) {
    
    # FUNCTION ARGUMENTS:
    # lst: an initial list file
    # excludeMain: variables to exclude from dfMain (symptoms are in a separate file)
    # excludeVax: variables to exclude from dfVax
    # excludeSymptoms: variables to exclude from dfSymptoms
    # makeSymptomMap: should dfSymptoms be made relational by moving symptom text to a separate file?
    #                 will replace symptom with the relevant code
    
    # Apply the symptom mapping if relevant
    if(isTRUE(makeSymptomMap)) {
        symptomMap <- lst[["dfSymptoms"]] %>% count(symptom) %>% mutate(symptomID=as.integer(row_number()))
        dfSymptoms <- lst[["dfSymptoms"]] %>%
            left_join(select(symptomMap, symptom, symptomID), by="symptom") %>%
            select(-symptom) %>%
            rename(symptom=symptomID)
    }
    else {
        symptomMap <- NULL
        dfSymptoms <- lst[["dfSymptoms"]]
    }
    
    # Create the skinny list
    list(dfMain=lst[["dfMain"]] %>% select(-all_of(excludeMain)), 
         dfSymptoms=dfSymptoms, 
         dfVax=lst[["dfVax"]] %>% select(-all_of(excludeVax)), 
         symptomMap=symptomMap
         )
    
}

tmpOrigSize <- object.size(processVAERS2020)
tmpNewSize <- object.size(makeSkinnyVAERSList(processVAERS2020))
cat("\nOld object size: ", 
    round(tmpOrigSize/1024/1024, 1), 
    " GB and new object size: ", 
    round(tmpNewSize/1024/1024, 1), 
    " GB (", 
    round(100*(1-tmpNewSize/tmpOrigSize), 1), 
    "% reduction)\n", 
    sep=""
    )

```
  
The 2019 data are loaded using skinny format:  
```{r, fig.height=9, fig.width=9}

mainLoc <- "./RInputFiles/Coronavirus/VAERS/Explore2019/2019VAERSDATA.csv"
sympLoc <- "./RInputFiles/Coronavirus/VAERS/Explore2019/2019VAERSSYMPTOMS.csv"
vaxLoc <- "./RInputFiles/Coronavirus/VAERS/Explore2019/2019VAERSVAX.csv"

skinnyVAERS2019 <- processRawVAERS(dfMain=readRawVAERSData(mainLoc, glimpseFile=FALSE), 
                                   dfSymptoms=readRawVAERSSymptoms(sympLoc, glimpseFile=FALSE), 
                                   dfVax=readRawVAERSVax(vaxLoc, glimpseFile=FALSE)
                                   ) %>%
    makeSkinnyVAERSList()

plotVAERSMonthly(skinnyVAERS2019, 
                 plotTitle="Adverse events by severity and month based on 2019 VAERS data", 
                 plotYLabel="Events reported in 2019 data"
                 )
plotVAERSMonthly(skinnyVAERS2019, 
                 vaxYears=2019, 
                 plotTitle="Adverse events by severity and month based on 2019 VAERS data", 
                 plotYLabel="Events in 2019 reported in 2019 data"
                 )
plotVAERSMonthly(skinnyVAERS2019, 
                 vaxYears=2019, 
                 segVars="SEX",
                 plotTitle="Adverse events by severity, sex, and month based on 2019 VAERS data", 
                 plotYLabel="Events in 2019 reported in 2019 data", 
                 returnDF=TRUE
                 )

```
  
