---
title: "Wild Card Game Analysis"
author: "davegoblue"
date: "2023-05-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This file builds on _v001 and explores card games that use a standard 52-card deck, optionally modified so that one or more cards are wild, as follows:  
  
* One or more jokers may be added (the card has no natural rank or suit but may take on the most favorable)  
* One or more natural cards may also be used as a wild (in deuces wild, the deuce has a natural suit and rank and can also take on the suit and rank of a more favorable card)  
  
The analysis treats cards 1-52 as being the standard deck, with suits being 1-13, 14-26, 27-39, and 40-52. Counting the first card of each suit as 1, ranks are 1=Ace, 2-10=same card as value, 11=Jack, 12=Queen, 13=King. The Ace almost always plays as the highest rank, but can be treated as low for making a 5432A straight
  
The core functions and setup steps are captured at the top of the file

## Methodology
### Basic Parameters and Core Functions
Several basic parameters of the game are established:  
```{r}

library(tidyverse) # tidyverse functionality is included throughout

source("./Generic_Added_Utility_Functions_202105_v001.R") # Basic functions

nCards <- 53 # This is a standard 52-card deck with a joker
nRanks <- 13 # There are 13 ranks A, 2-10, J, Q, K
nSuits <- 4 # There are four suits, each of 13 cards
perHand <- 5 # Number of cards drawn per hand
idxWild <- c(2, 15, 28, 41, 53) # The indices of the cards that can be considered wild (deuces and the joker)

# Check alignment of basic parameters
if(nSuits*nRanks > nCards) stop("\nerror, misaligned parameters for number of ranks, suits, and cards\n")
if(!isTRUE(setdiff(1:nCards, 1:(nRanks*nSuits)) %in% idxWild %>% reduce(.f=`&`)))
    stop("\nError, any cards outside the standard nRanks*nSuits must be included in idxWild\n")

# Announce the number of wild cards
cat("\nThe game will be played with a", nCards, "card deck with", length(idxWild), "cards considered wild")
cat("\nWild cards are of indices", paste0(idxWild, collapse=", "))

# Announce any cards of no suit or rank
if(!isTRUE(idxWild %in% 1:(nRanks*nSuits) %>% reduce(.f=`&`))) { 
    cat("\nWilds with no natural rank or suit at indices:", 
        paste0(idxWild[!(idxWild %in% 1:(nRanks*nSuits))], collapse=",")
        )
} else cat("\nAll wilds have a natural suit and rank")

# Announce wilds with natural ranks and suits
if(isTRUE(idxWild %in% 1:(nRanks*nSuits) %>% reduce(.f=`|`))) { 
    tmpIdxWild <- idxWild[idxWild %in% 1:(nRanks*nSuits)]
    cat("\nWilds with natural rank and suit at indices:", paste0(tmpIdxWild, collapse=", "))
    cat("\nThese are of suit-independent rank", paste0((tmpIdxWild-1)%%nRanks+1, collapse=", "), "\n")
} else cat("\nNo wilds have a natural suit and rank\n")

# Declaration of hand values
handValues <- tibble::tibble(handRank=c(1:11, 99), 
                             handDesc=c("5W", "RF", "5K", "SF", "4K", "FH", "FL", "ST", "3K", "2P", "1P", "HC"),
                             handLabel=paste0(handRank, " (", handDesc, ")"),
                             blindPay=c(2000, 100, 40, 25, 5, 4, 3, 2, 0, 0, 0, 0),
                             wildPay=c(2000, 90, 70, 25, 6, 5, 4, 3, 1, -1, -1, -1), 
                             naturalPay=c(NA, 1000, NA, 200, 60, 30, 25, 20, 6, -1, -1, -1)
                             )
handValues

# Create a mapping vector for type to description
vecMapHands <- handValues$handLabel %>% purrr::set_names(handValues$handRank)

```

The main functions are also captured:
```{r}

# Determine if a hand is a flush
isFlush <- function(suits, useWild=4) {
    apply(suits, 1, FUN=function(x) ifelse(min(x)==max(x), TRUE, (max(x[x!=useWild])==min(x[x!=useWild]))))
}

# Determine if a hand is a straight (hard-coded for 5-card hands in a 13-card deck, Ace high or low)
isStraight <- function(rankCounts, isEligible=TRUE) {
    # Exclusion matrix for straights, currently designed only for 13-card decks and A high or low
    strMatrix <- matrix(data=0L, nrow=13, ncol=10)
    strMatrix[c(1, 10, 11, 12, 13), 1] <- 1L
    for (intCtr in 1:9) { strMatrix[intCtr:(intCtr+4), intCtr+1] <- 1L }
    apply(rankCounts, 1, max)<=1 & rowSums(rankCounts%*%(strMatrix-1)==0)>=1 & isEligible
}

# Hard-coded for 52-card deck with 1 joker
findHandTypes <- function (aHands, retAll=FALSE) {
    
    # Track hands that contain the joker and total number of wilds
    aJoker <- rowSums(aHands==53)>0
    nWild <- aJoker + rowSums(aHands%%13==2)
    
    # 1. Calculate natural values
    # 1a. Calculate ranks and suits, counting the joker as having no "rank" and being of a fifth "suit"
    aRanks <- ifelse(aHands==53, 0, 1 + (aHands-1) %% 13)
    aSuits <- ifelse(aHands==53, 0, 1 + (aHands-1) %/% 13)

    # 1b. Find the counts by rank in each hand
    aRankCount <- matrix(data=-1L, nrow=nrow(aHands), ncol=13)
    for (intCtr in 1:13) { aRankCount[, intCtr] <- rowSums(aRanks == intCtr) }
        
    # 1c. Find the natural flushes and straights
    nFlush <- isFlush(aSuits, useWild=-1)
    nStraight <- isStraight(aRankCount, isEligible=!aJoker)

    # 1d. Find max and count of ranks
    nQuads <- rowSums(aRankCount == 4)
    nTrips <- rowSums(aRankCount == 3)
    nPairs <- rowSums(aRankCount == 2)

    # 1e. Score the natural hand values (default is that a hand has nothing of value)
    nType <- rep(0L, nrow(aHands))
    # Five wilds (cannot exist) <- 1
    nType[nFlush==1 & nStraight==1 & aRankCount[, 1]==1 & aRankCount[, 13]==1] <- 2 # Royal
    # Five of a kind (cannot exist as natural) <- 3
    nType[nFlush==1 & nStraight==1 & (aRankCount[, 1]!=1 | aRankCount[, 13]!=1)] <- 4 # Straight flush
    nType[nQuads==1] <- 5 # Four of a kind
    nType[nTrips==1 & nPairs==1] <- 6 # Full House
    nType[nFlush==1 & nStraight==0] <- 7 # Flush
    nType[nFlush==0 & nStraight==1] <- 8 # Straight
    nType[nTrips==1 & nPairs==0] <- 9 # Three of a Kind
    nType[nPairs==2] <- 10 # Two Pair
    nType[nTrips==0 & nPairs==1] <- 11 # Pair

    
    # 2. Calculate wild values
    # 2a. Calculate ranks and suits, counting the joker and 2's as having no "rank" and being of a fifth "suit"
    wRanks <- ifelse(aHands==53|(aHands%%13==2), 0, 1 + (aHands-1) %% 13)
    wSuits <- ifelse(aHands==53|(aHands%%13==2), 0, 1 + (aHands-1) %/% 13)

    # 2b. Find the counts by rank in each hand
    wRankCount <- matrix(data=-1L, nrow=nrow(aHands), ncol=13)
    for (intCtr in 1:13) { wRankCount[, intCtr] <- rowSums(wRanks == intCtr) }
        
    # 2c. Find the wild flushes and straights
    wFlush <- isFlush(wSuits, useWild=0)
    wStraight <- isStraight(wRankCount, isEligible=TRUE)
    
    # 2d. Find max and count of ranks
    wQuads <- rowSums(wRankCount == 4)
    wTrips <- rowSums(wRankCount == 3)
    wPairs <- rowSums(wRankCount == 2)

    # 2e. Score the wild hand values (default is that a hand has nothing of value)
    wType <- rep(0L, nrow(aHands))
    
    # 2e1. Calculate number of royal cards in hand
    nRoyal <- rowSums(wRankCount[, c(1, 10, 11, 12, 13)])
    
    # 2e2. Five wilds (always five wilds)
    wType[nWild==5] <- 1 # Five wild
    
    # 2e3. Four wilds (always either royal flush or five of a kind)
    wType[nWild==4 & nRoyal==1] <- 2 # Royal
    wType[nWild==4 & nRoyal==0] <- 3 # Five of a kind
    
    # 2e4. Three wilds (always either royal flush or five of a kind or straight flush or four of a kind)
    wType[nWild==3 & wFlush==1 & wStraight==1 & nRoyal==2] <- 2 # Royal
    wType[nWild==3 & wPairs==1] <- 3 # Five of a kind
    wType[nWild==3 & wFlush==1 & wStraight==1 & nRoyal<2] <- 4 # Straight Flush
    wType[nWild==3 & !(wFlush==1 & wStraight==1) & wPairs==0] <- 5 # Four of a kind
    
    # 2e5. Two wilds (always either royal, five of a kind, straight flush, four of a kind, flush, straight, trips)
    # Cannot be a full house as any 3-2 using two wilds could instead be 5-0 or 1-4 and thus a better hand
    wType[nWild==2 & wFlush==1 & wStraight==1 & nRoyal==3] <- 2 # Royal
    wType[nWild==2 & wTrips==1] <- 3 # Five of a kind
    wType[nWild==2 & wFlush==1 & wStraight==1 & nRoyal<3] <- 4 # Straight Flush
    wType[nWild==2 & wPairs==1] <- 5 # Four of a kind
    wType[nWild==2 & wFlush==1 & wStraight==0] <- 7 # Flush
    wType[nWild==2 & wFlush==0 & wStraight==1] <- 8 # Straight
    wType[nWild==2 & !(wFlush==1 | wStraight==1) & wPairs==0 & wTrips==0] <- 9 # Three of a kind
    
    # 2e6. One wild (can be anything from pair to royal, with the exception of never two pair)
    wType[nWild==1 & wFlush==1 & wStraight==1 & nRoyal==4] <- 2 # Royal
    wType[nWild==1 & wQuads==1] <- 3 # Five of a kind
    wType[nWild==1 & wFlush==1 & wStraight==1 & nRoyal<4] <- 4 # Straight Flush
    wType[nWild==1 & wTrips==1] <- 5 # Four of a kind
    wType[nWild==1 & wPairs==2] <- 6 # Full House
    wType[nWild==1 & wFlush==1 & wStraight==0] <- 7 # Flush
    wType[nWild==1 & wFlush==0 & wStraight==1] <- 8 # Straight
    wType[nWild==1 & wPairs==1] <- 9 # Three of a kind
    wType[nWild==1 & !(wFlush==1 | wStraight==1) & wPairs==0 & wTrips==0 & wQuads==0] <- 11 # Pair
    
    # 2e7. No wild (is identical to natural)
    wType[nWild==0] <- nType[nWild==0]
    
    # Return the requested data
    if (isTRUE(retAll)) {
        list(nType=nType, 
             wType=wType, 
             nWild=nWild,
             wRankCount=wRankCount,
             aRankCount=aRankCount
             )
    } else {
        list(nType=nType, wType=wType, nWild=nWild)
    }
    
}

# Function to get the cards in a straight
getStraightType <- function(x) {
    if(min(x[x>0]) >= 10) c(10, 11, 12, 13, 14)
    else if(min(x[x>0]) <= 5 & max(x)==14) c(2, 3, 4, 5, 14) # this treats Ace as the last card to resolve (Ace low)
    else seq(min(x[x>0]), min(x[x>0])+4)
}

# Function to get the cards in a flush
getFlushType <- function(x) {
    # Wild cards should be added as highest rank not already in hand (try A then K then Q then J then T)
    numAdd <- sum(x==0)
    if(numAdd > 0) {
        notThere <- (14:10)[!((14:10) %in% x)][1:numAdd]
        x <- c(x[x!=0], notThere)
    }
    sort(x, decreasing=TRUE)
}

# Function to get the cards in a four of a kind, full house, three of a kind, two pair, pair, or high card
getHighTypes <- function(rks, idx=1:14, outNum=c(1, 1, 1, 1, 1)) {
    vecOut <- rep(0L, length(outNum))
    for(x in seq_along(outNum)) {
        vecOut[x]<-max((rks==max(rks)) * idx)
        rks[match(vecOut[x], idx)] <- 0
    }
    rep(vecOut, times=outNum)
}

# Function to get probabilities
findHandProbs <- function(exRank, idxExclude=NULL, df=sortedHandRanks) {
    rkData <- if(is.null(idxExclude)) df$rank else df$rank[-idxExclude]
    qualData <- if(is.null(idxExclude)) df$qual else df$qual[-idxExclude]
    pLN <- mean(rkData < exRank & qualData=="N")
    pLQ <- mean(rkData < exRank & qualData=="Q")
    pTN <- mean(rkData == exRank & qualData=="N")
    pTQ <- mean(rkData == exRank & qualData=="Q")
    pWN <- mean(rkData > exRank & qualData=="N")
    pWQ <- mean(rkData > exRank & qualData=="Q")
    pWNoTie <- (pWN+pWQ)/(1-pTN-pTQ)
    c(pLN=pLN, pLQ=pLQ, pTN=pTN, pTQ=pTQ, pWN=pWN, pWQ=pWQ, pWNoTie=pWNoTie)
}

# Function to get key values
getKeyValues <- function(keyRank, exclCards=c(), lstExclude=cardInHand, dfSort=sortedHandRanks) {

    # Get the union of exclusions
    if(length(exclCards)==0) exclHands<-c()
    else exclHands <- purrr::reduce(.x=lstExclude[c(exclCards)], .f=union)

    dfSort %>%
        filter(!(handNum %in% exclHands)) %>%
        summarize(pLQ=mean(qual=="Q" & rank<keyRank),
                  pLN=mean(qual=="N" & rank<keyRank), 
                  pTQ=mean(qual=="Q" & rank==keyRank),
                  pTN=mean(qual=="N" & rank==keyRank), 
                  pWQ=mean(qual=="Q" & rank>keyRank),
                  pWN=mean(qual=="N" & rank>keyRank), 
                  n=n()
                  ) %>%
        mutate(pWNoTie=(pWQ+pWN)/(1-pTQ-pTN), rank=keyRank)
        
}

calculateResults <- function(useHandNums, dfRanks=sortedHandRanks, hands=mtxHands, bPay=blindPay) {

    map_dfr(useHandNums, 
            .f=function(x) {
                getKeyValues(keyRank=dfRanks$rank[x], exclCards=as.vector(hands[x,])) %>%
                    mutate(type=dfRanks$wType[x], 
                           handNum=x, 
                           betPlay=ifelse((4*pWQ+3*pWN+2*pTQ+2*pTN)<(pLQ+pLN), 0, ifelse(pWNoTie<0.5, 1, 2)), 
                           evAnte=ifelse(betPlay==0, -1, pWQ-pLQ-pLN), 
                           evPlay=betPlay*(pWQ+pWN-pLQ-pLN),
                           evBlind=ifelse(betPlay==0, -1, bPay[paste0("b_", type)]*(pWQ+pWN)-pLQ-pLN), 
                           evAll=evAnte+evPlay+evBlind
                           )
                }
            )
}

getHandOutcome <- function(pWQ, pWN, tie, lose, betPlay, blindPay) {
    if(betPlay==0) return(tibble::tibble(result="fold", ante=-1, blind=-1, play=0))
    result <- sample(c("win/qualify", "win/no qualify", "tie", "lose"), 
                     size=1, 
                     replace=TRUE, 
                     prob=c(pWQ, pWN, tie, lose)
                     )
    return(tibble::tibble(result=result, 
                          ante=case_when(result=="win/qualify" ~ 1, 
                                         result %in% c("win/no qualify", "tie") ~ 0, 
                                         result=="lose" ~ -1, 
                                         TRUE ~ -99
                                         ),
                          blind=case_when(result %in% c("win/qualify", "win/no qualify") ~ blindPay, 
                                          result=="tie" ~ 0, 
                                          result=="lose" ~ -1, 
                                          TRUE ~ -99
                                          ), 
                          play=case_when(result %in% c("win/qualify", "win/no qualify") ~ betPlay, 
                                         result=="tie" ~ 0, 
                                         result=="lose" ~ -betPlay, 
                                         TRUE ~ -99
                                         )
                          )
           )
    
}

# Function to check if file exists and run if so
runHandSequence <- function(handNums, 
                            rdsName, 
                            skipIfExists=TRUE, 
                            saveFile=TRUE, 
                            ovrWrite=FALSE, 
                            returnFile=!isTRUE(saveFile) | (!isTRUE(skipIfExists) & !isTRUE(ovrWrite))
                            ) {
    
    # FUNCTION ARGUMENTS:
    # handNums: sequence of hand numbers to evaluate
    # rdsName: name of the RDS file (if exists or to be saved - no directory, no .RDS extension)
    # skipIfExists: boolean, if TRUE, check whether the file is already saved, and skip running if so
    # saveFile: boolean, should the file be saved after running
    # ovrWrite: boolean, should the previous file be overwritten?
    # returnFile: boolean, should the file be returned after creation?
    
    # Manage the file existence and decision to run or not run
    keyLoc <- paste0(formals(readFromRDS)$dir, rdsName, formals(readFromRDS)$addSuffix)
    fileExists <- file.exists(keyLoc)
    if(isTRUE(skipIfExists) & isTRUE(fileExists)) {
        cat("\nFile", keyLoc, "already exists, exiting without running or overwriting\n")
        return(NULL)
    }
    
    # Run the process and report on the time
    t1 <- proc.time()
    df <- calculateResults(useHandNums=handNums)
    print(proc.time()-t1)
    
    # Save the file if requested (overwriting the existing file)
    if(isTRUE(saveFile)) {
        # If file is to be overwritten, delete it
        if(isTRUE(ovrWrite) & isTRUE(fileExists)) {
            cat("\nRemoving existing file", keyLoc, "so new file can be written\n")
            Sys.chmod(keyLoc, mode="0777", use_umask=FALSE)
            file.remove(keyLoc)
        }
        # Write the file
        saveToRDS(obj=df, file=paste0(rdsName, ".RDS"), ovrWriteError=FALSE)
    }
    
    # Return the file if requested
    if(isTRUE(returnFile)) return(df)
    
}

# Convert 14, 13, 12, 11, 10 to A, K, Q, J, T
numberToCard <- function(x) 
    case_when(x==14 ~ "A", x==13 ~ "K", x==12 ~ "Q", x==11 ~ "J", x==10 ~ "T", TRUE ~ as.character(x))

# Hand types possible given excluded cards
handTypesByExclusion <- function(exclCards=c(), 
                                 df=sortedHandRanks, 
                                 pivotWider=FALSE, 
                                 checkRank=NULL,
                                 keyVars="wType"
                                 ) {
    
    # Function arguments
    # exclCards: cards to be excluded
    # df: file containing hand ranks and types
    # pivotWider: boolean, whould outputs be pivoted wider (unique by keyVars)?
    # checkRank: rank for win/lose/tie (NULL means do not compare ranks)
    # keyVars: key variables to be included in the final output
    
    # Add rankResult as a variable if checkRank is not NULL
    if(!is.null(checkRank)) {
        if(!("rankResult" %in% keyVars)) keyVars <- c(keyVars, "rankResult")
    }
    
    # Create the excluded hand numbers
    exclHands <- purrr::reduce(.x=cardInHand[exclCards], .f=union, .init=c())
                               
    # Create the list of hand types
    df <- df %>% 
        mutate(isOut=ifelse(handNum %in% exclHands, "excl", "avail")) 
    if(!is.null(checkRank))  {
        df <- df %>% 
            mutate(rankResult=case_when(rank<checkRank~"lose", rank==checkRank~"tie", rank>checkRank~"win"))
    }
    df <- df %>% 
        group_by(across(.cols=c(keyVars, "isOut"))) %>%
        summarize(n=n(), .groups="drop")
    
    # Pivot wider if requested, otherwise return as-is
    if(isTRUE(pivotWider)) 
        df %>% pivot_wider(id_cols=all_of(keyVars), names_from="isOut", values_from="n", values_fill=0)
    else df
    
}

# Automates calls to handTypeExclusion()
compareHandExclusion <- function(c1, 
                                 c2, 
                                 keyVars=c("wType", "nWild"), 
                                 checkRank=NULL,
                                 labs=NULL
                                 ) {
    
    # FUNCTION ARGUMENTS
    # c1: card set number 1
    # c2: card set number 2
    # keyVars: variables passed to handTypesByExclusion
    # checkRank: the rank for the relevant hands - can be passed as length 1 or length 2
    #            (NULL means do not check win/lose/tie)
    # labs: labels for the relevant columns (must be length-2)
    
    # Get the relevant checkRank parameters
    if(is.null(checkRank)) {
        cr1 <- NULL
        cr2 <- NULL
    } else if(length(checkRank)==1) {
        cr1 <- checkRank
        cr2 <- checkRank
    } else if(length(checkRank)==2) {
        cr1 <- checkRank[1]
        cr2 <- checkRank[2]
    } else stop("\nParameter checkRank must be NULL or length-1 or length-2\n")
    
    # Check that labels is of appropriate length
    if(is.null(labs)) {
        lab1 <- "Cards 1"
        lab2 <- "Cards 2"
    } else if(length(labs)==2) {
        lab1 <- labs[1]
        lab2 <- labs[2]
    } else stop("\nlabs must be NULL or length-2\n")
    
    # Get the appropriate grouping variables
    groupVars <- c("src", setdiff(keyVars, "nWild"), "hasWild", "isOut")
    if(!is.null(checkRank)) groupVars <- c(groupVars, "rankResult")
    
    handTypesByExclusion(exclCards=c1, keyVars=keyVars, checkRank=cr1) %>%
        bind_rows(handTypesByExclusion(exclCards=c2, keyVars=keyVars, checkRank=cr2), .id="src") %>%
        mutate(src=c("1"=lab1, "2"=lab2)[src]) %>%
        filter(isOut=="avail") %>%
        mutate(hasWild=(nWild>0)) %>%
        group_by(across(all_of(groupVars))) %>%
        summarize(n=sum(n), .groups="drop") %>%
        pivot_wider(id_cols=setdiff(groupVars, "src"), 
                    names_from="src", 
                    values_from="n", 
                    values_fill=0
                    ) %>%
        mutate(delta=get(lab2)-get(lab1)) %>%
        select(setdiff(groupVars, "src"), all_of(lab1), all_of(lab2), "delta", everything()) %>%
        arrange(desc(hasWild), across(setdiff(keyVars, "nWild")))

}

```
  
Several key objects are also created:  
```{r, cache=TRUE}

# 1. Create the matrix of all possible hand types
startTime <- proc.time()

# Create a matrix of all possible hand indices
mtxHands <- t(combn(1:nCards, perHand))
str(mtxHands)

proc.time() - startTime


# 2. Calculate all hand types
tmpTime <- proc.time(); tmpHandTypes<-findHandTypes(mtxHands, retAll=TRUE); proc.time()-tmpTime


# 3. Create tibble of all possible hand types and numbers of wilds
tblAllTypes <- tibble::tibble(wType=tmpHandTypes$wType, 
                              nType=tmpHandTypes$nType, 
                              nWild=tmpHandTypes$nWild
                              ) %>%
    mutate(wType=ifelse(wType==0, 99, wType), 
           nType=ifelse(nType==0, 99, nType)
           )
tblAllTypes


# 4. Payout grid for all possible values
allValueGrid <- expand.grid(wildType=c(1:11, 99), naturalType=c(1:11, 99)) %>%
    tibble::tibble() %>%
    full_join(count(tblAllTypes, wType, nType), by=c("wildType"="wType", "naturalType"="nType")) %>%
    mutate(bestType=pmin(wildType, naturalType)) %>%
    left_join(select(handValues, handRank, wildPay), by=c("wildType"="handRank")) %>%
    left_join(select(handValues, handRank, naturalPay), by=c("naturalType"="handRank")) %>%
    left_join(select(handValues, handRank, blindPay), by=c("bestType"="handRank")) %>%
    mutate(bonusPay=pmax(wildPay, naturalPay, na.rm=TRUE))
allValueGrid


# 5. Numbers of wilds and ranks by hand
perHandNWild <- (rowSums(mtxHands==53)>0) + rowSums(mtxHands%%13==2)
perHandRanks <- ifelse(mtxHands==53, 0, 1 + (mtxHands-1) %% 13) # track joker as wild
perHandRanks[perHandRanks==2] <- 0 # track deuces as wild, ignore rank (place as best later)
perHandRanks[perHandRanks==1] <- 14 # count Aces as high
perHandRankCount <- matrix(data=-1L, nrow=nrow(mtxHands), ncol=14)
for (intCtr in 1:14) { perHandRankCount[, intCtr] <- rowSums(perHandRanks == intCtr) }

```
  
Hand ranks are also calculated:  
```{r cache=TRUE}

mtxTieBreak <- matrix(data=0L, nrow=nrow(mtxHands), ncol=13)

startTime <- proc.time()

# 1. Five wild (nothing to do)
# 2. Royal flush (make the ranks 1 in the last 5 columns)
perHandRanks[tmpHandTypes$wType==2, ] <- t(apply(perHandRanks[tmpHandTypes$wType==2, ], 1, FUN=getStraightType))
# 3. Five of a kind (make the ranks all the same as the only rank)
perHandRanks[tmpHandTypes$wType==3, ] <- t(apply(perHandRankCount[tmpHandTypes$wType==3, ], 
                                           1, 
                                           FUN=getHighTypes, 
                                           outNum=c(5)
                                           )
                                     )
# 4. Straight flush (make the ranks all the best straight possible)
perHandRanks[tmpHandTypes$wType==4, ] <- t(apply(perHandRanks[tmpHandTypes$wType==4, ], 1, FUN=getStraightType))
# 5. Four of a kind (make the ranks four of the most frequent card or the highest card, one of the other)
perHandRanks[tmpHandTypes$wType==5, ] <- t(apply(perHandRankCount[tmpHandTypes$wType==5, ], 
                                           1, 
                                           FUN=getHighTypes, 
                                           outNum=c(4, 1)
                                           )
                                     )
# 6. Full house (make the ranks three of the most frequent card, two of the other)
perHandRanks[tmpHandTypes$wType==6, ] <- t(apply(perHandRankCount[tmpHandTypes$wType==6, ], 
                                           1, 
                                           FUN=getHighTypes, 
                                           outNum=c(3, 2)
                                           )
                                     )
# 7. Flush (need to implement)
perHandRanks[tmpHandTypes$wType==7, ] <- t(apply(perHandRanks[tmpHandTypes$wType==7, ], 1, FUN=getFlushType))
# 8. Straight (make the ranks all the best straight possible)
perHandRanks[tmpHandTypes$wType==8, ] <- t(apply(perHandRanks[tmpHandTypes$wType==8, ], 1, FUN=getStraightType))
# 9. Three of a kind (make the ranks three of the most frequent or highest card, one each of the other)
perHandRanks[tmpHandTypes$wType==9, ] <- t(apply(perHandRankCount[tmpHandTypes$wType==9, ], 
                                           1, 
                                           FUN=getHighTypes, 
                                           outNum=c(3, 1, 1)
                                           )
                                     )
# 10. Two pair (make the ranks two of the highest pair, two of the lowest pair, one of the remainder)
perHandRanks[tmpHandTypes$wType==10, ] <- t(apply(perHandRankCount[tmpHandTypes$wType==10, ], 
                                            1, 
                                            FUN=getHighTypes, 
                                            outNum=c(2, 2, 1)
                                            )
                                      )
# 11. One pair (make the ranks two of the pair or highest card, one each of the remainder)
perHandRanks[tmpHandTypes$wType==11, ] <- t(apply(perHandRankCount[tmpHandTypes$wType==11, ], 
                                            1, 
                                            FUN=getHighTypes, 
                                            outNum=c(2, 1, 1, 1)
                                            )
                                      )
# 0. High card (make the ranks in order of the highest cards)
perHandRanks[tmpHandTypes$wType==0, ] <- t(apply(perHandRankCount[tmpHandTypes$wType==0, ], 
                                            1, 
                                            FUN=getHighTypes, 
                                            outNum=c(1, 1, 1, 1, 1)
                                            )
                                      )

proc.time() - startTime

# Create the hand ranking table
allHandRanks <- tibble::tibble(handNum=1:nrow(mtxHands), wType=tmpHandTypes$wType) %>% 
    bind_cols(perHandRanks) %>% 
    set_names(c("handNum", "wType", "tb1", "tb2", "tb3", "tb4", "tb5")) %>% 
    mutate(wType=ifelse(wType==0, 99, wType)) %>% 
    arrange(wType, desc(tb1), desc(tb2), desc(tb3), desc(tb4), desc(tb5)) %>% 
    mutate(chgRank=ifelse(row_number()==1 | wType!=lag(wType) | tb1 != lag(tb1) | tb2 != lag(tb2) | tb3 != lag(tb3) | tb4 != lag(tb4) | tb5 != lag(tb5), 1, 0), 
           rank=cumsum(chgRank)
           )

# Summary of hand ranks
allHandRanks %>%
    count(wType, rank, tb1, tb2, tb3, tb4, tb5) %>%
    arrange(-n, rank) %>%
    mutate(nvr=row_number()) %>%
    group_by(wType) %>%
    mutate(nty=row_number()) %>%
    ungroup() %>%
    filter(nvr<=10 | nty==1)

# Sort allHandRanks and ensure handNum and row_number() match
sortedHandRanks <- allHandRanks %>%
    mutate(qual=ifelse(wType<=11, "Q", "N")) %>%
    arrange(handNum) %>%
    mutate(nWild=perHandNWild)
if (sum((sortedHandRanks$handNum) != (1:nrow(sortedHandRanks)))>0) 
    stop("\nError in handNum order in sortedHandRanks\n")

# Get all exclusion indices once
startTime <- proc.time()
cardInHand <- lapply(1:53, FUN=function(x) which(rowSums(mtxHands==x)>0))
proc.time() - startTime

```

