---
title: "Open Meteo Weather Exploration"
author: "davegoblue"
date: "2024-12-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
Open-Meteo maintains an [API for historical weather](https://open-meteo.com/en/docs/historical-weather-api) that allows for non-commercial usage of historical weather data maintained by the website.

This file builds on _v001, _v002, _v003, and _v004 to run exploratory analysis on some historical weather data.

## Functions and Libraries
The exploration process uses tidyverse, ranger, several generic custom functions, and several functions specific to Open Meteo processing. First, tidyverse, ranger, and the generic functions are loaded:  
```{r}

library(tidyverse) # tidyverse functionality is included throughout
library(ranger) # predict() does not work on ranger objects unless ranger has been called

source("./Generic_Added_Utility_Functions_202105_v001.R") # Basic functions

```
  
Next, specific functions written in _v001, _v002, _v003, and _v004 are sourced:  
```{r}

source("./SimpleOneVar_Functions_202411_v001.R") # Functions for basic single variable analysis
source("./OpenMeteo_NextBest_202411_v001.R") # Functions for finding 'next best' predictor given existing model
source("./OpenMeteo_Functions_202411_v001.R") # Core functions for loading, processing, analysis of Open Meteo

```
  
Key mapping tables for available metrics are also copied:  
```{r}

hourlyMetrics <- "temperature_2m,relativehumidity_2m,dewpoint_2m,apparent_temperature,pressure_msl,surface_pressure,precipitation,rain,snowfall,cloudcover,cloudcover_low,cloudcover_mid,cloudcover_high,shortwave_radiation,direct_radiation,direct_normal_irradiance,diffuse_radiation,windspeed_10m,windspeed_100m,winddirection_10m,winddirection_100m,windgusts_10m,et0_fao_evapotranspiration,weathercode,vapor_pressure_deficit,soil_temperature_0_to_7cm,soil_temperature_7_to_28cm,soil_temperature_28_to_100cm,soil_temperature_100_to_255cm,soil_moisture_0_to_7cm,soil_moisture_7_to_28cm,soil_moisture_28_to_100cm,soil_moisture_100_to_255cm"
dailyMetrics <- "weathercode,temperature_2m_max,temperature_2m_min,apparent_temperature_max,apparent_temperature_min,precipitation_sum,rain_sum,snowfall_sum,precipitation_hours,sunrise,sunset,windspeed_10m_max,windgusts_10m_max,winddirection_10m_dominant,shortwave_radiation_sum,et0_fao_evapotranspiration"

hourlyDescription <- "Air temperature at 2 meters above ground\nRelative humidity at 2 meters above ground\nDew point temperature at 2 meters above ground\nApparent temperature is the perceived feels-like temperature combining wind chill factor, relative humidity and solar radiation\nAtmospheric air pressure reduced to mean sea level (msl) or pressure at surface. Typically pressure on mean sea level is used in meteorology. Surface pressure gets lower with increasing elevation.\nAtmospheric air pressure reduced to mean sea level (msl) or pressure at surface. Typically pressure on mean sea level is used in meteorology. Surface pressure gets lower with increasing elevation.\nTotal precipitation (rain, showers, snow) sum of the preceding hour. Data is stored with a 0.1 mm precision. If precipitation data is summed up to monthly sums, there might be small inconsistencies with the total precipitation amount.\nOnly liquid precipitation of the preceding hour including local showers and rain from large scale systems.\nSnowfall amount of the preceding hour in centimeters. For the water equivalent in millimeter, divide by 7. E.g. 7 cm snow = 10 mm precipitation water equivalent\nTotal cloud cover as an area fraction\nLow level clouds and fog up to 2 km altitude\nMid level clouds from 2 to 6 km altitude\nHigh level clouds from 6 km altitude\nShortwave solar radiation as average of the preceding hour. This is equal to the total global horizontal irradiation\nDirect solar radiation as average of the preceding hour on the horizontal plane and the normal plane (perpendicular to the sun)\nDirect solar radiation as average of the preceding hour on the horizontal plane and the normal plane (perpendicular to the sun)\nDiffuse solar radiation as average of the preceding hour\nWind speed at 10 or 100 meters above ground. Wind speed on 10 meters is the standard level.\nWind speed at 10 or 100 meters above ground. Wind speed on 10 meters is the standard level.\nWind direction at 10 or 100 meters above ground\nWind direction at 10 or 100 meters above ground\nGusts at 10 meters above ground of the indicated hour. Wind gusts in CERRA are defined as the maximum wind gusts of the preceding hour. Please consult the ECMWF IFS documentation for more information on how wind gusts are parameterized in weather models.\nET0 Reference Evapotranspiration of a well watered grass field. Based on FAO-56 Penman-Monteith equations ET0 is calculated from temperature, wind speed, humidity and solar radiation. Unlimited soil water is assumed. ET0 is commonly used to estimate the required irrigation for plants.\nWeather condition as a numeric code. Follow WMO weather interpretation codes. See table below for details. Weather code is calculated from cloud cover analysis, precipitation and snowfall. As barely no information about atmospheric stability is available, estimation about thunderstorms is not possible.\nVapor Pressure Deificit (VPD) in kilopascal (kPa). For high VPD (>1.6), water transpiration of plants increases. For low VPD (<0.4), transpiration decreases\nAverage temperature of different soil levels below ground.\nAverage temperature of different soil levels below ground.\nAverage temperature of different soil levels below ground.\nAverage temperature of different soil levels below ground.\nAverage soil water content as volumetric mixing ratio at 0-7, 7-28, 28-100 and 100-255 cm depths.\nAverage soil water content as volumetric mixing ratio at 0-7, 7-28, 28-100 and 100-255 cm depths.\nAverage soil water content as volumetric mixing ratio at 0-7, 7-28, 28-100 and 100-255 cm depths.\nAverage soil water content as volumetric mixing ratio at 0-7, 7-28, 28-100 and 100-255 cm depths."
dailyDescription <- "The most severe weather condition on a given day\nMaximum and minimum daily air temperature at 2 meters above ground\nMaximum and minimum daily air temperature at 2 meters above ground\nMaximum and minimum daily apparent temperature\nMaximum and minimum daily apparent temperature\nSum of daily precipitation (including rain, showers and snowfall)\nSum of daily rain\nSum of daily snowfall\nThe number of hours with rain\nSun rise and set times\nSun rise and set times\nMaximum wind speed and gusts on a day\nMaximum wind speed and gusts on a day\nDominant wind direction\nThe sum of solar radiaion on a given day in Megajoules\nDaily sum of ET0 Reference Evapotranspiration of a well watered grass field"

# Create tibble for hourly metrics
tblMetricsHourly <- tibble::tibble(metric=hourlyMetrics %>% str_split_1(","), 
                                   description=hourlyDescription %>% str_split_1("\n")
                                   )
tblMetricsHourly %>% 
    print(n=50)

# Create tibble for daily metrics
tblMetricsDaily <- tibble::tibble(metric=dailyMetrics %>% str_split_1(","), 
                                  description=dailyDescription %>% str_split_1("\n")
                                   )
tblMetricsDaily

```

A previously existing dataset is loaded, with key analysis variables defined in a vector:  
```{r}

# Load previous data
allCity <- readFromRDS("allCity_20241116")

# Get core training variables
varsTrain <- getVarsTrain(allCity)
varsTrain

# Assign default label
keyLabel <- genericKeyLabelOM()
keyLabel

```

The correlation heatmap is reproduced, with functions that borrowing from the recipe at [STHDA](http://www.sthda.com/english/wiki/ggplot2-quick-correlation-matrix-heatmap-r-software-and-data-visualization):
```{r, fig.height=9, fig.width=9}

# Default function
corVarsTrain <- makeHeatMap(allCity, vecSelect=varsTrain, returnData=TRUE)
corVarsTrain %>% filter(Var1!=Var2) %>% arrange(desc(value)) %>% print(n=20)

```

The correlation heatmap is produced for a single city:  
```{r, fig.height=9, fig.width=9}

corVarsTrainBOS <- makeHeatMap(allCity %>% filter(src=="Boston"), vecSelect=varsTrain, returnData=TRUE)
corVarsTrainBOS %>% filter(Var1!=Var2) %>% arrange(desc(value)) %>% print(n=20)

```

Some variables, such as surface pressure vs. MSL pressure, are much differently correlated if controlling for city. Differences are explored:  
```{r, fig.height=9, fig.width=9}

tstCorDelta <- corVarsTrain %>% 
    mutate(v1=pmin(as.character(Var1), as.character(Var2)), 
           v2=pmax(as.character(Var1), as.character(Var2))
           ) %>% 
    select(v1, v2, value_all=value) %>%
    full_join(corVarsTrainBOS %>% 
                  mutate(v1=pmin(as.character(Var1), as.character(Var2)), 
                         v2=pmax(as.character(Var1), as.character(Var2))
                         ) %>% 
                  select(v1, v2, value_bos=value), 
              by=c("v1", "v2")
              ) %>%
    mutate(delta=value_bos-value_all)

tstCorDelta
tstCorDelta %>% arrange(delta) %>% head(20)
tstCorDelta %>% arrange(desc(delta)) %>% head(20)

```

The process is repeated, using an aggregation of correlations in each of the seven cities:  
```{r, fig.height=9, fig.width=9}

keyCities <- allCity %>% pull(src) %>% unique()
keyCities

corVarsTrainEach <- keyCities %>% 
    map_dfr(.f=function(x) makeHeatMap(allCity %>% filter(src==x), 
                                       vecSelect=varsTrain[!varsTrain %in% ifelse(x=="Miami", "snowfall", "")],
                                       returnData=TRUE, 
                                       plotMap=FALSE
                                       ) %>% 
                mutate(v1=pmin(as.character(Var1), as.character(Var2)), 
                       v2=pmax(as.character(Var1), as.character(Var2))
                       ) %>% 
                select(v1, v2, value), .id="src") %>%
    mutate(city=keyCities[as.integer(src)]) %>%
    select(src, city, everything())

corVarsTrainEach
corVarsTrainEach %>% 
    arrange(desc(value)) %>% 
    filter(v1!=v2) %>% 
    slice(1:15, (nrow(.)-14):nrow(.)) %>%
    print(n=30)

```

Comparison of aggregated individual city correlations to the overall correlations:  
```{r, fig.height=9, fig.width=9}

corVarsTrainDelta <- corVarsTrain %>% 
    mutate(v1=pmin(as.character(Var1), as.character(Var2)), 
           v2=pmax(as.character(Var1), as.character(Var2))
    ) %>% 
    select(v1, v2, value_all=value) %>% 
    full_join(corVarsTrainEach %>% 
                  group_by(v1, v2) %>% 
                  summarize(value=mean(value), .groups="drop"), 
              by=c("v1", "v2")) %>% 
    mutate(delta=value-value_all) 

corVarsTrainDelta %>% 
    arrange(delta) %>% 
    slice(1:15, (nrow(.)-14):nrow(.)) %>%
    print(n=30)

```

Differences by variable are explored:  
```{r, fig.height=9, fig.width=9}

corVarsTrain %>% 
    mutate(v1=pmin(as.character(Var1), as.character(Var2)), 
           v2=pmax(as.character(Var1), as.character(Var2))
    ) %>% 
    select(v1, v2, value_all=value) %>% 
    full_join(corVarsTrainEach, by=c("v1", "v2")) %>% 
    mutate(delta=value-value_all) %>%
    filter(v1!=v2) %>%
    select(city, v1, v2, delta) %>%
    bind_rows(., ., .id="src") %>%
    mutate(vrbl=case_when(src=="1"~v1, src=="2"~v2, TRUE~NA)) %>%
    ggplot(aes(x=fct_reorder(vrbl, delta, .fun=function(x) diff(range(x))), y=delta)) + 
    geom_boxplot(fill="lightblue") + 
    coord_flip() + 
    geom_hline(yintercept=0, color="black", lty=2) +
    labs(title="Correlation delta (individual minus all-city aggregate)", 
         x=NULL, 
         y="Delta (individual minus all-city aggregate)", 
         subtitle="Calculated for each variable combination and city"
         )

```

Soil moisture, surface pressure, and dewpoint are among the variables with the highest changes in correlation when calculated n individual cities and the all-city aggregate

An example Simpson's paradox is MSL pressure vs. surface pressure:  
```{r, fig.height=9, fig.width=9}

allCity %>% 
    count(src, surface_pressure, pressure_msl) %>% 
    ggplot(aes(x=surface_pressure, pressure_msl)) + 
    geom_smooth(aes(weight=n, color=src), method="lm") + 
    geom_smooth(method="lm", lty=2, aes(weight=n), color="black") + 
    labs(title="Relationship between MSL pressure and surface pressure", 
         subtitle="Dashed black line is overall relationship"
         ) + 
    scale_color_discrete(NULL)

```

The process is converted to functional form:  
```{r, fig.height=9, fig.width=9}

tmpSmoothPlot <- function(df, x, y, xName=x, yName=y, printPlot=TRUE, returnPlot=!isTRUE(printPlot)) {
    
    # FUNCTION ARGUMENTS:
    # df: the data frame
    # x: x variable
    # y: y variable
    # xName: name to describe x variable
    # yName: name to describe y variable
    # printPlot: boolean, should plot be printed?
    # returnPlot: boolean, should plot object be returned?
    
    p1 <- df %>% 
        select(src, all_of(c(x, y))) %>%
        purrr::set_names(c("src", "x1", "y1")) %>%
        count(src, x1, y1) %>% 
        ggplot(aes(x=x1, y=y1)) + 
        geom_smooth(aes(weight=n, color=src), method="lm") + 
        geom_smooth(method="lm", lty=2, aes(weight=n), color="black") + 
        labs(title=paste0("Relationship between ", xName, " and ", yName), 
             subtitle="Dashed black line is overall relationship", 
             y=if(y!=yName) paste0(yName, "\n(", y, ")") else y,
             x=if(x!=xName) paste0(xName, "\n(", x, ")") else x
             ) + 
    scale_color_discrete(NULL)
    
    # Print plot if requested
    if(isTRUE(printPlot)) print(p1)
    
    # Return plot if requested
    if(isTRUE(returnPlot)) return(p1)
    
}

# Example function call
tmpSmoothPlot(allCity, 
              x="surface_pressure", 
              y="pressure_msl", 
              yName="MSL Pressure", 
              xName="Surface Pressure"
              )

```

The function is tested on two additional sets of metrics:  
```{r, fig.height=9, fig.width=9}

gridExtra::grid.arrange(tmpSmoothPlot(allCity %>% 
                                          mutate(across(c("apparent_temperature", "dewpoint_2m"), 
                                                        .fns=function(x) round(x)
                                                        )
                                                 ), 
                                      x="dewpoint_2m", 
                                      y="apparent_temperature", 
                                      xName="Dew Point", 
                                      yName="Apparent Temperature", 
                                      printPlot=FALSE
                                      ), 
                        tmpSmoothPlot(allCity %>% 
                                          mutate(across(c("surface_pressure", "dewpoint_2m"), 
                                                        .fns=function(x) round(x)
                                                        )
                                                 ), 
                                      x="dewpoint_2m", 
                                      y="surface_pressure", 
                                      xName="Dew Point", 
                                      yName="Surface Pressure", 
                                      printPlot=FALSE
                                      ), 
                        nrow=1
                        )

```

The function is updated to allow for rounding:  
```{r, fig.height=9, fig.width=9}

tmpSmoothPlot <- function(df, 
                          x, 
                          y, 
                          xRound=NULL, 
                          yRound=NULL,
                          xName=x, 
                          yName=y, 
                          printPlot=TRUE, 
                          returnPlot=!isTRUE(printPlot)
                          ) {
    
    # FUNCTION ARGUMENTS:
    # df: the data frame
    # x: x variable
    # y: y variable
    # {x,y}Round: rounding to apply to vector {x,y} using function autoRound()
    #             NULL means no rounding (default)
    #             -1L means make an estimate based on data (around 100 buckets created)
    #             a positive float or integer means round everything to the nearest multiple
    # xName: name to describe x variable
    # yName: name to describe y variable
    # printPlot: boolean, should plot be printed?
    # returnPlot: boolean, should plot object be returned?
    
    p1 <- df %>% 
        select(src, all_of(c(x, y))) %>%
        purrr::set_names(c("src", "x1", "y1")) %>%
        mutate(x1=autoRound(x1, rndTo=xRound), 
               y1=autoRound(y1, rndTo=yRound)
               ) %>%
        count(src, x1, y1) %>% 
        ggplot(aes(x=x1, y=y1)) + 
        geom_smooth(aes(weight=n, color=src), method="lm") + 
        geom_smooth(method="lm", lty=2, aes(weight=n), color="black") + 
        labs(title=paste0("Relationship between ", xName, " and ", yName), 
             subtitle="Dashed black line is overall relationship", 
             y=if(y!=yName) paste0(yName, "\n(", y, ")") else y,
             x=if(x!=xName) paste0(xName, "\n(", x, ")") else x
             ) + 
    scale_color_discrete(NULL)
    
    # Print plot if requested
    if(isTRUE(printPlot)) print(p1)
    
    # Return plot if requested
    if(isTRUE(returnPlot)) return(p1)
    
}

```

The updated function is tested on elements with many buckets:  
```{r, fig.height=9, fig.width=9}

# Example function call (no rounding)
t0 <- Sys.time()
system.time(
    tmpSmoothPlot(allCity, 
                  x="direct_normal_irradiance", 
                  y="shortwave_radiation", 
                  xName="Direct Solar Radiation", 
                  yName="Shortwave Solar Radiation"
                  )
    )
Sys.time() - t0

# Example function call (rounding)
t0 <- Sys.time()
system.time(
    tmpSmoothPlot(allCity, 
                  x="direct_normal_irradiance", 
                  y="shortwave_radiation", 
                  xRound=-1L, 
                  yRound=-1L,
                  xName="Direct Solar Radiation", 
                  yName="Shortwave Solar Radiation"
                  )
    )
Sys.time() - t0

```

A simple model is run to predict surface pressure as a function of dewpoint, without considering location:  
```{r, fig.height=9, fig.width=9}

# Create model
tstLM <- allCity %>%
    select(dewpoint_2m, surface_pressure) %>%
    lm(surface_pressure ~ dewpoint_2m, data=.)

# Summary
summary(tstLM)

# Review of predictions
allCity %>%
    select(src, dewpoint_2m, surface_pressure) %>%
    mutate(pred=predict(tstLM, newdata=.)) %>%
    mutate(across(c(surface_pressure, pred), .fns=function(x) autoRound(x))) %>%
    count(surface_pressure, pred) %>%
    ggplot(aes(x=pred, y=surface_pressure)) + 
    geom_point(aes(size=n), alpha=0.25) + 
    geom_smooth(aes(weight=n), method="lm") + 
    geom_abline(slope=1, intercept=0, lty=2, color="red") + 
    labs(x="Prediction", 
         y="Actual", 
         title="Predictions for Surface Pressure", 
         subtitle="Surface Pressure ~ Dewpoint"
         ) + 
    scale_size_continuous(NULL)

```

The model is updated to predict surface pressure as a function of dewpoint, considering location:  
```{r, fig.height=9, fig.width=9}

# Create model
tstLM_002 <- allCity %>%
    select(dewpoint_2m, surface_pressure, src) %>%
    lm(surface_pressure ~ dewpoint_2m:src + src, data=.)

# Summary
summary(tstLM_002)

# Review of predictions
allCity %>%
    select(src, dewpoint_2m, surface_pressure) %>%
    mutate(pred=predict(tstLM_002, newdata=.)) %>%
    mutate(across(c(surface_pressure, pred), .fns=function(x) autoRound(x))) %>%
    count(src, surface_pressure, pred) %>%
    ggplot(aes(x=pred, y=surface_pressure)) + 
    geom_point(aes(size=n, color=src), alpha=0.25) + 
    geom_smooth(aes(weight=n), method="lm") + 
    geom_abline(slope=1, intercept=0, lty=2, color="red") + 
    labs(x="Prediction", 
         y="Actual", 
         title="Predictions for Surface Pressure", 
         subtitle="Surface Pressure ~ Dewpoint:City + City"
         ) + 
    scale_size_continuous(NULL) + 
    scale_color_discrete(NULL)

```

Long term daily data is downloaded for Atlanta:  
```{r cache=TRUE}

# Daily data download for Atlanta, GA
testURLDaily <- helperOpenMeteoURL(cityName="Atlanta GA", 
                                   dailyIndices=1:nrow(tblMetricsDaily),
                                   startDate="1960-01-01", 
                                   endDate="2023-12-31", 
                                   tz="US/Eastern"
                                   )
testURLDaily

# Download file
if(!file.exists("testOM_daily_atl.json")) {
    fileDownload(fileName="testOM_daily_atl.json", url=testURLDaily)
} else {
    cat("\nFile testOM_daily_atl.json already exists, skipping download\n")
}

```

The daily dataset is loaded:  
```{r cache=TRUE}

# Read daily JSON file
atlOMDaily <- formatOpenMeteoJSON("testOM_daily_atl.json")

# Sample records of tibble
atlOMDaily$tblDaily %>% glimpse()

```

Variables are converted to proper data type:  
```{r}

dfDailyATL <- atlOMDaily$tblDaily %>%
    mutate(weathercode=factor(weathercode), 
           sunrise_chr=sunrise, 
           sunset_chr=sunset,
           sunrise=lubridate::ymd_hm(sunrise), 
           sunset=lubridate::ymd_hm(sunset), 
           fct_winddir=factor(winddirection_10m_dominant)
           )
glimpse(dfDailyATL)

```

