---
title: "Open Meteo Weather Exploration"
author: "davegoblue"
date: "2023-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
Open-Meteo maintains an [API for historical weather](https://open-meteo.com/en/docs/historical-weather-api) that allows for non-commercial usage of historical weather data maintained by the website.

This file runs exploratory analysis on some of the historical weather data.

## Exploratory Analysis
The exploration process uses tidyverse and several generic custom functions:  
```{r}

library(tidyverse) # tidyverse functionality is included throughout

source("./Generic_Added_Utility_Functions_202105_v001.R") # Basic functions

```
  
A sample of data for 365 days has been downloaded as a CSV. The downloaded data has three separate files included in a single tab, separated by a blank row. The first file is location data, the second file is hourly data, and the third file is daily data. For initial exploration, parameters specific to this file are used:  
```{r, fig.height=9, fig.width=9}

omFileLoc <- "./RInputFiles/openmeteo_20230612_example.csv"

# Location data
omLocation <- readr::read_csv(omFileLoc, n_max=1, skip=0) 
omLocation

# Hourly data 
# Elements: time, 2m temp (C), 2m dew point (C), 2m relative humidity (%), precip (mm), rain (mm), and snow (cm)
omHourlyRaw <- readr::read_csv(omFileLoc, n_max=8760, skip=3) 
omHourlyProcess <- omHourlyRaw %>%
    purrr::set_names(c("time", "temp2m_C", "relH2m", "dew2m_C", "precip_mm", "rain_mm", "snow_cm")) %>% 
    mutate(date=date(time))
omHourlyProcess
summary(omHourlyProcess)

# Daily data 
# Elements: date, sum of precip (mm), sum of rain (mm), and sum of snow (cm)
omDailyRaw <- readr::read_csv(omFileLoc, n_max=365, skip=8765) 
omDailyProcess <- omDailyRaw %>%
    purrr::set_names(c("date", "precip_mm", "rain_mm", "snow_cm"))
omDailyProcess
summary(omDailyProcess)

```
  
A function is written to read a portion of a CSV file:  
```{r, fig.height=9, fig.width=9}

partialCSVRead <- function(loc, firstRow=1L, lastRow=+Inf, col_names=TRUE, ...) {
    
    # FUNCTION arguments
    # loc: file location
    # firstRow: first row that is relevant to the partial file read (whether header line or data line)
    # last Row: last row that is relevant to the partial file read (+Inf means read until last line of file)
    # col_names: the col_names parameter passed to readr::read_csv
    #            TRUE means header=TRUE (get column names from file, read data starting on next line)
    #            FALSE means header=FALSE (auto-generate column names, read data starting on first line)
    #            character vector means use these as column names (read data starting on first line)
    # ...: additional arguments passed to read_csv

    # Read the file and return
    # skip: rows to be skipped are all those prior to firstRow
    # n_max: maximum rows read are lastRow-firstRow, with an additional data row when col_names is not TRUE
    readr::read_csv(loc, 
                    skip=firstRow-1, 
                    n_max=lastRow-firstRow+ifelse(isTRUE(col_names), 0, 1), 
                    ...
                    )
    
}

# Double check that data are the same
partialCSVRead(omFileLoc, firstRow=1L, lastRow=2L) %>% all.equal(omLocation)
partialCSVRead(omFileLoc, firstRow=4L, lastRow=8764L) %>% all.equal(omHourlyRaw)
partialCSVRead(omFileLoc, firstRow=8766L, lastRow=+Inf) %>% all.equal(omDailyRaw)

```
  
The blank lines are assessed, allowing for all tables to be read at the same time:  
```{r, fig.height=9, fig.width=9}

# Get the break points for gaps in a vector (e.g., 0, 3, 5:8, 20 has break points 0, 3, 5, 20 and 0, 3, 8, 30)
vecGaps <- function(x, addElements=c(), sortUnique=TRUE) {
    
    if(length(addElements)>0) x <- c(addElements, x)
    if(isTRUE(sortUnique)) x <- unique(sort(x))
    list("starts"=c(x[is.na(lag(x)) | x-lag(x)>1], +Inf), 
         "ends"=x[is.na(lead(x)) | lead(x)-x>1]
         )
    
}

vecGaps(c(3, 5:8, 20), addElements=0)

# Find the break points in a single file
flatFileGaps <- function(loc) {

    which(stringr::str_length(readLines(loc))==0) %>% vecGaps(addElements=0)
    
}

flatFileGaps(omFileLoc)


# Read all relevant data as CSV with header
readMultiCSV <- function(loc, col_names=TRUE, ...) {

    gaps <- flatFileGaps(loc)
    
    lapply(seq_along(gaps$ends), 
           FUN=function(x) partialCSVRead(loc, 
                                          firstRow=gaps$ends[x]+1, 
                                          lastRow=gaps$starts[x+1]-1, 
                                          col_names=col_names, 
                                          ...
                                          )
           )
    
}

tstMultiCSV <- readMultiCSV(omFileLoc)

all.equal(tstMultiCSV[[1]], omLocation)
all.equal(tstMultiCSV[[2]], omHourlyRaw)
all.equal(tstMultiCSV[[3]], omDailyRaw)

```
  
Data can also be downloaded through the Open-Meteo API, returning a JSON file. The data download has been completed off-line to minimize repeated hits against the server. The JSON file can then be read:  
```{r}

# Example download sequence
# download.file("https://archive-api.open-meteo.com/v1/archive?latitude=41.85&longitude=-87.65&start_date=2022-06-01&end_date=2023-06-08&hourly=temperature_2m,relativehumidity_2m,dewpoint_2m,precipitation,rain,snowfall&daily=precipitation_sum,rain_sum,snowfall_sum&timezone=America%2FChicago", "tempOM")

# Create hourly data tibble
jsonHourly <- jsonlite::read_json("tempOM", simplifyVector = TRUE)[["hourly"]] %>% 
    tibble::as_tibble() %>% 
    mutate(tm=lubridate::ymd_hm(time), date=date(tm))
jsonHourly

# Create daily data tibble
jsonDaily <- jsonlite::read_json("tempOM", simplifyVector = TRUE)[["daily"]] %>% 
    tibble::as_tibble()
jsonDaily

# Extract other elements
jsonNames <- jsonlite::read_json("tempOM", simplifyVector = TRUE) %>% names
for (jsonName in jsonNames[!(jsonNames %in% c("daily", "hourly", "daily_units", "hourly_units"))]) {
    cat("\n", jsonName, ":", jsonlite::read_json("tempOM", simplifyVector = TRUE)[[jsonName]])
}
for (jsonName in jsonNames[jsonNames %in% c("daily_units", "hourly_units")]) {
    cat("\n", jsonName, ":\n")
    print(jsonlite::read_json("tempOM", simplifyVector = TRUE)[[jsonName]] %>% tibble::as_tibble() %>% t())
}

```
  
Daily data read from JSON and CSV are compared:  
```{r}

# Convert variable names in JSON daily data
jsonDailyProcess <- jsonDaily %>%
    colRenamer(vecRename=c("precipitation_sum"="precip_mm", 
                           "rain_sum"="rain_mm", 
                           "snowfall_sum"="snow_cm", 
                           "time"="date"
                           )
               ) %>%
    mutate(date=as.Date(date))
jsonDailyProcess

# Check dates included
omDailyProcess %>% 
    select(date) %>% 
    mutate(inCSV=1) %>% 
    full_join(mutate(select(jsonDailyProcess, "date"), inJSON=1), by="date") %>%
    filter(!complete.cases(.))

# Check column names
all.equal(names(omDailyProcess), names(jsonDailyProcess))

# Check data elements from 2022-06-08 through 2023-06-04 (last full day of data)
all.equal(omDailyProcess %>% tibble::as_tibble() %>% filter(date>="2022-06-08", date<="2023-06-04"), 
          jsonDailyProcess %>% filter(date>="2022-06-08", date<="2023-06-04")
          )

```
  
Hourly data read from JSON and CSV are compared:  
```{r}

# Convert variable names in JSON hourly data
jsonHourlyProcess <- jsonHourly %>% 
    select(-time) %>%
    colRenamer(vecRename=c("temperature_2m"="temp2m_C", 
                           "relativehumidity_2m"="relH2m", 
                           "dewpoint_2m"="dew2m_C", 
                           "precipitation"="precip_mm", 
                           "rain"="rain_mm", 
                           "snowfall"="snow_cm",
                           "tm"="time"
                           )
               ) %>% 
    select(time, everything())
jsonHourlyProcess

# Check dates included
omHourlyProcess %>% 
    count(date, name="nCSV") %>% 
    full_join(count(jsonHourlyProcess, date, name="nJSON"), by="date") %>%
    filter(!complete.cases(.))

# Check column names
all.equal(names(omHourlyProcess), names(jsonHourlyProcess))

# Check data elements from 2022-06-08 through 2023-06-04 (last full day of data)
all.equal(omHourlyProcess %>% tibble::as_tibble() %>% filter(date>="2022-06-08", date<="2023-06-04"), 
          jsonHourlyProcess %>% filter(date>="2022-06-08", date<="2023-06-04")
          )

```
  
Metrics that can be reuested for hourly and daily data include:  
```{r}

hourlyMetrics <- "temperature_2m,relativehumidity_2m,dewpoint_2m,apparent_temperature,pressure_msl,surface_pressure,precipitation,rain,snowfall,cloudcover,cloudcover_low,cloudcover_mid,cloudcover_high,shortwave_radiation,direct_radiation,direct_normal_irradiance,diffuse_radiation,windspeed_10m,windspeed_100m,winddirection_10m,winddirection_100m,windgusts_10m,et0_fao_evapotranspiration,weathercode,vapor_pressure_deficit,soil_temperature_0_to_7cm,soil_temperature_7_to_28cm,soil_temperature_28_to_100cm,soil_temperature_100_to_255cm,soil_moisture_0_to_7cm,soil_moisture_7_to_28cm,soil_moisture_28_to_100cm,soil_moisture_100_to_255cm"
dailyMetrics <- "weathercode,temperature_2m_max,temperature_2m_min,apparent_temperature_max,apparent_temperature_min,precipitation_sum,rain_sum,snowfall_sum,precipitation_hours,sunrise,sunset,windspeed_10m_max,windgusts_10m_max,winddirection_10m_dominant,shortwave_radiation_sum,et0_fao_evapotranspiration"

hourlyDescription <- "Air temperature at 2 meters above ground\nRelative humidity at 2 meters above ground\nDew point temperature at 2 meters above ground\nApparent temperature is the perceived feels-like temperature combining wind chill factor, relative humidity and solar radiation\nAtmospheric air pressure reduced to mean sea level (msl) or pressure at surface. Typically pressure on mean sea level is used in meteorology. Surface pressure gets lower with increasing elevation.\nAtmospheric air pressure reduced to mean sea level (msl) or pressure at surface. Typically pressure on mean sea level is used in meteorology. Surface pressure gets lower with increasing elevation.\nTotal precipitation (rain, showers, snow) sum of the preceding hour. Data is stored with a 0.1 mm precision. If precipitation data is summed up to monthly sums, there might be small inconsistencies with the total precipitation amount.\nOnly liquid precipitation of the preceding hour including local showers and rain from large scale systems.\nSnowfall amount of the preceding hour in centimeters. For the water equivalent in millimeter, divide by 7. E.g. 7 cm snow = 10 mm precipitation water equivalent\nTotal cloud cover as an area fraction\nLow level clouds and fog up to 2 km altitude\nMid level clouds from 2 to 6 km altitude\nHigh level clouds from 6 km altitude\nShortwave solar radiation as average of the preceding hour. This is equal to the total global horizontal irradiation\nDirect solar radiation as average of the preceding hour on the horizontal plane and the normal plane (perpendicular to the sun)\nDirect solar radiation as average of the preceding hour on the horizontal plane and the normal plane (perpendicular to the sun)\nDiffuse solar radiation as average of the preceding hour\nWind speed at 10 or 100 meters above ground. Wind speed on 10 meters is the standard level.\nWind speed at 10 or 100 meters above ground. Wind speed on 10 meters is the standard level.\nWind direction at 10 or 100 meters above ground\nWind direction at 10 or 100 meters above ground\nGusts at 10 meters above ground of the indicated hour. Wind gusts in CERRA are defined as the maximum wind gusts of the preceding hour. Please consult the ECMWF IFS documentation for more information on how wind gusts are parameterized in weather models.\nET0 Reference Evapotranspiration of a well watered grass field. Based on FAO-56 Penman-Monteith equations ET0 is calculated from temperature, wind speed, humidity and solar radiation. Unlimited soil water is assumed. ET0 is commonly used to estimate the required irrigation for plants.\nWeather condition as a numeric code. Follow WMO weather interpretation codes. See table below for details. Weather code is calculated from cloud cover analysis, precipitation and snowfall. As barely no information about atmospheric stability is available, estimation about thunderstorms is not possible.\nVapor Pressure Deificit (VPD) in kilopascal (kPa). For high VPD (>1.6), water transpiration of plants increases. For low VPD (<0.4), transpiration decreases\nAverage temperature of different soil levels below ground.\nAverage temperature of different soil levels below ground.\nAverage temperature of different soil levels below ground.\nAverage temperature of different soil levels below ground.\nAverage soil water content as volumetric mixing ratio at 0-7, 7-28, 28-100 and 100-255 cm depths.\nAverage soil water content as volumetric mixing ratio at 0-7, 7-28, 28-100 and 100-255 cm depths.\nAverage soil water content as volumetric mixing ratio at 0-7, 7-28, 28-100 and 100-255 cm depths.\nAverage soil water content as volumetric mixing ratio at 0-7, 7-28, 28-100 and 100-255 cm depths."
dailyDescription <- "The most severe weather condition on a given day\nMaximum and minimum daily air temperature at 2 meters above ground\nMaximum and minimum daily air temperature at 2 meters above ground\nMaximum and minimum daily apparent temperature\nMaximum and minimum daily apparent temperature\nSum of daily precipitation (including rain, showers and snowfall)\nSum of daily rain\nSum of daily snowfall\nThe number of hours with rain\nSun rise and set times\nSun rise and set times\nMaximum wind speed and gusts on a day\nMaximum wind speed and gusts on a day\nDominant wind direction\nThe sum of solar radiaion on a given day in Megajoules\nDaily sum of ET0 Reference Evapotranspiration of a well watered grass field"

# Create tibble for hourly metrics
tblMetricsHourly <- tibble::tibble(metric=hourlyMetrics %>% str_split_1(","), 
                                   description=hourlyDescription %>% str_split_1("\n")
                                   )
tblMetricsHourly %>% 
    print(n=50)

# Create tibble for daily metrics
tblMetricsDaily <- tibble::tibble(metric=dailyMetrics %>% str_split_1(","), 
                                  description=dailyDescription %>% str_split_1("\n")
                                   )
tblMetricsDaily

```

Data can then be assembled into a string that is compatible with the Open-Meteo API format:  
```{r}

openMeteoURLCreate <- function(mainURL="https://archive-api.open-meteo.com/v1/archive", 
                               lat=45, 
                               lon=-90, 
                               startDate=paste(year(Sys.Date())-1, "01", "01", sep="-"), 
                               endDate=paste(year(Sys.Date())-1, "12", "31", sep="-"), 
                               hourlyMetrics=NULL, 
                               dailyMetrics=NULL,
                               tz="GMT", 
                               ...
                               ) {
    
    # Create formatted string
    fString <- paste0(mainURL, 
                      "?latitude=", 
                      lat, 
                      "&longitude=", 
                      lon, 
                      "&start_date=", 
                      startDate, 
                      "&end_date=", 
                      endDate
                      )
    if(!is.null(hourlyMetrics)) fString <- paste0(fString, "&hourly=", hourlyMetrics)
    if(!is.null(dailyMetrics)) fString <- paste0(fString, "&daily=", dailyMetrics)
    
    # Return the formatted string
    paste0(fString, "&timezone=", stringr::str_replace(tz, "/", "%2F"), ...)
    
}

# Blank example
openMeteoURLCreate()

# Matching previous CSV data pull
openMeteoURLCreate(lat=41.85, 
                   lon=-87.65, 
                   startDate="2022-06-01", 
                   endDate="2023-06-08", 
                   hourlyMetrics="temperature_2m,relativehumidity_2m,dewpoint_2m,precipitation,rain,snowfall", 
                   dailyMetrics="precipitation_sum,rain_sum,snowfall_sum", 
                   tz="America/Chicago"
                   )

```
  
A helper function is created to convert cities to lat/lon and to allow for selection of hourly and daily metrics by index number:  
```{r}

helperOpenMeteoURL <- function(cityName=NULL,
                               lat=NULL,
                               lon=NULL,
                               hourlyMetrics=NULL,
                               hourlyIndices=NULL,
                               hourlyDesc=tblMetricsHourly,
                               dailyMetrics=NULL,
                               dailyIndices=NULL,
                               dailyDesc=tblMetricsDaily,
                               startDate=NULL, 
                               endDate=NULL, 
                               tz=NULL,
                               ...
                               ) {
    
    # Convert city to lat/lon if lat/lon are NULL
    if(is.null(lat) | is.null(lon)) {
        if(is.null(cityName)) stop("\nMust provide lat/lon or city name available in maps::us.cities\n")
        cityData <- maps::us.cities %>% tibble::as_tibble() %>% filter(name==cityName)
        if(nrow(cityData)!=1) stop("\nMust provide city name that maps uniquely to maps::us.cities$name\n")
        lat <- cityData$lat[1]
        lon <- cityData$long[1]
    }
    
    # Get hourly metrics by index if relevant
    if(is.null(hourlyMetrics) & !is.null(hourlyIndices)) {
        hourlyMetrics <- hourlyDesc %>% slice(hourlyIndices) %>% pull(metric)
        hourlyMetrics <- paste0(hourlyMetrics, collapse=",")
        cat("\nHourly metrics created from indices:", hourlyMetrics, "\n\n")
    }
    
    # Get daily metrics by index if relevant
    if(is.null(dailyMetrics) & !is.null(dailyIndices)) {
        dailyMetrics <- dailyDesc %>% slice(dailyIndices) %>% pull(metric)
        dailyMetrics <- paste0(dailyMetrics, collapse=",")
        cat("\nDaily metrics created from indices:", dailyMetrics, "\n\n")
    }
    
    # Use default values from OpenMeteoURLCreate() for startDate, endDate, and tz if passed as NULL
    if(is.null(startDate)) startDate <- eval(formals(openMeteoURLCreate)$startDate)
    if(is.null(endDate)) endDate <- eval(formals(openMeteoURLCreate)$endDate)
    if(is.null(tz)) tz <- eval(formals(openMeteoURLCreate)$tz)
    
    # Create and return URL
    openMeteoURLCreate(lat=lat,
                       lon=lon, 
                       startDate=startDate, 
                       endDate=endDate, 
                       hourlyMetrics=hourlyMetrics, 
                       dailyMetrics=dailyMetrics, 
                       tz=tz,
                       ...
                       )
    
}

```
  
The URL is tested for file download, cached to avoid multiple hits to the server:  
```{r cache=TRUE}

testURL <- helperOpenMeteoURL(cityName="Chicago IL", 
                              hourlyIndices=c(1:3, 7:9),
                              dailyIndices=6:8,
                              startDate="2022-06-01", 
                              endDate="2023-06-08", 
                              tz="America/Chicago"
                              )
testURL

# Download file
if(!file.exists("notuse_testOM.json")) {
    fileDownload(fileName="notuse_testOM.json", url=testURL)
} else {
    cat("\nFile notuse_testOM.json already exists, skipping download\n")
}

```
  
Code is created to read the JSON return object:  
```{r}

readOpenMeteoJSON <- function(js) {
    
    # FUNCTION arguments: 
    # js: JSON list returned by download from Open-Meteo
    
    # Get the object and names
    jsObj <- jsonlite::read_json(js, simplifyVector = TRUE)
    nms <- jsObj %>% names()
    cat("\nObjects in JSON include:", paste(nms, collapse=", "), "\n\n")
    
    # Set default objects as NULL
    tblDaily <- NULL
    tblHourly <- NULL
    tblUnitsDaily <- NULL
    tblUnitsHourly <- NULL
    
    # Get daily and hourly as tibble if relevant
    if("daily" %in% nms) tblDaily <- jsObj$daily %>% tibble::as_tibble()
    if("hourly" %in% nms) tblHourly <- jsObj$hourly %>% tibble::as_tibble()
    
    # Helper function for unit conversions
    helperMetricUnit <- function(x, mapper, desc) {
        x %>% 
            tibble::as_tibble() %>% 
            pivot_longer(cols=everything()) %>% 
            left_join(mapper, by=c("name"="metric")) %>% 
            mutate(value=stringr::str_replace(value, "\u00b0", "deg ")) %>% 
            mutate(metricType=desc) %>% 
            select(metricType, everything())
    }
    
    # Get the unit descriptions
    if("daily_units" %in% nms) 
        tblUnitsDaily <- helperMetricUnit(jsObj$daily_units, tblMetricsDaily, desc="daily_units")
    if("hourly_units" %in% nms) 
        tblUnitsHourly <- helperMetricUnit(jsObj$hourly_units, tblMetricsHourly, desc="hourly_units")
    if(is.null(tblUnitsDaily) & !is.null(tblUnitsHourly)) tblUnits <- tblUnitsHourly
    else if(!is.null(tblUnitsDaily) & is.null(tblUnitsHourly)) tblUnits <- tblUnitsDaily
    else if(!is.null(tblUnitsDaily) & !is.null(tblUnitsHourly)) 
        tblUnits <- bind_rows(tblUnitsHourly, tblUnitsDaily)
    else tblUnits <- NULL
    
    # Put everything else together
    tblDescription <- jsObj[setdiff(nms, c("hourly", "hourly_units", "daily", "daily_units"))] %>%
        tibble::as_tibble()
    
    # Return the list objects
    list(tblDaily=tblDaily, tblHourly=tblHourly, tblUnits=tblUnits, tblDescription=tblDescription)
    
}

prettyOpenMeteoMeta <- function(df, extr="tblDescription") {
    if("list" %in% class(df)) df <- df[[extr]]
    for(name in names(df)) {
        cat("\n", name, ": ", df %>% pull(name), sep="")
    }
    cat("\n\n")
}


tmpOM <- readOpenMeteoJSON("notuse_testOM.json")
tmpOM
prettyOpenMeteoMeta(tmpOM)

```
  
Conversion functions are written for hourly and daily data:  
```{r}

omProcessDaily <- function(tbl, extr="tblDaily") {
    if("list" %in% class(tbl)) tbl <- tbl[[extr]]
    tbl %>% mutate(date=lubridate::ymd(time)) %>% select(date, everything())
}

omProcessHourly <- function(tbl, extr="tblHourly") {
    if("list" %in% class(tbl)) tbl <- tbl[[extr]]
    tbl %>% 
        mutate(origTime=time, 
               time=lubridate::ymd_hm(time), 
               date=lubridate::date(time), 
               hour=lubridate::hour(time)
               ) %>% 
        select(time, date, hour, everything())
}

omProcessDaily(tmpOM)
omProcessHourly(tmpOM)

```
  
Function readOpenMeteoJSON() is updated to automatically incorporate date conversions:  
```{r}

readOpenMeteoJSON <- function(js, mapDaily=tblMetricsDaily, mapHourly=tblMetricsHourly) {
    
    # FUNCTION arguments: 
    # js: JSON list returned by download from Open-Meteo
    # mapDaily: mapping file for daily metrics
    # mapHourly: mapping file for hourly metrics
    
    # Get the object and names
    jsObj <- jsonlite::read_json(js, simplifyVector = TRUE)
    nms <- jsObj %>% names()
    cat("\nObjects in JSON include:", paste(nms, collapse=", "), "\n\n")
    
    # Set default objects as NULL
    tblDaily <- NULL
    tblHourly <- NULL
    tblUnitsDaily <- NULL
    tblUnitsHourly <- NULL
    
    # Get daily and hourly as tibble if relevant
    if("daily" %in% nms) tblDaily <- jsObj$daily %>% tibble::as_tibble() %>% omProcessDaily()
    if("hourly" %in% nms) tblHourly <- jsObj$hourly %>% tibble::as_tibble() %>% omProcessHourly()
    
    # Helper function for unit conversions
    helperMetricUnit <- function(x, mapper, desc=NULL) {
        if(is.null(desc)) 
            desc <- as.list(match.call())$x %>% 
                deparse() %>% 
                stringr::str_replace_all(pattern=".*\\$", replacement="")
        x %>% 
            tibble::as_tibble() %>% 
            pivot_longer(cols=everything()) %>% 
            left_join(mapper, by=c("name"="metric")) %>% 
            mutate(value=stringr::str_replace(value, "\u00b0", "deg ")) %>% 
            mutate(metricType=desc) %>% 
            select(metricType, everything())
    }
    
    # Get the unit descriptions
    if("daily_units" %in% nms) tblUnitsDaily <- helperMetricUnit(jsObj$daily_units, mapDaily)
    if("hourly_units" %in% nms) tblUnitsHourly <- helperMetricUnit(jsObj$hourly_units, mapHourly)
    if(is.null(tblUnitsDaily) & !is.null(tblUnitsHourly)) tblUnits <- tblUnitsHourly
    else if(!is.null(tblUnitsDaily) & is.null(tblUnitsHourly)) tblUnits <- tblUnitsDaily
    else if(!is.null(tblUnitsDaily) & !is.null(tblUnitsHourly)) 
        tblUnits <- bind_rows(tblUnitsHourly, tblUnitsDaily)
    else tblUnits <- NULL
    
    # Put everything else together
    tblDescription <- jsObj[setdiff(nms, c("hourly", "hourly_units", "daily", "daily_units"))] %>%
        tibble::as_tibble()
    
    # Return the list objects
    list(tblDaily=tblDaily, tblHourly=tblHourly, tblUnits=tblUnits, tblDescription=tblDescription)
    
}

tmpOM2 <- readOpenMeteoJSON("notuse_testOM.json")
tmpOM2
prettyOpenMeteoMeta(tmpOM2)

identical(tmpOM$tblUnits, tmpOM2$tblUnits)
identical(tmpOM$tblDescription, tmpOM2$tblDescription)
identical(tmpOM$tblDaily %>% omProcessDaily(), tmpOM2$tblDaily)
identical(tmpOM$tblHourly %>% omProcessHourly(), tmpOM2$tblHourly)

```
  
The daily data is tested for file download, cached to avoid multiple hits to the server:  
```{r cache=TRUE}

testURLDaily <- helperOpenMeteoURL(cityName="Chicago IL", 
                                   dailyIndices=1:nrow(tblMetricsDaily),
                                   startDate="2010-01-01", 
                                   endDate="2023-06-15", 
                                   tz="America/Chicago"
                                   )
testURLDaily

# Download file
if(!file.exists("notuse_testOM_daily.json")) {
    fileDownload(fileName="notuse_testOM_daily.json", url=testURLDaily)
} else {
    cat("\nFile notuse_testOM_daily.json already exists, skipping download\n")
}

```
  
Data are read and stored as a list:  
```{r, fig.height=9, fig.width=9}

tmpOMDaily <- readOpenMeteoJSON("notuse_testOM_daily.json")
tmpOMDaily
prettyOpenMeteoMeta(tmpOMDaily)

# Exploration of precipitation hours by day
tmpOMDaily$tblDaily %>% count(precipitation_hours) %>% print(n=30)
tmpOMDaily$tblDaily %>%
    filter(lubridate::year(date)<=2022) %>%
    ggplot(aes(x=precipitation_hours)) + 
    geom_density(aes(group=lubridate::year(date), color=as.factor(lubridate::year(date)))) + 
    scale_color_discrete("Year") + 
    labs(title="Hours of Precipitation per Day", x="Hours of Precipitation", y="Annual density")
tmpOMDaily$tblDaily %>%
    filter(lubridate::year(date)<=2022) %>%
    ggplot(aes(x=precipitation_hours)) + 
    geom_histogram(aes(fill=as.factor(lubridate::year(date))), bins=25) + 
    scale_fill_discrete("Year") + 
    facet_wrap(~lubridate::year(date)) +
    labs(title="Hours of Precipitation per Day", x="Hours of Precipitation", y="# Days")

```
  
Precipitation by month is explored:  
```{r, fig.height=9, fig.width=9}

dfPrecip <- tmpOMDaily$tblDaily %>%
    filter(lubridate::year(date)<=2022) %>%
    select(date, precipitation_sum, rain_sum, snowfall_sum) %>%
    mutate(month=factor(month.abb[lubridate::month(date)], levels=month.abb), 
           yyyymm=customYYYYMM(date)
           ) %>%
    group_by(yyyymm, month) %>%
    summarize(across(where(is.numeric), sum), n=n(), .groups="drop")
dfPrecip

# Boxplot of precipitation by month
dfPrecip %>%
    select(-n) %>%
    pivot_longer(-c(yyyymm, month)) %>%
    ggplot(aes(x=month, y=ifelse(name=="snowfall_sum", 10*value, value))) + 
    geom_boxplot(fill="lightblue") + 
    facet_wrap(~name, scales="free_y") + 
    labs(title="Precipitation by month (2010-2022)", y="Precipitation (mm)", x=NULL) + 
    theme(axis.text.x = element_text(angle = 90)) + 
    lims(y=c(0, NA))

# Mean precipitation by month
dfPrecip %>%
    group_by(month) %>%
    summarize(across(where(is.numeric), mean)) %>%
    ggplot(aes(x=month)) + 
    geom_col(aes(y=precipitation_sum), fill="green") + 
    geom_col(aes(y=rain_sum), fill="lightblue") + 
    geom_text(aes(y=rain_sum/2, label=round(rain_sum))) +
    geom_text(aes(y=rain_sum/2 + precipitation_sum/2, 
                  label=ifelse(precipitation_sum>rain_sum+3, round(precipitation_sum-rain_sum), "")
                  )
              ) + 
    geom_text(aes(y=precipitation_sum+5, label=round(precipitation_sum))) +
    labs(x=NULL, 
         y="Precipitation (mm)", 
         title="Mean precipitation by month (2010-2022)", 
         subtitle="Light blue is mm falling as rain, green is liquid equivalent of other"
         )

```
  
Average temperatures by month are also explored:  
```{r, fig.height=9, fig.width=9}

dfTemp <- tmpOMDaily$tblDaily %>%
    filter(lubridate::year(date)<=2022) %>%
    select(date, 
           temperature_2m_max, 
           temperature_2m_min, 
           apparent_temperature_max, 
           apparent_temperature_min
           ) %>%
    mutate(month=factor(month.abb[lubridate::month(date)], levels=month.abb), 
           yyyymm=customYYYYMM(date)
           ) %>%
    group_by(yyyymm, month) %>%
    summarize(across(where(is.numeric), mean), n=n(), .groups="drop")
dfTemp

# Boxplot of precipitation by month
dfTemp %>%
    select(-n) %>%
    pivot_longer(-c(yyyymm, month)) %>%
    ggplot(aes(x=month, y=value)) + 
    geom_boxplot(fill="lightblue") + 
    facet_wrap(~name) + 
    labs(title="Average temperature by month (2010-2022)", y="Average temperature (C)", x=NULL) + 
    theme(axis.text.x = element_text(angle = 90))

# Mean temperatures by month
dfTemp %>%
    select(-n) %>%
    group_by(month) %>%
    summarize(across(where(is.numeric), mean)) %>%
    pivot_longer(cols=-c(month)) %>%
    mutate(measType=stringr::str_replace(name, ".*_", ""), 
           meas=ifelse(str_detect(name, "apparent"), "apparent", "actual")
           ) %>%
    select(-name) %>%
    pivot_wider(id_cols=c(month, meas), names_from="measType", values_from="value") %>%
    ggplot(aes(x=month)) + 
    geom_tile(aes(y=(max+min)/2, height=max-min), width=0.5, fill="lightblue") +
    geom_text(aes(y=max+1, label=round(max, 1)), size=2.5) +
    geom_text(aes(y=min-1, label=round(min, 1)), size=2.5) +
    labs(x=NULL, 
         y="Temperature (C)", 
         title="Mean high and low temperature by month (2010-2022)", 
         subtitle="Actual temperature and apparent temperature"
         ) + 
    facet_wrap(~meas)

```
  
Sunrise and sunset times are explored:  
```{r, fig.height=9, fig.width=9}

dfSun <- tmpOMDaily$tblDaily %>%
    filter(lubridate::year(date)<=2022) %>%
    select(date, sunrise, sunset) %>%
    mutate(month=factor(month.abb[lubridate::month(date)], levels=month.abb), 
           yyyymm=customYYYYMM(date),
           across(c(sunrise, sunset), lubridate::ymd_hm), 
           sr=hms::as_hms(sunrise), 
           ss=hms::as_hms(sunset), 
           doy=lubridate::yday(date), 
           year=lubridate::year(date)
           ) 
dfSun

# Plot of sunrise and sunset by day of year
dfSun %>%
    select(date, month, year, doy, sr, ss) %>%
    ggplot(aes(x=doy, group=factor(year), color=factor(year))) + 
    geom_line(aes(y=sr)) + 
    geom_line(aes(y=ss)) + 
    geom_line(aes(y=(ss+sr)/2)) +
    labs(x="Day of year", y="Time (always on DST)", title="Sunrise, sunset, and solar noon by day of year") + 
    scale_color_discrete("Year")

# Plot of minutes gained from earliest/latest
dfSun %>%
    select(date, month, year, doy, sr, ss) %>%
    group_by(year) %>%
    mutate(dsr=max(sr)-sr, dss=ss-min(ss)) %>%
    ungroup() %>%
    rename(sunrise_change=dsr, sunset_change=dss) %>%
    pivot_longer(cols=c(sunrise_change, sunset_change)) %>%
    ggplot(aes(x=doy)) + 
    geom_point(aes(y=as.numeric(value)/60, color=name), size=0.5) +
    labs(x="Day of year", y="Minutes", title="Delta from latest sunrise / earliest sunset") + 
    scale_color_discrete("Metric")

```
  
Wind data is explored:  
```{r, fig.height=9, fig.width=9}

dfWind <- tmpOMDaily$tblDaily %>% 
    select(date, 
           dir=winddirection_10m_dominant, 
           spd=windspeed_10m_max, 
           gst=windgusts_10m_max
           ) %>% 
    mutate(month=lubridate::month(date), 
           year=lubridate::year(date), 
           dir10=round(dir/10)*10, 
           spd5=round(spd/5)*5, 
           gst5=round(gst/5)*5
           ) 
dfWind

# Plot of wind direction and speed
dfWind %>%
    ggplot(aes(x=dir, y=spd)) + 
    geom_point(alpha=0.2, size=0.5) + 
    coord_polar() + 
    facet_wrap(~factor(month.abb[month], levels=month.abb), nrow=2) + 
    geom_vline(xintercept=c(0, 90, 180, 270), lty=2, color="red") + 
    labs(title="Maximum wind speed and predominant direction (measured daily)", 
         y="Maximum Wind speed (km/h)", 
         x="Predominant Wind direction"
         ) + 
    scale_x_continuous(breaks=c(0, 90, 180, 270))

dfWind %>%
    filter(lubridate::year(date)<=2022) %>%
    count(month, dir10, spd5) %>%
    ggplot(aes(x=dir10, y=spd5)) + 
    geom_point(aes(size=n), alpha=0.2) + 
    coord_polar() + 
    facet_wrap(~factor(month.abb[month], levels=month.abb)) + 
    geom_vline(xintercept=c(0, 90, 180, 270), lty=2, color="red") + 
    labs(title="Maximum wind speed and predominant direction (measured daily)", 
         subtitle="Wind speed rounded to nearest 5 km/h, wind direction rounded to nearest 10 degrees",
         y="Maximum Wind speed (km/h)", 
         x="Predominant Wind direction"
         ) + 
    scale_x_continuous(breaks=c(0, 90, 180, 270))

# Plot of predominant wind direction
dfWind %>% 
    ggplot(aes(x=dir10)) + 
    geom_histogram(binwidth=10) + 
    facet_wrap(~factor(month.abb[month], levels=month.abb)) + 
    geom_vline(xintercept=c(0, 90, 180, 270, 360), lty=2, color="red") + 
    labs(title="Predominant wind direction (measured daily)", 
         y="# Days", 
         x="Predominant wind direction (rounded to nearest 10 degrees)"
         ) + 
    scale_x_continuous(breaks=c(0, 90, 180, 270, 360))

# Plot of maximum wind speed
dfWind %>% 
    ggplot(aes(x=spd5)) + 
    geom_histogram(binwidth=5) + 
    facet_wrap(~factor(month.abb[month], levels=month.abb)) + 
    labs(title="Maximum wind speed (measured daily)", 
         y="# Days", 
         x="Maximum wind speed (km/h, rounded to nearest 5 km/h)"
         )

# Mean maximum wind speed by month
dfWind %>%
    filter(year<=2022) %>%
    select(date, month, year, spd, gst) %>%
    pivot_longer(cols=-c(date, month, year)) %>%
    ggplot(aes(x=factor(month.abb[month], levels=month.abb), y=value)) + 
    geom_boxplot(fill="lightblue") + 
    facet_wrap(~c("gst"="2. Maximum wind gust", "spd"="1. Maximum wind speed")[name]) + 
    labs(title="Wind speed measured daily (2010-2022)", y="Wind speed (km/h)", x=NULL) + 
    theme(axis.text.x = element_text(angle = 90)) + 
    lims(y=c(0, NA))

```
  
Weather codes, radiation, and evapotranspiration are explored:  
```{r, fig.height=9, fig.width=9}

dfOther <- tmpOMDaily$tblDaily %>% 
    select(date, wc=weathercode, sw=shortwave_radiation_sum, et=et0_fao_evapotranspiration) %>%
    mutate(wc=factor(wc, levels=sort(unique(wc))), 
           year=lubridate::year(date), 
           month=factor(month.abb[lubridate::month(date)], levels=month.abb), 
           yyyymm=customYYYYMM(date)
           )
dfOther

# Histogram of weather code
dfOther %>%
    filter(year<=2022) %>%
    ggplot(aes(x=wc)) + 
    geom_bar() + 
    facet_wrap(~month) + 
    labs(title="Weather codes by month (2010-2022)", y="Count", x="Weather code") + 
    theme(axis.text.x = element_text(angle = 90))

# Mean radiation and evapotranspiration by month
dfOther %>%
    select(-year) %>%
    group_by(month) %>%
    summarize(across(where(is.numeric), mean)) %>%
    pivot_longer(cols=-c(month)) %>%
    ggplot(aes(x=month)) + 
    geom_point(aes(y=value)) +
    geom_line(aes(y=value, group=1)) +
    labs(x=NULL, 
         y=NULL, 
         title="Mean radiation and evapotranspiration by month (2010-2022)",
         subtitle="Evapotranspiration (mm) and Radiation (MegaJoules)"
         ) + 
    facet_wrap(~c("et"="Evapotranspiration (mm)", "sw"="Radiation (MJ)")[name], scales="free_y") + 
    lims(y=c(0, NA))

# Boxplot for radiation and evapotranspiration by month
dfOther %>%
    select(date, month, sw, et) %>%
    pivot_longer(-c(date, month)) %>%
    ggplot(aes(x=month)) + 
    geom_boxplot(aes(y=value), fill="lightblue") +
    labs(x=NULL, 
         y=NULL, 
         title="Daily radiation and evapotranspiration (2010-2022)",
         subtitle="Evapotranspiration (mm) and Radiation (MegaJoules)"
         ) + 
    facet_wrap(~c("et"="Evapotranspiration (mm)", "sw"="Radiation (MJ)")[name], scales="free_y") + 
    lims(y=c(0, NA))

```
  
The hourly data is tested for file download, cached to avoid multiple hits to the server:  
```{r cache=TRUE}

testURLHourly <- helperOpenMeteoURL(cityName="Chicago IL", 
                                    hourlyIndices=1:nrow(tblMetricsHourly),
                                    startDate="2010-01-01", 
                                    endDate="2023-06-15", 
                                    tz="America/Chicago"
                                    )
testURLHourly

# Download file
if(!file.exists("notuse_testOM_hourly.json")) {
    fileDownload(fileName="notuse_testOM_hourly.json", url=testURLHourly)
} else {
    cat("\nFile notuse_testOM_hourly.json already exists, skipping download\n")
}

```
  
Data are read and stored as a list:  
```{r, fig.height=9, fig.width=9}

tmpOMHourly <- readOpenMeteoJSON("notuse_testOM_hourly.json")
tmpOMHourly
prettyOpenMeteoMeta(tmpOMHourly)

```
  
Consistency of data between daily and hourly is explored:  
```{r}

# Variables where maximum of hourly should be created
vrblMax <- c("weathercode", "temperature_2m", "apparent_temperature", "windspeed_10m", "windgusts_10m")

# Variables where minimum of hourly should be created
vrblMin <- c("temperature_2m", "apparent_temperature")

# Variables where sum of hourly should be created
vrblSum <- c("precipitation", "rain", "snowfall", "shortwave_radiation", "et0_fao_evapotranspiration")

# Variables in daily not to explore
# date, time, sunrise, sunset

# Variables that require a different approach
# winddirection_10m_dominant, precipitation_hours

# Check that all variables are included in hourly data
c(vrblMax, vrblMin, vrblSum) %in% (tmpOMHourly$tblHourly %>% names)

# Create daily data from hourly
dfDailyFromHourly <- tmpOMHourly$tblHourly %>%
    group_by(date) %>%
    summarize(across(.cols=all_of(vrblMax), .fns=max, .names="{.col}_max"),
              across(.cols=all_of(vrblMin), .fns=min, .names="{.col}_min"), 
              across(.cols=all_of(vrblSum), .fns=sum, .names="{.col}_sum"), 
              precipitation_hours=sum(precipitation>0)
              ) %>%
    rename(weathercode=weathercode_max, et0_fao_evapotranspiration=et0_fao_evapotranspiration_sum)
dfDailyFromHourly
names(dfDailyFromHourly)
names(dfDailyFromHourly) %in% names(tmpOMDaily$tblDaily)

# Check data consistency
for (colName in names(dfDailyFromHourly)) {
    cat("\n", 
        colName, 
        ":", 
        all.equal(dfDailyFromHourly %>% pull(colName), tmpOMDaily$tblDaily %>% pull(colName))
        )
}

# Plot for differences in radiation
dfRadiation <- dfDailyFromHourly %>%
    select(date, shortwave_radiation_sum) %>%
    bind_rows(select(tmpOMDaily$tblDaily, date, shortwave_radiation_sum), .id="src") %>%
    mutate(src=c("1"="Daily from Hourly", "2"="Daily as Reported")[src])
dfRadiation %>%
    ggplot(aes(x=date, y=shortwave_radiation_sum)) + 
    geom_line(aes(group=src, color=src)) + 
    labs(x=NULL, y="Sum of radiation", title="Comparison of shortwave radiation by day by source")

# Exploration of units
tmpOMDaily$tblUnits %>% filter(name=="shortwave_radiation_sum")
tmpOMHourly$tblUnits %>% filter(name=="shortwave_radiation")

# Conversion of Watts per hour to MegaJoules
# 0.0036 megajoules/watt-hour
dfRadiation %>%
    ggplot(aes(x=date, y=ifelse(src=="Daily from Hourly", 0.0036, 1)*shortwave_radiation_sum)) + 
    geom_line(aes(group=src, color=src)) + 
    labs(x=NULL, 
         y="Sum of radiation", 
         title="Comparison of shortwave radiation by day by source", 
         subtitle="Summed from hourly multiplied by 0.0036 to convert Watt-hours to MegaJoules"
         )
dfRadiation %>%
    pivot_wider(id_cols="date", names_from="src", values_from="shortwave_radiation_sum") %>%
    mutate(rat=`Daily as Reported`/`Daily from Hourly`) %>%
    summary()

```
  
With the exception of radiation (reported in different units causing slight rounding differences), the reported daily data matches the expected aggregate of the reported hourly data
  
Precipitation by hour is explored:  
```{r, fig.height=9, fig.width=9}

# Hourly precipitation data
dfHourlyPrecip <- tmpOMHourly$tblHourly %>%
    select(time, hour, precipitation, snowfall, rain) %>%
    mutate(year=lubridate::year(time), 
           month=factor(month.abb[lubridate::month(time)], levels=month.abb)
           ) %>%
    pivot_longer(cols=-c(time, year, month, hour))
dfHourlyPrecip

# Nil precipitation percent
dfNilPrecip <- dfHourlyPrecip %>%
    group_by(month, name) %>%
    summarize(pctNil=mean(value==0), .groups="drop")
dfNilPrecip

# Graphs of precipitation amount by month
for(metric in unique(dfHourlyPrecip$name)) {
    p1 <- dfHourlyPrecip %>%
        filter(name==metric, value>0, year<=2022) %>%
        ggplot() + 
        geom_histogram(aes(x=value), bins = 50) + 
        facet_wrap(~month) + 
        labs(x=NULL, 
             y=NULL, 
             title=paste0(metric, 
                          ": hourly total (", 
                          tmpOMHourly$tblUnits %>% filter(name==metric) %>% pull(value), 
                          ") from 2010-2022"
                          )
             ) + 
        geom_text(data=dfNilPrecip %>% filter(name==metric), 
                  aes(x=Inf, 
                      y=Inf, 
                      label=paste0("Excludes ", round(100*pctNil, 1), "%\nof observations at 0")
                      ), 
                  size=2.5, 
                  hjust=1, 
                  vjust=1
                  )
    print(p1)
}

```
  
Temperature by hour is explored:  
```{r, fig.height=9, fig.width=9}

# Hourly temperature data
dfHourlyTemp <- tmpOMHourly$tblHourly %>%
    select(time, hour, temperature_2m, apparent_temperature, dewpoint_2m) %>%
    mutate(year=lubridate::year(time), 
           month=factor(month.abb[lubridate::month(time)], levels=month.abb)
           ) %>%
    pivot_longer(cols=-c(time, year, month, hour))
dfHourlyTemp

# Graphs of precipitation amount by month
for(metric in unique(dfHourlyTemp$name)) {
    p1 <- dfHourlyTemp %>%
        filter(name==metric, year<=2022) %>%
        ggplot() + 
        geom_boxplot(aes(x=factor(hour), y=value), fill = "lightblue") + 
        facet_wrap(~month) + 
        labs(x=NULL, 
             y=NULL, 
             title=paste0(metric, 
                          ": hourly boxplot (", 
                          tmpOMHourly$tblUnits %>% filter(name==metric) %>% pull(value), 
                          ") from 2010-2022"
                          )
             )
    print(p1)
    
}

# Spread of temperature by day
dfHourlyTemp %>%
    mutate(date=lubridate::date(time)) %>%
    group_by(year, month, date, name) %>%
    summarize(maxValue=max(value), minValue=min(value), mdnValue=median(value), .groups="drop") %>%
    mutate(spd=maxValue-minValue) %>%
    group_by(month, name) %>%
    summarize(across(where(is.numeric), mean), .groups="drop") %>%
    ggplot(aes(x=fct_rev(month), y=spd)) + 
    geom_point() + 
    coord_flip() +
    facet_wrap(~name) + 
    labs(title="Average high/low spread of key metrics by month (deg C)", x=NULL, y="deg C") + 
    lims(y=c(0, NA))

```
  
Hours with maximum/minimum temperature and precipitation are explored:  
```{r, fig.height=9, fig.width=9}

# Create temperature and precipitation data
dfHourlyTempPrecip <- dfHourlyTemp %>%
    bind_rows(dfHourlyPrecip) %>%
    mutate(date=lubridate::date(time)) %>% 
    arrange(time, name) 
dfHourlyTempPrecip

# Limit to temperature, dewpoint, and precipitation
# Limit precipitation to only days with precipitation > 0
tmpDF <- dfHourlyTempPrecip %>%
    filter(name %in% c("dewpoint_2m", "precipitation", "temperature_2m")) %>%
    group_by(date, name) %>% 
    filter(name!="precipitation" | sum(value)>0) %>%
    mutate(isMax=ifelse(value==max(value), 1, 0), isMin=ifelse(value==min(value), 1, 0)) %>% 
    group_by(name, month, hour) %>% 
    summarize(across(c(isMax, isMin), mean), .groups="drop") %>% 
    pivot_longer(-c(name, month, hour), names_to="metric")
tmpDF

# Plot max/min for metric
for(keyMetric in unique(tmpDF$name)) {

    p1 <- tmpDF %>% 
        filter(name==keyMetric) %>% 
        ggplot(aes(x=hour, y=value)) + 
        geom_line(aes(color=metric, group=metric)) + 
        facet_wrap(~month) + 
        labs(x="Hour of day", 
             y="% of time as max/min", 
             title=paste0(keyMetric, ": maximum and minimum by hour"), 
             subtitle=paste0("Ties included as full value", 
                             ifelse(keyMetric=="precipitation", " (days with no precipitation excluded)", "")
                             )
             ) + 
        scale_color_discrete("Metric:")
    print(p1)
}

# Plot percent of hours with precipitation
dfHourlyTempPrecip %>% 
    filter(name=="precipitation") %>% 
    group_by(month, hour) %>%
    summarize(pct0=mean(value>0), pct05=mean(value>=0.5), .groups="drop") %>%
    pivot_longer(-c(month, hour)) %>%
    mutate(name=ifelse(name=="pct0", ">=0.1 mm", ">=0.5 mm")) %>%
    ggplot(aes(x=hour, y=value)) + 
    geom_line(aes(group=name, color=name)) + 
    facet_wrap(~month) + 
    labs(x="Hour of day", 
         y="% at/above precipitation hurdle", 
         title=paste0("% of observations with precipitation in past hour")
         ) + 
    lims(y=c(0, NA))

```
  
Wind by hour is explored:  
```{r, fig.height=9, fig.width=9}

# Hourly wind data
dfHourlyWind <- tmpOMHourly$tblHourly %>%
    select(time, hour, windspeed_10m, windgusts_10m, winddirection_10m) %>%
    mutate(year=lubridate::year(time), 
           month=factor(month.abb[lubridate::month(time)], levels=month.abb)
           ) %>%
    pivot_longer(cols=-c(time, year, month, hour))
dfHourlyWind

# Graphs of wind speed/gust by month
for(metric in setdiff(unique(dfHourlyWind$name), "winddirection_10m")) {
    p1 <- dfHourlyWind %>%
        filter(name==metric, year<=2022) %>%
        ggplot() + 
        geom_boxplot(aes(x=factor(hour), y=value), fill = "lightblue") + 
        facet_wrap(~month) + 
        labs(x=NULL, 
             y=NULL, 
             title=paste0(metric, 
                          ": hourly boxplot (", 
                          tmpOMHourly$tblUnits %>% filter(name==metric) %>% pull(value), 
                          ") from 2010-2022"
                          )
             )
    print(p1)
    
}

# Average wind speed and gust by hour
dfHourlyWind %>%
    filter(!(name %in% c("winddirection_10m")), year<=2022) %>%
    group_by(month, hour, name) %>%
    summarize(across("value", mean), .groups="drop") %>%
    ggplot(aes(x=factor(hour), y=value)) + 
    geom_point(aes(group=name, color=name)) + 
    facet_wrap(~month) + 
    labs(title="Average wind speed and gust (km/h)", x=NULL, y="km/h") + 
    lims(y=c(0, NA))

# Average wind direction by hour
dfHourlyWind %>%
    filter((name %in% c("winddirection_10m")), year<=2022) %>%
    mutate(preDom=case_when(value<45|value>=315~"N", 
                            value<135~"E", 
                            value<225~"S", 
                            value<315~"W", 
                            TRUE~"error"
                            )
           ) %>%
    count(month, hour, preDom) %>%
    ggplot(aes(x=factor(hour), y=n)) + 
    geom_col(aes(fill=factor(preDom, levels=c("N", "W", "S", "E"))), position="fill") + 
    facet_wrap(~month) + 
    labs(title="Distribution of wind direction", x=NULL, y="Wind direction (%)") + 
    scale_fill_discrete("")

```
  
Wind by N/S and E/W is also explored:  
```{r, fig.height=9, fig.width=9}

# Average wind direction by hour
tmpWindDir <- dfHourlyWind %>%
    filter((name %in% c("winddirection_10m")), year<=2022) %>%
    mutate(ew=case_when(value>30&value<150~"E", 
                        value>210&value<=330~"W", 
                        TRUE~"none"
                        ), 
           ns=case_when(value>300|value<=60~"N", 
                        value>120&value<=240~"S", 
                        TRUE~"none"
                        )
    ) 
tmpWindDir

tmpWindDir %>%
    count(month, hour, ew) %>%
    group_by(month, hour) %>%
    mutate(pct=n/sum(n)) %>%
    ungroup() %>%
    ggplot(aes(x=hour, y=pct)) + 
    geom_line(aes(color=factor(ew, levels=c("W", "none", "E")))) + 
    facet_wrap(~month) + 
    labs(title="Wind direction", 
         x="Hour of day", 
         y="% of observations", 
         subtitle="(030-150 deg defined as East, 210-330 deg defined as West)"
         ) + 
    scale_color_discrete("") + 
    lims(y=c(0, NA))

tmpWindDir %>%
    count(month, hour, ns) %>%
    group_by(month, hour) %>%
    mutate(pct=n/sum(n)) %>%
    ungroup() %>%
    ggplot(aes(x=hour, y=pct)) + 
    geom_line(aes(color=factor(ns, levels=c("S", "none", "N")))) + 
    facet_wrap(~month) + 
    labs(title="Wind direction", 
         x="Hour of day", 
         y="% of observations", 
         subtitle="(300-060 deg defined as North, 120-240 deg defined as South)"
         ) + 
    scale_color_discrete("") + 
    lims(y=c(0, NA))

```
  
Hourly wind directions are averaged using arctan. The formula is arctan2(y=sum-of-sin, x=sum-of-cos):  
```{r, fig.height=9, fig.width=9}

# Unweighted by wind speed
tmpWindatan_uw <- tmpWindDir %>%
    mutate(date=lubridate::date(time), cosine=cos(2*pi*value/360), sine=sin(2*pi*value/360)) %>%
    group_by(date) %>%
    summarize(across(c(cosine, sine), sum), .groups="drop") %>%
    mutate(arctangent=atan(sine/cosine), 
           arctangent2=atan2(y=sine, x=cosine), 
           avgdir=((arctangent2/2)*(360/pi)), 
           avgdir=ifelse(avgdir<0, 360+avgdir, avgdir)
           ) %>%
    left_join(select(tmpOMDaily$tblDaily, date, wdd=winddirection_10m_dominant), by="date")
tmpWindatan_uw

tmpWindatan_uw %>%
    mutate(delta=avgdir-wdd) %>%
    summary()

tmpWindatan_uw %>%
    count(wdd, awdd=round(avgdir)) %>%
    ggplot(aes(x=wdd, y=awdd)) + 
    geom_point(aes(size=n)) + 
    labs(x="Reported dominant wind direction (daily data)", 
         y="Calculated dominant wind direction (hourly data)", 
         title="Relationship between reported and calculated dominant wind direction", 
         subtitle="Unweighted by wind speed"
         ) + 
    scale_size_continuous("# days")

# Weighted by wind speed
tmpWindatan_wtd <- tmpOMHourly$tblHourly %>%
    select(date, time, wd=winddirection_10m, ws=windspeed_10m) %>%
    mutate(cosine=cos(2*pi*wd/360), sine=sin(2*pi*wd/360)) %>%
    group_by(date) %>%
    summarize(across(c(cosine, sine), .fns=function(x) sum(x*ws)/sum(ws)), sws=sum(ws), .groups="drop") %>%
    mutate(arctangent=atan(sine/cosine), 
           arctangent2=atan2(y=sine, x=cosine), 
           avgdir=((arctangent2/2)*(360/pi)), 
           avgdir=ifelse(avgdir<0, 360+avgdir, avgdir), 
           avgspd=sws/24, 
           dist=sws*sqrt(cosine**2+sine**2)
           ) %>%
    left_join(select(tmpOMDaily$tblDaily, date, wdd=winddirection_10m_dominant), by="date")
tmpWindatan_wtd

tmpWindatan_wtd %>%
    mutate(delta=avgdir-wdd) %>%
    summary()

tmpWindatan_wtd %>%
    count(wdd, awdd=round(avgdir)) %>%
    ggplot(aes(x=wdd, y=awdd)) + 
    geom_point(aes(size=n)) + 
    labs(x="Reported dominant wind direction (daily data)", 
         y="Calculated dominant wind direction (hourly data)", 
         title="Relationship between reported and calculated dominant wind direction", 
         subtitle="Weighted by wind speed"
         ) + 
    scale_size_continuous("# days")

tmpWindatan_wtd %>%
    count(rdist=round(dist), rspd=round(avgspd)) %>%
    ggplot(aes(x=rspd, y=rdist/24)) + 
    geom_point(aes(size=n)) + 
    labs(title="Average wind speed (total and weighted by direction) by day", 
         x="Average wind speed per day", 
         y="Weighted average wind speed\n(total distance on average angle, divided by 24)"
         ) + 
    geom_abline(slope=1, intercept=0, lty=2) +
    scale_size_continuous("# days")

tmpWindatan_wtd %>%
    mutate(rdist=round(dist), rspd=round(avgspd)) %>%
    ggplot(aes(x=rspd)) + 
    geom_boxplot(aes(y=dist/24/rspd, group=rspd), fill="lightblue") + 
    labs(title="Average wind speed (total and weighted by direction) by day", 
         x="Average wind speed per day", 
         y="Average weighted wind speed\n(as ratio of gross average)"
         )

tmpWindatan_wtd %>%
    filter(abs(avgdir-wdd)>1)

```
  
The wind direction averaging of hourly data, weighted by wind speed, is consistent with the reported dominant wind direction in the daily data.
  
Weather codes are explored:  
```{r, fig.height=9, fig.width=9}

# Hourly weather codes, evapotranspiration, and shortwave
dfHourlyCode <- tmpOMHourly$tblHourly %>%
    select(time, hour, wc=weathercode, et=et0_fao_evapotranspiration, sw=shortwave_radiation) %>%
    mutate(year=lubridate::year(time), 
           month=factor(month.abb[lubridate::month(time)], levels=month.abb)
           ) %>%
    pivot_longer(cols=-c(time, year, month, hour))
dfHourlyCode

# Exploration of weather codes overall
dfHourlyCode %>%
    filter(name=="wc", year<=2022) %>%
    count(value) %>%
    ggplot(aes(x=fct_rev(factor(value)), y=n/1000)) + 
    geom_col(fill="lightblue") + 
    geom_text(aes(label=round(n/1000, 1)), hjust=0, size=3) +
    labs(title="Weather codes in hourly data (2010-2022)", y="Count (000)", x="Weather Code") + 
    coord_flip()

# Exploration of weather codes by month
dfHourlyCode %>%
    filter(name=="wc", year<=2022) %>%
    count(month, value) %>%
    group_by(month) %>%
    mutate(pct=n/sum(n)) %>%
    ungroup() %>%
    ggplot(aes(x=fct_rev(factor(value)), y=pct)) + 
    geom_col(fill="lightblue") + 
    geom_text(aes(label=paste0(round(100*pct, 1), "%")), hjust=0, size=3) +
    labs(title="Weather codes in hourly data (2010-2022)", y="Frequency", x="Weather Code") + 
    coord_flip() + 
    facet_wrap(~month)

# Exploration of weather codes by month
dfHourlyCode %>%
    filter(name=="wc", year<=2022) %>%
    mutate(wType=case_when(value<=3~"Dry", 
                           value>=51 & value<=55~"Drizzle", 
                           value>=61 & value<=65~"Rain", 
                           value>=71 & value<=75~"Snow", 
                           TRUE~"error"
                           )
           ) %>%
    count(month, wType) %>%
    group_by(month) %>%
    mutate(pct=n/sum(n)) %>%
    ungroup() %>%
    ggplot(aes(x=month, y=pct)) + 
    geom_col(aes(fill=factor(wType, levels=c("Dry", "Drizzle", "Rain", "Snow"))), position="stack") + 
    labs(title="Precipitation types in hourly data (2010-2022)", y="Frequency", x=NULL) + 
    coord_flip() + 
    scale_fill_discrete("")

# Weather codes from WMO Code Table 4677 (select examples)
# 00	Cloud development not observed or not observable
# 01	Clouds generally dissolving or becoming less developed
# 02	State of sky on the whole unchanged
# 03	Clouds generally forming or developing
# 51	Drizzle, not freezing, continuous (slight)
# 53	Drizzle, not freezing, continuous (moderate)
# 55	Drizzle, not freezing, continuous (heavy)
# 61	Rain, not freezing, continuous (slight)
# 63	Rain, not freezing, continuous (moderate)
# 65	Rain, not freezing, continuous (heavy)
# 71	Continuous fall of snowflakes (slight)
# 73	Continuous fall of snowflakes (moderate)
# 75	Continuous fall of snowflakes (heavy)

```
  
Shortwave radiation is explored:  
```{r, fig.height=9, fig.width=9}

dfHourlyCode %>%
    filter(name=="sw", year<=2022) %>%
    ggplot(aes(x=factor(hour), y=value)) + 
    geom_boxplot(fill="lightblue") + 
    facet_wrap(~month) + 
    labs(title="Average shortwave solar radiation over the past hour", x=NULL, y="Watts per sqaure meter")

dfHourlyCode %>%
    filter(name=="sw", year<=2022) %>%
    group_by(hour, month) %>%
    summarize(value=mean(value), .groups="drop") %>%
    ggplot(aes(x=factor(hour), y=value)) + 
    geom_point() + 
    facet_wrap(~month) + 
    labs(title="Average hourly shortwave solar radiation by hour and month (2010-2022)", 
         x=NULL, 
         y="Watts per sqaure meter"
         )

dfHourlyCode %>%
    filter(name %in% c("sw"), year<=2022) %>%
    mutate(date=lubridate::date(time)) %>%
    group_by(date, name) %>% 
    mutate(isMax=ifelse(value==max(value), 1, 0), isMin=ifelse(value==min(value), 1, 0)) %>% 
    group_by(name, month, hour) %>% 
    summarize(across(c(isMax, isMin), mean), .groups="drop") %>% 
    pivot_longer(-c(name, month, hour), names_to="metric") %>%
    ggplot(aes(x=hour, y=value)) + 
    geom_line(aes(color=metric, group=metric)) + 
    facet_wrap(~month) + 
    labs(x="Hour of day", 
         y="% of time as max/min", 
         title=paste0("Shortwave radiation", ": maximum and minimum by hour"), 
         subtitle=paste0("Ties included as full value")
         ) + 
    scale_color_discrete("Metric:")

```
  

Evapotranspiration is explored:  
```{r, fig.height=9, fig.width=9}

dfHourlyCode %>%
    filter(name=="et", year<=2022) %>%
    ggplot(aes(x=factor(hour), y=value)) + 
    geom_boxplot(fill="lightblue") + 
    facet_wrap(~month) + 
    labs(title="Evapotranspiration of a well-watered grass field", x="Hour of day", y="mm")

dfHourlyCode %>%
    filter(name=="et", year<=2022) %>%
    group_by(hour, month) %>%
    summarize(value=mean(value), .groups="drop") %>%
    ggplot(aes(x=factor(hour), y=value)) + 
    geom_point() + 
    facet_wrap(~month) + 
    labs(title="Mean evapotranspiration of a well-watered grass field by hour and month (2010-2022)", 
         x="Hour of day", 
         y="mm"
         )

dfHourlyCode %>%
    filter(name %in% c("et"), year<=2022) %>%
    mutate(date=lubridate::date(time)) %>%
    group_by(date, name) %>% 
    mutate(isMax=ifelse(value==max(value), 1, 0), isMin=ifelse(value==min(value), 1, 0)) %>% 
    group_by(name, month, hour) %>% 
    summarize(across(c(isMax, isMin), mean), .groups="drop") %>% 
    pivot_longer(-c(name, month, hour), names_to="metric") %>%
    ggplot(aes(x=hour, y=value)) + 
    geom_line(aes(color=metric, group=metric)) + 
    facet_wrap(~month) + 
    labs(x="Hour of day", 
         y="proportion of time as max/min", 
         title=paste0("Evapotranspiration", ": maximum and minimum by hour"), 
         subtitle=paste0("Ties included as full value")
         ) + 
    scale_color_discrete("Metric:")

```
  
Cloud cover is explored:  
```{r, fig.height=9, fig.width=9}

# Create cloud cover data
dfHourlyCloud <- tmpOMHourly$tblHourly %>% 
    select(time, hour, contains("cloud")) %>% 
    mutate(year=lubridate::year(time),
           month=factor(month.abb[lubridate::month(time)], levels=month.abb)
           ) %>%
    pivot_longer(cols=-c(time, year, month, hour))
dfHourlyCloud

# Boxplot for cloud cover types
for(keyMetric in unique(dfHourlyCloud$name)) {

    p1 <- dfHourlyCloud %>% 
        filter(name==keyMetric, year<=2022) %>% 
        ggplot(aes(x=factor(hour), y=value)) + 
        geom_boxplot(fill="lightblue") + 
        facet_wrap(~month) + 
        labs(x="Hour of day", 
             y="% sky covered with cloud", 
             title=paste0(keyMetric, ": % sky covered with cloud")
             )
    print(p1)

}

# Create max/min for metric
tmpDFCloud <- dfHourlyCloud %>%
    filter(year<=2022) %>%
    mutate(date=lubridate::date(time)) %>%
    group_by(date, name) %>% 
    mutate(isMax=ifelse(value==max(value), 1, 0), isMin=ifelse(value==min(value), 1, 0)) %>% 
    group_by(name, month, hour) %>% 
    summarize(across(c(isMax, isMin), mean), .groups="drop") %>% 
    pivot_longer(-c(name, month, hour), names_to="metric")
tmpDFCloud

# Plot max/min for metric
for(keyMetric in unique(tmpDFCloud$name)) {

    p1 <- tmpDFCloud %>% 
        filter(name==keyMetric) %>% 
        ggplot(aes(x=hour, y=value)) + 
        geom_line(aes(color=metric, group=metric)) + 
        facet_wrap(~month) + 
        labs(x="Hour of day", 
             y="% of time as max/min", 
             title=paste0(keyMetric, ": maximum and minimum by hour"), 
             subtitle=paste0("Ties included as full value")
             ) + 
        scale_color_discrete("Metric:")
    print(p1)
    
}

```
  
Atmospheric pressure is explored:  
```{r, fig.height=9, fig.width=9}

# Create pressure data
dfHourlyPressure <- tmpOMHourly$tblHourly %>% 
    select(time, hour, contains("pressure")) %>% 
    mutate(year=lubridate::year(time),
           month=factor(month.abb[lubridate::month(time)], levels=month.abb)
           ) %>%
    pivot_longer(cols=-c(time, year, month, hour))
dfHourlyPressure

# Boxplot for pressure types
for(keyMetric in unique(dfHourlyPressure$name)) {

    tmpUnits <- tmpOMHourly$tblUnits %>% filter(name==keyMetric) %>% pull(value)
    
    p1 <- dfHourlyPressure %>% 
        filter(name==keyMetric, year<=2022) %>% 
        ggplot(aes(x=factor(hour), y=value)) + 
        geom_boxplot(fill="lightblue") + 
        facet_wrap(~month) + 
        labs(x="Hour of day", 
             y=paste0(keyMetric, " (", tmpUnits, ")"), 
             title=paste0(keyMetric, ": ", tmpUnits)
             )
    print(p1)

}

dfHourlyPressure %>%
    pivot_wider(id_cols=c(time, hour, year, month)) %>%
    count(pressure_msl, surface_pressure) %>%
    ggplot(aes(x=pressure_msl, y=surface_pressure)) + 
    geom_point(aes(size=n)) + 
    geom_smooth(aes(weight=n), method="lm") +
    labs(title="Surface pressure vs. MSL", x="MSL", y="Surface Pressure")

# Create max/min for metric
tmpDFPressure <- dfHourlyPressure %>%
    filter(year<=2022) %>%
    mutate(date=lubridate::date(time)) %>%
    group_by(date, name) %>% 
    mutate(isMax=ifelse(value==max(value), 1, 0), isMin=ifelse(value==min(value), 1, 0)) %>% 
    group_by(name, month, hour) %>% 
    summarize(across(c(isMax, isMin), mean), .groups="drop") %>% 
    pivot_longer(-c(name, month, hour), names_to="metric")
tmpDFPressure

# Plot max/min for metric
for(keyMetric in unique(tmpDFPressure$name)) {

    p1 <- tmpDFPressure %>% 
        filter(name==keyMetric) %>% 
        ggplot(aes(x=hour, y=value)) + 
        geom_line(aes(color=metric, group=metric)) + 
        facet_wrap(~month) + 
        labs(x="Hour of day", 
             y="% of time as max/min", 
             title=paste0(keyMetric, ": maximum and minimum by hour"), 
             subtitle=paste0("Ties included as full value")
             ) + 
        scale_color_discrete("Metric:")
    print(p1)
    
}

```
  
Soil temperature is explored:  
```{r, fig.height=9, fig.width=9}

# Create soil temperature data
dfHourlySoilTemp <- tmpOMHourly$tblHourly %>% 
    select(time, date, hour, starts_with("soil_temp")) %>% 
    mutate(year=lubridate::year(time),
           month=factor(month.abb[lubridate::month(time)], levels=month.abb)
           ) %>%
    pivot_longer(cols=-c(time, date, year, month, hour))
dfHourlySoilTemp

# Boxplot for soil temperature
for(keyMetric in unique(dfHourlySoilTemp$name)) {

    tmpUnits <- tmpOMHourly$tblUnits %>% filter(name==keyMetric) %>% pull(value)
    
    p1 <- dfHourlySoilTemp %>% 
        filter(name==keyMetric, year<=2022) %>% 
        ggplot(aes(x=factor(hour), y=value)) + 
        geom_boxplot(fill="lightblue") + 
        facet_wrap(~month) + 
        labs(x="Hour of day", 
             y=paste0(keyMetric, " (", tmpUnits, ")"), 
             title=paste0(keyMetric, ": ", tmpUnits)
             )
    print(p1)

}

# Mean and standard deviation by month
dfHourlySoilTemp %>%
    group_by(date, name) %>%
    summarize(across(value, .fns=list(mu=mean, sigma=sd)), .groups="drop") %>%
    mutate(doy=lubridate::yday(date)) %>%
    group_by(doy, name) %>%
    summarize(across(starts_with("value"), .fns=list(mu=mean)), .groups="drop") %>%
    pivot_longer(cols=-c(doy, name), names_to="metric") %>%
    ggplot(aes(x=doy, y=value)) + 
    geom_line(aes(group=name, color=stringr::str_replace(name, "soil_temperature_", ""))) + 
    facet_wrap(~c("value_mu_mu"="Daily mean", "value_sigma_mu"="Mean daily standard deviation")[metric], 
               nrow=2, 
               scales="free_y"
               ) + 
    labs(x="Day of Year", 
         y="Degrees (C)", 
         title="Soil temperature mean and average daily standard deviation"
         ) + 
    scale_color_discrete("Soil depth")

# Create max/min for metric
tmpDFSoilTemp <- dfHourlySoilTemp %>%
    filter(year<=2022) %>%
    mutate(date=lubridate::date(time)) %>%
    group_by(date, name) %>% 
    mutate(isMax=ifelse(value==max(value), 1, 0), isMin=ifelse(value==min(value), 1, 0)) %>% 
    group_by(name, month, hour) %>% 
    summarize(across(c(isMax, isMin), mean), .groups="drop") %>% 
    pivot_longer(-c(name, month, hour), names_to="metric")
tmpDFSoilTemp

# Plot max/min for metric
for(keyMetric in unique(tmpDFSoilTemp$name)) {

    p1 <- tmpDFSoilTemp %>% 
        filter(name==keyMetric) %>% 
        ggplot(aes(x=hour, y=value)) + 
        geom_line(aes(color=metric, group=metric)) + 
        facet_wrap(~month) + 
        labs(x="Hour of day", 
             y="% of time as max/min", 
             title=paste0(keyMetric, ": maximum and minimum by hour"), 
             subtitle=paste0("Ties included as full value")
             ) + 
        scale_color_discrete("Metric:")
    print(p1)
    
}

```
  
Soil moisture is explored:  
```{r, fig.height=9, fig.width=9}

# Create soil moisture data
dfHourlySoilMoist <- tmpOMHourly$tblHourly %>% 
    select(time, date, hour, starts_with("soil_moist")) %>% 
    mutate(year=lubridate::year(time),
           month=factor(month.abb[lubridate::month(time)], levels=month.abb)
           ) %>%
    pivot_longer(cols=-c(time, date, year, month, hour))
dfHourlySoilMoist

# Boxplot for soil moisture
for(keyMetric in unique(dfHourlySoilMoist$name)) {

    tmpUnits <- tmpOMHourly$tblUnits %>% filter(name==keyMetric) %>% pull(value)
    
    p1 <- dfHourlySoilMoist %>% 
        filter(name==keyMetric, year<=2022) %>% 
        ggplot(aes(x=factor(hour), y=value)) + 
        geom_boxplot(fill="lightblue") + 
        facet_wrap(~month) + 
        labs(x="Hour of day", 
             y=paste0(keyMetric, " (", tmpUnits, ")"), 
             title=paste0(keyMetric, ": ", tmpUnits)
             )
    print(p1)

}

# Mean and standard deviation by month
dfHourlySoilMoist %>%
    group_by(date, name) %>%
    summarize(across(value, .fns=list(mu=mean, sigma=sd)), .groups="drop") %>%
    mutate(doy=lubridate::yday(date)) %>%
    group_by(doy, name) %>%
    summarize(across(starts_with("value"), .fns=list(mu=mean)), .groups="drop") %>%
    pivot_longer(cols=-c(doy, name), names_to="metric") %>%
    ggplot(aes(x=doy, y=value)) + 
    geom_line(aes(group=name, color=stringr::str_replace(name, "soil_moisture_", ""))) + 
    facet_wrap(~c("value_mu_mu"="Daily mean", "value_sigma_mu"="Mean daily standard deviation")[metric], 
               nrow=2, 
               scales="free_y"
               ) + 
    labs(x="Day of Year", 
         y="cubic meters per cubic meter\n(volumetric mixing ratio)", 
         title="Soil moisture mean and average daily standard deviation"
         ) + 
    scale_color_discrete("Soil depth")

# Create max/min for metric
tmpDFSoilMoist <- dfHourlySoilMoist %>%
    filter(year<=2022) %>%
    mutate(date=lubridate::date(time)) %>%
    group_by(date, name) %>% 
    mutate(isMax=ifelse(value==max(value), 1, 0), isMin=ifelse(value==min(value), 1, 0)) %>% 
    group_by(name, month, hour) %>% 
    summarize(across(c(isMax, isMin), mean), .groups="drop") %>% 
    pivot_longer(-c(name, month, hour), names_to="metric")
tmpDFSoilMoist

# Plot max/min for metric
for(keyMetric in unique(tmpDFSoilMoist$name)) {

    p1 <- tmpDFSoilMoist %>% 
        filter(name==keyMetric) %>% 
        ggplot(aes(x=hour, y=value)) + 
        geom_line(aes(color=metric, group=metric)) + 
        facet_wrap(~month) + 
        labs(x="Hour of day", 
             y="% of time as max/min", 
             title=paste0(keyMetric, ": maximum and minimum by hour"), 
             subtitle=paste0("Ties included as full value")
             ) + 
        scale_color_discrete("Metric:")
    print(p1)
    
}

```
  
Metrics are explored for their variation over months and over hours of day:  
```{r, fig.height=9, fig.width=9}

# Sample database
tmpTemp <- tmpOMHourly$tblHourly %>%
    select(time, date, temperature_2m) %>%
    mutate(month=factor(month.abb[lubridate::month(date)], levels=month.abb), 
           hour=lubridate::hour(time), 
           fct_hour=factor(hour), 
           rndTemp=round(2*temperature_2m, 0)/2
           )
tmpTemp

# Simple predictive model for temperature/month
prdTemp <- tmpTemp %>%
    count(rndTemp, month) %>%
    arrange(rndTemp, desc(n)) %>%
    group_by(rndTemp) %>%
    mutate(corr=row_number()==1, pred=first(month)) %>%
    ungroup()
prdTemp

# Confusion matrix and accuracy
prdTemp %>%
    count(month, corr, wt=n) %>%
    pivot_wider(id_cols=month, names_from=corr, values_from=n, values_fill=0) %>%
    bind_rows(summarize(., across(where(is.numeric), sum)) %>% 
                  mutate(month="All") %>% 
                  select(month, everything())
              ) %>%
    mutate(n=`TRUE`+`FALSE`, 
           pctCorrect=`TRUE`/n, 
           pctNaive=ifelse(month=="All", 1/(nrow(.)-1), 2*n/sum(n)), 
           lift=pctCorrect/pctNaive
           )

prdTemp %>%
    count(month, pred, corr, wt=n) %>%
    ggplot(aes(x=month, y=pred)) + 
    labs(x="Actual month", y="Predicted month", title="Actual vs. predicted month using temperature") + 
    geom_text(aes(label=n)) + 
    geom_tile(aes(fill=corr), alpha=0.25)

```
  
The simple predictive model is converted to functional form:  
```{r, fig.height=9, fig.width=9}

simpleOneVarPredict <- function(df, 
                                tgt, 
                                prd, 
                                nPrint=30, 
                                showPlot=TRUE, 
                                returnData=TRUE
                                ) {
    
    # FUNCTION ARGUMENTS:
    # df: data frame or tibble with key elements
    # tgt: target variable
    # prd: predictor variable
    # nPrint: maximum number of lines of confusion matrix to print
    #         0 means do not print any summary statistics
    # showPlot: boolean, should overlap plot be created and shown?
    
    # Counts of predictor to target variable
    dfPred <- df %>%
        group_by(across(all_of(c(prd, tgt)))) %>%
        summarize(n=n(), .groups="drop") %>%
        arrange(across(all_of(prd)), desc(n)) %>%
        group_by(across(all_of(prd))) %>%
        mutate(correct=row_number()==1, predicted=first(get(tgt))) %>%
        ungroup()

    # Confusion matrix and accuracy
    dfConf <- dfPred %>%
        group_by(across(all_of(c(tgt, "correct")))) %>%
        summarize(n=sum(n), .groups="drop") %>%
        pivot_wider(id_cols=tgt, names_from=correct, values_from=n, values_fill=0) %>%
        mutate(n=`TRUE`+`FALSE`, 
               pctCorrect=`TRUE`/n, 
               pctNaive=1/(nrow(.)), 
               lift=pctCorrect/pctNaive-1
               )
    
    # Overall confusion matrix
    dfConfAll <- dfConf %>%
        summarize(nMax=max(n), across(c(`FALSE`, `TRUE`, "n"), sum)) %>%
        mutate(pctCorrect=`TRUE`/n, 
               pctNaive=nMax/n, 
               lift=pctCorrect/pctNaive-1, 
               nBucket=length(unique(dfPred[[prd]]))
               )
    
    # Print confusion matrices
    if(nPrint > 0) {
        cat("\nAccuracy by target subgroup:\n")
        dfConf %>% print(n=nPrint)
        cat("\nOverall Accuracy:\n")
        dfConfAll %>% print(n=nPrint)
    }
    
    # Plot of overlaps
    if(isTRUE(showPlot)) {
        p1 <- dfPred %>%
            group_by(across(c(all_of(tgt), "predicted", "correct"))) %>%
            summarize(n=sum(n), .groups="drop") %>%
            ggplot(aes(x=get(tgt), y=predicted)) + 
            labs(x="Actual", 
                 y="Predicted", 
                 title=paste0("Actual vs. predicted ", tgt), 
                 subtitle=paste0("(using ", prd, ")")
                 ) + 
            geom_text(aes(label=n)) + 
            geom_tile(aes(fill=correct), alpha=0.25)
        print(p1)
    }
    
    # Return data if requested
    if(isTRUE(returnData)) list(dfPred=dfPred, dfConf=dfConf, dfConfAll=dfConfAll)
    
}

tstFunc <- simpleOneVarPredict(tmpTemp, tgt="month", prd="rndTemp")
tstFunc

all.equal(tstFunc$dfPred, rename(prdTemp, correct=corr, predicted=pred))

```
  
The function is tested for predictive power of numeric variables, bucketed in to percentiles, on month:  
```{r, fig.height=9, fig.width=9}

# Create percentiles for numeric variables
tmpTemp <- tmpOMHourly$tblHourly %>%
    mutate(month=factor(month.abb[lubridate::month(date)], levels=month.abb), 
           hour=lubridate::hour(time), 
           fct_hour=factor(hour), 
           across(where(is.numeric), .fns=function(x) round(100*percent_rank(x)), .names="pct_{.col}")
           )
tmpTemp

# Get key variable names
tmpNames <- tmpTemp %>% 
    select(starts_with("pct")) %>% 
    names()
tmpNames

# Get the key predictive metrics
tmpDFR <- map_dfr(.x=tmpNames, 
                  .f=function(x) simpleOneVarPredict(tmpTemp, tgt="month", prd=x, nPrint=0, showPlot=FALSE)$dfConfAll
                  ) %>% 
    mutate(vrbl=tmpNames) %>% 
    arrange(desc(lift))

# Print and plot lift by variable
tmpDFR %>% 
    print(n=50)
tmpDFR %>% 
    ggplot(aes(x=fct_reorder(stringr::str_replace_all(vrbl, "pct_", ""), lift), y=lift)) + 
    geom_col(fill="lightblue") + 
    coord_flip() + 
    labs(x=NULL, y="lift", title="Lift by hourly variable percentile in predicting month")

# Example for soil temperature and high clouds
simpleOneVarPredict(tmpTemp, tgt="month", prd="pct_soil_temperature_100_to_255cm", returnData=FALSE)
simpleOneVarPredict(tmpTemp, tgt="month", prd="pct_cloudcover_high", returnData=FALSE)

```
  
The function is tested for predictive power of numeric variables, bucketed in to percentiles, on hour:  
```{r, fig.height=9, fig.width=9}

# Remove hour from tmpNames
tmpNamesNoHour <- setdiff(tmpNames, "pct_hour")
tmpNamesNoHour

# Get the key predictive metrics
tmpDFRHour <- map_dfr(.x=tmpNamesNoHour, 
                  .f=function(x) simpleOneVarPredict(tmpTemp, tgt="fct_hour", prd=x, nPrint=0, showPlot=FALSE)$dfConfAll
                  ) %>% 
    mutate(vrbl=tmpNamesNoHour) %>% 
    arrange(desc(lift))

# Print and plot lift by variable
tmpDFRHour %>% 
    print(n=50)
tmpDFRHour %>% 
    ggplot(aes(x=fct_reorder(stringr::str_replace_all(vrbl, "pct_", ""), lift), y=lift)) + 
    geom_col(fill="lightblue") + 
    coord_flip() + 
    labs(x=NULL, y="lift", title="Lift by hourly variable percentile in predicting hour")

# Example for diffuse radiation and soil moisture
simpleOneVarPredict(tmpTemp, tgt="fct_hour", prd="pct_diffuse_radiation", returnData=FALSE)
simpleOneVarPredict(tmpTemp, tgt="fct_hour", prd="pct_soil_moisture_100_to_255cm", returnData=FALSE)

```
  
Random variables, split equally 0-5, 0-25, and 0-100, are included as an example null state:  
```{r, fig.height=9, fig.width=9}

# Add random variables
set.seed(23072413)
tmpTemp <- tmpTemp %>%
    mutate(rnd005=sample(0:5, size=n(), replace=TRUE), 
           rnd025=sample(0:25, size=n(), replace=TRUE), 
           rnd100=sample(0:100, size=n(), replace=TRUE)
           )

# Get key variable names
tmpNames_v2 <- tmpTemp %>% 
    select(starts_with("pct"), starts_with("rnd")) %>% 
    names()
tmpNames_v2

# Get the key predictive metrics
tmpDFR_v2 <- map_dfr(.x=tmpNames_v2, 
                     .f=function(x) simpleOneVarPredict(tmpTemp, tgt="month", prd=x, nPrint=0, showPlot=FALSE)$dfConfAll
                     ) %>% 
    mutate(vrbl=tmpNames_v2) %>% 
    arrange(desc(lift))

# Print and plot lift by variable
tmpDFR_v2 %>% 
    print(n=50)
tmpDFR_v2 %>% 
    mutate(fillColor=ifelse(str_detect(vrbl, pattern="pct_"), "lightblue", "red")) %>%
    ggplot(aes(x=fct_reorder(stringr::str_replace_all(vrbl, "pct_", ""), lift), y=lift)) + 
    geom_col(aes(fill=fillColor)) + 
    coord_flip() + 
    labs(x=NULL, y="lift", title="Lift by hourly variable percentile in predicting month") + 
    scale_fill_identity()


# Remove hour from tmpNames
tmpNamesNoHour_v2 <- setdiff(tmpNames_v2, "pct_hour")
tmpNamesNoHour_v2

# Get the key predictive metrics
tmpDFRHour_v2 <- map_dfr(.x=tmpNamesNoHour_v2, 
                         .f=function(x) simpleOneVarPredict(tmpTemp, tgt="fct_hour", prd=x, nPrint=0, showPlot=FALSE)$dfConfAll
                  ) %>% 
    mutate(vrbl=tmpNamesNoHour_v2) %>% 
    arrange(desc(lift))

# Print and plot lift by variable
tmpDFRHour_v2 %>% 
    print(n=50)
tmpDFRHour_v2 %>% 
    mutate(fillColor=ifelse(str_detect(vrbl, pattern="pct_"), "lightblue", "red")) %>%
    ggplot(aes(x=fct_reorder(stringr::str_replace_all(vrbl, "pct_", ""), lift), y=lift)) + 
    geom_col(aes(fill=fillColor)) + 
    coord_flip() + 
    labs(x=NULL, y="lift", title="Lift by hourly variable percentile in predicting hour") + 
    scale_fill_identity()

```
  
Function simpleOneVarPredict() is updated to allow for test-train:  
```{r, fig.height=9, fig.width=9}

simpleOneVarPredict <- function(df, 
                                tgt, 
                                prd, 
                                dfTest=NULL,
                                nPrint=30, 
                                showPlot=TRUE, 
                                returnData=TRUE
                                ) {
    
    # FUNCTION ARGUMENTS:
    # df: data frame or tibble with key elements (training data set)
    # tgt: target variable
    # prd: predictor variable
    # dfTest: test dataset for applying predictions
    # nPrint: maximum number of lines of confusion matrix to print
    #         0 means do not print any summary statistics
    # showPlot: boolean, should overlap plot be created and shown?
    
    # Counts of predictor to target variable
    dfPred <- df %>%
        group_by(across(all_of(c(prd, tgt)))) %>%
        summarize(n=n(), .groups="drop") %>%
        arrange(across(all_of(prd)), desc(n)) %>%
        group_by(across(all_of(prd))) %>%
        mutate(correct=row_number()==1, predicted=first(get(tgt))) %>%
        ungroup()

    # Confusion matrix and accuracy
    dfConf <- dfPred %>%
        group_by(across(all_of(c(tgt, "correct")))) %>%
        summarize(n=sum(n), .groups="drop") %>%
        pivot_wider(id_cols=tgt, names_from=correct, values_from=n, values_fill=0) %>%
        mutate(n=`TRUE`+`FALSE`, 
               pctCorrect=`TRUE`/n, 
               pctNaive=1/(nrow(.)), 
               lift=pctCorrect/pctNaive-1
               )
    
    # Overall confusion matrix
    dfConfAll <- dfConf %>%
        summarize(nMax=max(n), across(c(`FALSE`, `TRUE`, "n"), sum)) %>%
        mutate(pctCorrect=`TRUE`/n, 
               pctNaive=nMax/n, 
               lift=pctCorrect/pctNaive-1, 
               nBucket=length(unique(dfPred[[prd]]))
               )
    
    # Print confusion matrices
    if(nPrint > 0) {
        cat("\nAccuracy by target subgroup (training data):\n")
        dfConf %>% print(n=nPrint)
        cat("\nOverall Accuracy (training data):\n")
        dfConfAll %>% print(n=nPrint)
    }
    
    # Plot of overlaps
    if(isTRUE(showPlot)) {
        p1 <- dfPred %>%
            group_by(across(c(all_of(tgt), "predicted", "correct"))) %>%
            summarize(n=sum(n), .groups="drop") %>%
            ggplot(aes(x=get(tgt), y=predicted)) + 
            labs(x="Actual", 
                 y="Predicted", 
                 title=paste0("Training data - Actual vs. predicted ", tgt), 
                 subtitle=paste0("(using ", prd, ")")
                 ) + 
            geom_text(aes(label=n)) + 
            geom_tile(aes(fill=correct), alpha=0.25)
        print(p1)
    }
    
    # Create metrics for test dataset if requested
    if(!is.null(dfTest)) {
        # Get maximum category from training data
        mostPredicted <- count(dfPred, predicted, wt=n) %>% slice(1) %>% pull(predicted)
        # Get mapping of metric to prediction
        dfPredict <- dfPred %>% 
            group_by(across(all_of(c(prd, "predicted")))) %>% 
            summarize(n=sum(n), .groups="drop")
        # Create predictions for test data
        dfPredTest <- dfTest %>%
            select(all_of(c(prd, tgt))) %>%
            left_join(select(dfPredict, -n)) %>%
            replace_na(list(predicted=mostPredicted)) %>%
            group_by(across(all_of(c(prd, tgt, "predicted")))) %>%
            summarize(n=n(), .groups="drop") %>%
            mutate(correct=(get(tgt)==predicted))
        # Create confusion statistics for test data
        dfConfTest <- dfPredTest %>%
            group_by(across(all_of(c(tgt, "correct")))) %>%
            summarize(n=sum(n), .groups="drop") %>%
            pivot_wider(id_cols=tgt, names_from=correct, values_from=n, values_fill=0) %>%
            mutate(n=`TRUE`+`FALSE`, 
                   pctCorrect=`TRUE`/n, 
                   pctNaive=1/(nrow(.)), 
                   lift=pctCorrect/pctNaive-1
                   )
        # Overall confusion matrix for test data
        dfConfAllTest <- dfConfTest %>%
            summarize(nMax=max(n), across(c(`FALSE`, `TRUE`, "n"), sum)) %>%
            mutate(pctCorrect=`TRUE`/n, 
                   pctNaive=nMax/n, 
                   lift=pctCorrect/pctNaive-1, 
                   nBucket=length(unique(dfConfTest[[prd]]))
               )
        # Print confusion matrices
        if(nPrint > 0) {
            cat("\nAccuracy by target subgroup (testing data):\n")
            dfConfTest %>% print(n=nPrint)
            cat("\nOverall Accuracy (testing data):\n")
            dfConfAllTest %>% print(n=nPrint)
            }
    } else {
        dfPredTest <- NULL
        dfConfTest <- NULL
        dfConfAllTest <- NULL
        
    }
    
    # Return data if requested
    if(isTRUE(returnData)) list(dfPred=dfPred, 
                                dfConf=dfConf, 
                                dfConfAll=dfConfAll, 
                                dfPredTest=dfPredTest, 
                                dfConfTest=dfConfTest, 
                                dfConfAllTest=dfConfAllTest
                                )
    
}

# Original format
simpleOneVarPredict(tmpTemp, tgt="month", prd="pct_soil_temperature_100_to_255cm", showPlot=FALSE)

# Train-test format
set.seed(23072514)
idxTrain <- sample(1:nrow(tmpTemp), size=round(.8*nrow(tmpTemp)), replace=FALSE)
simpleOneVarPredict(tmpTemp[idxTrain,], 
                    tgt="month", 
                    prd="pct_soil_temperature_100_to_255cm", 
                    showPlot=FALSE, 
                    dfTest=tmpTemp[-idxTrain,]
                    )

```
  
The function is split into components for better modularity:  
```{r, fig.height=9, fig.width=9}

# Fit a single predictor to a single categorical variable
simpleOneVarFit <- function(df, 
                            tgt, 
                            prd, 
                            rankType="last", 
                            naMethod=TRUE
                            ) {
    
    # FUNCTION ARGUMENTS:
    # df: data frame or tibble with key elements (training data set)
    # tgt: target variable
    # prd: predictor variable
    # rankType: method for breaking ties of same n, passed to base::rank as ties.method=
    # naMethod: method for handling NA in ranks, passed to base::rank as na.last=
    
    # Counts of predictor to target variable, and associated predictions
    df %>%
        group_by(across(all_of(c(prd, tgt)))) %>%
        summarize(n=n(), .groups="drop") %>%
        arrange(across(all_of(prd)), desc(n), across(all_of(tgt))) %>%
        group_by(across(all_of(prd))) %>%
        mutate(rankN=n()+1-rank(n, ties.method=rankType, na.last=naMethod)) %>%
        arrange(across(all_of(prd)), rankN) %>%
        ungroup()

}

# Test that results are the same for a variable with many ties, and a variable with fewer ties
tstFit <- simpleOneVarFit(tmpTemp, tgt="month", prd="pct_snowfall")
tstOrig <- simpleOneVarPredict(tmpTemp, tgt="month", prd="pct_snowfall", nPrint=0, showPlot=FALSE)$dfPred
all.equal(tstOrig %>% select(-correct, -predicted), tstFit %>% select(-rankN))

tstFit <- simpleOneVarFit(tmpTemp, tgt="month", prd="pct_soil_temperature_100_to_255cm")
tstOrig <- simpleOneVarPredict(tmpTemp, 
                               tgt="month", 
                               prd="pct_soil_temperature_100_to_255cm", 
                               nPrint=0, 
                               showPlot=FALSE
                               )$dfPred
all.equal(tstOrig %>% select(-correct, -predicted), tstFit %>% select(-rankN))

```
  
A prediction mapper is created, along with the mapping for anything not in the data:  
```{r, fig.height=9, fig.width=9}

simpleOneVarMapper <- function(df, tgt, prd) {
    
    # FUNCTION ARGUMENTS:
    # df: data frame or tibble from SimpleOneVarFit()
    # tgt: target variable
    # prd: predictor variable
    
    # Get the most common actual results
    dfCommon <- df %>% count(across(all_of(tgt)), wt=n, sort=TRUE)
    
    # Get the predictions
    dfPredictor <- df %>%
        group_by(across(all_of(prd))) %>%
        filter(row_number()==1) %>%
        select(all_of(c(prd, tgt))) %>%
        ungroup()
    
    list(dfPredictor=dfPredictor, dfCommon=dfCommon)
    
}

tstMapper <- simpleOneVarMapper(tstFit, tgt="month", prd="pct_soil_temperature_100_to_255cm")
tstMapper

```
  
A function to apply the prediction mapper is created:  
```{r, fig.height=9, fig.width=9}

simpleOneVarApplyMapper <- function(df, 
                                    tgt,
                                    prd, 
                                    mapper, 
                                    mapperDF="dfPredictor", 
                                    mapperDefault="dfCommon",
                                    prdName="predicted"
                                    ) {
    
    # FUNCTION ARGUMENTS:
    # df: data frame containing prd for predicting tgt
    # tgt: target variable in df
    # prd: predictor variable in df
    # mapper: mapping list from sinpleOneVarMapper()
    # mapperDF: element that can be used to merge mappings
    # mapperDefault: element that can be used for NA resulting from merging mapperDF
    # prdName: name for the prediction variable
    
    # Extract the mapper and default value
    vecRename <- c(prdName) %>% purrr::set_names(tgt)
    dfMap <- mapper[[mapperDF]] %>% select(all_of(c(prd, tgt))) %>% colRenamer(vecRename=vecRename)
    chrDefault <- mapper[[mapperDefault]] %>% slice(1) %>% pull(tgt)
    
    # Merge mappings to df
    df %>%
        left_join(dfMap, by=prd) %>%
        replace_na(list("predicted"=chrDefault))
    
}

# Example with mutated variable
tmpMutated <- tmpTemp %>% 
    select(date, hour, month, pct_soil_temperature_100_to_255cm) %>%
    mutate(pct_soil_temperature_100_to_255cm=ifelse(hour==0, -10, pct_soil_temperature_100_to_255cm))
tstApplied <- simpleOneVarApplyMapper(tmpMutated,
                                      tgt="month", 
                                      prd="pct_soil_temperature_100_to_255cm", 
                                      mapper=tstMapper
                                      )
tstApplied

# Example using tstFit to confirm same outputs
tstApplied <- simpleOneVarApplyMapper(tstFit,
                                      tgt="month", 
                                      prd="pct_soil_temperature_100_to_255cm", 
                                      mapper=tstMapper
                                      )
tstApplied
all.equal(tstOrig %>% select(-correct, correct), 
          tstApplied %>% select(-rankN) %>% mutate(correct=month==predicted)
          )

```
  
A function to create the confusion matrix data is written:  
```{r, fig.height=9, fig.width=9}

simpleOneVarConfusionData <- function(df, 
                                      tgtOrig,
                                      tgtPred, 
                                      otherVars=c(),
                                      weightBy="n"
                                      ) {
    
    # FUNCTION ARGUMENTS:
    # df: data frame from simpleOneVarApplyMapper()
    # tgtOrig: original target variable name in df
    # tgtPred: predicted target variable name in df
    # otherVars: other variables to be kept (will be grouping variables)
    # weightBy: weighting variable for counts in df (NULL means count each row of df as 1)
    
    # Confusion matrix data creation
    df %>%
        group_by(across(all_of(c(tgtOrig, tgtPred, otherVars)))) %>%
        summarize(n=if(!is.null(weightBy)) sum(get(weightBy)) else n(), .groups="drop") %>%
        mutate(correct=get(tgtOrig)==get(tgtPred))
    
}

# Example with and without weighting
simpleOneVarConfusionData(tstApplied, tgtOrig="month", tgtPred="predicted", weightBy=NULL)
simpleOneVarConfusionData(tstApplied, tgtOrig="month", tgtPred="predicted")

```
  
A function to report the confusion matrix data is written:  
```{r, fig.height=9, fig.width=9}

simpleOneVarConfusionReport <- function(df, 
                                        tgtOrig,
                                        tgtPred, 
                                        otherVars=c(), 
                                        printConf=TRUE,
                                        printConfOrig=printConf, 
                                        printConfPred=printConf,
                                        printConfOverall=printConf, 
                                        plotConf=TRUE, 
                                        plotDesc="",
                                        nBucket=NA, 
                                        predictorVarName="", 
                                        returnData=FALSE
                                        ) {
    
    # FUNCTION ARGUMENTS:
    # df: data frame from simpleOneVarConfusionData()
    # tgtOrig: original target variable name in df
    # tgtPred: predicted target variable name in df
    # otherVars: other variables to be kept (will be grouping variables) - NOT IMPLEMENTED
    # printConf: boolean, should confusion matrix data be printed? Applies to all three
    # printConfOrig: boolean, should confusion data be printed based on original target variable?
    # printConfPred: boolean, should confusion data be printed based on predicted target variable?
    # printConfOverall: boolean, should overall confusion data be printed?
    # plotConf: boolean, should confusion overlap data be plotted?
    # plotDesc: descriptive label to be included in front of plot title
    # nBucket: number of buckets used for prediction (pass from previous data)
    # predictorVarName: variable name to be included in chart description
    # returnData: boolean, should the confusion matrices be returned?
    
    # Confusion data based on original target variable
    if(isTRUE(printConfOrig) | isTRUE(returnData)) {
        dfConfOrig <- df %>%
            group_by(across(all_of(c(tgtOrig)))) %>%
            summarize(right=sum(n*correct), wrong=sum(n)-right, n=sum(n), .groups="drop") %>%
            mutate(pctRight=right/n, pctNaive=n/(sum(n)), lift=pctRight/pctNaive-1)
    }

    # Confusion data based on predicted target variable
    if(isTRUE(printConfPred) | isTRUE(returnData)) {
        dfConfPred <- df %>%
            group_by(across(all_of(c(tgtPred)))) %>%
            summarize(right=sum(n*correct), wrong=sum(n)-right, n=sum(n), .groups="drop") %>%
            mutate(pctRight=right/n)
    }

    # Overall confusion data
    if(isTRUE(printConfOverall) | isTRUE(returnData)) {
        maxNaive <- df %>%
            group_by(across(all_of(tgtOrig))) %>%
            summarize(n=sum(n), .groups="drop") %>%
            arrange(desc(n)) %>%
            slice(1) %>%
            pull(n)
        dfConfOverall <- df %>%
            summarize(right=sum(n*correct), wrong=sum(n)-right, n=sum(n), .groups="drop") %>%
            mutate(maxN=maxNaive, pctRight=right/n, pctNaive=maxN/n, lift=pctRight/pctNaive-1, nBucket=nBucket)
    }
    
    # Confusion report based on original target variable
    if(isTRUE(printConfOrig)) {
        cat("\nConfusion data based on original target variable:", tgtOrig, "\n")
        dfConfOrig %>%
            print(n=50)
    }

    # Confusion report based on predicted target variable
    if(isTRUE(printConfPred)) {
        cat("\nConfusion data based on predicted target variable:", tgtPred, "\n")
        dfConfPred %>%
            print(n=50)
    }
    
    # Overall confusion matrix
    if(isTRUE(printConfOverall)) {
        cat("\nOverall confusion matrix\n")
        dfConfOverall %>%
            print(n=50)
    }
    
    # Plot of overlaps
    if(isTRUE(plotConf)) {
        p1 <- df %>%
            group_by(across(all_of(c(tgtOrig, tgtPred, "correct")))) %>%
            summarize(n=sum(n), .groups="drop") %>%
            ggplot(aes(x=get(tgtOrig), y=get(tgtPred))) + 
            labs(x="Actual", 
                 y="Predicted", 
                 title=paste0(plotDesc, "Actual vs. predicted ", tgtOrig), 
                 subtitle=paste0("(using ", predictorVarName, ")")
                 ) + 
            geom_text(aes(label=n)) + 
            geom_tile(aes(fill=correct), alpha=0.25)
        print(p1)
    }
    
    # Return data if requested
    if(isTRUE(returnData)) list(dfConfOrig=dfConfOrig, dfConfPred=dfConfPred, dfConfOverall=dfConfOverall)
    
}

# Example with weighting
simpleOneVarConfusionData(tstApplied, tgtOrig="month", tgtPred="predicted") %>%
    simpleOneVarConfusionReport(tgtOrig="month", 
                                tgtPred="predicted", 
                                nBucket=length(unique(tstApplied$pct_soil_temperature_100_to_255cm)), 
                                predictorVarName=names(tstApplied)[1]
                                )

# Example with weighting and data returned without reporting
simpleOneVarConfusionData(tstApplied, tgtOrig="month", tgtPred="predicted") %>%
    simpleOneVarConfusionReport(tgtOrig="month", 
                                tgtPred="predicted", 
                                nBucket=length(unique(tstApplied$pct_soil_temperature_100_to_255cm)), 
                                predictorVarName=names(tstApplied)[1], 
                                printConf=FALSE, 
                                plotConf=FALSE,
                                returnData=TRUE
                                )

```
  
The functions are run together:  
```{r, fig.height=9, fig.width=9}

# Create fitting data
dfFit <- simpleOneVarFit(tmpTemp, tgt="month", prd="pct_temperature_2m") 

# Apply and report on fitting data
dfFit %>%
    simpleOneVarMapper(tgt="month", prd="pct_temperature_2m") %>%
    simpleOneVarApplyMapper(df=dfFit, tgt="month", prd="pct_temperature_2m", mapper=.) %>%
    simpleOneVarConfusionData(tgtOrig="month", tgtPred="predicted") %>%
    simpleOneVarConfusionReport(tgtOrig="month", 
                                tgtPred="predicted", 
                                nBucket=length(unique(dfFit$pct_temperature_2m))
                                )

```
  
A function is written to chain all of the functions:  
```{r, fig.height=9, fig.width=9}

simpleOneVarChain <- function(df,
                              tgt,
                              prd,
                              mapper=NULL, 
                              rankType="last", 
                              naMethod=TRUE, 
                              printReport=TRUE, 
                              plotDesc="",
                              returnData=TRUE, 
                              includeConfData=FALSE
                              ) {

    # FUNCTION ARGUMENTS:
    # df: data frame or tibble with key elements (training or testing data set)
    # tgt: target variable
    # prd: predictor variable
    # mapper: mapping file to be applied for predictions (NULL means create from simpleOneVarApply())
    # rankType: method for breaking ties of same n, passed to base::rank as ties.method=
    # naMethod: method for handling NA in ranks, passed to base::rank as na.last=    
    # printReport: boolean, should the confusion report data and plot be printed?
    # plotDesc: descriptive label to be included in front of plot title
    # returnData: boolean, should data elements be returned?
    # includeConfData: boolean, should confusion data be returned?
    
    # Create the summary of predictor-target-n
    dfFit <- simpleOneVarFit(df, tgt=tgt, prd=prd, rankType=rankType, naMethod=naMethod)     

    # Create the mapper if it does not already exist
    if(is.null(mapper)) mapper <- simpleOneVarMapper(dfFit, tgt=tgt, prd=prd)
    
    # Apply mapper to data
    dfApplied <- simpleOneVarApplyMapper(dfFit, tgt=tgt, prd=prd, mapper=mapper)

    # Create confusion data
    dfConfusion <- simpleOneVarConfusionData(dfApplied, tgtOrig=tgt, tgtPred="predicted")
    
    # Create confusion report if requested
    if(isTRUE(printReport) | isTRUE(includeConfData)) {
        dfConfReport <- simpleOneVarConfusionReport(df=dfConfusion, 
                                                    tgtOrig=tgt, 
                                                    tgtPred="predicted", 
                                                    nBucket=length(unique(dfApplied[[prd]])), 
                                                    predictorVarName=prd, 
                                                    printConf=printReport, 
                                                    plotConf=printReport,
                                                    plotDesc=plotDesc,
                                                    returnData=includeConfData
                                                    )
    }
    
    # Return data if requested
    if(isTRUE(returnData)) {
        ret <- list(dfFit=dfFit, mapper=mapper, dfApplied=dfApplied, dfConfusion=dfConfusion)
        if(isTRUE(includeConfData)) ret<-c(ret, list(dfConfData=dfConfReport))
        ret
    }
    
}

# Full process
tmpChain <- simpleOneVarChain(tmpTemp, tgt="month", prd="pct_temperature_2m")
str(tmpChain)

# Plots only
simpleOneVarChain(tmpTemp, tgt="month", prd="pct_temperature_2m", returnData=FALSE)

# Data only
tmpChain_v2 <- simpleOneVarChain(tmpTemp, tgt="month", prd="pct_temperature_2m", printReport=FALSE)
identical(tmpChain_v2, tmpChain)

# Data only using a mapper
tmpChain_v3 <- simpleOneVarChain(tmpTemp, 
                                 tgt="month", 
                                 prd="pct_temperature_2m", 
                                 mapper=tmpChain$mapper,
                                 printReport=FALSE
                                 )
identical(tmpChain_v3, tmpChain)

# Return confusion data
tmpChain_v4 <- simpleOneVarChain(tmpTemp, 
                                 tgt="month", 
                                 prd="pct_temperature_2m", 
                                 mapper=tmpChain$mapper,
                                 printReport=FALSE, 
                                 includeConfData=TRUE
                                 )
identical(tmpChain_v4[1:4], tmpChain)
tmpChain_v4$dfConfData

```
  
The process is then run on a train-test basis for a single variable:  
```{r, fig.height=9, fig.width=9}

# Add random variables to dataset, then split in to test and train
set.seed(23080412)
tmpTempRand <- tmpTemp %>%
    mutate(pct_0005=sample(0:5, size=nrow(.), replace=TRUE),
           pct_0025=sample(0:25, size=nrow(.), replace=TRUE), 
           pct_0100=sample(0:100, size=nrow(.), replace=TRUE), 
           pct_0250=sample(0:250, size=nrow(.), replace=TRUE),
           pct_0500=sample(0:500, size=nrow(.), replace=TRUE), 
           pct_1000=sample(0:1000, size=nrow(.), replace=TRUE), 
           pct_2500=sample(0:2500, size=nrow(.), replace=TRUE), 
           pct_5000=sample(0:5000, size=nrow(.), replace=TRUE), 
           )
idxTrain <- sort(sample(1:nrow(tmpTempRand), size=round(0.75*nrow(tmpTempRand)), replace=FALSE))
tmpTempTrain <- tmpTempRand[idxTrain, ]
tmpTempTest <- tmpTempRand[-idxTrain, ]

# Full process run on training data
tmpChainTrain <- simpleOneVarChain(tmpTempTrain, 
                                   tgt="month", 
                                   prd="pct_temperature_2m", 
                                   includeConfData=TRUE,
                                   plotDesc="Training data: "
                                   )

# Diagnostics run on testing data
tmpChainTest <- simpleOneVarChain(tmpTempTest, 
                                  tgt="month", 
                                  prd="pct_temperature_2m", 
                                  mapper=tmpChainTrain$mapper,
                                  includeConfData=TRUE,
                                  plotDesc="Testing data: "
                                  )

# Overall confusion
tmpChainTrain$dfConfData$dfConfOverall
tmpChainTest$dfConfData$dfConfOverall

```
  
The process for running train-test is created as a function:  
```{r, fig.height=9, fig.width=9}

simpleOneVarTrainTest <- function(dfTrain,
                                  dfTest,
                                  tgt,
                                  prd,
                                  rankType="last", 
                                  naMethod=TRUE, 
                                  printReport=FALSE, 
                                  includeConfData=TRUE, 
                                  returnData=TRUE
                              ) {

    # FUNCTION ARGUMENTS:
    # dfTrain: data frame or tibble with key elements (training data set)
    # dfTest: data frame or tibble with key elements (testing data set)
    # tgt: target variable
    # prd: predictor variable
    # rankType: method for breaking ties of same n, passed to base::rank as ties.method=
    # naMethod: method for handling NA in ranks, passed to base::rank as na.last=    
    # printReport: boolean, should the confusion report data and plot be printed?
    # includeConfData: boolean, should confusion data be returned?
    # returnData: boolean, should data elements be returned?
    
    # Fit the training data
    tmpTrain <- simpleOneVarChain(df=dfTrain, 
                                  tgt=tgt, 
                                  prd=prd,
                                  rankType=rankType,
                                  naMethod=naMethod,
                                  printReport=printReport,
                                  plotDesc="Training data: ",
                                  returnData=TRUE,
                                  includeConfData=includeConfData
                                  )
    
    # Fit the testing data
    tmpTest <- simpleOneVarChain(df=dfTest, 
                                 tgt=tgt, 
                                 prd=prd,
                                 mapper=tmpTrain$mapper,
                                 rankType=rankType,
                                 naMethod=naMethod,
                                 printReport=printReport,
                                 plotDesc="Testing data: ",
                                 returnData=TRUE,
                                 includeConfData=includeConfData
                                 )
    
    # Return data if requested
    if(isTRUE(returnData)) list(tmpTrain=tmpTrain, tmpTest=tmpTest)
    
}

# Full process without plotting
tmpVTT <- simpleOneVarTrainTest(dfTrain=tmpTempTrain, 
                                dfTest=tmpTempTest, 
                                tgt="month", 
                                prd="pct_temperature_2m"
                                )
str(tmpVTT)
# Extracting key elements of prediction accuracy
map_dfr(.x=tmpVTT, .f=function(x) x$dfConfData$dfConfOverall) %>%
    mutate(dataType=names(tmpVTT), 
           tgt=names(tmpVTT[[1]]$mapper$dfPredictor)[2], 
           prd=names(tmpVTT[[1]]$mapper$dfPredictor)[1]
           ) %>%
    select(dataType, tgt, prd, everything())

```
  
Predictive power for each variable on month is explored:  
```{r, fig.height=9, fig.width=9}

# Get all pct variables
pctVars <- tmpTempTrain %>% 
    select(starts_with("pct")) %>%
    names()
pctVars

# Run each variable and combine as dfr
tmpLiftPct <- map_dfr(.x=pctVars, 
        .f=function(x) {
            tmp <- simpleOneVarTrainTest(dfTrain=tmpTempTrain, dfTest=tmpTempTest, tgt="month", prd=x)
            map_dfr(.x=tmp, .f=function(y) y$dfConfData$dfConfOverall) %>%
                mutate(dataType=names(tmp), 
                       tgt=names(tmp[[1]]$mapper$dfPredictor)[2], 
                       prd=names(tmp[[1]]$mapper$dfPredictor)[1]
                       ) %>%
                select(dataType, tgt, prd, everything())
            }
        )
tmpLiftPct

```
  
Variables are plotted based on explanatory power on month:  
```{r, fig.height=9, fig.width=9}

tmpLiftPct %>% 
    ggplot(aes(x=fct_reorder(prd, lift, min), y=lift)) + 
    geom_point() + 
    coord_flip() + 
    facet_wrap(~c("tmpTest"="1. Test data", "tmpTrain"="2. Training data")[dataType]) + 
    geom_hline(yintercept=0, lty=2) +
    labs(y="Lift (percent correct divided by percent of most frequent month, minus 1)", 
         x=NULL, 
         title="Explanatory power of variable on month"
         )

```
  
Predictive power for each variable on hour (as factor) is explored:  
```{r, fig.height=9, fig.width=9}

# Run each variable and combine as dfr (pctVars derived previously for month)
tmpLiftHourPct <- map_dfr(.x=pctVars, 
                          .f=function(x) {
                              tmp <- simpleOneVarTrainTest(dfTrain=tmpTempTrain, 
                                                           dfTest=tmpTempTest, 
                                                           tgt="fct_hour", 
                                                           prd=x
                                                           )
                              map_dfr(.x=tmp, .f=function(y) y$dfConfData$dfConfOverall) %>%
                                  mutate(dataType=names(tmp), 
                                         tgt=names(tmp[[1]]$mapper$dfPredictor)[2], 
                                         prd=names(tmp[[1]]$mapper$dfPredictor)[1]
                                         ) %>%
                                  select(dataType, tgt, prd, everything())
                              }
                          )
tmpLiftHourPct

```
  
Variables are plotted based on explanatory power on hour:  
```{r, fig.height=9, fig.width=9}

tmpLiftHourPct %>% 
    filter(prd != "pct_hour") %>%
    ggplot(aes(x=fct_reorder(prd, lift, min), y=lift)) + 
    geom_point() + 
    coord_flip() + 
    facet_wrap(~c("tmpTest"="1. Test data", "tmpTrain"="2. Training data")[dataType]) + 
    geom_hline(yintercept=0, lty=2) +
    labs(y="Lift (percent correct divided by percent of most frequent hour, minus 1)", 
         x=NULL, 
         title="Explanatory power of variable on hour (as factor)"
         )

```
  
Data are converted to add aggregate elements:  
  
* Season (Dec-Feb=Winter, Mar-May=Spring, Jun-Aug=Summer, Sep-Nov=Fall)  
* Night/Day (07h-18h=Day, 00h-06h and 19h-23h=Night)  
* Combined Season and Night/Day  
  
```{r, fig.height=9, fig.width=9}

# Add random variables to dataset, then split in to test and train
set.seed(23080412) # Same seed as above
tmpTempRand <- tmpTemp %>%
    mutate(pct_0005=sample(0:5, size=nrow(.), replace=TRUE),
           pct_0025=sample(0:25, size=nrow(.), replace=TRUE), 
           pct_0100=sample(0:100, size=nrow(.), replace=TRUE), 
           pct_0250=sample(0:250, size=nrow(.), replace=TRUE),
           pct_0500=sample(0:500, size=nrow(.), replace=TRUE), 
           pct_1000=sample(0:1000, size=nrow(.), replace=TRUE), 
           pct_2500=sample(0:2500, size=nrow(.), replace=TRUE), 
           pct_5000=sample(0:5000, size=nrow(.), replace=TRUE), 
           tod=ifelse(hour>=7 & hour<=18, "Day", "Night"), 
           season=case_when(month %in% c("Mar", "Apr", "May") ~ "Spring", 
                            month %in% c("Jun", "Jul", "Aug") ~ "Summer", 
                            month %in% c("Sep", "Oct", "Nov") ~ "Fall", 
                            month %in% c("Dec", "Jan", "Feb") ~ "Winter", 
                            TRUE~"typo"
                            ), 
           todSeason=paste0(season, "-", tod), 
           tod=factor(tod, levels=c("Day", "Night")), 
           season=factor(season, levels=c("Spring", "Summer", "Fall", "Winter")), 
           todSeason=factor(todSeason, 
                            levels=paste0(rep(c("Spring", "Summer", "Fall", "Winter"), each=2), 
                                          "-", 
                                          c("Day", "Night")
                                          )
                            )
           )
tmpTempRand %>% count(tod)
tmpTempRand %>% count(season)
tmpTempRand %>% count(todSeason)
idxTrain <- sort(sample(1:nrow(tmpTempRand), size=round(0.75*nrow(tmpTempRand)), replace=FALSE))
tmpTempTrain <- tmpTempRand[idxTrain, ]
tmpTempTest <- tmpTempRand[-idxTrain, ]

# Example process for season
simpleOneVarTrainTest(dfTrain=tmpTempTrain, 
                      dfTest=tmpTempTest, 
                      tgt="season", 
                      prd="pct_temperature_2m", 
                      printReport=TRUE, 
                      returnData=FALSE
                      )

```
  
Predictive power for each variable on season is explored:  
```{r, fig.height=9, fig.width=9}

# Get all pct variables
pctVars <- tmpTempTrain %>% 
    select(starts_with("pct")) %>%
    names()
pctVars

# Run each variable and combine as dfr
tmpLiftPctSeason <- map_dfr(.x=pctVars, 
        .f=function(x) {
            tmp <- simpleOneVarTrainTest(dfTrain=tmpTempTrain, dfTest=tmpTempTest, tgt="season", prd=x)
            map_dfr(.x=tmp, .f=function(y) y$dfConfData$dfConfOverall) %>%
                mutate(dataType=names(tmp), 
                       tgt=names(tmp[[1]]$mapper$dfPredictor)[2], 
                       prd=names(tmp[[1]]$mapper$dfPredictor)[1]
                       ) %>%
                select(dataType, tgt, prd, everything())
            }
        )
tmpLiftPctSeason

```
  
Variables are plotted based on explanatory power on season:  
```{r, fig.height=9, fig.width=9}

tmpLiftPctSeason %>% 
    ggplot(aes(x=fct_reorder(prd, lift, min), y=lift)) + 
    geom_point() + 
    coord_flip() + 
    facet_wrap(~c("tmpTest"="1. Test data", "tmpTrain"="2. Training data")[dataType]) + 
    geom_hline(yintercept=0, lty=2) +
    labs(y="Lift (percent correct divided by percent of most frequent season, minus 1)", 
         x=NULL, 
         title="Explanatory power of variable on season"
         )

tmpLiftPctSeason %>% 
    ggplot(aes(x=fct_reorder(prd, lift, min), y=lift)) + 
    geom_point(aes(color=c("tmpTest"="1. Test data", "tmpTrain"="2. Training data")[dataType])) + 
    coord_flip() + 
    geom_hline(yintercept=0, lty=2) +
    labs(y="Lift (percent correct divided by percent of most frequent season, minus 1)", 
         x=NULL, 
         title="Explanatory power of variable on season"
         ) + 
    scale_color_discrete(NULL)

```
  
Predictive power for each variable on day-night is explored:  
```{r, fig.height=9, fig.width=9}

# Get all pct variables (exclude hour)
pctVars <- tmpTempTrain %>% 
    select(starts_with("pct")) %>%
    select(-pct_hour) %>%
    names()
pctVars

# Run each variable and combine as dfr
tmpLiftPctDayNight <- map_dfr(.x=pctVars, 
        .f=function(x) {
            tmp <- simpleOneVarTrainTest(dfTrain=tmpTempTrain, dfTest=tmpTempTest, tgt="tod", prd=x)
            map_dfr(.x=tmp, .f=function(y) y$dfConfData$dfConfOverall) %>%
                mutate(dataType=names(tmp), 
                       tgt=names(tmp[[1]]$mapper$dfPredictor)[2], 
                       prd=names(tmp[[1]]$mapper$dfPredictor)[1]
                       ) %>%
                select(dataType, tgt, prd, everything())
            }
        )
tmpLiftPctDayNight

```
  
Variables are plotted based on explanatory power on night-day:  
```{r, fig.height=9, fig.width=9}

tmpLiftPctDayNight %>% 
    ggplot(aes(x=fct_reorder(prd, lift, min), y=lift)) + 
    geom_point() + 
    coord_flip() + 
    facet_wrap(~c("tmpTest"="1. Test data", "tmpTrain"="2. Training data")[dataType]) + 
    geom_hline(yintercept=0, lty=2) +
    labs(y="Lift (percent correct divided by percent of most frequent day-night, minus 1)", 
         x=NULL, 
         title="Explanatory power of variable on day-night"
         )

tmpLiftPctDayNight %>% 
    ggplot(aes(x=fct_reorder(prd, lift, min), y=lift)) + 
    geom_point(aes(color=c("tmpTest"="1. Test data", "tmpTrain"="2. Training data")[dataType])) + 
    coord_flip() + 
    geom_hline(yintercept=0, lty=2) +
    labs(y="Lift (percent correct divided by percent of most frequent day-night, minus 1)", 
         x=NULL, 
         title="Explanatory power of variable on day-night"
         ) + 
    scale_color_discrete(NULL)

```
  
Lift by Variable is plotted for season and night-day:  
```{r, fig.height=9, fig.width=9}

tmpLiftPctSeason %>% 
    filter(prd != "pct_hour") %>%
    bind_rows(tmpLiftPctDayNight) %>%
    filter(dataType=="tmpTest") %>%
    select(tgt, prd, lift) %>%
    pivot_wider(id_cols="prd", names_from="tgt", values_from="lift") %>%
    mutate(prdType=case_when(str_detect(prd, "_\\d{4}$")~"4. Random", 
                             str_detect(prd, "radia")|str_detect(prd, "evapotrans")~"2. Radiation/et0",
                             str_detect(prd, "temper|dewp|vapor|soil")~"1. Temp/Dew/Vapor/Soil",
                             TRUE ~ "3. Other"
                             )
           ) %>%
    ggplot(aes(x=season, y=tod)) + 
    geom_point(aes(color=prdType)) + 
    labs(x="Lift (season)", 
         y="Lift (night-day)", 
         title="Lift on test data of single-variable predictor"
         ) +
    geom_hline(yintercept=c(0, 0.25), lty=2) +
    geom_vline(xintercept=c(0, 0.5), lty=2) +
    facet_wrap(~prdType) + 
    scale_color_discrete("Type")

```
  
Predictive power for each variable on day-night-season is explored:  
```{r, fig.height=9, fig.width=9}

# Get all pct variables (exclude hour)
pctVars <- tmpTempTrain %>% 
    select(starts_with("pct")) %>%
    select(-pct_hour) %>%
    names()
pctVars

# Run each variable and combine as dfr
tmpLiftPctDayNightSeason <- map_dfr(.x=pctVars, 
        .f=function(x) {
            tmp <- simpleOneVarTrainTest(dfTrain=tmpTempTrain, dfTest=tmpTempTest, tgt="todSeason", prd=x)
            map_dfr(.x=tmp, .f=function(y) y$dfConfData$dfConfOverall) %>%
                mutate(dataType=names(tmp), 
                       tgt=names(tmp[[1]]$mapper$dfPredictor)[2], 
                       prd=names(tmp[[1]]$mapper$dfPredictor)[1]
                       ) %>%
                select(dataType, tgt, prd, everything())
            }
        )
tmpLiftPctDayNightSeason

```
  
Variables are plotted based on explanatory power on night-day-season:  
```{r, fig.height=9, fig.width=9}

tmpLiftPctDayNightSeason %>% 
    ggplot(aes(x=fct_reorder(prd, lift, min), y=lift)) + 
    geom_point() + 
    coord_flip() + 
    facet_wrap(~c("tmpTest"="1. Test data", "tmpTrain"="2. Training data")[dataType]) + 
    geom_hline(yintercept=0, lty=2) +
    labs(y="Lift (percent correct divided by percent of most frequent day-night-season, minus 1)", 
         x=NULL, 
         title="Explanatory power of variable on day-night-season"
         )

tmpLiftPctDayNightSeason %>% 
    ggplot(aes(x=fct_reorder(prd, lift, min), y=lift)) + 
    geom_point(aes(color=c("tmpTest"="1. Test data", "tmpTrain"="2. Training data")[dataType])) + 
    coord_flip() + 
    geom_hline(yintercept=0, lty=2) +
    labs(y="Lift (percent correct divided by percent of most frequent day-night-season, minus 1)", 
         x=NULL, 
         title="Explanatory power of variable on day-night-season"
         ) + 
    scale_color_discrete(NULL)

# Example process for season
simpleOneVarTrainTest(dfTrain=tmpTempTrain, 
                      dfTest=tmpTempTest, 
                      tgt="season", 
                      prd="pct_temperature_2m", 
                      printReport=TRUE, 
                      returnData=FALSE
                      )

```
  
The top-performing Variable for night-day-season is plotted:  
```{r, fig.height=9, fig.width=9}

# Example process for night-day-season
simpleOneVarTrainTest(dfTrain=tmpTempTrain, 
                      dfTest=tmpTempTest, 
                      tgt="todSeason", 
                      prd="pct_soil_temperature_0_to_7cm", 
                      printReport=TRUE, 
                      returnData=FALSE
                      )

```
  
